### 2019-12-17 paper/paper.tex
Move

%%
%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[conference]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage{multirow}
\usepackage[inline]{enumitem}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{subfig}

\usepackage[nomain, toc, acronym]{glossaries}
\glsdisablehyper

\newcommand\note[2]{{\color{#1}#2}}
\newcommand\todo[1]{{\note{red}{TODO: #1}}}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{An in-depth study of Explainability and Adversarial Robustness for RNNs}

\author{\IEEEauthorblockN{Alexander Hartl,
Maximilian Bachl, Joachim Fabini, Tanja Zseby}
\IEEEauthorblockA{Technische Universität Wien\\
firstname.lastname@tuwien.ac.at}}

%\date{\today}

% \IEEEoverridecommandlockouts
% \IEEEpubid{\begin{minipage}[t]{\textwidth}\ \\[10pt]
%         \centering\normalsize{xxx-x-xxxx-xxxx-x/xx/\$31.00 \copyright 2018 IEEE}
% \end{minipage}}

% \renewcommand*{\bibfont}{\footnotesize}

\maketitle%

\thispagestyle{plain}
\pagestyle{plain}

\newacronym{ml}{ML}{Machine Learning}
\newacronym{dl}{DL}{Deep Learning}
\newacronym{aml}{AML}{Adversarial Machine Learning}
\newacronym{ids}{IDS}{Intrusion Detection System}
\newacronym{rnn}{RNN}{Recurrent Neural Network}
\newacronym{fgsm}{FGSM}{Fast Gradient Sign Method}
\newacronym{cw}{CW}{Carlini-Wagner method}
\newacronym{pgd}{PGD}{Projected Gradient Descent}
\newacronym{pdp}{PDP}{Partial Dependence Plot}
\newacronym{ars}{ARS}{Adversarial Robustness Score}
\newacronym{ttl}{TTL}{Time-to-Live}
\newacronym{dos}{DoS}{Denial-of-Service}
\newacronym{iat}{IAT}{Interarrival time}
\newacronym{tcp}{TCP}{Transmission control protocol}

\begin{abstract}

\glspl{rnn} yield attractive properties for %performing intrusion detection on network data.
constructing \glspl{ids} for network data.
With the rise of ubiquitous \gls{ml} systems, malicious actors have been catching up quickly to find new ways to exploit \gls{ml} for profit. Recently developed adversarial \gls{ml} techniques focus on computer vision and their applicability to network traffic is not straightforward: Network packets expose fewer features than an image, are sequential and impose several constraints on their features. %For example, destination ports cannot be modified since that could impair communication.

We show that despite these completely different characteristics, adversarial samples can be reliably generated for \glspl{rnn}.
To understand a classifier's potential for misclassification, we extend existing explainability techniques and propose new ones, suitable particularly % specifically crafted %AH most can also be used for non-sequential models, so lets not restrict applicability too much
for sequential data. Applying them shows that already the first packets of a communication flow are of crucial importance and are likely to be targeted by attackers. Feature importance methods show that even relatively unimportant features can be effectively abused to generate adversarial samples.
Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids}, capturing a common notion of adversarial robustness, and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}. %as a countermeasure. % AH. the score alone is not yet a countermeasure
%As a countermeasure, we develop an easy-to-use adversarial training procedure %and can confirm that it
%which significantly and successfully reduces the attack surface and improves \gls{ars}.

\end{abstract}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
%\begin{abstract}
%  A clear and well-documented \LaTeX\ document is presented as an
%  article formatted for publication by ACM in a conference proceedings
%  or journal publication. Based on the ``acmart'' document class, this
%  article presents and explains many of the common variations, as well
%  as many of the formatting elements an author may use in the
%  preparation of the documentation of their work.
%\end{abstract}

\maketitle

\section{Introduction}

There is a significant body of scientific work focusing on the detection of unwanted behavior in networks. In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content. More recently, with the rise of omnipresent encryption, available features have changed. Since payload is no longer available, the focus now lies on features that are always available to network monitoring equipment such as port numbers, protocol flags or packet sizes.

Network communication is usually grouped into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties, between two applications potentially being located at different physical locations.
When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets. % as well as their size. %AH: don't see why pkt size is not available without flows
%For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one to the attacker.
Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards an anomaly can be detected. Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature selection procedures while providing any available information to the classifier.
We show that our system has similar performance to other flow-based anomaly detection systems but can commonly detect anomalies \textit{before} the flow is over.

However, for practical use, high accuracy is not enough. With the recent rise of interest in \gls{aml} techniques, also \gls{aml} for \glspl{ids} has been investigated. For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids}.
In this research, we investigate whether adversarial samples can be found for our \gls{rnn}-based model, i.e. minimally different attack flows which are classified as benign. This is not a trivially answerable question since adversarial samples have mostly been analyzed in the context of computer vision. Our scenario significantly deviates from this in several ways because \begin{enumerate*}
\item the number of features is significantly smaller and
\item only certain features can be manipulated by an adversary (for example the destination port cannot be manipulated)
\end{enumerate*}.

Surprisingly, we can confirm that adversarial samples can be found even when considering these tight constraints. % mentioned before.
Thus, we argue that traditional methods like, e.g., accuracy are not sufficient for evaluating security-sensitive \gls{ml} systems.
We propose the \acrfull{ars} as a new performance score, which precisely captures the notion of how easily an adversary can generate adversarial samples.

Due to the threat of adversarial attacks, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans. This recently stirred up increased interest in explainability methods. Also in this case, common methods are designed to work with images or tabular data but not with sequences. Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data. %These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
Astonishingly, feature importance methods reveal that
features manipulable for adversarial flows,
%the that are manipulable features, that are left for modification by adversaries
are not even particularly important for the \gls{rnn}'s classification outcome. %Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}.

%Since common explainability methods are designed to work with images or tabular data but not sequences, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data. These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.

Based on the insights gained by the feature importance and explainability methods, we propose two defence methods.
%develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
On the one hand, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Furthermore,
by deploying an adversarial training procedure,
%AH. compared to other research about adversarial training, I dont feel that out approach is so special
%develop a custom adversarial training procedure which especially focuses on ease of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low. With this novel strategy
we can reduce the attack surface and harden the resulting \gls{ids} while retaining similar classification performance. % like for the original \gls{ml} model.

\section{An \gls{rnn}-based classifier}

% \begin{table}[b]
% \caption{occurrence frequency of attack types.} \label{tab:occurrence}
% \centering
% \begin{tabular}{l r} \toprule
% Attack type & Proportion \\
% \midrule
% Normal                                                         & $0.747468$ \\
% DoS / DDoS:DoS Hulk                                            & $0.101014$ \\
% PortScan:PortScan - Firewall off                               & $0.069020$ \\
% DDoS:LOIT                                                      & $0.040784$ \\
% Infiltration:Dropbox download                                  & $0.032982$ \\
% DoS / DDoS:DoS GoldenEye                                       & $0.003224$ \\
% DoS / DDoS:DoS Slowhttptest                                    & $0.001819$ \\
% DoS / DDoS:DoS slowloris                                       & $0.001680$ \\
% Brute Force:SSH-Patator                                        & $0.001107$ \\
% Botnet:ARES                                                    & $0.000327$ \\
% Web Attack:XSS                                                 & $0.000293$ \\
% PortScan:PortScan - Firewall on                                & $0.000165$ \\
% Brute Force:FTP-Patator                                        & $0.000110$ \\
% Web Attack:Sql Injection                                       & $0.000006$ \\
% DoS / DDoS:Heartbleed                                          & $0.000001$ \\
% \bottomrule
% \end{tabular}
% \label{tab:occurrence}
% \end{table}

\begin{table}[b]
\caption{Occurrence frequency of attack types.}
\label{tab:occurrence}
\centering
\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
DoS Hulk	&	10.10\%	\\
PortScan, Firewall	&	6.90\%	\\
DDoS LOIT	&	4.08\%	\\
Infiltration	&	3.30\%	\\
DoS GoldenEye	&	0.32\%	\\
DoS SlowHTTPTest	&	0.18\%	\\
DoS Slowloris	&	0.17\%	\\
Brute-force SSH	&	0.11\%	\\
Botnet ARES	&	0.03\%	\\
XSS attack	&	0.03\%	\\
PortScan, no Fw.	&	0.02\%	\\
Brute-force FTP	&	0.01\%	\\
SQL injection	&	$<$0.01\%	\\
Heartbleed	&	$<$0.01\%	\\
\bottomrule
\end{tabular}
}{}
\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
]{
\begin{tabular}{l r}
\toprule
Attack type & \hspace*{-4mm}Proportion \\ \midrule
Exploits	&	1.42\%	\\
Fuzzers	&	1.01\%	\\
Reconnaissance	&	0.58\%	\\
Generic	&	0.21\%	\\
DoS	&	0.19\%	\\
Shellcode	&	0.08\%	\\
Analysis	&	0.03\%	\\
Backdoors	&	0.02\%	\\
Worms	&	0.01\%	\\
\bottomrule
\end{tabular}
}{}
\end{table}

We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (if packet goes there or back) and all \gls{tcp} flags (0 if the flow is not \gls{tcp}).
% \begin{itemize*}
% \item source port
% \item destination port
% \item protocol identifier
% \item packet sizes
% \item \glspl{iat} of packets
% \item packet directions (if packet goes there or back)
% \item all \gls{tcp} flags (0 if the flow is not \gls{tcp})
% \end{itemize*}.
We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.  Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
During flow extraction we used a usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
We use Z-score normalization and a train/test split of 2:1.

For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks. Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
\autoref{tab:performance_results} shows that our classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}. However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before they are over.

\begin{table}
% Evaluated using
% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
% on Dec. 4th
\caption{Performance metrics per packet and per flow.} \label{tab:performance_results}
\centering
\begin{tabular}{l r r r r} \toprule
& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
	&	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
Accuracy	&	99.1\%	&	99.7\%	&	99.5\%	&	98.3\%	\\
Precision	&	97.0\%	&	99.7\%	&	83.4\%	&	78.6\%	\\
Recall	&	97.8\%	&	99.1\%	&	87.6\%	&	72.6\%	\\
F1	&	97.4\%	&	99.4\%	&	85.5\%	&	75.5\%	\\
Youden	&	97.2\%	&	99.0\%	&	87.3\%	&	71.9\%	\\
\bottomrule
\end{tabular}
\end{table}

\section{Adversarial attacks}
\label{sec:adv}
We now investigate whether known \gls{aml} methods can be used for generating adversarial flows for our \gls{rnn}. \cite{hashemi_towards_2019} previously studied the behavior of several Network \glspl{ids} under adversarial attacks but unlike us did not investigate \glspl{rnn}.

A network packet contains significantly less features than, e.g., an image and, furthermore, most features such as \gls{tcp} flags cannot be easily manipulated, as their manipulation might violate the protocol specifications and thus cause communication to fail. We identify the packet length and the \gls{iat} as features which are most likely to be exploited, but even these features cannot be manipulated freely. Beside the non-negativity constraint,  there are also more problem-specific constraints:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Only packets can be manipulated which are transmitted by the attacker.
Except for botnet and backdoor traffic, which is entirely controlled by an attacker, thus only packets travelling in one direction can be manipulated.
 %This means the attacker can only control one direction (except for botnets where both sides are attackers).
\item Packets must not be smaller than initially, as otherwise less information could be transmitted or packets might be rendered syntactically invalid.
\item \glspl{iat} must not decrease, as otherwise the physical speed of data transmission can be violated in some cases. An in-depth analysis of when the reduction of \glspl{iat} is possible, would come with substantial effort, so we generally disallowed it.
\end{itemize}

Several \gls{aml} methods have been proposed in the recent literature, achieving different speed-quality tradeoffs:
\cite{szegedy_intriguing_2014} shows that it is possible to create adversarial samples for an image, which look similar to the original image but are classified wrongly. \cite{goodfellow_explaining_2015} develops the \gls{fgsm}, which makes easy generation of adversarial samples possible. \cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks import \gls{aml} methods, which we consider in this paper.

% Max: But FGSM is high-speed... So I removed this paragraph altogether
%As finding adversarial flows is particularly difficult in our case, % for the reasons we discussed above,
%we preferred methods  producing high-quality adversarial samples while not being as fast as other methods. We implemented the following \gls{aml} methods.

\subsection{Carlini-Wagner}
We implemented the \gls{cw} \cite{carlini_towards_2017}, performing gradient descent on the optimization objective
\begin{equation} \label{eq:carliniWagner}
d(X,\tilde X) + c  \max(Z(\tilde X), \delta).
\end{equation}
Here, $d(\cdot)$ is a distance metric and $c \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow. Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow under optimization.

$\delta \in \mathbb R$ is a parameter that determines how far an adversary wants to exceed the decision boundary.
In the original publication $\delta=0$, meaning that the network's decision for adversarial samples is just between attack and benign traffic. In the present context, we need to make sure that the classifier's prediction would actually be benign, even though a certain level of noise will be added to \glspl{iat} due to the network between attacker and victim. Hence, we introduced $\delta=-0.2$, corresponding to a prediction confidence of 55\% for the sample to be benign.

We used $L_1$ distance as distance metric, as we consider it to represent practically relevant differences of network flows best. We used \gls{pgd} for meeting the real-world constraints discussed above.

\subsection{$L_\infty$-bounded Projected Gradient Descent}
\cite{madry_towards_2018} uses a method for generating adversarial samples which deploys \gls{pgd}  to maximize the network's loss function while constraining the achieved $L_\infty$ distance from the original samples. Hence, an important difference to \gls{cw} is that for this method the network's loss instead of its logit output is considered.
We consider this method an interesting combination of \gls{cw} and \gls{fgsm}.
Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be of similar quality.

\subsection{Fast Gradient Sign Method}
Finally, we also tested the \gls{fgsm} % which was initially proposed in
 \cite{goodfellow_explaining_2015}, which perhaps is the most frequently used method for generating adversarial samples. FGSM can be considered a single pass of \gls{pgd} on the loss function with an equality constraint on the $L_\infty$ distance, i.e. the adversarial sample is found as
\begin{equation}
\tilde X = X + \epsilon \text{sgn}( \nabla_X L(X)),
\end{equation}
where $L(X)$ denotes the network's loss function and $\epsilon$ denotes the achieved $L_\infty$ distance.

\begin{figure}[t]
\includegraphics[width=\columnwidth]{../plots/adv_comparison/adv_comparison_17.pdf}
\caption{Attack success ratios for both datasets and per attack family for CIC-IDS-2017. Depicted are flow detection accuracies for adversarial flows and for unmodified attack flows.}
\label{fig:adv_per_family}
\end{figure}

\subsection{Evaluation}
\autoref{fig:adv_per_family} depicts the performance for the investigated algorithms for both datasets and for each attack family in CIC-IDS-2017.
We used \gls{cw} with a tradeoff of 1 and, to provide a fair comparison, for \gls{fgsm} and \gls{pgd} we used an average $L_\infty$ distance as observed for \gls{cw}.

As expected, \gls{cw} delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
%\autoref{fig:adv_per_family} depicts how successful the attack is for individual attack families.
The figure shows significant differences for detecting different attack families in the first place. Also for generating adversarial flows some attack families are more susceptible than others. However, the results match to a high degree with our expectations. For example, SQL injection attacks apparently are closer to benign traffic than \gls{dos} attacks and, hence, finding adversarial samples should be easier.

Interestingly, any increase of the $L_\infty$ bound for \gls{pgd} did not yield significantly improved performance. 
\gls{cw} thus generally achieved superior results to $L_\infty$-bounded \gls{pgd}
and we can confirm the recommendation by \cite{carlini_towards_2017} to perform gradient descent on the logit output instead of the loss for good results.

\autoref{fig:adv_cw} depicts the success ratio and average distance for \gls{cw} for different tradeoffs $c$.
%\autoref{fig:adv_cw} depicts the ratio of attack flows which were successfully misclassified as benign with respect to the final packet along with the resulting distance from the original sample.
Evidently, the larger $c$, the better the attack works, but also distances from original flows become higher. With acceptable distances, we were able to generate adversarial samples for about 80\% of attack samples.

Note, however, that for successful adversarial samples, the second term in \autoref{eq:carliniWagner} becomes constant and, hence, is no longer relevant for optimization. Hence, $c$ is irrelevant for the achieved distance, but governs if for one sample an adversarial sample is found.
However, to avoid overly large steps when using a large $c$ in \autoref{eq:carliniWagner}, it is necessary to reduce the gradient descent learning rate. Since the distance term is not affected by $c$, it is furthermore necessary to scale the iteration count inversely with the learning rate. \gls{cw} with a large $c$ therefore needs more time than with a small $c$.

% The figure suggests that a tradeoff value between 0.5 and 1 would be ideal to find adversarial samples with a low enough distance from real samples.
% This conclusion might be misleading, however, as due to the objective function \eqref{eq:carliniWagner} the distance cannot become larger anymore for one sample as soon as an adversarial sample has been found. %, when performing optimization as depicted by the objective function .
% Hence, the depicted distance increase presumably is due to samples for which no adversarial sample could be found for lower $c$ values.

\begin{figure}[b]
\includegraphics[width=\columnwidth]{../plots/adv_comparison/tradeoff.pdf}
% \includegraphics[width=0.49\textwidth]{adv_plots/cwl1.pdf}
% \includegraphics[width=0.49\textwidth]{adv_plots/cwl2.pdf}
\caption{Performance of \gls{cw} with different $c$ values.}
\label{fig:adv_cw}
\end{figure}

\section{Adversarial Robustness Metric}

As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}. High accuracy for a particular attack category does not imply that adversarial attacks are  likely to fail for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a new metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
Such a metric should be easy to compute, easy to interpret and robust with respect to outliers. We thus propose the \gls{ars} as the average distance of the adversarial samples to the original samples for the 50\% of the samples which an adversary could successfully modified and which have the smallest distances. Since \gls{cw} is able to find adversarial samples with minimum distance, it can be used for finding the \gls{ars} efficiently.

Since \gls{cw} becomes slow for large $c$, the \gls{ars} can be efficiently found as follows: Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type. If, of a total of $N$ samples,
 more than $N/2$  are wrongly classified, and
 the smallest distance of non-adversarial samples is not smaller than  the $\lceil\frac{N}{2}\rceil$th smallest distance of adversarial samples,
stop. Otherwise increase $c$ and repeat. If after a predefined number of iterations -- e.g. 100 -- still not more than half of the samples are adversarial, also break.

The result is a number of adversarial samples and some that could not be made adversarial. If at least half of the samples are misclassified, we sort the adversarial samples by ascending distance to the original samples. Then we take samples by ascending distance until we took 50\% of all samples. Of these samples, the distance of each adversarial sample to the original sample is taken and averaged. The resulting average distance is the \gls{ars}. The larger it is the better since then an adversary needs to modify the samples more to make them adversarial. If not more than half of the samples could be made adversarial, our metric is $\infty$ since then apparently it is not possible for an adversary to reliably conduct adversarial attacks on the majority of samples. Then the ratio of samples that could be made adversarial ($<$50\%) can be a useful metric to determine the exact extent of the adversarial threat. Setting the threshold for the \gls{ars} to 50\% is arbitrary, but
 % reasonable as this way the \gls{ars} approximates the median of the distances of all samples and, additionally, marks the distance where an adversarial attack is more likely to be successful than unsuccessful.
 reasonable as outliers with very large distances are ignored and if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.
%we think that it is reasonable since if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.

\begin{figure}[h]
\includegraphics[width=\columnwidth]{../plots/ars_original.pdf}
\caption{Recall for each attack type and \gls{ars} for CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and that \gls{ars} is $\infty$.}
\label{fig:recall_ars}
\end{figure}

\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly easy for an attacker to perform adversarial modifications to make them look benign.

\section{Explaining Recurrent Neural Networks}
From a naive perspective, one might be tempted to reuse explainability methods for \glspl{rnn} by considering a flow the sum of its packet features.
We identify several difficulties which occur when trying to explain decisions made by \glspl{rnn}.

\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item
\textbf{Feature quantity.}
The number of features that is fed into an \gls{rnn} is found as the product of packet features with the total length of the flow. For long flows, the total number of inputs can therefore grow to a substantial size.

\item
\textbf{Variable sequence lengths.}
The length of different flows might differ tremendously. Hence, features which are important for the network's outcome for one flow might not even exist for other flows. Furthermore, the question arises how the distance between flows of different length can be computed.

\item
\textbf{Lack of a distance measure.}
However, even if we restricted the analysis to flows of a constant length, a flow is different from the plain concatenation of its packet features.
If we, for example, consider a classification problem as in this document, we expect that after a certain amount of time steps the classifier is able to perform the classification with a certain confidence. Arguably, features at the beginning of a flow are thus more important for the classifier. Giving feature differences at the end of a flow the same weight as differences at its beginning, therefore seems incorrect.

\item
\textbf{Multiple prediction outputs.}
Often an \gls{rnn} produces an output at each time step. When  applying explainability methods, the question arises which output to consider for the method. The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation. Not only does this approach require a substantially lower amount of computational resources compared to predictions occurring at a later time step, but we also expect the immediately occurring prediction outcome to exhibit a stronger dependence of the investigated feature, compared to a prediction outcome occurring many time steps later.
However, due to the complex decision processes of deep neural networks,
this does not hold true and a feature might influence a decision many time steps later.
\end{itemize}

Many explainability methods proposed recently are local and thus provide explanations for a particular data sample \cite{shapley_value_1953,lundberg_unified_2017,dhurandhar_model_2018,ribeiro_why_2016}. However, for the particular problem of network traffic, due to the high number of flows and the characteristics of data, analyzing individual samples is of low interest. Explainability methods presented in this paper therefore aim to understand a model by analyzing which features are important, at which time step they are important and which feature values lead to classification as attack.

% where does this come from :)
%\subfloat[Throughput of the fq \gls{qdisc}\label{fig:fqThroughput}
%]{
%\includegraphics[width=\columnwidth]{{{plots/throughput_1_fq_cubic_10_20_60_0.5_bw_1571240937211.pcap}}}
%}{}

\subsection{Feature Importance Metrics}
%\begin{table}
%\caption{Accuracy drop for:}
%\label{tab:feat_importance}
%\centering
%\subfloat[Random replacement method\label{fig:featRandom}
%]{
%\begin{tabular}{l r}
%\toprule
%Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
%Protocol & 0.207 \\
%Packet Length & 0.165 \\
%SYN Flag & 0.099 \\
%ACK Flag & 0.084 \\
%Direction & 0.073 \\
%Destination port & 0.071 \\
%Source port & 0.060 \\
%RST Flag & 0.057 \\
%PSH Flag & 0.056 \\
%Interarrival time & 0.024 \\
%FIN Flag & 0.012 \\
%URG Flag & 0 \\
%ECE Flag & 0 \\
%CWR Flag & 0 \\
%NS Flag & 0 \\
%\bottomrule
%\end{tabular}
%}{}
%\subfloat[Feature dropout\label{fig:featDropout}
%]{
%\begin{tabular}{l r}
%\toprule
%Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
%Destination port & 0.025 \\
%Source port & 0.003 \\
%RST Flag & 0.001 \\
%ACK Flag & 0.001 \\
%Protocol & 0.001 \\
%Packet Length & 0.001 \\
%Direction & 0.001 \\
%SYN Flag & 0 \\
%Interarrival time & 0 \\
%FIN Flag & 0 \\
%ECE Flag & 0 \\
%URG Flag & 0 \\
%CWR Flag & 0 \\
%NS Flag & 0 \\
%PSH Flag & 0 \\
%\bottomrule
%\end{tabular}
%}{}
%\end{table}
As a first step to understanding the neural network's decisions, we  estimate how important individual features are for the model's predictions.
When investigating an \gls{ml} classifier, it is natural to ask which inputs have a large influence on the classifier's prediction. We feel the need to distinguish metrics for this purpose based on their main aim:

A large amount of research has been spent on finding \emph{feature importance} metrics, which allow the selection  of high-importance features, providing a reasonably good classification performance while resulting in a more light-weight classifier.

On the other hand, both adversarial machine learning and explainable machine learning bring up the question to what extent individual features are able to change the prediction outcome. While being seemingly similar at first sight, traditional feature importance can yield markedly wrong results for this case.  To distinguish, we propose the term \emph{feature sensitivity} for such metrics.
To analyze features for our network, we use the following approaches.

\subsubsection{Neural Network Weights}
In previous works~\cite{olden_accurate_2004}, a simple method for determining feature importance in neural networks, has been summing up neural network weights leading from a certain input to the prediction outcome. The weights method can be considered for both feature importance and feature sensitivity, however, clearly, especially in the case of complex network architectures, this method is likely to provide wrong results. Hence, we provide results for the weights method mainly for comparison. Note that an LSTM cell alone has four weights leading from one input to an output.

\subsubsection{Input Perturbation}

 In general, the most commonly deployed feature importance techniques, used by practitioners for \glspl{rnn} \cite{stackexchange_cross_validated_neural_2019} and \gls{dl} \cite{molnar_interpretable_2019,stackexchange_cross_validated_feature_2016,olden_accurate_2004}, are based on adding noise to a particular feature and evaluating the drop in accuracy that occurs. We argue that it is hard to determine the ``correct'' intensity of noise to add. %, we slightly vary the proposed approach: We do not add noise drawn from a probability distribution but instead
 Hence, we sample the value for a feature from the distribution of all values of this feature in the dataset. This makes the method non-parametric since the noise distribution  does not need to be chosen. We ensured that features which stay constant for a flow, i.e. source port, destination port and protocol, also stay constant throughout the flow when randomizing the feature.

\subsubsection{Feature Dropout}
While the perturbation method %of adding noise to features/randomizing them
is convenient since it is easy to implement and understand, we argue that it has some shortcomings:
The \gls{rnn} was never trained for dealing with ``garbage'' values that the randomization creates. For example, completely unrealistic feature combinations could occur that were never observed during training. Furthermore, sequences of features could occur that cannot occur in reality.

%We think that the ``ideal'' feature importance method for our scenario are Shapley values \cite{shapley_value_1953}: To analyze feature importance with them, the feature under investigation is completely omitted from the dataset and a model is trained without it. The resulting accuracy is then compared to the baseline accuracy with all features. The effort required to do this is high both in terms of storage space as well as training time since for $n$ features $n$ models need to be separately trained and stored. While this seems tractable, when also analyzing the feature importance of combinations of features, the number of models required becomes even larger: Out of $n$ features, $2^n$ unique subsets can be chosen, meaning that the computational demand rises exponentially.

To evaluate true feature importance, we thus develop a more sound method %compared to input perturbation
called \textit{feature dropout}: When training a model, for each sample, we leave out each feature  with independent probability $\frac{1}{n}$, $n \in \mathbb N$ being the number of features, by setting it to zero. On average, one feature gets zeroed out but it is also possible that none or more than one are left out. This procedure is equivalent to using dropout \cite{srivastava_dropout:_2014} with probability $\frac{1}{n}$ before the first layer.

An important implementation detail is that for each feature we add another input which is 1 if the feature is suppressed and 0 otherwise. This is necessary for the neural network to be able to distinguish between a feature missing and a feature genuinely being zero. The overall outcome is a classifier being able to deal with missing features.
The results in \autoref{fig:feat_imp} show that, unlike input perturbation, feature dropout  does not vastly overestimate features' importance. With feature dropout, it becomes apparent that only very few features  actually contain unique information, affecting accuracy when left out: the destination port and the source port.

A model trained with feature dropout typically yields slightly lower accuracy than a regularly trained model, even if no features are left out (flow accuracy of 99.43\% vs.99.65\%). We thus recommend training two models: One regular one and one with feature dropout to use for the feature importance.

Another method that uses dropout for feature importance is \cite{chang_dropout_2017}, applying a technique called \textit{Variational Dropout} to learn the optimal dropout rate for each feature. It tries to leave out as many features as possible and at the same time keep accuracy high. Thus important features are going to be left out less often and one can then extract the dropout probabilities for each feature and assess their significance based on them.
%Feature dropout can also be used as a building block for computing the accuracy of datapoints with missing features that is necessary to compute Shapley values \cite{shapley_value_1953} or KernelSHAP \cite{lundberg_unified_2017}.

As it can also happen that several features are left out for a sample, our method can also be used to analyze the effect of multiple features missing and can thus show possibly correlated features or -- more general -- features that contain common information, relevant for the classification task:
%Feature dropout also allows us to find correlated features or -- more general -- features that contain common information:
\begin{equation}
\text{score} = \frac{\text{acc}_{\text{base}}-\text{acc}_{\text{-both\_features}}}{\left(\text{acc}_{\text{base}}-\text{acc}_{\text{-feature\_1}}\right) + \left(\text{acc}_{\text{base}}-\text{acc}_{\text{-feature\_2}}\right)}
\end{equation}
With $\text{acc}_{\text{base}}$ we denote the accuracy of the classifier with all features included, with $\text{acc}_{\text{-feature\_i}}$ the accuracy if feature $i$ is omitted and with $\text{acc}_{\text{-both\_features}}$ the accuracy if both features are omitted. Assuming a non-negative accuracy drop when omitting a feature, the resulting score is $\ge 0.5$. The higher it is, the larger the information that both features share.

%AH. As it can become <1, I would say it this way
%If the score is one,  both features contain no joint information.
For example, the score between RST and the protocol identifier is 8.5; the highest of all pairs of features. While this may be unintuitive at the first glance, it likely stems from the fact that if the protocol identifier (\gls{tcp} or UDP) is missing, then 
%if RST is 1 at some point of a flow, it has to be \gls{tcp}.
RST being 1 at some point indicates that the flow is \gls{tcp}.

Feature dropout might constitute a building block for methods based on Shapley values \cite{shapley_value_1953} like KernelSHAP \cite{lundberg_unified_2017}. 
%Due to space limitations, we leave this for future work.
\subsubsection{Mutual Information}
\begin{figure}
\includegraphics[width=\columnwidth]{../plots/mutinfo_example.pdf}
\caption{Two distributions yielding an identical accuracy drop.}
\label{fig:mutinfo_example}
\end{figure}
Input perturbation and feature dropout mainly address feature importance.
For example, assuming that the test dataset is representative for production use, for feature importance it is reasonable to equate the distribution for perturbing a feature with the feature distribution itself. However, when evaluating feature sensitivity, e.g. for analyzing potential for adversarial samples, the attacker is not limited by this distribution and frequently is able to choose arbitrary values in the feature space.

Furthermore, let $f(A,x)$  denote the joint probability
%\todo{Max: Ist joint probability nicht immer wenn man mehrere Ereignisse hat? Ich hätte hier probability density function geschrieben...}
%AH. changed notation and wording
for classification as attack and a feature value $x$. Accuracy then is $\int f(A,x) dx$ for attacks and $1-\int f(A,x) dx$ for benign traffic.
\autoref{fig:mutinfo_example} shows two different distributions
yielding the same accuracy, but clearly the right-side distribution has a larger influence on the prediction, as in the right-side case the prediction deterministically depends on the feature value.
To capture such dependencies, we propose to use mutual information %instead of accuracy drop
to determine feature sensitivity.
Mutual Information between two random variables $X,Y$ is defined as
\begin{equation}
I_{X,Y} = \mathbb E \left\{ \log\left(\frac{f_{X,Y}(x,y)}{f_X(x)f_Y(y)}\right) \right\},
\end{equation}
with $f_X(x), f_Y(y)$ and $f_{X,Y}(x,y)$ denoting the distribution of $X,Y$ and their joint distribution, respectively. To obtain feature sensitivity, we compute mutual information between an input variable and the prediction output for one flow at one particular time step, averaging over the result for the test set.

\subsubsection{Comparison}
\begin{figure*}
\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2015.pdf}
\caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
\label{fig:feat_imp}
\end{figure*}
\autoref{fig:feat_imp} shows the results, which match to a large extent with domain-specific expectations for classifying network flows.
In particular, rarely used \gls{tcp} flags like NS or URG are unimportant for the classifier. On the other hand, destination port and protocol are essential for characterizing flows by hinting at the type of traffic. \gls{iat} and packet length are important for estimating the amount of transferred information and several flags hint at certain attacks like \gls{dos} attacks.

The weights method is able to reveal features with a very low importance to a certain degree, but disagrees with the other methods to a large extent. Less anticipated, however, also input perturbation does not exhibit a considerable correlation with feature dropout. Considering its functioning, feature dropout is the most reliable method for evaluating importance for removing features. It is remarkable that none of the other methods is able to depict the distinct peak of importance for the destination port.

It is not surprising that mutual information disagrees with feature dropout, since both their aim and their functioning are substantially different. For example, mutual information shows that the protocol field can have a substantial impact on the classification even though an identical accuracy can be achieved when omitting it.

% In particular, the most important features turn out to be the used protocol, packet length and the SYN TCP flag. While the first two features are essential for characterizing flows, a high number of packets with set SYN flag is an important characteristic for port scanning or certain DoS attacks. Also the fact that most of TCP flags do not play an important role in our dataset was expected. It is interesting to note, however, that the interarrival time of packets does not play a crucial role for the detection of attacks.

Metrics for UNSW-NB15 differ substantially from CIC-IDS-2017. However, due to the large number of different network attacks and network configurations, it is easily possible that relevant features are very different. We consider the question of model transferability of substantial importance for \gls{ids} applications, but out of scope for the present research.

\subsection{Explainability Plots}
Knowing which features to investigate, it is important to analyze which feature values lead to classification as attack.
In literature, the use of Partial Dependency Plots (\gls{pdp}) has been proposed~\cite{friedman_greedy_2001}. To inspect attack types in detail, in this research we use a \textit{conditional} form of the \gls{pdp}. If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space, $f(\boldsymbol X) \in [0,1]$ the neural network's prediction and $c$ the attack class, we define the conditional \gls{pdp} for feature $i$  as
% \begin{equation} \label{eq:pdp}
% \text{\gls{pdp}}_i(y) = \mathbb E_{\boldsymbol X}\left(f(X_1,\ldots,X_{i-1},y,X_{i+1},\ldots X_n)\right)
% \end{equation}

% To inspect attack types in detail, in this research we experiment with a modified definition of the \gls{pdp}  (with C being the attack classes):
\begin{equation} \label{eq:pdp_conditional}
\text{\gls{pdp}}_{c,i}(w) = \mathbb E_{\boldsymbol X | C} \Big(f \left( X_1,\ldots,X_{i-1},w,X_{i+1},\ldots X_n \right) | c\Big),
\end{equation}
empirically approximating the conditional distribution by the observed subset that belongs to a certain class.

By using a classical \gls{pdp} we would be likely to lose information due to the averaging over very different types of traffic. However, for network traffic in particular, investigating each sample individually is not possible. Hence, the conditional \gls{pdp} provides the ideal compromise for our application.

Due to the use of a 5-tuple flow key, port numbers and the protocol identifier are constant for all packets in one flow, which allows us to use \glspl{pdp}, which have been proposed for non-recurrent classification methods.
The results (omitted for brevity) show that for some attack classes the port numbers play an important role. When looking at the class of benign traffic %(non-attack traffic)
it becomes apparent that traffic destined to a high destination port is generally indicative of an attack. We argue that this is because most services that regular users use have low port numbers.

\subsection{Plots for Sequences}

%As apparent from the above discussion, trying to explain how packet features influence the model's predictions, poses various difficulties. As a first step to explaining our model's predictions, we investigate how important features at different time steps are for the model's decisions.
Intuitively, features at the beginning of a flow should be the most important while the classifier's predictions should not vary significantly anymore, as soon as it has come to a decision.
\begin{figure}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_5_DoS - DDoS-DoS Hulk"}.pdf}
\includegraphics[width=\columnwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_15_All samples"}.pdf}
% \includegraphics[width=\columnwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_6_DoS - DDoS-DoS Slowhttptest"}.pdf}
% \includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_full_no_ttl_pred_plots_outcomes_0_3.pickle_14_Web Attack-XSS"}.pdf}
% \includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_full_no_ttl_pred_plots_outcomes_0_3.pickle_11_PortScan-PortScan - Firewall off"}.pdf}
\caption{Classifier confidence per time step for CIC-IDS-2017. % and first and third quartiles.
For the majority of attack types, confidence increases in the first few steps and then stays almost constant at 1.}
\label{fig:confidence}
\end{figure}
To evaluate this hypothesis, we plot the classifier's prediction confidence for each time step. % for each attack class.
%averaging over all flows. %AH. there's no averaging happening
% and also show the first and third quartiles.
\autoref{fig:confidence} shows that %this hypothesis is correct for most attack classes.
while at the first couple of packets the confidence is not very high, towards the end of the flow it reaches values close to 1 and stays there. Hence, not only is the classifier able to make a reasonable classification after just a few packets, the figure also suggests that indeed later packets have a neglectable influence on the prediction.

For investigating in more detail how features influence the prediction outcome, we extend \glspl{pdp} to the sequential setting.
Denoting as $X=\{\boldsymbol X^1, \ldots, \boldsymbol X^m\}$ the series of packet features $\boldsymbol X^t \in \mathbb R^n$ and $h_t(X)$ the network's hidden state after time step $t$, we define the sequential \gls{pdp} as
\begin{align}
\text{se}&\text{qPDP}_{c,i}(t,w)= \\ &\mathbb E_{X | C} \Big(f \left(h_{t-1}( X), X_1^t,\ldots,X_{i-1}^t,w,X_{i+1}^t,\ldots X_n^t \right) | c\Big). \nonumber
\end{align}

\autoref{fig:pred_plots2} shows an example together with the mean values for both unmodified samples and adversarial samples. Also in this figure we notice that mainly the first few packets are able to influence the prediction outcome and modifying features at a later time step does not change its confidence any longer. In many cases, the adversarial sample generation indeed moves packet features to areas where the network is less likely to be classified as attacks, confirming the effectiveness of \glspl{pdp}. In other cases, however, we did not observe an agreement between \gls{pdp} and adversarial modifications, hinting at dependencies which cannot be presented by \glspl{pdp}.

\begin{figure}[b]
%\includegraphics[width=0.48\textwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_0_Botnet-ARES"}.pdf}
\includegraphics[width=\columnwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_7_DoS - DDoS-DoS slowloris"}.pdf}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_11_PortScan-PortScan - Firewall off"}.pdf}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_1_Brute Force-FTP-Patator"}.pdf}

\caption{Exemplary sequential PD plot and adversarial flows for the DoS Slowloris attack in CIC-IDS-2017. The lines show the feature's mean values. % for all flows of the attack type.
The shaded region shows the change in confidence that occurs when the feature is varied.}
%\todo{IAT plot is wrong: There shouldn't be step 0.} %AH. guess this is fine now
\label{fig:pred_plots2}
\end{figure}

% AH. I guess this can only be properly formulated with a formula
%Next, we want to investigate if our classifier just memorized its inputs. The most interesting packet features that can be used to visualize this in our use case are the \gls{iat} between packets in a flows and the packet length. To depict how the model comes to a decision, we determined the total minimum and maximum for these features that ever occur for all flows. For a given flow $X$ we then varied feature values at each time step between minimum and maximum and plot how the confidence changes when the feature varies (\autoref{fig:pred_plots2}) depict the averaged predictions for different attack families.

Furthermore, it seems clear that the classifier did not just memorize the inputs but actually learned appropriate space representations, as there is always a smooth decrease of confidence when modifying feature values. When memorizing, we would  instead expect a sharp decline of confidence. 
% to occur.
%Another interesting observation is that after the first couple of packets the classifier is certain about its decision and changing features doesn't change its confidence any longer. This allows the conclusion that if an attacker is able to change the first few packets, she could change the decision of the classifier. This is consistent with our observations in \autoref{sec:adv}. %We will further investigate this question in \autoref{sec:adv}.

\begin{figure}
%AH. I think the 'Firewall off' plot is not a good example as it shows only 4 packets
\includegraphics[width=\columnwidth]{{"../plots/plot_features/flows_pred_plots2_outcomes_0_3.pickle_2_Brute Force-SSH-Patator"}.pdf}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot_features/flows_full_no_ttl_pred_plots2_outcomes_0_3.pickle_6_DoS - DDoS-DoS Slowhttptest"}.pdf}
%\includegraphics[width=\columnwidth]{{"../plots/plot_features/flows_full_no_ttl_pred_plots2_outcomes_0_3.pickle_11_PortScan-PortScan - Firewall off"}.pdf}
%\includegraphics[width=0.48\textwidth]{{"../plots/plot_features/flows_full_no_ttl_pred_plots2_outcomes_0_3.pickle_7_DoS - DDoS-DoS slowloris"}.pdf}

\caption{\gls{iat} and packet length for SSH brute-force attacks in CIC-IDS-2017.}% The solid line shows the median value of the feature for all flows of a specific attack type. The shaded region are the first and third quartile.} % think legend is enough for unterstanding
\label{fig:plot_features}
\end{figure}
Finally, we investigated whether our classification task involves recognizing complex patterns in the feature space.
As example, \autoref{fig:plot_features} shows that
%Looking at the plots of the features that vary over time (\autoref{fig:plot_features}) we see that certain
attack types indeed have a characteristic pattern in which they send packets by which they are easily recognizable.
Other attack families similarly show characteristic patterns.
%We want to determine how the pattern of packet sizes and \glspl{iat} is significant for the classification. \autoref{fig:plot_features} clearly shows that some attack types have very characteristic patterns of packet sizes and \glspl{iat}.

\section{Defences}
We now propose two approaches to increase robustness of the \gls{rnn} against adversarial attacks.
\subsection{Reducing features}
The most obvious defence, which is only possible because of the nature of the dataset, is to simply omitting features an attacker can manipulate. We try two different approaches:
\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
\item Leaving out all manipulable features: packet size and \gls{iat}
\item Leaving out manipulable features when the attacker can manipulate them: only in the direction from the attacker to the victim (This, however does not prevent adversarial samples for botnets, for which both sides are malicious)
\end{itemize}

Both approaches lead to complete resistance to adversarial samples (as we left out all manipulable features), except for botnets, which can still operate when only leaving out manipulable features in one direction. The results show that -- surprisingly -- there is only a small difference in classification performance when only computing the accuracy for the last packet of each flow (flow accuracy): Flow accuracy stays over 99.93\%. However, packet accuracy is only 98\% when leaving out the features in one direction and 96.7\% when leaving them out in both directions, Thus, apparently the \gls{iat} and packet size are especially important for determining whether a flow is malicious in the first packets of a flow.

\subsection{Adversarial training}
As alternative to omitting manipulable features, we attempted to make the classifier more robust against adversarial samples by augmenting the training set by adversarial flows generated using  \gls{cw}, labeled as additional attacks.
For this purpose, we added one adversarial sample per attack sample to the training set. Since \gls{cw} is deterministic, this is the maximum number of adversarial samples possible. With this augmented training set we then alternated retraining of the neural network and optimization of the adversarial samples. % using gradient descent.

A question which occurs in this process, is how many CPU cycles to spend on network training and adversarial sample optimization. As a natural approach, we decided to use as many backpropagation steps for neural network training as we use for adversarial sample optimization.  For each adversarial sample optimization step, we performed 10 iterations of gradient descent. Hence, all adversarial samples were optimized each 10 epochs of neural network training.

\autoref{fig:avdistance} shows the \gls{ars} throughout  adversarial training  for several attack categories and clearly indicates that adversarial training is effective, as the distance gradually increases. Hence, an attacker would have to modify attack samples more and more, eventually rendering the attack unpractical.

%However, the figure also shows that an increase of distance can only be achieved after a considerable amount of training. It is, hence, questionable if the gained increase in robustness is enough for reliable attack detection.

For both datasets, after a small number of epochs, it was no longer possible to create a significant number of adversarial samples for most attack categories. %For the datasets we considered, adversarial training left classification performance unaffected.
%\todo{Max: Insert results that show that accuracy on normal data is still ok after adversarial training.}
%\autoref{tab:adv_training_performance} depicts classification on non-adversarial data after 50 epochs of adversarial training.
After 50 epochs of training, accuracy is essentially identical to the results presented in \autoref{tab:performance_results}. Recall and Youden increased and precision slightly decreased. However, this is due to the higher proportion of attack samples in training data and, hence, expected.
We conclude that adversarial training is effective for reducing the attack surface for adversarial attacks in our scenario.

% \begin{table}
% \caption{Detection performance after adversarial training.} \label{tab:adv_training_performance}
% \todo{Probably no space for table}
% \centering
% \begin{tabular}{l r r r r} \toprule
% & \multicolumn{2}{c}{CIC-IDS-17} & \multicolumn{2}{c}{UNSW-NB-15} \\
	% &	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
% Accuracy	&	0.991	&	0.997	&	0.994	&	0.981	\\
% Precision	&	0.971	&	0.996	&	0.761	&	0.704	\\
% Recall	&	0.979	&	0.992	&	0.970	&	0.824	\\
% F1	&	0.975	&	0.994	&	0.853	&	0.759	\\
% Youden	&	0.973	&	0.991	&	0.965	&	0.811	\\

% \bottomrule
% \end{tabular}
% \end{table}

% AH. guess this is no longer true
%However, against our expectations, this approach did not lead to satisfactory results in the sense that the network would learn to classify adversarial samples as attacks. In fact, classification accuracy stayed about constant during the training of the network.

%A possible explanation for this observation might be that the adversarial samples assumed too much the shape of normal flows, so that the network could not reliable tell them apart.

%\begin{table*}
%\caption{Comparing performance with all features, only safe features (\gls{iat} and packet size removed) and safe features in one direction (\gls{iat} and packet size removed for packets coming from source). \textit{packets} means that we consider the neural network output of all packets of the flow while \textit{flows} means that we only consider the last packet. At the last packet the classifier is usually already a lot more confident than at the first packets of a flow and thus metrics are always better in this case. \todo{Max: Is this that interesting? Remove tables and inline the most interesting results?}}  \label{tab:performance_results_no_manipulable}
%\newcommand{\cmidrulespace}{6pt}
%\centering
%\begin{tabular}{l l l l l l l} \toprule
%& \multicolumn{2}{l}{all} & \multicolumn{2}{l}{safe features} & \multicolumn{2}{l}{safe features one direction} \\
%\cmidrule(r){2-3} \cmidrule(lr){4-5} \cmidrule(l){6-7}
%& packets & flows & packets & flows & packets & flows \\
%\midrule
%Accuracy & 0.9857054 & 0.99472587 & 0.96678023 & 0.99337596 & 0.98056073 & 0.99353386 \\
%Precision & 0.94963342 & 0.9894422 & 0.87175275 & 0.98798596 & 0.91359787 & 0.98994595 \\
%Recall & 0.96735895 & 0.98972051 & 0.94374958 & 0.98581328 & 0.97836672 & 0.9844478 \\
%F1 & 0.95841423 & 0.98958134 & 0.90632359 & 0.98689842 & 0.94487366 & 0.98718922 \\
%Youden & 0.95682951 & 0.98614231 & 0.91525628 & 0.98175163 & 0.95937772 & 0.9810602 \\
%\bottomrule
%\end{tabular}
%\end{table*}

\section{Conclusion}

We have implemented a recurrent classifier based on LSTMs to detect network attacks, %While attack detection performance is not necessarily better than that of non-recurrent approaches,
which can already detect attacks before they are over. Furthermore, the recurrent approach allows us to inspect the influence of single packets on the detection performance and shows which packets are \textit{characteristic} for attacks.
Even though the interpretation of \glspl{rnn} poses several difficulties, we have demonstrated methods for gaining insights into the model's functioning.

We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified. We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.
Deploying an adversarial training procedure, we could significantly reduce the adversarial threat.

\section*{Acknowledgements}
The Titan Xp used for this research was donated by the NVIDIA Corporation.

\begin{figure}
\centering
\includegraphics[width=0.98\columnwidth]{../plots/ars/ars_15.pdf}
\caption{\gls{ars} during adversarial training for UNSW-NB15.}
\label{fig:avdistance}
\end{figure}

\renewcommand*{\bibfont}{\small}
\bibliographystyle{ieeetr}
\bibliography{bibliography}

\end{document}
\endinput
%%
%% End of file `sample-sigconf.tex'.

### 2019-12-17 paper/paper.tex
Fit to 8 pages
Add url

@@
 we can reduce the attack surface and harden the resulting \gls{ids} while retaining similar classification performance. % like for the original \gls{ml} model.
+
+For reproducibility, we make our code, data and figures publicly available at \url{https://github.com/CN-TU/adversarial-recurrent-ids}.
 
@@
 Then we take samples by ascending distance until we took 50\% of all samples.
-Of these samples, the distance of each adversarial sample to the original sample is taken and averaged.
-The resulting average distance is the \gls{ars}.
+Of these samples, the distance of each adversarial sample to the original sample is taken and averaged, resulting in the \gls{ars}.
 The larger it is the better since then an adversary needs to modify the samples more to make them adversarial.

### 2019-12-17 paper/paper.tex
Minor text changes

@@
 \item the number of features is significantly smaller and
-\item only certain features can be manipulated by an adversary (for example the destination port cannot be manipulated)
-\end{enumerate*}.
+\item only certain features can be manipulated by an adversary. % (for example the destination port cannot be manipulated)
+\end{enumerate*}
 
@@
 This recently stirred up increased interest in explainability methods.
-Also in this case, common methods are designed to work with images or tabular data but not with sequences.
+Also in this case, common methods are designed to work with images or tabular data, but not with sequences.
 Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data.
@@
 Astonishingly, feature importance methods reveal that
-features manipulable for adversarial flows,
+features, manipulated for successful adversarial flows,
 %the that are manipulable features, that are left for modification by adversaries
@@
 
-Note, however, that for successful adversarial samples, the second term in \autoref{eq:carliniWagner} becomes constant and, hence, is no longer relevant for optimization.
+%Note, however, that f
+For successful adversarial samples, the second term in \autoref{eq:carliniWagner} becomes constant and, hence, is no longer relevant for optimization.
 Hence, $c$ is irrelevant for the achieved distance, but governs if for one sample an adversarial sample is found.
-However, to avoid overly large steps when using a large $c$ in \autoref{eq:carliniWagner}, it is necessary to reduce the gradient descent learning rate.
+However, when using a large $c$ in \autoref{eq:carliniWagner}, to avoid overly large steps that impede convergence, it is necessary to reduce the gradient descent learning rate.
 Since the distance term is not affected by $c$, it is furthermore necessary to scale the iteration count inversely with the learning rate.
@@
 However, due to the complex decision processes of deep neural networks,
-this does not hold true and a feature might influence a decision many time steps later.
+this does not always hold true and a feature might influence a decision many time steps later.
 \end{itemize}
@@
 \item Leaving out all manipulable features: packet size and \gls{iat}
-\item Leaving out manipulable features when the attacker can manipulate them: only in the direction from the attacker to the victim (This, however does not prevent adversarial samples for botnets, for which both sides are malicious)
+\item Leaving out manipulable features only when the attacker can manipulate them, i.e.
+only in the direction from the attacker to the victim.
+This, however, does not prevent adversarial samples for botnets, for which both sides are malicious.
 \end{itemize}
@@
 The results show that -- surprisingly -- there is only a small difference in classification performance when only computing the accuracy for the last packet of each flow (flow accuracy): Flow accuracy stays over 99.93\%.
-However, packet accuracy is only 98\% when leaving out the features in one direction and 96.7\% when leaving them out in both directions, Thus, apparently the \gls{iat} and packet size are especially important for determining whether a flow is malicious in the first packets of a flow.
+However, packet accuracy is only 98\% when leaving out the features in one direction and 96.7\% when leaving them out in both directions.
+Thus, apparently the \gls{iat} and packet size are especially important for determining whether a flow is malicious in the first packets of a flow.
 

### 2019-12-17 paper/paper.tex
Fix typo in ARS caption
Update link to GitHub

@@
 
-For reproducibility, we make our code, data and figures publicly available at \url{https://github.com/CN-TU/adversarial-recurrent-ids}.
+We make all the source code, data and figures freely available to enable reproducibility and encourage experimentation at \url{https://github.com/CN-TU/adversarial-recurrent-ids}.
 
@@
 \includegraphics[width=\columnwidth]{../plots/ars_original.pdf}
-\caption{Recall for each attack type and \gls{ars} for CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and that \gls{ars} is $\infty$.}
+\caption{Recall for each attack type and \gls{ars} for CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and the \gls{ars} is $\infty$.}
 \label{fig:recall_ars}

### 2019-12-18 paper/paper.tex
Change title. Hopefully improve ARS definition. Only define ARS once and not twice. Make it more clear that only IAT and packet length are manipulated

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{An in-depth study of Explainability and Adversarial Robustness for RNNs}
+\title{Explainability and Adversarial Robustness for RNNs}
 
@@
 A network packet contains significantly less features than, e.g., an image and, furthermore, most features such as \gls{tcp} flags cannot be easily manipulated, as their manipulation might violate the protocol specifications and thus cause communication to fail.
-We identify the packet length and the \gls{iat} as features which are most likely to be exploited, but even these features cannot be manipulated freely.
+We identify the packet length and the \gls{iat} as features which are most likely to be exploited and thus choose them to be modified by adversary.
+But even these features cannot be manipulated freely.
 Beside the non-negativity constraint,  there are also more problem-specific constraints:
@@
 Such a metric should be easy to compute, easy to interpret and robust with respect to outliers.
-We thus propose the \gls{ars} as the average distance of the adversarial samples to the original samples for the 50\% of the samples which an adversary could successfully modified and which have the smallest distances.
+We consider an adversary successful if they manage more than 50\% of all samples to be misclassified.
+In this case, the \gls{ars} is the average of the distances of the adversarial samples to the original samples for the 50\% of all samples which have the smallest distances.
+If the adversary doesn't manage to manipulate more than 50\% then the \gls{ars} is $\infty$.
 Since \gls{cw} is able to find adversarial samples with minimum distance, it can be used for finding the \gls{ars} efficiently.
@@
  the smallest distance of non-adversarial samples is not smaller than the $\lceil\frac{N}{2}\rceil$th smallest distance of adversarial samples,
-stop. Otherwise increase $c$ and repeat. If after a predefined number of iterations -- e.g. 100 -- still not more than half of the samples are adversarial, also break.
+stop and compute the \gls{ars}.
+Otherwise increase $c$ and repeat.
+If after a predefined number of iterations -- e.g.
+100 -- still not more than half of the samples are adversarial, also break and set the \gls{ars} to $\infty$.
 
-The result is a number of adversarial samples and some that could not be made adversarial.
-If at least half of the samples are misclassified, we sort the adversarial samples by ascending distance to the original samples.
-Then we take samples by ascending distance until we took 50\% of all samples.
-Of these samples, the distance of each adversarial sample to the original sample is taken and averaged, resulting in the \gls{ars}.
-The larger it is the better since then an adversary needs to modify the samples more to make them adversarial.
+The larger the \gls{ars} is the better since then an adversary needs to modify the samples more to make them adversarial.
 If not more than half of the samples could be made adversarial, our metric is $\infty$ since then apparently it is not possible for an adversary to reliably conduct adversarial attacks on the majority of samples.

### 2019-12-18 learn.py
Added distance check to ars

@@ -1156,7 +1156,6 @@ def adv_until_less_than_half():
 	MAX = 10
 	SCALING = 0.5
 	THRESHOLD = 0.5
-	# FIXME: Max: I guess it should be < THRESHOLD not <= THRESHOLD as it is currently in the code
 	i = 0
 	prev_results = []
 	prev_flows = []
@@ -1167,19 +1166,47 @@ def adv_until_less_than_half():
 	tradeoff = 0
 	lr = opt.lr
 	iterations = 1
+
 	while True:
 		results_dict = list(adv_internal(False, tradeoff, lr, iterations, next_filter))[0]
 		modified_flows_by_attack, modified_results_by_attack, original_flows_by_attack, original_results_by_attack = results_dict["modified_flows_by_attack_number"], results_dict["results_by_attack_number"], results_dict["orig_flows_by_attack_number"], results_dict["orig_results_by_attack_number"]
 		ratio_modified_by_attack_number = np.array([np.mean(np.round(numpy_sigmoid(np.array([item[-1] for item in modified_results])))) if len(modified_results)>0 else -float("inf") for modified_results in modified_results_by_attack])
-		next_filter = [index for index, item in enumerate(ratio_modified_by_attack_number) if item <= THRESHOLD or (len(prev_ratios) > 0 and prev_ratios[-1][index] <= THRESHOLD)]
 		if i==0:
 			orig_results = original_results_by_attack
 			orig_flows = original_flows_by_attack
+			distances_packets = [None]*len(orig_results)
+			min_non_adv_distance = [None]*len(orig_results)
+			max_distance_packets = [None]*len(orig_results)
+			distances_flows = [None]*len(orig_results)
+			max_distance_flows = [None]*len(orig_results)
+			final_ratios = [None]*len(orig_results)
 		prev_results.append(modified_results_by_attack)
 		prev_flows.append(modified_flows_by_attack)
 		prev_ratios.append(ratio_modified_by_attack_number)
+		
+		for attack_index, ratio in enumerate(prev_ratios[i]):
+			if ratio > -float("inf"):
+				print(f"Looking at attack {attack_index} with a ratio of {ratio}")
+				final_ratios[attack_index] = ratio
+
+				successfully_changed_flows_mask = (np.round(numpy_sigmoid(np.array([item[-1] for item in prev_results[i][attack_index]]))) == 0).flatten()
+				# guess it makes sense to use the same distance metric as in training
[…]

### 2019-12-19 paper/paper.tex
Modified ARS section

@@
 High accuracy for a particular attack category does not imply that adversarial attacks are  likely to fail for these attacks.
-Thus we argue that beside the classical metrics such as accuracy, false positives etc., a new metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
-Such a metric should be easy to compute, easy to interpret and robust with respect to outliers.
+Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
+
+Such a metric should be easy to compute and easy to interpret.
+While in general, adversarial robustness for \gls{ml} models is frequently quantified as the average or median of minimum distances of adversarial samples \cite{carlini_evaluating_2019}, we found no generally agreed-upon metric for \glspl{ids} in the literature.
+If we consider flows, for which no adversarial sample can be found, to have infinite distance, the median has the advantage of ignoring such unsuccessful samples and outliers.
+On the other hand, the median might depict expected distances badly if they have a very uneven distribution.
+We propose the \gls{ars} as 
+\begin{equation}
+% \text{ARS} = \text{mean}(\{d \in \mathcal D \mid d < Q_{0.5}(\mathcal D)\}),
+\text{ARS} = \frac{1}{\lfloor N/2 \rfloor} \sum\nolimits_{d \in \{d \in \mathcal D \mid d < Q_{0.5}(\mathcal D)\}} d,
+\end{equation}
+denoting by $Q_{0.5}(\cdot)$ the median, by $N = | \mathcal D | \in \mathbb N$ the total number of samples and by $\mathcal D$ the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$.
+The \gls{ars} is thus the average of distances smaller than the median distance.
+
+%Such a metric should be easy to compute, easy to interpret and robust with respect to outliers.
 We consider an adversary successful if they manage more than 50\% of all samples to be misclassified.
@@
 If the adversary doesn't manage to manipulate more than 50\% then the \gls{ars} is $\infty$.
-Since \gls{cw} is able to find adversarial samples with minimum distance, it can be used for finding the \gls{ars} efficiently.
 
-Since \gls{cw} becomes slow for large $c$, the \gls{ars} can be efficiently found as follows: Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type.
-If, of a total of $N$ samples,
+\gls{cw} is able to find adversarial samples with minimum distance, but becomes slow for large $c$. Hence, it can be used for finding the \gls{ars} efficiently as follows: 
+%Since \gls{cw} becomes slow for large $c$, the \gls{ars} can be efficiently found as follows:
+Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type. If %, of a total of $N$ samples,
  more than $N/2$  are wrongly classified, and

### 2019-12-19 paper/paper.tex
Added Tanja's changes

@@
 In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content.
-More recently, with the rise of omnipresent encryption, available features have changed.
-Since payload is no longer available, the focus now lies on features that are always available to network monitoring equipment such as port numbers, protocol flags or packet sizes.
+More recently, with the rise of omnipresent encryption, %available features have changed.
+Since payload is no longer available,
+the focus now lies on features that are always available to network monitoring equipment like port numbers, protocol flags or packet sizes when encrypting above the transport layer.
 
-Network communication is usually grouped into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties, between two applications potentially being located at different physical locations.
+Network communication is usually grouped into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties.
+%, between two applications potentially being located at different physical locations.
 When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets.
@@
 Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
-While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards an anomaly can be detected.
+While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows.
 Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
-At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature selection procedures while providing any available information to the classifier.
+At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
 We show that our system has similar performance to other flow-based anomaly detection systems but can commonly detect anomalies \textit{before} the flow is over.
 
-However, for practical use, high accuracy is not enough.
+However, for practical use, high detection accuracy is not enough.
 With the recent rise of interest in \gls{aml} techniques, also \gls{aml} for \glspl{ids} has been investigated.
@@
 This is not a trivially answerable question since adversarial samples have mostly been analyzed in the context of computer vision.
-Our scenario significantly deviates from this in several ways because \begin{enumerate*}
+Our scenario significantly deviates from computer vision because \begin{enumerate*}
 \item the number of features is significantly smaller and
@@
 Surprisingly, we can confirm that adversarial samples can be found even when considering these tight constraints. % mentioned before.
-Thus, we argue that traditional methods like, e.g., accuracy are not sufficient for evaluating security-sensitive \gls{ml} systems.
-We propose the \acrfull{ars} as a new performance score, which precisely captures the notion of how easily an adversary can generate adversarial samples.
+Thus, we argue that traditional performance metrics like, e.g., accuracy are not sufficient for evaluating security-sensitive \gls{ml} systems.
+We propose the \acrfull{ars} as a new performance score, which captures the notion of how easily an adversary can generate adversarial samples.
 
@@
 Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
-During flow extraction we used a usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
+During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
 We use Z-score normalization and a train/test split of 2:1.
@@
 Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
-\autoref{tab:performance_results} shows that our classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}.
-However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before they are over.
+\autoref{tab:performance_results} shows that our \gls{rnn}-based classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}.
+However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before the attack flows are over.
 
@@
 A network packet contains significantly less features than, e.g., an image and, furthermore, most features such as \gls{tcp} flags cannot be easily manipulated, as their manipulation might violate the protocol specifications and thus cause communication to fail.
-We identify the packet length and the \gls{iat} as features which are most likely to be exploited and thus choose them to be modified by adversary.
+We identify the packet length and the \gls{iat} as features which are most likely to be exploited and thus choose them to be modified by the adversary.
 But even these features cannot be manipulated freely.
@@
 \item \glspl{iat} must not decrease, as otherwise the physical speed of data transmission can be violated in some cases.
-An in-depth analysis of when the reduction of \glspl{iat} is possible, would come with substantial effort, so we generally disallowed it.
+An in-depth analysis of cases where the reduction of \glspl{iat} might be possible, would come with substantial effort, so we generally disallowed \gls{iat} reductions.
 \end{itemize}
@@
 \cite{goodfellow_explaining_2015} develops the \gls{fgsm}, which makes easy generation of adversarial samples possible.
-\cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks import \gls{aml} methods, which we consider in this paper.
+\cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks important \gls{aml} methods, which we consider in this paper.
 
@@
 \begin{equation} \label{eq:carliniWagner}
-d(X,\tilde X) + c  \max(Z(\tilde X), \delta).
+d(X,\tilde X) + \kappa  \max(Z(\tilde X), \delta).
 \end{equation}
-Here, $d(\cdot)$ is a distance metric and $c \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow.
+Here, $d(\cdot)$ is a distance metric and $\kappa \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow.
 Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow under optimization.
@@
 We consider this method an interesting combination of \gls{cw} and \gls{fgsm}.
-Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be of similar quality.
+Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be similarly close the the original samples.
 
@@
 \begin{equation}
-\tilde X = X + \epsilon \text{sgn}( \nabla_X L(X)),
+\tilde X = X + \epsilon \, \text{sgn}( \nabla_X L(X)),
 \end{equation}
@@
 
-\autoref{fig:adv_cw} depicts the success ratio and average distance for \gls{cw} for different tradeoffs $c$.
+\autoref{fig:adv_cw} depicts the success ratio and average distance for \gls{cw} for different tradeoffs $\kappa$.
 %\autoref{fig:adv_cw} depicts the ratio of attack flows which were successfully misclassified as benign with respect to the final packet along with the resulting distance from the original sample.
-Evidently, the larger $c$, the better the attack works, but also distances from original flows become higher.
+Evidently, the larger $\kappa$, the better the attack works, but also distances from original flows become higher.
 With acceptable distances, we were able to generate adversarial samples for about 80\% of attack samples.
@@
 For successful adversarial samples, the second term in \autoref{eq:carliniWagner} becomes constant and, hence, is no longer relevant for optimization.
-Hence, $c$ is irrelevant for the achieved distance, but governs if for one sample an adversarial sample is found.
-However, when using a large $c$ in \autoref{eq:carliniWagner}, to avoid overly large steps that impede convergence, it is necessary to reduce the gradient descent learning rate.
-Since the distance term is not affected by $c$, it is furthermore necessary to scale the iteration count inversely with the learning rate.
-\gls{cw} with a large $c$ therefore needs more time than with a small $c$.
+Hence, $\kappa$ is irrelevant for the achieved distance, but governs if for one sample an adversarial sample is found.
+However, when using a large $\kappa$ in \autoref{eq:carliniWagner}, to avoid overly large steps that impede convergence, it is necessary to reduce the gradient descent learning rate.
+Since the distance term is not affected by $\kappa$, it is furthermore necessary to scale the iteration count inversely with the learning rate.
+\gls{cw} with a large $\kappa$ therefore needs more time than with a small $\kappa$.
 
@@
 %, when performing optimization as depicted by the objective function .
-% Hence, the depicted distance increase presumably is due to samples for which no adversarial sample could be found for lower $c$ values.
+% Hence, the depicted distance increase presumably is due to samples for which no adversarial sample could be found for lower $\kappa$ values.
 
@@
 % \includegraphics[width=0.49\textwidth]{adv_plots/cwl2.pdf}
-\caption{Performance of \gls{cw} with different $c$ values.}
+\caption{Performance of \gls{cw} with different $\kappa$ values.}
 \label{fig:adv_cw}
@@
 
-\gls{cw} is able to find adversarial samples with minimum distance, but becomes slow for large $c$. Hence, it can be used for finding the \gls{ars} efficiently as follows: 
-%Since \gls{cw} becomes slow for large $c$, the \gls{ars} can be efficiently found as follows:
+\gls{cw} is able to find adversarial samples with minimum distance, but becomes slow for large $\kappa$. Hence, it can be used for finding the \gls{ars} efficiently as follows: 
+%Since \gls{cw} becomes slow for large $\kappa$, the \gls{ars} can be efficiently found as follows:
 Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type. If %, of a total of $N$ samples,
@@
 stop and compute the \gls{ars}.
-Otherwise increase $c$ and repeat.
+Otherwise increase $\kappa$ and repeat.
 If after a predefined number of iterations -- e.g.
@@
 
-The larger the \gls{ars} is the better since then an adversary needs to modify the samples more to make them adversarial.
+The larger the \gls{ars} is, the more robust is the model, as then an adversary needs to modify the samples more to make them adversarial.
 If not more than half of the samples could be made adversarial, our metric is $\infty$ since then apparently it is not possible for an adversary to reliably conduct adversarial attacks on the majority of samples.
-Then the ratio of samples that could be made adversarial ($<$50\%) can be a useful metric to determine the exact extent of the adversarial threat.
+In this case, the ratio of samples that could be made adversarial can be a useful metric to determine the exact extent of the adversarial threat.
+
 Setting the threshold for the \gls{ars} to 50\% is arbitrary, but
@@
 The length of different flows might differ tremendously.
-Hence, features which are important for the network's outcome for one flow might not even exist for other flows.
+Hence, features at one particular time step might be important for the network's outcome for one flow, but not even exist for other flows.
 Furthermore, the question arises how the distance between flows of different length can be computed.
@@
 To inspect attack types in detail, in this research we use a \textit{conditional} form of the \gls{pdp}.
-If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space, $f(\boldsymbol X) \in [0,1]$ the neural network's prediction and $c$ the attack class, we define the conditional \gls{pdp} for feature $i$  as
+If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space, $f(\boldsymbol X) \in [0,1]$ the neural network's prediction and $\kappa$ the attack class, we define the conditional \gls{pdp} for feature $i$  as
 % \begin{equation} \label{eq:pdp}
@@
 
-Due to the use of a 5-tuple flow key, port numbers and the protocol identifier are constant for all packets in one flow, which allows us to use \glspl{pdp}, which have been proposed for non-recurrent classification methods.
+Due to the use of a 5-tuple flow key, port numbers and the protocol identifier are constant for all packets in one flow.
+Hence, we can consider the \gls{rnn} a regular classifier and reuse 
+%which allows us to use
+\glspl{pdp}, which have been proposed for non-recurrent classification methods, by plotting the \gls{rnn}'s flow prediction outcome over one of these  features.
 The results (omitted for brevity) show that for some attack classes the port numbers play an important role. When looking at the class of benign traffic %(non-attack traffic)
@@
 \end{figure}
-To evaluate this hypothesis, we plot the classifier's prediction confidence for each time step. % for each attack class.
+To evaluate this hypothesis, \autoref{fig:confidence} depicts the classifier's prediction confidence for each time step, along with the number of samples having at least this length, which were used for evaluating the figure.
+% for each attack class.
 %averaging over all flows. %AH. there's no averaging happening
 % and also show the first and third quartiles.
-\autoref{fig:confidence} shows that %this hypothesis is correct for most attack classes.
-while at the first couple of packets the confidence is not very high, towards the end of the flow it reaches values close to 1 and stays there.
+%\autoref{fig:confidence} shows that %this hypothesis is correct for most attack classes.
+While at the first couple of packets the confidence is not very high, towards the end of the flow it reaches values close to 1 and stays there.
 Hence, not only is the classifier able to make a reasonable classification after just a few packets, the figure also suggests that indeed later packets have a neglectable influence on the prediction.

### 2019-12-19 auxplot.py
Changed Fig.5 size

@@ -33,7 +33,7 @@ values = np.array(values, dtype=float)
 
 def importance():
 	global values
-	plt.figure(figsize=(5,4))
+	plt.figure(figsize=(5,3))
 	values[values < 0] = 0
 	values /= np.sum(values,axis=0)
 

### 2019-12-19 paper/paper.tex
Update ARS formula
Small changes to ARS section

@@
 Such a metric should be easy to compute and easy to interpret.
-While in general, adversarial robustness for \gls{ml} models is frequently quantified as the average or median of minimum distances of adversarial samples \cite{carlini_evaluating_2019}, we found no generally agreed-upon metric for \glspl{ids} in the literature.
+While in general adversarial robustness for \gls{ml} models is frequently quantified as the average or median of minimum distances of adversarial samples \cite{carlini_evaluating_2019}, we found no generally agreed-upon metric for \glspl{ids} in the literature.
 If we consider flows, for which no adversarial sample can be found, to have infinite distance, the median has the advantage of ignoring such unsuccessful samples and outliers.
@@
 % \text{ARS} = \text{mean}(\{d \in \mathcal D \mid d < Q_{0.5}(\mathcal D)\}),
-\text{ARS} = \frac{1}{\lfloor N/2 \rfloor} \sum\nolimits_{d \in \{d \in \mathcal D \mid d < Q_{0.5}(\mathcal D)\}} d,
+\text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \{d \in \mathcal D \mid d \leq Q_{0.5}(\mathcal D)\}} d,
 \end{equation}
 denoting by $Q_{0.5}(\cdot)$ the median, by $N = | \mathcal D | \in \mathbb N$ the total number of samples and by $\mathcal D$ the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$.
-The \gls{ars} is thus the average of distances smaller than the median distance.
+The \gls{ars} is thus the average of distances smaller than or equal to the median distance.
 
 %Such a metric should be easy to compute, easy to interpret and robust with respect to outliers.
-We consider an adversary successful if they manage more than 50\% of all samples to be misclassified.
+We consider an adversary successful if they cause at least 50\% of all samples to be misclassified.
 In this case, the \gls{ars} is the average of the distances of the adversarial samples to the original samples for the 50\% of all samples which have the smallest distances.
-If the adversary doesn't manage to manipulate more than 50\% then the \gls{ars} is $\infty$.
+If the adversary doesn't manage to manipulate enough samples, the \gls{ars} is $\infty$.
 
@@
 Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type. If %, of a total of $N$ samples,
- more than $N/2$  are wrongly classified, and
+ at least $N/2$  are wrongly classified, and
  the smallest distance of non-adversarial samples is not smaller than the $\lceil\frac{N}{2}\rceil$th smallest distance of adversarial samples,
@@
 
-The larger the \gls{ars} is, the more robust is the model, as then an adversary needs to modify the samples more to make them adversarial.
+The larger the \gls{ars} is, the more robust the model is, as then an adversary needs to modify the samples more to make them adversarial.
 If not more than half of the samples could be made adversarial, our metric is $\infty$ since then apparently it is not possible for an adversary to reliably conduct adversarial attacks on the majority of samples.
@@
  % reasonable as this way the \gls{ars} approximates the median of the distances of all samples and, additionally, marks the distance where an adversarial attack is more likely to be successful than unsuccessful.
- reasonable as outliers with very large distances are ignored and if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.
+ reasonable as outliers with very large distances are ignored and because if an attacker can make the majority of samples adversarial, we argue that its attack is ``successful''.
 %we think that it is reasonable since if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.
@@
 
-\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly easy for an attacker to perform adversarial modifications to make them look benign.
+\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly vulnerable to adversarial modifications by an attacker to make them look benign.
 

### 2019-12-19 paper/paper.tex
Typos and text changes

@@
 \newacronym{iat}{IAT}{Interarrival time}
-\newacronym{tcp}{TCP}{Transmission control protocol}
 
@@
 At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-We show that our system has similar performance to other flow-based anomaly detection systems but can commonly detect anomalies \textit{before} the flow is over.
+We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow is over.
 
@@
 With the recent rise of interest in \gls{aml} techniques, also \gls{aml} for \glspl{ids} has been investigated.
-For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids}.
+For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids} and \cite{hashemi_towards_2019} investigate adversarial robustness of common flow- and packet-based \gls{ids}.
 In this research, we investigate whether adversarial samples can be found for our \gls{rnn}-based model, i.e.
@@
 Astonishingly, feature importance methods reveal that
-features, manipulated for successful adversarial flows,
+features, which are manipulated for successful adversarial flows,
 %the that are manipulable features, that are left for modification by adversaries
@@
 %develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
-On the one hand, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Furthermore,
+First, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Second,
 by deploying an adversarial training procedure,
@@
 We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
-source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (if packet goes there or back) and all \gls{tcp} flags (0 if the flow is not \gls{tcp}).
+source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (if packet goes there or back) and all TCP flags (0 if the flow is not TCP).
 % \begin{itemize*}
@@
 % \item packet directions (if packet goes there or back)
-% \item all \gls{tcp} flags (0 if the flow is not \gls{tcp})
+% \item all TCP flags (0 if the flow is not TCP)
 % \end{itemize*}.
@@
 
-A network packet contains significantly less features than, e.g., an image and, furthermore, most features such as \gls{tcp} flags cannot be easily manipulated, as their manipulation might violate the protocol specifications and thus cause communication to fail.
+A network packet contains significantly less features than, e.g., an image and, furthermore, most features such as TCP flags cannot be easily manipulated, as their manipulation might violate the protocol specifications and thus cause communication to fail.
 We identify the packet length and the \gls{iat} as features which are most likely to be exploited and thus choose them to be modified by the adversary.
-But even these features cannot be manipulated freely.
-Beside the non-negativity constraint,  there are also more problem-specific constraints:
+But even these features cannot be manipulated due to problem-specific constraints:
 \begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
-\item Only packets can be manipulated which are transmitted by the attacker.
-Except for botnet and backdoor traffic, which is entirely controlled by an attacker, thus only packets travelling in one direction can be manipulated.
+\item Only packets can be manipulated which are transmitted by the attacker, except for botnet and backdoor traffic, which is entirely controlled by an attacker and thus only packets travelling in one direction can be manipulated.
  %This means the attacker can only control one direction (except for botnets where both sides are attackers).
-\item Packets must not be smaller than initially, as otherwise less information could be transmitted or packets might be rendered syntactically invalid.
+\item Packets must not be smaller than initially, as otherwise less information could be transmitted.
 \item \glspl{iat} must not decrease, as otherwise the physical speed of data transmission can be violated in some cases.
-An in-depth analysis of cases where the reduction of \glspl{iat} might be possible, would come with substantial effort, so we generally disallowed \gls{iat} reductions.
+An in-depth analysis of cases in which reduction of \glspl{iat} is legitimate is complex, so we generally disallowed \gls{iat} reductions.
 \end{itemize}
@@
 Finally, we also tested the \gls{fgsm} % which was initially proposed in
- \cite{goodfellow_explaining_2015}, which perhaps is the most frequently used method for generating adversarial samples.
+ \cite{goodfellow_explaining_2015}, which is the first method for generating adversarial samples.
 FGSM can be considered a single pass of \gls{pgd} on the loss function with an equality constraint on the $L_\infty$ distance, i.e.
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 \includegraphics[width=\columnwidth]{../plots/adv_comparison/adv_comparison_17.pdf}
@@
 
-\begin{figure}[b]
+\begin{figure}[h]
 \includegraphics[width=\columnwidth]{../plots/adv_comparison/tradeoff.pdf}
@@
 \textbf{Feature quantity.}
-The number of features that is fed into an \gls{rnn} is found as the product of packet features with the total length of the flow.
-For long flows, the total number of inputs can therefore grow to a substantial size.
+The number of features fed into an \gls{rnn} is number of packet features times the length of the flow. For long flows, the total number of inputs can become very large.
 
@@
 \textbf{Variable sequence lengths.}
-The length of different flows might differ tremendously.
-Hence, features at one particular time step might be important for the network's outcome for one flow, but not even exist for other flows.
-Furthermore, the question arises how the distance between flows of different length can be computed.
+The length of different flows might differ tremendously. Hence, features at one particular time step might be important for the network's outcome for one flow, but not even exist for other flows.
 
@@
 However, even if we restricted the analysis to flows of a constant length, a flow is different from the plain concatenation of its packet features.
-If we, for example, consider a classification problem as in this document, we expect that after a certain amount of time steps the classifier is able to perform the classification with a certain confidence.
-Arguably, features at the beginning of a flow are thus more important for the classifier.
-Giving feature differences at the end of a flow the same weight as differences at its beginning, therefore seems incorrect.
+For example, in a sentence, which is sequence of words, parts can be rearranged, giving a different sequence with possibly the exact same meaning.
 
@@
 When  applying explainability methods, the question arises which output to consider for the method.
-The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation.
-Not only does this approach require a substantially lower amount of computational resources compared to predictions occurring at a later time step, but we also expect the immediately occurring prediction outcome to exhibit a stronger dependence of the investigated feature, compared to a prediction outcome occurring many time steps later.
+The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation: This approach requires is less complex compared to considering also features of all earlier time steps.
+Also, we expect the current prediction outcome to more dependent on the current feature, compared to a feature from 100 steps ago.
 However, due to the complex decision processes of deep neural networks,
-this does not always hold true and a feature might influence a decision many time steps later.
+this is not always true and a feature might influence a decision many time steps later. 
 \end{itemize}
@@
 On the other hand, both adversarial machine learning and explainable machine learning bring up the question to what extent individual features are able to change the prediction outcome.
-While being seemingly similar at first sight, traditional feature importance can yield markedly wrong results for this case.
+While appearing similar, traditional feature importance can yield markedly wrong results for this case.
 To distinguish, we propose the term \emph{feature sensitivity} for such metrics.
-To analyze features for our network, we use the following approaches.
+To analyze features, we use the following approaches:
 
 \subsubsection{Neural Network Weights}
-In previous works~\cite{olden_accurate_2004}, a simple method for determining feature importance in neural networks, has been summing up neural network weights leading from a certain input to the prediction outcome.
+In previous works~\cite{olden_accurate_2004}, a simple method for determining feature importance in neural networks has been summing up neural network weights leading from a certain input to the prediction outcome.
 The weights method can be considered for both feature importance and feature sensitivity, however, clearly, especially in the case of complex network architectures, this method is likely to provide wrong results.
@@
 
- In general, the most commonly deployed feature importance techniques, used by practitioners for \glspl{rnn} \cite{stackexchange_cross_validated_neural_2019} and \gls{dl} \cite{molnar_interpretable_2019,stackexchange_cross_validated_feature_2016,olden_accurate_2004}, are based on adding noise to a particular feature and evaluating the drop in accuracy that occurs.
+The most commonly deployed feature importance techniques, used by practitioners for \glspl{rnn} \cite{stackexchange_cross_validated_neural_2019} and \gls{dl} \cite{molnar_interpretable_2019,stackexchange_cross_validated_feature_2016,olden_accurate_2004}, are based on adding noise to a particular feature and evaluating the drop in accuracy that occurs.
 We argue that it is hard to determine the ``correct'' intensity of noise to add.
@@
 
-As it can also happen that several features are left out for a sample, our method can also be used to analyze the effect of multiple features missing and can thus show possibly correlated features or -- more general -- features that contain common information, relevant for the classification task:
+For feature dropout, it can also happen that several features are left out for a sample and so our method can also be used to analyze the effect of multiple features missing and can thus show possibly correlated features or -- more general -- features that contain common information, relevant for the classification task:
 %Feature dropout also allows us to find correlated features or -- more general -- features that contain common information:
@@
 For example, the score between RST and the protocol identifier is 8.5; the highest of all pairs of features.
-While this may be unintuitive at the first glance, it likely stems from the fact that if the protocol identifier (\gls{tcp} or UDP) is missing, then
-%if RST is 1 at some point of a flow, it has to be \gls{tcp}.
-RST being 1 at some point indicates that the flow is \gls{tcp}.
+While this may be unintuitive at the first glance, it likely stems from the fact that if the protocol identifier (TCP or UDP) is missing, then
+%if RST is 1 at some point of a flow, it has to be TCP.
+RST being 1 at some point indicates that the flow is TCP.
 
@@
 \autoref{fig:feat_imp} shows the results, which match to a large extent with domain-specific expectations for classifying network flows.
-In particular, rarely used \gls{tcp} flags like NS or URG are unimportant for the classifier.
+In particular, rarely used TCP flags like NS or URG are unimportant for the classifier.
 On the other hand, destination port and protocol are essential for characterizing flows by hinting at the type of traffic.
@@
 To inspect attack types in detail, in this research we use a \textit{conditional} form of the \gls{pdp}.
-If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space, $f(\boldsymbol X) \in [0,1]$ the neural network's prediction and $\kappa$ the attack class, we define the conditional \gls{pdp} for feature $i$  as
+If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space, $f(\boldsymbol X) \in [0,1]$ the neural network's prediction and $c$ the attack class, we define the conditional \gls{pdp} for feature $i$  as
 % \begin{equation} \label{eq:pdp}
@@
 In other cases, however, we did not observe an agreement between \gls{pdp} and adversarial modifications, hinting at dependencies which cannot be presented by \glspl{pdp}.
+Since the \gls{iat} is undefined for the first packet, we always set it to 0.
+Still, interestingly, the plot shows, that the classifier considers packets with a higher IAT to be more likely to be attacks than those with a smaller one.
 
-\begin{figure}[b]
+\begin{figure}[h]
 %\includegraphics[width=0.48\textwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_0_Botnet-ARES"}.pdf}
@@
 
-Furthermore, it seems clear that the classifier did not just memorize the inputs but actually learned appropriate space representations, as there is always a smooth decrease of confidence when modifying feature values.
+%Furthermore, it seems clear that the classifier did not just memorize the inputs but actually learned appropriate space representations, as there is always a smooth decrease of confidence when modifying feature values.
 When memorizing, we would  instead expect a sharp decline of confidence.

### 2019-12-19 auxplot.py
Change fig to kappa

@@ -152,7 +152,7 @@ def adv():
 	plt.figure(figsize=(5,2))
 	x_values = [ float(value) for value in group_names ]
 	lines = plt.plot(x_values, values[:,:2])
-	plt.xlabel('Tradeoff $c$')
+	plt.xlabel('Tradeoff $\kappa$')
 	plt.ylabel('Success ratio')
 	plt.gca().set_ylabel_legend(Line2D([0],[0], color='gray'), handlelength=1.4)
 	plt.twinx()

### 2019-12-19 paper/paper.tex
Small text changes

@@
 
-Network communication is usually grouped into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties.
+Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties.
 %, between two applications potentially being located at different physical locations.
@@
 With this novel strategy
-we can reduce the attack surface and harden the resulting \gls{ids} while retaining similar classification performance. % like for the original \gls{ml} model.
+we can reduce the attack surface and harden the resulting \gls{ids} while retaining all features and similar classification performance. % like for the original \gls{ml} model.
 
@@
 Here, $d(\cdot)$ is a distance metric and $\kappa \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow.
-Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow under optimization.
+Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow optimized by \gls{cw}.
 
@@
 In the present context, we need to make sure that the classifier's prediction would actually be benign, even though a certain level of noise will be added to \glspl{iat} due to the network between attacker and victim.
-Hence, we introduced $\delta=-0.2$, corresponding to a prediction confidence of 55\% for the sample to be benign.
+Hence, we introduced $\delta=-0.2$, corresponding to a prediction confidence of 55\% for the sample to be benign after the sigmoid activation function.
 
-We used $L_1$ distance as distance metric, as we consider it to represent practically relevant differences of network flows best.
+We used $L_1$ as distance metric, as we consider $L_1$ distance to represent practically relevant differences of network flows best.
 We used \gls{pgd} for meeting the real-world constraints discussed above.
@@
 
-As expected, \gls{cw} delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
+%As expected,
+\gls{cw} delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
 %\autoref{fig:adv_per_family} depicts how successful the attack is for individual attack families.
@@
 
-\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly vulnerable to adversarial modifications by an attacker to make them look benign.
+\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly vulnerable to adversarial modifications by an attacker aiming to make them look benign.
 
@@
 \textbf{Feature quantity.}
-The number of features fed into an \gls{rnn} is number of packet features times the length of the flow. For long flows, the total number of inputs can become very large.
+The number of features fed into an \gls{rnn} is the number of packet features times the length of the flow. For long flows, the total number of inputs can become very large.
 

### 2019-12-19 paper/paper.tex
Small text changes

@@
 constructing \glspl{ids} for network data.
-With the rise of ubiquitous \gls{ml} systems, malicious actors have been catching up quickly to find new ways to exploit \gls{ml} for profit.
+With the rise of ubiquitous \gls{ml} systems, malicious actors have been catching up quickly to find new ways to exploit \gls{ml} vulnerabilities for profit.
 Recently developed adversarial \gls{ml} techniques focus on computer vision and their applicability to network traffic is not straightforward: Network packets expose fewer features than an image, are sequential and impose several constraints on their features.
@@
 
-We show that despite these completely different characteristics, adversarial samples can be reliably generated for \glspl{rnn}.
+We show that despite these completely different characteristics, adversarial samples can be  generated reliably for \glspl{rnn}.
 To understand a classifier's potential for misclassification, we extend existing explainability techniques and propose new ones, suitable particularly % specifically crafted %AH most can also be used for non-sequential models, so lets not restrict applicability too much
@@
 In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content.
-More recently, with the rise of omnipresent encryption, %available features have changed.
+More recently, with the increasing deployment of encryption, %available features have changed.
 Since payload is no longer available,
@@
 With the recent rise of interest in \gls{aml} techniques, also \gls{aml} for \glspl{ids} has been investigated.
-For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids} and \cite{hashemi_towards_2019} investigate adversarial robustness of common flow- and packet-based \gls{ids}.
+For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids} and \cite{hashemi_towards_2019} investigate adversarial robustness of common network \glspl{ids}.
 In this research, we investigate whether adversarial samples can be found for our \gls{rnn}-based model, i.e.
@@
 \item the number of features is significantly smaller and
-\item only certain features can be manipulated by an adversary. % (for example the destination port cannot be manipulated)
+\item only certain features can be manipulated if the flow should remain valid. % (for example the destination port cannot be manipulated)
 \end{enumerate*}
@@
 \cite{goodfellow_explaining_2015} develops the \gls{fgsm}, which makes easy generation of adversarial samples possible.
-\cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks important \gls{aml} methods, which we consider in this paper.
-
+\cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks important \gls{aml} methods.
+%, which we consider in this paper.
 % Max: But FGSM is high-speed... So I removed this paragraph altogether
 %As finding adversarial flows is particularly difficult in our case, % for the reasons we discussed above,
-%we preferred methods  producing high-quality adversarial samples while not being as fast as other methods. We implemented the following \gls{aml} methods.
+%we preferred methods  producing high-quality adversarial samples while not being as fast as other methods.
+We implemented the following \gls{aml} methods.
 

### 2019-12-19 paper/paper.tex
Fixed ARS definition

@@
 On the other hand, the median might depict expected distances badly if they have a very uneven distribution.
-We propose the \gls{ars} as 
+% We propose the \gls{ars} as 
+% \begin{equation}
+% \text{ARS} = \text{mean}(\{d \in \mathcal D \mid d < Q_{0.5}(\mathcal D)\}),
+% \text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \{d \in \mathcal D \mid d \leq Q_{0.5}(\mathcal D)\}} d,
+% \end{equation}
+% denoting by $Q_{0.5}(\cdot)$ the median, by $N = | \mathcal D | \in \mathbb N$ the total number of samples and by $\mathcal D$ the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$.
+
+We thus propose the \gls{ars} as follows.
+Let $\mathcal D$ denote the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$, $N = | \mathcal D | \in \mathbb N$ the total number of samples and $Q_{0.5}(\cdot)$ the median function.
+We define the \gls{ars} as 
 \begin{equation}
-% \text{ARS} = \text{mean}(\{d \in \mathcal D \mid d < Q_{0.5}(\mathcal D)\}),
-\text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \{d \in \mathcal D \mid d \leq Q_{0.5}(\mathcal D)\}} d,
+\text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \tilde {\mathcal D}} d,
 \end{equation}
-denoting by $Q_{0.5}(\cdot)$ the median, by $N = | \mathcal D | \in \mathbb N$ the total number of samples and by $\mathcal D$ the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$.
-The \gls{ars} is thus the average of distances smaller than or equal to the median distance.
+with $\tilde {\mathcal D} \subset \mathcal D$ denoting the set of cardinality $|\tilde {\mathcal D}| =\lceil N/2 \rceil$, so that for all $d \in  \tilde {\mathcal D}, d \le Q_{0.5}(\mathcal D)$.
+The \gls{ars} is thus approximately the average of distances smaller than or equal to the median distance.
 
@@
 
+\begin{figure*}
+\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
+\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2015.pdf}
+\caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
+\label{fig:feat_imp}
+\end{figure*}
 Furthermore, let $f(A,x)$  denote the joint probability
@@
 \subsubsection{Comparison}
-\begin{figure*}
-\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
-\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2015.pdf}
-\caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
-\label{fig:feat_imp}
-\end{figure*}
 \autoref{fig:feat_imp} shows the results, which match to a large extent with domain-specific expectations for classifying network flows.
@@
 % \end{equation}
-
 % To inspect attack types in detail, in this research we experiment with a modified definition of the \gls{pdp}  (with C being the attack classes):

### 2019-12-20 paper/paper.tex
Add some glue text

@@
 \section{Explaining Recurrent Neural Networks}
-From a naive perspective, one might be tempted to reuse explainability methods for \glspl{rnn} by considering a flow the sum of its packet features.
+Having verified the effectiveness of \gls{aml}  for our \glspl{rnn}, we now  investigate how the classifiers come to a decision.
+From a naive perspective, one might be tempted to reuse existing explainability methods for \glspl{rnn} by considering a flow the sum of its packet features.
 We identify several difficulties which occur when trying to explain decisions made by \glspl{rnn}.
@@
 \section{Defences}
-We now propose two approaches to increase robustness of the \gls{rnn} against adversarial attacks.
+We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
 \subsection{Reducing features}

### 2019-12-20 paper/paper.tex
Remove cardinality

--- 2019-12-19 paper/paper.tex
@@
 \end{equation}
-with $\tilde {\mathcal D} \subset \mathcal D$ denoting the set of cardinality $|\tilde {\mathcal D}| =\lceil N/2 \rceil$, so that for all $d \in  \tilde {\mathcal D}, d \le Q_{0.5}(\mathcal D)$.
+with $\tilde {\mathcal D} \subset \mathcal D$ denoting the set of size $|\tilde {\mathcal D}| =\lceil N/2 \rceil$, so that for all $d \in  \tilde {\mathcal D}, d \le Q_{0.5}(\mathcal D)$.
 The \gls{ars} is thus approximately the average of distances smaller than or equal to the median distance.

### 2019-12-20 paper/paper.tex
Fix 'attack type' inconsistencies

--- 2019-12-20 paper/paper.tex
@@
 \includegraphics[width=\columnwidth]{../plots/adv_comparison/adv_comparison_17.pdf}
-\caption{Attack success ratios for both datasets and per attack family for CIC-IDS-2017. Depicted are flow detection accuracies for adversarial flows and for unmodified attack flows.}
+\caption{Attack success ratios for both datasets and per attack type for CIC-IDS-2017. Depicted are flow detection accuracies for adversarial flows and for unmodified attack flows.}
 \label{fig:adv_per_family}
@@
 \subsection{Evaluation}
-\autoref{fig:adv_per_family} depicts the performance for the investigated algorithms for both datasets and for each attack family in CIC-IDS-2017.
+\autoref{fig:adv_per_family} depicts the performance for the investigated algorithms for both datasets and for each attack type in CIC-IDS-2017.
 We used \gls{cw} with a tradeoff of 1 and, to provide a fair comparison, for \gls{fgsm} and \gls{pgd} we used an average $L_\infty$ distance as observed for \gls{cw}.
@@
 %Such a metric should be easy to compute, easy to interpret and robust with respect to outliers.
-We consider an adversary successful if they cause at least 50\% of all samples to be misclassified.
+We consider an adversary successful if he can cause at least 50\% of all samples to be misclassified.
 In this case, the \gls{ars} is the average of the distances of the adversarial samples to the original samples for the 50\% of all samples which have the smallest distances.
@@
 
-A model trained with feature dropout typically yields slightly lower accuracy than a regularly trained model, even if no features are left out (flow accuracy of 99.43\% vs.99.65\%).
+A model trained with feature dropout typically yields slightly lower accuracy than a regularly trained model, even if no features are left out (flow accuracy of 99.43\% vs.
+99.65\%).
 We thus recommend training two models: One regular one and one with feature dropout to use for the feature importance.
@@
 To inspect attack types in detail, in this research we use a \textit{conditional} form of the \gls{pdp}.
-If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space, $f(\boldsymbol X) \in [0,1]$ the neural network's prediction and $c$ the attack class, we define the conditional \gls{pdp} for feature $i$  as
+If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space, $f(\boldsymbol X) \in [0,1]$ the neural network's prediction and $c$ the attack type, we define the conditional \gls{pdp} for feature $i$  as
 % \begin{equation} \label{eq:pdp}
@@
 \end{equation}
-empirically approximating the conditional distribution by the observed subset that belongs to a certain class.
+empirically approximating the conditional distribution by the observed subset that belongs to a certain attack type.
 
@@
 \glspl{pdp}, which have been proposed for non-recurrent classification methods, by plotting the \gls{rnn}'s flow prediction outcome over one of these  features.
-The results (omitted for brevity) show that for some attack classes the port numbers play an important role. When looking at the class of benign traffic %(non-attack traffic)
+The results (omitted for brevity) show that for some attack types the port numbers play an important role. When looking at the \gls{pdp} for benign traffic samples, %(non-attack traffic)
 it becomes apparent that traffic destined to a high destination port is generally indicative of an attack. We argue that this is because most services that regular users use have low port numbers.

### 2019-12-20 plot.py
Fixed 'time step' inconsistencies

@@ -144,7 +144,7 @@ for attack_type, seqs in enumerate(results_by_attack_number):
 	if SHOW_TITLE:
 		plt.title(reverse_mapping[attack_type])
 	# plt.legend(all_legends)
-	ax1.set_xlabel('Sequence index')
+	ax1.set_xlabel('Time step $t$')
 	plt.xlim((-0.5,max(x_values)+0.5))
 	plt.tight_layout()
 	# plt.xticks(range(medians.shape[0]))

### 2019-12-20 plot2_adv.py

@@ -161,7 +161,7 @@ for attack_type, (results_by_attack_number_item, flows_by_attack_number_item, re
 		print("feature_index", feature_index)
 		plt.subplot("{}{}{}".format(len(FEATURE_NAMES), 1, feature_index_from_zero+1))
 		if feature_index_from_zero == len(FEATURE_NAMES)-1:
-			plt.xlabel('Sequence index')
+			plt.xlabel('Time step $t$')
 		plt.ylabel(feature_name)
 
 		legend = "{}, {}".format(ORDERING[0], feature_name)

### 2019-12-20 plot.py
Remove fractional time steps from plot.py figure

@@ -141,6 +141,9 @@ for attack_type, seqs in enumerate(results_by_attack_number):
 	ax2.set_ylabel_legend(all_legends[0])
 	ax1.set_ylabel_legend(Rectangle((0,0), 1, 1, fc='gray', alpha=0.2), handlelength=0.7, handletextpad=0.4)
 
+	ticks = plt.xticks()
+	plt.xticks([ tick for tick in ticks[0][1:-1] if tick.is_integer() ])
+	
 	if SHOW_TITLE:
 		plt.title(reverse_mapping[attack_type])
 	# plt.legend(all_legends)

### 2019-12-20 paper/paper.tex
Tanja's changes

--- 2019-12-20 paper/paper.tex
@@
 
-On the other hand, both adversarial machine learning and explainable machine learning bring up the question to what extent individual features are able to change the prediction outcome.
+Conversely, both adversarial machine learning and explainable machine learning bring up the question to what extent individual features are able to change the prediction outcome.
 While appearing similar, traditional feature importance can yield markedly wrong results for this case.
@@
 Thus important features are going to be left out less often and one can then extract the dropout probabilities for each feature and assess their significance based on them.
+While this method looks seemingly related to feature dropout, it is significantly more complex and does not aim to show the accuracy drop that occurs when omitting a feature but instead returns a unique feature importance value between 0 and 1.
 %Feature dropout can also be used as a building block for computing the accuracy of datapoints with missing features that is necessary to compute Shapley values \cite{shapley_value_1953} or KernelSHAP \cite{lundberg_unified_2017}.
@@
 Less anticipated, however, also input perturbation does not exhibit a considerable correlation with feature dropout.
-Considering its functioning, feature dropout is the most reliable method for evaluating importance for removing features.
+Considering its functioning of completely removing individual features, feature dropout is the most reliable method for evaluating importance for removing features.
 It is remarkable that none of the other methods is able to depict the distinct peak of importance for the destination port.

### 2019-12-20 paper/paper.tex
State that change is optional

--- 2019-12-20 paper/paper.tex
@@
 Thus important features are going to be left out less often and one can then extract the dropout probabilities for each feature and assess their significance based on them.
+%
+% Max: I'd say that the following sentence is optional...
+While this method looks seemingly related to feature dropout, it is significantly more complex and does not aim to show the accuracy drop that occurs when omitting a feature but instead returns a unique feature importance value between 0 and 1.
 %Feature dropout can also be used as a building block for computing the accuracy of datapoints with missing features that is necessary to compute Shapley values \cite{shapley_value_1953} or KernelSHAP \cite{lundberg_unified_2017}.

### 2019-12-20 paper/paper.tex
Tanjas/Joachims suggestions

--- 2019-12-20 paper/paper.tex
@@
 At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow is over.
+We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates.
 
@@
 \begin{table}[b]
-\caption{Occurrence frequency of attack types.}
+\caption{Flow occurrence frequency of attack types.}
 \label{tab:occurrence}
@@
 We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
-source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (if packet goes there or back) and all TCP flags (0 if the flow is not TCP).
+source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e.
+forward or reverse path) and all TCP flags (0 if the flow is not TCP).
 % \begin{itemize*}
@@
 As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}.
-High accuracy for a particular attack category does not imply that adversarial attacks are  likely to fail for these attacks.
-Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
+High recall for a particular attack type does not imply that adversarial samples are hard to find  %are  likely to fail
+for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
 
@@
 \includegraphics[width=\columnwidth]{../plots/ars_original.pdf}
-\caption{Recall for each attack type and \gls{ars} for CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and the \gls{ars} is $\infty$.}
+\caption{Recall for unmodified flows and \gls{ars} for each attack type in CIC-IDS-2017.
+For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and the \gls{ars} is $\infty$.}
 \label{fig:recall_ars}
@@
 
-By using a classical \gls{pdp} we would be likely to lose information due to the averaging over very different types of traffic.
+By using a classical \gls{pdp} we would likely lose information due to the averaging over very different types of traffic.
 However, for network traffic in particular, investigating each sample individually is not possible.

### 2019-12-20 paper/paper.tex
Small changes

@@
 \includegraphics[width=\columnwidth]{../plots/ars_original.pdf}
-\caption{Recall for unmodified flows and \gls{ars} for each attack type in CIC-IDS-2017.
-For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and the \gls{ars} is $\infty$.}
+\caption{Recall for unmodified flows and \gls{ars} for attacks in CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls below 66\%: The attack does not succeed and the \gls{ars} is $\infty$.}
 \label{fig:recall_ars}
@@
 When  applying explainability methods, the question arises which output to consider for the method.
-The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation: This approach requires is less complex compared to considering also features of all earlier time steps.
-Also, we expect the current prediction outcome to more dependent on the current feature, compared to a feature from 100 steps ago.
+The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation: This approach is less complex compared to considering also features of all earlier time steps.
+Also, we expect the current prediction outcome to more dependent on the current feature, compared to a feature from many steps ago.
 However, due to the complex decision processes of deep neural networks,

### 2019-12-20 paper/paper.tex
Fix leaving out features section

--- 2019-12-20 paper/paper.tex
@@
 At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates.
+We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow is over.
 
@@
 \begin{table}[b]
-\caption{Flow occurrence frequency of attack types.}
+\caption{Occurrence frequency of attack types.}
 \label{tab:occurrence}
@@
 We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
-source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e.
-forward or reverse path) and all TCP flags (0 if the flow is not TCP).
+source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (if packet goes there or back) and all TCP flags (0 if the flow is not TCP).
 % \begin{itemize*}
@@
 As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}.
-High recall for a particular attack type does not imply that adversarial samples are hard to find  %are  likely to fail
-for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
+High accuracy for a particular attack category does not imply that adversarial attacks are  likely to fail for these attacks.
+Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
 
@@
 \includegraphics[width=\columnwidth]{../plots/ars_original.pdf}
-\caption{Recall for unmodified flows and \gls{ars} for each attack type in CIC-IDS-2017.
-For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and the \gls{ars} is $\infty$.}
+\caption{Recall for each attack type and \gls{ars} for CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls under 66\%: The adversary does not succeed and the \gls{ars} is $\infty$.}
 \label{fig:recall_ars}
@@
 
-By using a classical \gls{pdp} we would likely lose information due to the averaging over very different types of traffic.
+By using a classical \gls{pdp} we would be likely to lose information due to the averaging over very different types of traffic.
 However, for network traffic in particular, investigating each sample individually is not possible.
@@
 
-Both approaches lead to complete resistance to adversarial samples (as we left out all manipulable features), except for botnets, which can still operate when only leaving out manipulable features in one direction.
-The results show that -- surprisingly -- there is only a small difference in classification performance when only computing the accuracy for the last packet of each flow (flow accuracy): Flow accuracy stays over 99.93\%.
+Both approaches lead to complete resistance to adversarial samples, except for botnets, which can still operate when only leaving out manipulable features in one direction.
+The results show that -- surprisingly -- there is only a small difference in classification performance when looking at flow accuracy: It is slightly lower at 99.3\% compared to 99.7\% originally for CIC-IDS-2017.
 However, packet accuracy is only 98\% when leaving out the features in one direction and 96.7\% when leaving them out in both directions.

### 2019-12-20 paper/paper.tex
Tiny changes

--- 2019-12-20 paper/paper.tex
@@
 
-\begin{figure}[h]
+\begin{figure}[t]
 \includegraphics[width=\columnwidth]{../plots/adv_comparison/tradeoff.pdf}
@@
 We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
-\subsection{Reducing features}
+\subsection{Reducing Features}
 The most obvious defence, which is only possible because of the nature of the dataset, is to simply omitting features an attacker can manipulate. We try two different approaches:
@@
 
-\subsection{Adversarial training}
+\subsection{Adversarial Training}
 As alternative to omitting manipulable features, we attempted to make the classifier more robust against adversarial samples by augmenting the training set by adversarial flows generated using  \gls{cw}, labeled as additional attacks.

### 2019-12-20 paper/paper.tex
Fixed ARS definition again

@@
 We thus propose the \gls{ars} as follows.
-Let $\mathcal D$ denote the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$, $N = | \mathcal D | \in \mathbb N$ the total number of samples and $Q_{0.5}(\cdot)$ the median function.
+Let $\mathcal S$ denote the set of samples, $N = | \mathcal S | \in \mathbb N$ the total number of samples and $d_s \in \mathbb R_0^+$ the distance of an adversarial flow from the original flow for a sample $s\in \mathcal S$, assigning unsuccessful samples a distance of $\infty$.
+% and $Q_{0.5}(\cdot)$ the median function.
 We define the \gls{ars} as 
 \begin{equation}
-\text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \tilde {\mathcal D}} d,
+\text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{\tilde s \in \tilde {\mathcal S}} d_{\tilde s},
 \end{equation}
-with $\tilde {\mathcal D} \subset \mathcal D$ denoting the set of size $|\tilde {\mathcal D}| =\lceil N/2 \rceil$, so that for all $d \in  \tilde {\mathcal D}, d \le Q_{0.5}(\mathcal D)$.
-The \gls{ars} is thus approximately the average of distances smaller than or equal to the median distance.
+with $\tilde {\mathcal S} \subset \mathcal S$ denoting a set of size $|\tilde {\mathcal S}| =\lceil N/2 \rceil$, so that $d_{\tilde s} \le d_s$ for all $\tilde s \in  \tilde {\mathcal S}, s \in   {\mathcal S} \setminus \tilde {\mathcal S} $.
+% We thus propose the \gls{ars} as follows.
+Let $\mathcal D$ denote the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$, $N = | \mathcal D | \in \mathbb N$ the total number of samples and $Q_{0.5}(\cdot)$ the median function.
+% We define the \gls{ars} as 
+% \begin{equation}
+% \text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \tilde {\mathcal D}} d,
+% \end{equation}
+% with $\tilde {\mathcal D} \subset \mathcal D$ denoting the set of size $|\tilde {\mathcal D}| =\lceil N/2 \rceil$, so that for all $d \in  \tilde {\mathcal D}, d \le Q_{0.5}(\mathcal D)$.
+The \gls{ars} is thus approximately the average of distances not larger than the median distance.
 

### 2019-12-20 paper/paper.tex
Streamlined to US english

@@
 
-Based on the insights gained by the feature importance and explainability methods, we propose two defence methods.
+Based on the insights gained by the feature importance and explainability methods, we propose two defense methods.
 %develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
@@
 % \end{itemize*}.
-We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.
+We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behavior \cite{bachl_walling_2019}.
 Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
@@
 We consider this method an interesting combination of \gls{cw} and \gls{fgsm}.
-Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be similarly close the the original samples.
+Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be similarly close to the original samples.
 
@@
 
-\section{Defences}
+\section{Defenses}
 We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
 \subsection{Reducing Features}
-The most obvious defence, which is only possible because of the nature of the dataset, is to simply omitting features an attacker can manipulate. We try two different approaches:
+The most obvious defense, which is only possible because of the nature of the dataset, is to simply omitting features an attacker can manipulate. We try two different approaches:
 \begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]

### 2019-12-20 paper/paper.tex
Tiny changes

@@
 
-\section{An \gls{rnn}-based classifier}
+\section{An \gls{rnn}-based Classifier}
 
@@
 \autoref{tab:performance_results} shows that our \gls{rnn}-based classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}.
-However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before the attack flows are over.
+However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before the attack flows terminate.
 
@@
 
-\section{Adversarial attacks}
+\section{Adversarial Attacks}
 \label{sec:adv}

### 2019-12-20 plot.py
FLow number instead of sample number

@@ -109,7 +109,7 @@ for attack_type, seqs in enumerate(results_by_attack_number):
 	ax2 = ax1.twinx()
 
 	ax2.set_ylabel('Confidence')
-	ax1.set_ylabel("Sample number")
+	ax1.set_ylabel("Flow number")
 
 	ax1.yaxis.tick_right()
 	ax1.yaxis.set_label_position("right")

### 2019-12-20 paper/paper.tex
Add separate plots directory

@@
 \begin{figure}[h]
-\includegraphics[width=\columnwidth]{../plots/adv_comparison/adv_comparison_17.pdf}
+\includegraphics[width=\columnwidth]{./plots/adv_comparison/adv_comparison_17.pdf}
 \caption{Attack success ratios for both datasets and per attack type for CIC-IDS-2017. Depicted are flow detection accuracies for adversarial flows and for unmodified attack flows.}
@@
 \begin{figure}[t]
-\includegraphics[width=\columnwidth]{../plots/adv_comparison/tradeoff.pdf}
-% \includegraphics[width=0.49\textwidth]{adv_plots/cwl1.pdf}
-% \includegraphics[width=0.49\textwidth]{adv_plots/cwl2.pdf}
+\includegraphics[width=\columnwidth]{./plots/adv_comparison/tradeoff.pdf}
 \caption{Performance of \gls{cw} with different $\kappa$ values.}
@@
 \begin{figure}[h]
-\includegraphics[width=\columnwidth]{../plots/ars_original.pdf}
+\includegraphics[width=\columnwidth]{./plots/ars_original.pdf}
 \caption{Recall for unmodified flows and \gls{ars} for attacks in CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls below 66\%: The attack does not succeed and the \gls{ars} is $\infty$.}
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{../plots/mutinfo_example.pdf}
+\includegraphics[width=\columnwidth]{./plots/mutinfo_example.pdf}
 \caption{Two distributions yielding an identical accuracy drop.}
@@
 \begin{figure*}
-\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
-\includegraphics[width=.49\textwidth]{../plots/importance/feat_imp_flow_2015.pdf}
+\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
+\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2015.pdf}
 \caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
@@
 \begin{figure}
-%\includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_5_DoS - DDoS-DoS Hulk"}.pdf}
-\includegraphics[width=\columnwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_15_All samples"}.pdf}
-% \includegraphics[width=\columnwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_6_DoS - DDoS-DoS Slowhttptest"}.pdf}
-% \includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_full_no_ttl_pred_plots_outcomes_0_3.pickle_14_Web Attack-XSS"}.pdf}
-% \includegraphics[width=0.48\textwidth]{{"../plots/plot/flows_full_no_ttl_pred_plots_outcomes_0_3.pickle_11_PortScan-PortScan - Firewall off"}.pdf}
+\includegraphics[width=\columnwidth]{{"./plots/plot/flows_pred_plots2_outcomes_0_3.pickle_15_All_samples"}.pdf}
 \caption{Classifier confidence per time step for CIC-IDS-2017. % and first and third quartiles.
@@
 \begin{figure}[h]
-%\includegraphics[width=0.48\textwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_0_Botnet-ARES"}.pdf}
-\includegraphics[width=\columnwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_7_DoS - DDoS-DoS slowloris"}.pdf}
-%\includegraphics[width=0.48\textwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_11_PortScan-PortScan - Firewall off"}.pdf}
-%\includegraphics[width=0.48\textwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_1_Brute Force-FTP-Patator"}.pdf}
+
+\includegraphics[width=\columnwidth]{{"./plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_7_DoS_-_DDoS-DoS_slowloris"}.pdf}
 
@@
 %AH. I think the 'Firewall off' plot is not a good example as it shows only 4 packets
-\includegraphics[width=\columnwidth]{{"../plots/plot_features/flows_pred_plots2_outcomes_0_3.pickle_2_Brute Force-SSH-Patator"}.pdf}
-%\includegraphics[width=0.48\textwidth]{{"../plots/plot_features/flows_full_no_ttl_pred_plots2_outcomes_0_3.pickle_6_DoS - DDoS-DoS Slowhttptest"}.pdf}
-%\includegraphics[width=\columnwidth]{{"../plots/plot_features/flows_full_no_ttl_pred_plots2_outcomes_0_3.pickle_11_PortScan-PortScan - Firewall off"}.pdf}
-%\includegraphics[width=0.48\textwidth]{{"../plots/plot_features/flows_full_no_ttl_pred_plots2_outcomes_0_3.pickle_7_DoS - DDoS-DoS slowloris"}.pdf}
-
+\includegraphics[width=\columnwidth]{{"./plots/plot_features/flows_pred_plots2_outcomes_0_3.pickle_2_Brute_Force-SSH-Patator"}.pdf}
 \caption{\gls{iat} and packet length for SSH brute-force attacks in CIC-IDS-2017.}% The solid line shows the median value of the feature for all flows of a specific attack type.
@@
 \centering
-\includegraphics[width=0.98\columnwidth]{../plots/ars/ars_15.pdf}
+\includegraphics[width=0.98\columnwidth]{./plots/ars/ars_15.pdf}
 \caption{\gls{ars} during adversarial training for UNSW-NB15.}

### 2019-12-20 learn.py

@@ -200,6 +200,8 @@ def train():
 	fold = opt.fold
 	lstm_module.train()
 
+	if opt.rl:
+
 	train_indices, _ = get_nth_split(dataset, n_fold, fold)
 	train_data = torch.utils.data.Subset(dataset, train_indices)
 	if opt.advTraining:
@@ -221,20 +223,12 @@ def train():
 			writer.add_scalar('adv_avdistance', av_distance, i)
 
 		for input_data, labels, flow_categories in train_loader:
-			# print("iterating")
-			# samples += len(input_data)
 			optimizer.zero_grad()
 			batch_size = input_data.sorted_indices.shape[0]
 			assert batch_size <= opt.batchSize, "batch_size: {}, opt.batchSize: {}".format(batch_size, opt.batchSize)
 			lstm_module.init_hidden(batch_size)
 
-			# actual_input = torch.FloatTensor(input_tensor[:,:,:-1]).to(device)
-
-			# print("input_data.data.shape", input_data.data.shape)
 			output, seq_lens = lstm_module(input_data)
-
-			# torch.set_printoptions(profile="full")
-			# print("output", output.detach().squeeze().transpose(1,0))
 
 			samples += output.shape[1]
 
@@ -244,8 +238,6 @@ def train():
 
 			mask = (index_tensor <= selection_tensor).byte().to(device)
 			mask_exact = (index_tensor == selection_tensor).byte().to(device)
-			# torch.set_printoptions(profile="full")
-			# print("mask", mask.squeeze())
 			labels, _ = torch.nn.utils.rnn.pad_packed_sequence(labels)
 			flow_categories, _ = torch.nn.utils.rnn.pad_packed_sequence(flow_categories)
 
@@ -254,9 +246,6 @@ def train():
 
 			optimizer.step()
 
-			# print("masked_output_shape", torch.round(output.detach()[mask]).shape, "masked_labels_shape", labels[mask].shape)
-			# print("exact_masked_output_shape", torch.round(output.detach()[mask_exact]).shape, "exact_masked_labels_shape", labels[mask_exact].shape)
-			# print("output.shape", output.shape, "labels.shape", labels.shape)
 			assert output.shape == labels.shape
 			writer.add_scalar("loss", loss.item(), samples)
 			sigmoided_output = torch.sigmoid(output.detach())
@@ -264,11 +253,6 @@ def train():
 			writer.add_scalar("accuracy", accuracy, samples)
[…]

### 2020-01-19 learn.py
Implement another sampling strategy

@@ -207,6 +207,64 @@ def custom_collate(seqs, things=(True, True, True)):
 			current_mask = torch.zeros(current_seq_len, dtype=torch.float32)
 			current_mask[:int(allocated_per_seq_discrete[seq_index])] = 1
 			masks.append(current_mask)
+	elif opt.sampling=="first_n_equal":
+		total_length = sum([len(item) for item in seqs])
+		n_seqs = len(seqs)
+		chosen_packets = round(opt.samplingProbability*total_length)
+
+		len_of_each_seq = torch.tensor([len(seq) for seq in seqs], dtype=torch.float32)
+
+		sorted_by_length, sorted_by_length_indices = torch.sort(len_of_each_seq, descending=True)
+		# cumulative_lens_for_each_step
+		cumulative_packets = 0
+
+		# TODO:
+		last_full_length_packets = 0
+		final_length = 0
+		# print("int(max(len_of_each_seq))+1", int(max(len_of_each_seq))+1)
+		for current_len in range(1, int(max(len_of_each_seq))+1):
+			# print("beginning", current_len, "cumulative_packets", cumulative_packets, "chosen_packets", chosen_packets, "final_length", final_length)
+			if cumulative_packets >= chosen_packets:
+				# print("first break")
+				break
+			for seq_len in sorted_by_length:
+				# print("cumulative_packets incremented", cumulative_packets)
+
+				# print("seq_len", seq_len)
+				if seq_len < current_len:
+					# print("second break")
+					break
+				if cumulative_packets >= chosen_packets:
+					# print("third_break")
+					break
+				cumulative_packets += 1
+			if cumulative_packets < chosen_packets:
+				final_length = current_len
+				last_full_length_packets = cumulative_packets
+				# print("last_full_length_packets", last_full_length_packets)
+
+
+		assert last_full_length_packets <= chosen_packets
+		difference = chosen_packets - last_full_length_packets
+		# print("difference", difference)
+
+		positions_that_could_be_filled = sorted_by_length >= final_length+1
+		assert torch.sum(positions_that_could_be_filled) >= difference
+		indices_that_could_be_filled = torch.nonzero(positions_that_could_be_filled).squeeze()
+		chosen_indices = indices_that_could_be_filled[-difference:]
+
[…]

### 2020-01-19 learn.py
Small fix

@@ -210,7 +210,7 @@ def custom_collate(seqs, things=(True, True, True)):
 	elif opt.sampling=="first_n_equal":
 		total_length = sum([len(item) for item in seqs])
 		n_seqs = len(seqs)
-		chosen_packets = round(opt.samplingProbability*total_length)
+		chosen_packets = min(round(opt.samplingProbability*total_length), n_seqs)
 
 		len_of_each_seq = torch.tensor([len(seq) for seq in seqs], dtype=torch.float32)
 

### 2020-01-19 learn.py
Experiment with new sparsity penalty

@@ -669,6 +669,7 @@ def train_rl():
 						new_already_found_packets_per_sample[seq_index] += 1
 						already_skipped_packets_per_sample[seq_index] += torch.min(chosen_indices[step_index+1][seq_index] if step_index+1 < len(chosen_indices) else torch.tensor(float("inf")), orig_seq_lens[seq_index]) - chosen_indices[step_index][seq_index] - 1
 				rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
+				rewards_sparsity.append((already_skipped_packets_per_sample-overshoot)/(already_skipped_packets_per_sample+already_found_packets_per_sample))
 				overshoot = torch.FloatTensor([0] * batch_size).to(device)
 				already_found_packets_per_sample = new_already_found_packets_per_sample
 

### 2020-01-19 learn.py
Make some things more efficient

@@ -587,7 +587,7 @@ def train_rl():
 
 			# assert batch_size <= opt.batchSize
 
-			seq_index = torch.LongTensor([0] * batch_size).to(device)
+			seq_index = torch.zeros((batch_size), dtype=torch.int64).to(device)
 			chosen_indices = [seq_index]
 
 			lstm_module.init_hidden(batch_size)
@@ -652,11 +652,11 @@ def train_rl():
 			rewards_classification = []
 			rewards_sparsity = []
 
-			already_found_packets_per_sample = torch.FloatTensor([0] * batch_size).to(device)
-			new_already_found_packets_per_sample = torch.FloatTensor([0] * batch_size).to(device)
-			leave_out_last = torch.FloatTensor([0] * batch_size).to(device)
-			overshoot = torch.FloatTensor([0] * batch_size).to(device)
-			already_skipped_packets_per_sample = torch.FloatTensor([0] * batch_size).to(device)
+			already_found_packets_per_sample = torch.zeros((batch_size), dtype=torch.float32).to(device)
+			new_already_found_packets_per_sample = torch.zeros((batch_size), dtype=torch.float32).to(device)
+			leave_out_last = torch.zeros((batch_size), dtype=torch.float32).to(device)
+			overshoot = torch.zeros((batch_size), dtype=torch.float32).to(device)
+			already_skipped_packets_per_sample = torch.zeros((batch_size), dtype=torch.float32).to(device)
 
 			for step_index in reversed(range(len(chosen_indices)-1)):
 				for seq_index in range(len(chosen_indices[step_index])):
@@ -668,9 +668,9 @@ def train_rl():
 							overshoot[seq_index] = chosen_indices[step_index+1][seq_index] - orig_seq_lens[seq_index]
 						new_already_found_packets_per_sample[seq_index] += 1
 						already_skipped_packets_per_sample[seq_index] += torch.min(chosen_indices[step_index+1][seq_index] if step_index+1 < len(chosen_indices) else torch.tensor(float("inf")), orig_seq_lens[seq_index]) - chosen_indices[step_index][seq_index] - 1
-				rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
+				# rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
[…]

### 2020-01-20 learn.py
Fix test function for rl

@@ -311,6 +311,7 @@ def custom_collate(seqs, things=(True, True, True)):
 		input_data = seqs
 
 		orig_seq_lens = torch.LongTensor([len(item) for item in input_data]).to(device)
+		orig_seq_lens_minus_one = orig_seq_lens - 1
 		remaining_seq_lens = orig_seq_lens
 		batch_size = len(input_data)
 
@@ -323,18 +324,12 @@ def custom_collate(seqs, things=(True, True, True)):
 			lstm_module_rl_actor.eval()
 			lstm_module_rl_actor.init_hidden(batch_size)
 
-		# outputs_rl_actor = []
-		# outputs_rl_actor_decisions = []
+		padded_input_data = torch.nn.utils.rnn.pad_sequence(input_data).to(device)
+		batch_indices = torch.arange(batch_size)
 
 		while (remaining_seq_lens > torch.zeros_like(remaining_seq_lens)).any():
-			current_slice = [item[index:index+1] for index, item in zip(seq_index, input_data)]
-			# print("remaining_seq_lens", remaining_seq_lens, "len(current_slice)", len(current_slice))
-			# for index, item in enumerate(current_slice):
-			# 	assert item.shape[0] >= 0 and item.shape[1] > 0
-			# 	item[:,-1] = chosen_indices[-1][index]-chosen_indices[-2][index]-1 if len(chosen_indices) > 1 else 0
-			current_collated_slice = torch.nn.utils.rnn.pad_sequence(current_slice).to(device)
+			current_collated_slice = padded_input_data[torch.min(seq_index, orig_seq_lens_minus_one), batch_indices, :].unsqueeze(0)
 			currently_skipped = chosen_indices[-1]-chosen_indices[-2]-1 if len(chosen_indices) > 1 else chosen_indices[0]
-			assert (currently_skipped >= 0).all()
 			current_collated_slice[0,:,-1] = currently_skipped
 
 			if not opt.shareNet:
@@ -346,11 +341,7 @@ def custom_collate(seqs, things=(True, True, True)):
 				decisions = torch.argmax(current_slice_action_probs, dim=-1)
 			else:
 				current_slice_action_probs = torch.nn.functional.softplus(current_slice_action_probs)
-				# locs, _ = compute_normal_distribution_from_log_normal(current_slice_action_probs[:,:,0], current_slice_action_probs[:,:,1])
 				decisions = current_slice_action_probs[:,:,0]
[…]

### 2020-01-20 learn.py
Bugfix in test function

@@ -391,9 +391,10 @@ def bernoullize_seq(seq, p):
 
 global_chosen_packets = 0
 global_skipped_packets = 0
+global_total = 0
 
 def collate_things(seqs, is_seqs=False, sampling_masks=None):
-	global global_chosen_packets, global_skipped_packets
+	global global_chosen_packets, global_skipped_packets, global_total
 	if is_seqs and opt.averageFeaturesToPruneDuringTraining!=-1:
 		assert not opt.function=="dropout_feature_importance" or not lstm_module.training
 		assert not opt.function=="test" or not lstm_module.training
@@ -407,34 +408,25 @@ def collate_things(seqs, is_seqs=False, sampling_masks=None):
 		skipped = []
 		if is_seqs:
 			for mask in sampling_masks:
-				# skipped.append([])
-				# skipped_counter = 0
-				# for index in range(len(mask)):
-				# 	skipped[-1].append(skipped_counter)
-				# 	if not mask[index]:
-				# 		skipped_counter += 1
-				# 	else:
-				# 		skipped_counter = 0
-				# skipped[-1] = torch.tensor(skipped[-1], dtype=torch.float32)
 				seq_range = torch.arange(len(mask))[mask]
 				skipped.append(torch.cat((torch.tensor([0]), seq_range[1:]-seq_range[:-1]-1)))
-				# print("mask", mask, "skipped", skipped[-1], "seq_range", seq_range)
-				# skipped.append([seq_range[index]-seq_range[index-1] if index>0 else 0 for index in range(len(seq_range))])
+				assert (skipped[-1] >= 0).all()
 
-		for index in range(len(sampling_masks)):
-			assert len(sampling_masks[index]) == len(seqs[index])
-			chosen_ones = int(torch.sum(sampling_masks[index] == 1))
-			global_chosen_packets += chosen_ones
-			global_skipped_packets += len(sampling_masks[index]) - chosen_ones
+			for index in range(len(sampling_masks)):
+				assert len(sampling_masks[index]) == len(seqs[index])
+				assert len(seqs[index]) <= opt.maxLength
+				chosen_ones = int(torch.sum(sampling_masks[index] == 1))
+				global_chosen_packets += chosen_ones
+				skipped_ones = len(sampling_masks[index]) - chosen_ones
+				global_skipped_packets += skipped_ones
+				global_total += len(seqs[index])
[…]

### 2020-01-20 learn.py
Fixes for sampling

@@ -169,8 +169,8 @@ def custom_collate(seqs, things=(True, True, True)):
 		total_length = sum([len(item) for item in seqs])
 		n_seqs = len(seqs)
 		p = ((opt.samplingProbability*total_length)-n_seqs)/(total_length-n_seqs)
-		if p > 1 or p < 0:
-			print("opt.samplingProbability", opt.samplingProbability, "total_length", total_length, "n_seqs", n_seqs, "p", p)
+		# if p > 1 or p < 0:
+		# 	print("opt.samplingProbability", opt.samplingProbability, "total_length", total_length, "n_seqs", n_seqs, "p", p)
 		p = max(min(p, 1.0), 0.0)
 
 		masks = []
@@ -199,7 +199,7 @@ def custom_collate(seqs, things=(True, True, True)):
 		missing[missing==0] = 1
 		those_which_get_an_extra_packet = torch.argsort(missing)[:packets_left_to_distribute]
 		allocated_per_seq_discrete[those_which_get_an_extra_packet] += 1
-		assert torch.sum(allocated_per_seq_discrete) == chosen_packets
+		assert torch.sum(allocated_per_seq_discrete) == chosen_packets or chosen_packets < n_seqs
 
 		masks = []
 		for seq_index in range(len(seqs)):
@@ -210,43 +210,30 @@ def custom_collate(seqs, things=(True, True, True)):
 	elif opt.sampling=="first_n_equal":
 		total_length = sum([len(item) for item in seqs])
 		n_seqs = len(seqs)
-		chosen_packets = min(round(opt.samplingProbability*total_length), n_seqs)
+		chosen_packets = max(round(opt.samplingProbability*total_length), n_seqs)
 
 		len_of_each_seq = torch.tensor([len(seq) for seq in seqs], dtype=torch.float32)
 
 		sorted_by_length, sorted_by_length_indices = torch.sort(len_of_each_seq, descending=True)
-		# cumulative_lens_for_each_step
 		cumulative_packets = 0
 
-		# TODO:
 		last_full_length_packets = 0
 		final_length = 0
-		# print("int(max(len_of_each_seq))+1", int(max(len_of_each_seq))+1)
 		for current_len in range(1, int(max(len_of_each_seq))+1):
-			# print("beginning", current_len, "cumulative_packets", cumulative_packets, "chosen_packets", chosen_packets, "final_length", final_length)
 			if cumulative_packets >= chosen_packets:
-				# print("first break")
 				break
[…]

### 2020-01-21 paper_rl/paper.tex
Add paper

--- 2019-12-20 paper/paper.tex
@@
 \documentclass[conference]{IEEEtran}
+%% IEEE CNS addition:
+\makeatletter
+\def\ps@headings{%
+\def\@oddhead{\mbox{}\scriptsize\rightmark \hfil \thepage}%
+\def\@evenhead{\scriptsize\thepage \hfil \leftmark\mbox{}}%
+\def\@oddfoot{}%
+\def\@evenfoot{}}
+\makeatother
+\pagestyle{empty}
 
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Explainability and Adversarial Robustness for RNNs}
+\title{SparseIDS: Learning Packet Sampling with Reinforcement Learning}
 
-\author{\IEEEauthorblockN{Alexander Hartl,
-Maximilian Bachl, Joachim Fabini, Tanja Zseby}
+\author{\IEEEauthorblockN{Maximilian Bachl, Fares Meghdouri, Joachim Fabini, Tanja Zseby}
 \IEEEauthorblockA{Technische Universität Wien\\
@@
 
-\thispagestyle{plain}
-\pagestyle{plain}
+%\thispagestyle{plain}
+%\pagestyle{plain}
 
@@
 \newacronym{iat}{IAT}{Interarrival time}
+\newacronym{rl}{RL}{Reinforcement Learning}
+
+\newcommand{\ours}{SparseIDS}
 
@@
 
-\glspl{rnn} yield attractive properties for %performing intrusion detection on network data.
-constructing \glspl{ids} for network data.
-With the rise of ubiquitous \gls{ml} systems, malicious actors have been catching up quickly to find new ways to exploit \gls{ml} vulnerabilities for profit.
-Recently developed adversarial \gls{ml} techniques focus on computer vision and their applicability to network traffic is not straightforward: Network packets expose fewer features than an image, are sequential and impose several constraints on their features.
-%For example, destination ports cannot be modified since that could impair communication.
-
-We show that despite these completely different characteristics, adversarial samples can be  generated reliably for \glspl{rnn}.
-To understand a classifier's potential for misclassification, we extend existing explainability techniques and propose new ones, suitable particularly % specifically crafted %AH most can also be used for non-sequential models, so lets not restrict applicability too much
-for sequential data.
-Applying them shows that already the first packets of a communication flow are of crucial importance and are likely to be targeted by attackers.
-Feature importance methods show that even relatively unimportant features can be effectively abused to generate adversarial samples.
-Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids}, capturing a common notion of adversarial robustness, and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}.
-%as a countermeasure.
-% AH.
-the score alone is not yet a countermeasure
-%As a countermeasure, we develop an easy-to-use adversarial training procedure %and can confirm that it
-%which significantly and successfully reduces the attack surface and improves \gls{ars}.
+\glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data.
+They allow analyzing network flows and already determining if a flow is malicious or not before it is over, making it possible to take action immediately.
+However, considering the large number of packets that have to be inspected, the question of computational as well as energy efficiency arises.
+We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high.
+Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy with the same number of sampled packets.
+A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of packets like current approaches but can possibly skip arbitrarily many packets, saving even more computational resources for long sequences.
+Finally we build an automatic steering mechanism that can guide \ours{} to achieve a desired level of sparsity.
 
@@
 In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content.
-More recently, with the increasing deployment of encryption, %available features have changed.
-Since payload is no longer available,
-the focus now lies on features that are always available to network monitoring equipment like port numbers, protocol flags or packet sizes when encrypting above the transport layer.
+More recently, with the increasing deployment of encryption, the focus now lies on features that are always available to network monitoring equipment like port numbers, protocol flags or packet sizes when encrypting above the transport layer.
 
 Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties.
-%, between two applications potentially being located at different physical locations.
 When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets.
-% as well as their size.
-%AH: don't see why pkt size is not available without flows
-%For example, a specific attack might require the attacker to send three small packets and the victim to reply with two large packets followed by a small one to the attacker.
 Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
@@
 At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates.
+\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
 
-However, for practical use, high detection accuracy is not enough.
-With the recent rise of interest in \gls{aml} techniques, also \gls{aml} for \glspl{ids} has been investigated.
-For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids} and \cite{hashemi_towards_2019} investigate adversarial robustness of common network \glspl{ids}.
-In this research, we investigate whether adversarial samples can be found for our \gls{rnn}-based model, i.e.
-minimally different attack flows which are classified as benign.
-This is not a trivially answerable question since adversarial samples have mostly been analyzed in the context of computer vision.
-Our scenario significantly deviates from computer vision because \begin{enumerate*}
-\item the number of features is significantly smaller and
-\item only certain features can be manipulated if the flow should remain valid. % (for example the destination port cannot be manipulated)
-\end{enumerate*}
+However, for practical use, not only classification accuracy and security are important but also computational efficiency.
+This follows from the fact that a rationally-acting company will only deploy an \gls{ids} if the cost that is saved by avoiding additional security breaches is higher than the cost of the \gls{ids} itself.
+The cost of the \gls{ids} can be split in the cost of purchase/installation, the cost of operating it and the cost of maintaining it.
+The goal we pursue here is to develop sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data.
+For this purpose, we develop the \gls{rl}-based \gls{ids} \textit{\ours{}} that has a tunable parameter that allows trading off accuracy for sparsity of the data.
+We show that a significant number of packets can be skipped while accuracy doesn't drop significantly.
+Compared to other common sampling strategies, \ours{} performs better when being trained on the same number of packets.
 
-Surprisingly, we can confirm that adversarial samples can be found even when considering these tight constraints. % mentioned before.
-Thus, we argue that traditional performance metrics like, e.g., accuracy are not sufficient for evaluating security-sensitive \gls{ml} systems.
-We propose the \acrfull{ars} as a new performance score, which captures the notion of how easily an adversary can generate adversarial samples.
+As \ours{} uses a parameter to trade off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
+However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.
 
-Due to the threat of adversarial attacks, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans.
-This recently stirred up increased interest in explainability methods.
-Also in this case, common methods are designed to work with images or tabular data, but not with sequences.
-Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data.
-%These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
-Astonishingly, feature importance methods reveal that
-features, which are manipulated for successful adversarial flows,
-%the that are manipulable features, that are left for modification by adversaries
-are not even particularly important for the \gls{rnn}'s classification outcome. %Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}.
-
-%Since common explainability methods are designed to work with images or tabular data but not sequences, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data.
-These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
-
-Based on the insights gained by the feature importance and explainability methods, we propose two defense methods.
-%develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
-First, by simply leaving out manipulable features, we obtain a classifier which is slightly less accurate but is still useful to be deployed in a real scenario. Second,
-by deploying an adversarial training procedure,
-%AH. compared to other research about adversarial training, I dont feel that out approach is so special
-%develop a custom adversarial training procedure which especially focuses on ease of use and low resource consumption and aims to keep the number of additionally introduced training hyperparameters low.
-With this novel strategy
-we can reduce the attack surface and harden the resulting \gls{ids} while retaining all features and similar classification performance. % like for the original \gls{ml} model.
-
-We make all the source code, data and figures freely available to enable reproducibility and encourage experimentation at \url{https://github.com/CN-TU/adversarial-recurrent-ids}.
-
-\section{An \gls{rnn}-based Classifier}
+\section{Setup}
 
@@
 \centering
-\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
-]{
+%\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
+%]{
 \begin{tabular}{l r}
@@
 \end{tabular}
-}{}
-\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
-]{
-\begin{tabular}{l r}
-\toprule
-Attack type & \hspace*{-4mm}Proportion \\ \midrule
-Exploits	&	1.42\%	\\
-Fuzzers	&	1.01\%	\\
-Reconnaissance	&	0.58\%	\\
-Generic	&	0.21\%	\\
-DoS	&	0.19\%	\\
-Shellcode	&	0.08\%	\\
-Analysis	&	0.03\%	\\
-Backdoors	&	0.02\%	\\
-Worms	&	0.01\%	\\
-\bottomrule
-\end{tabular}
-}{}
+%}{}
+%\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
+%]{
+%\begin{tabular}{l r}
+%\toprule
+%Attack type & \hspace*{-4mm}Proportion \\ \midrule
+%Exploits	&	1.42\%	\\
+%Fuzzers	&	1.01\%	\\
+%Reconnaissance	&	0.58\%	\\
+%Generic	&	0.21\%	\\
+%DoS	&	0.19\%	\\
+%Shellcode	&	0.08\%	\\
+%Analysis	&	0.03\%	\\
+%Backdoors	&	0.02\%	\\
+%Worms	&	0.01\%	\\
+%\bottomrule
+%\end{tabular}
+%}{}
 \end{table}
 
-We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
+We implemented a three-layer LSTM-based classifier with 128 neurons at each layer. As the input features we use
 source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e.
 forward or reverse path) and all TCP flags (0 if the flow is not TCP).
-% \begin{itemize*}
-% \item source port
-% \item destination port
-% \item protocol identifier
-% \item packet sizes
-% \item \glspl{iat} of packets
-% \item packet directions (if packet goes there or back)
-% \item all TCP flags (0 if the flow is not TCP)
-% \end{itemize*}.
-We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behavior \cite{bachl_walling_2019}.
+We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.
 Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
@@
 
-For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+%For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+
 Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
-\autoref{tab:performance_results} shows that our \gls{rnn}-based classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}.
-However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before the attack flows terminate.
 
-\begin{table}
-% Evaluated using
-% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
-% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
-% on Dec. 4th
-\caption{Performance metrics per packet and per flow.} \label{tab:performance_results}
-\centering
-\begin{tabular}{l r r r r} \toprule
-& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
-	&	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
-Accuracy	&	99.1\%	&	99.7\%	&	99.5\%	&	98.3\%	\\
-Precision	&	97.0\%	&	99.7\%	&	83.4\%	&	78.6\%	\\
-Recall	&	97.8\%	&	99.1\%	&	87.6\%	&	72.6\%	\\
-F1	&	97.4\%	&	99.4\%	&	85.5\%	&	75.5\%	\\
-Youden	&	97.2\%	&	99.0\%	&	87.3\%	&	71.9\%	\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\section{Adversarial Attacks}
-\label{sec:adv}
-We now investigate whether known \gls{aml} methods can be used for generating adversarial flows for our \gls{rnn}.
-\cite{hashemi_towards_2019} previously studied the behavior of several Network \glspl{ids} under adversarial attacks but unlike us did not investigate \glspl{rnn}.
-
-A network packet contains significantly less features than, e.g., an image and, furthermore, most features such as TCP flags cannot be easily manipulated, as their manipulation might violate the protocol specifications and thus cause communication to fail.
-We identify the packet length and the \gls{iat} as features which are most likely to be exploited and thus choose them to be modified by the adversary.
-But even these features cannot be manipulated due to problem-specific constraints:
-\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
-\item Only packets can be manipulated which are transmitted by the attacker, except for botnet and backdoor traffic, which is entirely controlled by an attacker and thus only packets travelling in one direction can be manipulated.
- %This means the attacker can only control one direction (except for botnets where both sides are attackers).
-\item Packets must not be smaller than initially, as otherwise less information could be transmitted.
-\item \glspl{iat} must not decrease, as otherwise the physical speed of data transmission can be violated in some cases.
-An in-depth analysis of cases in which reduction of \glspl{iat} is legitimate is complex, so we generally disallowed \gls{iat} reductions.
-\end{itemize}
-
-Several \gls{aml} methods have been proposed in the recent literature, achieving different speed-quality tradeoffs:
-\cite{szegedy_intriguing_2014} shows that it is possible to create adversarial samples for an image, which look similar to the original image but are classified wrongly.
-\cite{goodfellow_explaining_2015} develops the \gls{fgsm}, which makes easy generation of adversarial samples possible.
-\cite{papernot_crafting_2016} explores the use of \gls{aml} for \glspl{rnn}, but lacks important \gls{aml} methods.
-%, which we consider in this paper.
-% Max: But FGSM is high-speed... So I removed this paragraph altogether
-%As finding adversarial flows is particularly difficult in our case, % for the reasons we discussed above,
-%we preferred methods  producing high-quality adversarial samples while not being as fast as other methods.
-We implemented the following \gls{aml} methods.
-
-\subsection{Carlini-Wagner}
-We implemented the \gls{cw} \cite{carlini_towards_2017}, performing gradient descent on the optimization objective
-\begin{equation} \label{eq:carliniWagner}
-d(X,\tilde X) + \kappa  \max(Z(\tilde X), \delta).
-\end{equation}
-Here, $d(\cdot)$ is a distance metric and $\kappa \in \mathbb R^+$ is a parameter governing the tradeoff achieved between attack success and distance from the original flow.
-Furthermore, $Z(\cdot)$ denotes the neural network's logit output, $X$ denotes the original flow and $\tilde X$ the adversarial flow optimized by \gls{cw}.
-
-$\delta \in \mathbb R$ is a parameter that determines how far an adversary wants to exceed the decision boundary.
-In the original publication $\delta=0$, meaning that the network's decision for adversarial samples is just between attack and benign traffic.
-In the present context, we need to make sure that the classifier's prediction would actually be benign, even though a certain level of noise will be added to \glspl{iat} due to the network between attacker and victim.
-Hence, we introduced $\delta=-0.2$, corresponding to a prediction confidence of 55\% for the sample to be benign after the sigmoid activation function.
-
-We used $L_1$ as distance metric, as we consider $L_1$ distance to represent practically relevant differences of network flows best.
-We used \gls{pgd} for meeting the real-world constraints discussed above.
-
-\subsection{$L_\infty$-bounded Projected Gradient Descent}
-\cite{madry_towards_2018} uses a method for generating adversarial samples which deploys \gls{pgd}  to maximize the network's loss function while constraining the achieved $L_\infty$ distance from the original samples.
-Hence, an important difference to \gls{cw} is that for this method the network's loss instead of its logit output is considered.
-We consider this method an interesting combination of \gls{cw} and \gls{fgsm}.
-Since its functioning is somewhat similar to \gls{cw}, we expected the generated adversarial samples to be similarly close to the original samples.
-
-\subsection{Fast Gradient Sign Method}
-Finally, we also tested the \gls{fgsm} % which was initially proposed in
- \cite{goodfellow_explaining_2015}, which is the first method for generating adversarial samples.
-FGSM can be considered a single pass of \gls{pgd} on the loss function with an equality constraint on the $L_\infty$ distance, i.e.
-the adversarial sample is found as
-\begin{equation}
-\tilde X = X + \epsilon \, \text{sgn}( \nabla_X L(X)),
-\end{equation}
-where $L(X)$ denotes the network's loss function and $\epsilon$ denotes the achieved $L_\infty$ distance.
-
-\begin{figure}[h]
-\includegraphics[width=\columnwidth]{./plots/adv_comparison/adv_comparison_17.pdf}
-\caption{Attack success ratios for both datasets and per attack type for CIC-IDS-2017. Depicted are flow detection accuracies for adversarial flows and for unmodified attack flows.}
-\label{fig:adv_per_family}
-\end{figure}
-
-\subsection{Evaluation}
-\autoref{fig:adv_per_family} depicts the performance for the investigated algorithms for both datasets and for each attack type in CIC-IDS-2017.
-We used \gls{cw} with a tradeoff of 1 and, to provide a fair comparison, for \gls{fgsm} and \gls{pgd} we used an average $L_\infty$ distance as observed for \gls{cw}.
-
-%As expected,
-\gls{cw} delivers the best performance and FGSM, while being the fastest of all investigated algorithms, delivers the worst performance.
-%\autoref{fig:adv_per_family} depicts how successful the attack is for individual attack families.
-The figure shows significant differences for detecting different attack families in the first place.
-Also for generating adversarial flows some attack families are more susceptible than others.
-However, the results match to a high degree with our expectations.
-For example, SQL injection attacks apparently are closer to benign traffic than \gls{dos} attacks and, hence, finding adversarial samples should be easier.
-
-Interestingly, any increase of the $L_\infty$ bound for \gls{pgd} did not yield significantly improved performance. 
-\gls{cw} thus generally achieved superior results to $L_\infty$-bounded \gls{pgd}
-and we can confirm the recommendation by \cite{carlini_towards_2017} to perform gradient descent on the logit output instead of the loss for good results.
-
-\autoref{fig:adv_cw} depicts the success ratio and average distance for \gls{cw} for different tradeoffs $\kappa$.
-%\autoref{fig:adv_cw} depicts the ratio of attack flows which were successfully misclassified as benign with respect to the final packet along with the resulting distance from the original sample.
-Evidently, the larger $\kappa$, the better the attack works, but also distances from original flows become higher.
-With acceptable distances, we were able to generate adversarial samples for about 80\% of attack samples.
-
-%Note, however, that f
-For successful adversarial samples, the second term in \autoref{eq:carliniWagner} becomes constant and, hence, is no longer relevant for optimization.
-Hence, $\kappa$ is irrelevant for the achieved distance, but governs if for one sample an adversarial sample is found.
-However, when using a large $\kappa$ in \autoref{eq:carliniWagner}, to avoid overly large steps that impede convergence, it is necessary to reduce the gradient descent learning rate.
-Since the distance term is not affected by $\kappa$, it is furthermore necessary to scale the iteration count inversely with the learning rate.
-\gls{cw} with a large $\kappa$ therefore needs more time than with a small $\kappa$.
-
-% The figure suggests that a tradeoff value between 0.5 and 1 would be ideal to find adversarial samples with a low enough distance from real samples.
-% This conclusion might be misleading, however, as due to the objective function \eqref{eq:carliniWagner} the distance cannot become larger anymore for one sample as soon as an adversarial sample has been found.
-%, when performing optimization as depicted by the objective function .
-% Hence, the depicted distance increase presumably is due to samples for which no adversarial sample could be found for lower $\kappa$ values.
-
-\begin{figure}[t]
-\includegraphics[width=\columnwidth]{./plots/adv_comparison/tradeoff.pdf}
-\caption{Performance of \gls{cw} with different $\kappa$ values.}
-\label{fig:adv_cw}
-\end{figure}
-
-\section{Adversarial Robustness Metric}
-
-As \autoref{fig:adv_per_family} shows, for most attack types, most samples can be modified by an adversary to be classified as benign by using \gls{cw}.
-High recall for a particular attack type does not imply that adversarial samples are hard to find  %are  likely to fail
-for these attacks. Thus we argue that beside the classical metrics such as accuracy, false positives etc., a metric is needed, which quantifies how easily an \gls{ids} classifier can be fooled.
-
-Such a metric should be easy to compute and easy to interpret.
-While in general adversarial robustness for \gls{ml} models is frequently quantified as the average or median of minimum distances of adversarial samples \cite{carlini_evaluating_2019}, we found no generally agreed-upon metric for \glspl{ids} in the literature.
-If we consider flows, for which no adversarial sample can be found, to have infinite distance, the median has the advantage of ignoring such unsuccessful samples and outliers.
-On the other hand, the median might depict expected distances badly if they have a very uneven distribution.
-% We propose the \gls{ars} as 
-% \begin{equation}
-% \text{ARS} = \text{mean}(\{d \in \mathcal D \mid d < Q_{0.5}(\mathcal D)\}),
-% \text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \{d \in \mathcal D \mid d \leq Q_{0.5}(\mathcal D)\}} d,
-% \end{equation}
-% denoting by $Q_{0.5}(\cdot)$ the median, by $N = | \mathcal D | \in \mathbb N$ the total number of samples and by $\mathcal D$ the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$.
-
-We thus propose the \gls{ars} as follows.
-Let $\mathcal S$ denote the set of samples, $N = | \mathcal S | \in \mathbb N$ the total number of samples and $d_s \in \mathbb R_0^+$ the distance of an adversarial flow from the original flow for a sample $s\in \mathcal S$, assigning unsuccessful samples a distance of $\infty$.
-% and $Q_{0.5}(\cdot)$ the median function.
-We define the \gls{ars} as 
-\begin{equation}
-\text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{\tilde s \in \tilde {\mathcal S}} d_{\tilde s},
-\end{equation}
-with $\tilde {\mathcal S} \subset \mathcal S$ denoting a set of size $|\tilde {\mathcal S}| =\lceil N/2 \rceil$, so that $d_{\tilde s} \le d_s$ for all $\tilde s \in  \tilde {\mathcal S}, s \in   {\mathcal S} \setminus \tilde {\mathcal S} $.
-% We thus propose the \gls{ars} as follows.
-Let $\mathcal D$ denote the set of distances of adversarial flows from original flows, assigning unsuccessful samples a distance of $\infty$, $N = | \mathcal D | \in \mathbb N$ the total number of samples and $Q_{0.5}(\cdot)$ the median function.
-% We define the \gls{ars} as 
-% \begin{equation}
-% \text{ARS} = \frac{1}{\lceil N/2 \rceil} \sum\nolimits_{d \in \tilde {\mathcal D}} d,
-% \end{equation}
-% with $\tilde {\mathcal D} \subset \mathcal D$ denoting the set of size $|\tilde {\mathcal D}| =\lceil N/2 \rceil$, so that for all $d \in  \tilde {\mathcal D}, d \le Q_{0.5}(\mathcal D)$.
-The \gls{ars} is thus approximately the average of distances not larger than the median distance.
-
-%Such a metric should be easy to compute, easy to interpret and robust with respect to outliers.
-We consider an adversary successful if he can cause at least 50\% of all samples to be misclassified.
-In this case, the \gls{ars} is the average of the distances of the adversarial samples to the original samples for the 50\% of all samples which have the smallest distances.
-If the adversary doesn't manage to manipulate enough samples, the \gls{ars} is $\infty$.
-
-\gls{cw} is able to find adversarial samples with minimum distance, but becomes slow for large $\kappa$. Hence, it can be used for finding the \gls{ars} efficiently as follows: 
-%Since \gls{cw} becomes slow for large $\kappa$, the \gls{ars} can be efficiently found as follows:
-Use \gls{cw} with a small tradeoff, trying to generate adversarial samples for an attack type. If %, of a total of $N$ samples,
- at least $N/2$  are wrongly classified, and
- the smallest distance of non-adversarial samples is not smaller than the $\lceil\frac{N}{2}\rceil$th smallest distance of adversarial samples,
-stop and compute the \gls{ars}.
-Otherwise increase $\kappa$ and repeat.
-If after a predefined number of iterations -- e.g.
-100 -- still not more than half of the samples are adversarial, also break and set the \gls{ars} to $\infty$.
-
-The larger the \gls{ars} is, the more robust the model is, as then an adversary needs to modify the samples more to make them adversarial.
-If not more than half of the samples could be made adversarial, our metric is $\infty$ since then apparently it is not possible for an adversary to reliably conduct adversarial attacks on the majority of samples.
-In this case, the ratio of samples that could be made adversarial can be a useful metric to determine the exact extent of the adversarial threat.
-
-Setting the threshold for the \gls{ars} to 50\% is arbitrary, but
- % reasonable as this way the \gls{ars} approximates the median of the distances of all samples and, additionally, marks the distance where an adversarial attack is more likely to be successful than unsuccessful.
- reasonable as outliers with very large distances are ignored and because if an attacker can make the majority of samples adversarial, we argue that its attack is ``successful''.
-%we think that it is reasonable since if an attacker can make the majority of samples adversarial, we think that its attack is ``successful''.
-
-\begin{figure}[h]
-\includegraphics[width=\columnwidth]{./plots/ars_original.pdf}
-\caption{Recall for unmodified flows and \gls{ars} for attacks in CIC-IDS-2017. For ``PortScan, Firewall'' recall never falls below 66\%: The attack does not succeed and the \gls{ars} is $\infty$.}
-\label{fig:recall_ars}
-\end{figure}
-
-\autoref{fig:recall_ars} shows that certain attack types, like ``DoS SlowHTTPTest'', are easy to classify for an IDS but still are also surprisingly vulnerable to adversarial modifications by an attacker aiming to make them look benign.
-
-\section{Explaining Recurrent Neural Networks}
-Having verified the effectiveness of \gls{aml}  for our \glspl{rnn}, we now  investigate how the classifiers come to a decision.
-From a naive perspective, one might be tempted to reuse existing explainability methods for \glspl{rnn} by considering a flow the sum of its packet features.
-We identify several difficulties which occur when trying to explain decisions made by \glspl{rnn}.
-
-\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
-\item
-\textbf{Feature quantity.}
-The number of features fed into an \gls{rnn} is the number of packet features times the length of the flow. For long flows, the total number of inputs can become very large.
-
-\item
-\textbf{Variable sequence lengths.}
-The length of different flows might differ tremendously. Hence, features at one particular time step might be important for the network's outcome for one flow, but not even exist for other flows.
-
-\item
-\textbf{Lack of a distance measure.}
-However, even if we restricted the analysis to flows of a constant length, a flow is different from the plain concatenation of its packet features.
-For example, in a sentence, which is sequence of words, parts can be rearranged, giving a different sequence with possibly the exact same meaning.
-
-\item
-\textbf{Multiple prediction outputs.}
-Often an \gls{rnn} produces an output at each time step.
-When  applying explainability methods, the question arises which output to consider for the method.
-The natural choice is to base the methods on the prediction output which occurs at the same time step as the feature under investigation: This approach is less complex compared to considering also features of all earlier time steps.
-Also, we expect the current prediction outcome to more dependent on the current feature, compared to a feature from many steps ago.
-However, due to the complex decision processes of deep neural networks,
-this is not always true and a feature might influence a decision many time steps later. 
-\end{itemize}
-
-Many explainability methods proposed recently are local and thus provide explanations for a particular data sample \cite{shapley_value_1953,lundberg_unified_2017,dhurandhar_model_2018,ribeiro_why_2016}.
-However, for the particular problem of network traffic, due to the high number of flows and the characteristics of data, analyzing individual samples is of low interest.
-Explainability methods presented in this paper therefore aim to understand a model by analyzing which features are important, at which time step they are important and which feature values lead to classification as attack.
-
-% where does this come from :)
-%\subfloat[Throughput of the fq \gls{qdisc}\label{fig:fqThroughput}
-%]{
-%\includegraphics[width=\columnwidth]{{{plots/throughput_1_fq_cubic_10_20_60_0.5_bw_1571240937211.pcap}}}
-%}{}
-
-\subsection{Feature Importance Metrics}
 %\begin{table}
-%\caption{Accuracy drop for:}
-%\label{tab:feat_importance}
+%% Evaluated using
+%% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
+%% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
+%% on Dec. 4th
+%\caption{Performance metrics per packet and per flow.} \label{tab:performance_results}
 %\centering
-%\subfloat[Random replacement method\label{fig:featRandom}
-%]{
-%\begin{tabular}{l r}
-%\toprule
-%Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
-%Protocol & 0.207 \\
-%Packet Length & 0.165 \\
-%SYN Flag & 0.099 \\
-%ACK Flag & 0.084 \\
-%Direction & 0.073 \\
-%Destination port & 0.071 \\
-%Source port & 0.060 \\
-%RST Flag & 0.057 \\
-%PSH Flag & 0.056 \\
-%Interarrival time & 0.024 \\
-%FIN Flag & 0.012 \\
-%URG Flag & 0 \\
-%ECE Flag & 0 \\
-%CWR Flag & 0 \\
-%NS Flag & 0 \\
+%\begin{tabular}{l r r r r} \toprule
+%& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
+%	&	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
+%Accuracy	&	99.1\%	&	99.7\%	&	99.5\%	&	98.3\%	\\
+%Precision	&	97.0\%	&	99.7\%	&	83.4\%	&	78.6\%	\\
+%Recall	&	97.8\%	&	99.1\%	&	87.6\%	&	72.6\%	\\
+%F1	&	97.4\%	&	99.4\%	&	85.5\%	&	75.5\%	\\
+%Youden	&	97.2\%	&	99.0\%	&	87.3\%	&	71.9\%	\\
 %\bottomrule
 %\end{tabular}
-%}{}
-%\subfloat[Feature dropout\label{fig:featDropout}
-%]{
-%\begin{tabular}{l r}
-%\toprule
-%Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
-%Destination port & 0.025 \\
-%Source port & 0.003 \\
-%RST Flag & 0.001 \\
-%ACK Flag & 0.001 \\
-%Protocol & 0.001 \\
-%Packet Length & 0.001 \\
-%Direction & 0.001 \\
-%SYN Flag & 0 \\
-%Interarrival time & 0 \\
-%FIN Flag & 0 \\
-%ECE Flag & 0 \\
-%URG Flag & 0 \\
-%CWR Flag & 0 \\
-%NS Flag & 0 \\
-%PSH Flag & 0 \\
-%\bottomrule
-%\end{tabular}
-%}{}
 %\end{table}
-As a first step to understanding the neural network's decisions, we  estimate how important individual features are for the model's predictions.
-When investigating an \gls{ml} classifier, it is natural to ask which inputs have a large influence on the classifier's prediction.
-We feel the need to distinguish metrics for this purpose based on their main aim:
-
-A large amount of research has been spent on finding \emph{feature importance} metrics, which allow the selection  of high-importance features, providing a reasonably good classification performance while resulting in a more light-weight classifier.
-
-Conversely, both adversarial machine learning and explainable machine learning bring up the question to what extent individual features are able to change the prediction outcome.
-While appearing similar, traditional feature importance can yield markedly wrong results for this case.
-To distinguish, we propose the term \emph{feature sensitivity} for such metrics.
-To analyze features, we use the following approaches:
-
-\subsubsection{Neural Network Weights}
-In previous works~\cite{olden_accurate_2004}, a simple method for determining feature importance in neural networks has been summing up neural network weights leading from a certain input to the prediction outcome.
-The weights method can be considered for both feature importance and feature sensitivity, however, clearly, especially in the case of complex network architectures, this method is likely to provide wrong results.
-Hence, we provide results for the weights method mainly for comparison.
-Note that an LSTM cell alone has four weights leading from one input to an output.
-
-\subsubsection{Input Perturbation}
-
-The most commonly deployed feature importance techniques, used by practitioners for \glspl{rnn} \cite{stackexchange_cross_validated_neural_2019} and \gls{dl} \cite{molnar_interpretable_2019,stackexchange_cross_validated_feature_2016,olden_accurate_2004}, are based on adding noise to a particular feature and evaluating the drop in accuracy that occurs.
-We argue that it is hard to determine the ``correct'' intensity of noise to add.
-%, we slightly vary the proposed approach: We do not add noise drawn from a probability distribution but instead
- Hence, we sample the value for a feature from the distribution of all values of this feature in the dataset.
-This makes the method non-parametric since the noise distribution  does not need to be chosen.
-We ensured that features which stay constant for a flow, i.e.
-source port, destination port and protocol, also stay constant throughout the flow when randomizing the feature.
-
-\subsubsection{Feature Dropout}
-While the perturbation method %of adding noise to features/randomizing them
-is convenient since it is easy to implement and understand, we argue that it has some shortcomings:
-The \gls{rnn} was never trained for dealing with ``garbage'' values that the randomization creates.
-For example, completely unrealistic feature combinations could occur that were never observed during training.
-Furthermore, sequences of features could occur that cannot occur in reality.
-
-%We think that the ``ideal'' feature importance method for our scenario are Shapley values \cite{shapley_value_1953}: To analyze feature importance with them, the feature under investigation is completely omitted from the dataset and a model is trained without it.
-The resulting accuracy is then compared to the baseline accuracy with all features.
-The effort required to do this is high both in terms of storage space as well as training time since for $n$ features $n$ models need to be separately trained and stored.
-While this seems tractable, when also analyzing the feature importance of combinations of features, the number of models required becomes even larger: Out of $n$ features, $2^n$ unique subsets can be chosen, meaning that the computational demand rises exponentially.
-
-To evaluate true feature importance, we thus develop a more sound method %compared to input perturbation
-called \textit{feature dropout}: When training a model, for each sample, we leave out each feature  with independent probability $\frac{1}{n}$, $n \in \mathbb N$ being the number of features, by setting it to zero.
-On average, one feature gets zeroed out but it is also possible that none or more than one are left out.
-This procedure is equivalent to using dropout \cite{srivastava_dropout:_2014} with probability $\frac{1}{n}$ before the first layer.
-
-An important implementation detail is that for each feature we add another input which is 1 if the feature is suppressed and 0 otherwise.
-This is necessary for the neural network to be able to distinguish between a feature missing and a feature genuinely being zero.
-The overall outcome is a classifier being able to deal with missing features.
-The results in \autoref{fig:feat_imp} show that, unlike input perturbation, feature dropout  does not vastly overestimate features' importance.
-With feature dropout, it becomes apparent that only very few features  actually contain unique information, affecting accuracy when left out: the destination port and the source port.
-
-A model trained with feature dropout typically yields slightly lower accuracy than a regularly trained model, even if no features are left out (flow accuracy of 99.43\% vs.
-99.65\%).
-We thus recommend training two models: One regular one and one with feature dropout to use for the feature importance.
-
-Another method that uses dropout for feature importance is \cite{chang_dropout_2017}, applying a technique called \textit{Variational Dropout} to learn the optimal dropout rate for each feature.
-It tries to leave out as many features as possible and at the same time keep accuracy high.
-Thus important features are going to be left out less often and one can then extract the dropout probabilities for each feature and assess their significance based on them.
-%
-% Max: I'd say that the following sentence is optional...
-While this method looks seemingly related to feature dropout, it is significantly more complex and does not aim to show the accuracy drop that occurs when omitting a feature but instead returns a unique feature importance value between 0 and 1.
-%Feature dropout can also be used as a building block for computing the accuracy of datapoints with missing features that is necessary to compute Shapley values \cite{shapley_value_1953} or KernelSHAP \cite{lundberg_unified_2017}.
-
-For feature dropout, it can also happen that several features are left out for a sample and so our method can also be used to analyze the effect of multiple features missing and can thus show possibly correlated features or -- more general -- features that contain common information, relevant for the classification task:
-%Feature dropout also allows us to find correlated features or -- more general -- features that contain common information:
-\begin{equation}
-\text{score} = \frac{\text{acc}_{\text{base}}-\text{acc}_{\text{-both\_features}}}{\left(\text{acc}_{\text{base}}-\text{acc}_{\text{-feature\_1}}\right) + \left(\text{acc}_{\text{base}}-\text{acc}_{\text{-feature\_2}}\right)}
-\end{equation}
-With $\text{acc}_{\text{base}}$ we denote the accuracy of the classifier with all features included, with $\text{acc}_{\text{-feature\_i}}$ the accuracy if feature $i$ is omitted and with $\text{acc}_{\text{-both\_features}}$ the accuracy if both features are omitted.
-Assuming a non-negative accuracy drop when omitting a feature, the resulting score is $\ge 0.5$.
-The higher it is, the larger the information that both features share.
-
-%AH. As it can become <1, I would say it this way
-%If the score is one,  both features contain no joint information.
-For example, the score between RST and the protocol identifier is 8.5; the highest of all pairs of features.
-While this may be unintuitive at the first glance, it likely stems from the fact that if the protocol identifier (TCP or UDP) is missing, then
-%if RST is 1 at some point of a flow, it has to be TCP.
-RST being 1 at some point indicates that the flow is TCP.
-
-Feature dropout might constitute a building block for methods based on Shapley values \cite{shapley_value_1953} like KernelSHAP \cite{lundberg_unified_2017}. 
-%Due to space limitations, we leave this for future work.
-\subsubsection{Mutual Information}
-\begin{figure}
-\includegraphics[width=\columnwidth]{./plots/mutinfo_example.pdf}
-\caption{Two distributions yielding an identical accuracy drop.}
-\label{fig:mutinfo_example}
-\end{figure}
-Input perturbation and feature dropout mainly address feature importance.
-For example, assuming that the test dataset is representative for production use, for feature importance it is reasonable to equate the distribution for perturbing a feature with the feature distribution itself.
-However, when evaluating feature sensitivity, e.g.
-for analyzing potential for adversarial samples, the attacker is not limited by this distribution and frequently is able to choose arbitrary values in the feature space.
-
-\begin{figure*}
-\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
-\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2015.pdf}
-\caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
-\label{fig:feat_imp}
-\end{figure*}
-Furthermore, let $f(A,x)$  denote the joint probability
-%\todo{Max: Ist joint probability nicht immer wenn man mehrere Ereignisse hat? Ich hätte hier probability density function geschrieben...}
-%AH. changed notation and wording
-for classification as attack and a feature value $x$. Accuracy then is $\int f(A,x) dx$ for attacks and $1-\int f(A,x) dx$ for benign traffic.
-\autoref{fig:mutinfo_example} shows two different distributions
-yielding the same accuracy, but clearly the right-side distribution has a larger influence on the prediction, as in the right-side case the prediction deterministically depends on the feature value.
-To capture such dependencies, we propose to use mutual information %instead of accuracy drop
-to determine feature sensitivity.
-Mutual Information between two random variables $X,Y$ is defined as
-\begin{equation}
-I_{X,Y} = \mathbb E \left\{ \log\left(\frac{f_{X,Y}(x,y)}{f_X(x)f_Y(y)}\right) \right\},
-\end{equation}
-with $f_X(x), f_Y(y)$ and $f_{X,Y}(x,y)$ denoting the distribution of $X,Y$ and their joint distribution, respectively.
-To obtain feature sensitivity, we compute mutual information between an input variable and the prediction output for one flow at one particular time step, averaging over the result for the test set.
-
-\subsubsection{Comparison}
-\autoref{fig:feat_imp} shows the results, which match to a large extent with domain-specific expectations for classifying network flows.
-In particular, rarely used TCP flags like NS or URG are unimportant for the classifier.
-On the other hand, destination port and protocol are essential for characterizing flows by hinting at the type of traffic.
-\gls{iat} and packet length are important for estimating the amount of transferred information and several flags hint at certain attacks like \gls{dos} attacks.
-
-The weights method is able to reveal features with a very low importance to a certain degree, but disagrees with the other methods to a large extent.
-Less anticipated, however, also input perturbation does not exhibit a considerable correlation with feature dropout.
-Considering its functioning of completely removing individual features, feature dropout is the most reliable method for evaluating importance for removing features.
-It is remarkable that none of the other methods is able to depict the distinct peak of importance for the destination port.
-
-It is not surprising that mutual information disagrees with feature dropout, since both their aim and their functioning are substantially different.
-For example, mutual information shows that the protocol field can have a substantial impact on the classification even though an identical accuracy can be achieved when omitting it.
-
-% In particular, the most important features turn out to be the used protocol, packet length and the SYN TCP flag.
-While the first two features are essential for characterizing flows, a high number of packets with set SYN flag is an important characteristic for port scanning or certain DoS attacks.
-Also the fact that most of TCP flags do not play an important role in our dataset was expected.
-It is interesting to note, however, that the interarrival time of packets does not play a crucial role for the detection of attacks.
-
-Metrics for UNSW-NB15 differ substantially from CIC-IDS-2017.
-However, due to the large number of different network attacks and network configurations, it is easily possible that relevant features are very different.
-We consider the question of model transferability of substantial importance for \gls{ids} applications, but out of scope for the present research.
-
-\subsection{Explainability Plots}
-Knowing which features to investigate, it is important to analyze which feature values lead to classification as attack.
-In literature, the use of Partial Dependency Plots (\gls{pdp}) has been proposed~\cite{friedman_greedy_2001}.
-To inspect attack types in detail, in this research we use a \textit{conditional} form of the \gls{pdp}.
-If $\boldsymbol X \in \mathbb R ^n$ denotes a random vector drawn from feature space, $f(\boldsymbol X) \in [0,1]$ the neural network's prediction and $c$ the attack type, we define the conditional \gls{pdp} for feature $i$  as
-% \begin{equation} \label{eq:pdp}
-% \text{\gls{pdp}}_i(y) = \mathbb E_{\boldsymbol X}\left(f(X_1,\ldots,X_{i-1},y,X_{i+1},\ldots X_n)\right)
-% \end{equation}
-% To inspect attack types in detail, in this research we experiment with a modified definition of the \gls{pdp}  (with C being the attack classes):
-\begin{equation} \label{eq:pdp_conditional}
-\text{\gls{pdp}}_{c,i}(w) = \mathbb E_{\boldsymbol X | C} \Big(f \left( X_1,\ldots,X_{i-1},w,X_{i+1},\ldots X_n \right) | c\Big),
-\end{equation}
-empirically approximating the conditional distribution by the observed subset that belongs to a certain attack type.
-
-By using a classical \gls{pdp} we would likely lose information due to the averaging over very different types of traffic.
-However, for network traffic in particular, investigating each sample individually is not possible.
-Hence, the conditional \gls{pdp} provides the ideal compromise for our application.
-
-Due to the use of a 5-tuple flow key, port numbers and the protocol identifier are constant for all packets in one flow.
-Hence, we can consider the \gls{rnn} a regular classifier and reuse 
-%which allows us to use
-\glspl{pdp}, which have been proposed for non-recurrent classification methods, by plotting the \gls{rnn}'s flow prediction outcome over one of these  features.
-The results (omitted for brevity) show that for some attack types the port numbers play an important role. When looking at the \gls{pdp} for benign traffic samples, %(non-attack traffic)
-it becomes apparent that traffic destined to a high destination port is generally indicative of an attack. We argue that this is because most services that regular users use have low port numbers.
-
-\subsection{Plots for Sequences}
-
-%As apparent from the above discussion, trying to explain how packet features influence the model's predictions, poses various difficulties.
-As a first step to explaining our model's predictions, we investigate how important features at different time steps are for the model's decisions.
-Intuitively, features at the beginning of a flow should be the most important while the classifier's predictions should not vary significantly anymore, as soon as it has come to a decision.
-\begin{figure}
-\includegraphics[width=\columnwidth]{{"./plots/plot/flows_pred_plots2_outcomes_0_3.pickle_15_All_samples"}.pdf}
-\caption{Classifier confidence per time step for CIC-IDS-2017. % and first and third quartiles.
-For the majority of attack types, confidence increases in the first few steps and then stays almost constant at 1.}
-\label{fig:confidence}
-\end{figure}
-To evaluate this hypothesis, \autoref{fig:confidence} depicts the classifier's prediction confidence for each time step, along with the number of samples having at least this length, which were used for evaluating the figure.
-% for each attack class.
-%averaging over all flows. %AH. there's no averaging happening
-% and also show the first and third quartiles.
-%\autoref{fig:confidence} shows that %this hypothesis is correct for most attack classes.
-While at the first couple of packets the confidence is not very high, towards the end of the flow it reaches values close to 1 and stays there.
-Hence, not only is the classifier able to make a reasonable classification after just a few packets, the figure also suggests that indeed later packets have a neglectable influence on the prediction.
-
-For investigating in more detail how features influence the prediction outcome, we extend \glspl{pdp} to the sequential setting.
-Denoting as $X=\{\boldsymbol X^1, \ldots, \boldsymbol X^m\}$ the series of packet features $\boldsymbol X^t \in \mathbb R^n$ and $h_t(X)$ the network's hidden state after time step $t$, we define the sequential \gls{pdp} as
-\begin{align}
-\text{se}&\text{qPDP}_{c,i}(t,w)= \\ &\mathbb E_{X | C} \Big(f \left(h_{t-1}( X), X_1^t,\ldots,X_{i-1}^t,w,X_{i+1}^t,\ldots X_n^t \right) | c\Big). \nonumber
-\end{align}
-
-\autoref{fig:pred_plots2} shows an example together with the mean values for both unmodified samples and adversarial samples.
-Also in this figure we notice that mainly the first few packets are able to influence the prediction outcome and modifying features at a later time step does not change its confidence any longer.
-In many cases, the adversarial sample generation indeed moves packet features to areas where the network is less likely to be classified as attacks, confirming the effectiveness of \glspl{pdp}.
-In other cases, however, we did not observe an agreement between \gls{pdp} and adversarial modifications, hinting at dependencies which cannot be presented by \glspl{pdp}.
-Since the \gls{iat} is undefined for the first packet, we always set it to 0.
-Still, interestingly, the plot shows, that the classifier considers packets with a higher IAT to be more likely to be attacks than those with a smaller one.
-
-\begin{figure}[h]
-
-\includegraphics[width=\columnwidth]{{"./plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_7_DoS_-_DDoS-DoS_slowloris"}.pdf}
-
-\caption{Exemplary sequential PD plot and adversarial flows for the DoS Slowloris attack in CIC-IDS-2017. The lines show the feature's mean values. % for all flows of the attack type.
-The shaded region shows the change in confidence that occurs when the feature is varied.}
-%\todo{IAT plot is wrong: There shouldn't be step 0.} %AH. guess this is fine now
-\label{fig:pred_plots2}
-\end{figure}
-
-% AH. I guess this can only be properly formulated with a formula
-%Next, we want to investigate if our classifier just memorized its inputs.
-The most interesting packet features that can be used to visualize this in our use case are the \gls{iat} between packets in a flows and the packet length.
-To depict how the model comes to a decision, we determined the total minimum and maximum for these features that ever occur for all flows.
-For a given flow $X$ we then varied feature values at each time step between minimum and maximum and plot how the confidence changes when the feature varies (\autoref{fig:pred_plots2}) depict the averaged predictions for different attack families.
-
-%Furthermore, it seems clear that the classifier did not just memorize the inputs but actually learned appropriate space representations, as there is always a smooth decrease of confidence when modifying feature values.
-When memorizing, we would  instead expect a sharp decline of confidence.
-% to occur.
-%Another interesting observation is that after the first couple of packets the classifier is certain about its decision and changing features doesn't change its confidence any longer.
-This allows the conclusion that if an attacker is able to change the first few packets, she could change the decision of the classifier.
-This is consistent with our observations in \autoref{sec:adv}.
-%We will further investigate this question in \autoref{sec:adv}.
-
-\begin{figure}
-%AH. I think the 'Firewall off' plot is not a good example as it shows only 4 packets
-\includegraphics[width=\columnwidth]{{"./plots/plot_features/flows_pred_plots2_outcomes_0_3.pickle_2_Brute_Force-SSH-Patator"}.pdf}
-\caption{\gls{iat} and packet length for SSH brute-force attacks in CIC-IDS-2017.}% The solid line shows the median value of the feature for all flows of a specific attack type.
-The shaded region are the first and third quartile.} % think legend is enough for unterstanding
-\label{fig:plot_features}
-\end{figure}
-Finally, we investigated whether our classification task involves recognizing complex patterns in the feature space.
-As example, \autoref{fig:plot_features} shows that
-%Looking at the plots of the features that vary over time (\autoref{fig:plot_features}) we see that certain
-attack types indeed have a characteristic pattern in which they send packets by which they are easily recognizable.
-Other attack families similarly show characteristic patterns.
-%We want to determine how the pattern of packet sizes and \glspl{iat} is significant for the classification.
-\autoref{fig:plot_features} clearly shows that some attack types have very characteristic patterns of packet sizes and \glspl{iat}.
-
-\section{Defenses}
-We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
-\subsection{Reducing Features}
-The most obvious defense, which is only possible because of the nature of the dataset, is to simply omitting features an attacker can manipulate. We try two different approaches:
-\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
-\item Leaving out all manipulable features: packet size and \gls{iat}
-\item Leaving out manipulable features only when the attacker can manipulate them, i.e.
-only in the direction from the attacker to the victim.
-This, however, does not prevent adversarial samples for botnets, for which both sides are malicious.
-\end{itemize}
-
-Both approaches lead to complete resistance to adversarial samples, except for botnets, which can still operate when only leaving out manipulable features in one direction.
-The results show that -- surprisingly -- there is only a small difference in classification performance when looking at flow accuracy: It is slightly lower at 99.3\% compared to 99.7\% originally for CIC-IDS-2017.
-However, packet accuracy is only 98\% when leaving out the features in one direction and 96.7\% when leaving them out in both directions.
-Thus, apparently the \gls{iat} and packet size are especially important for determining whether a flow is malicious in the first packets of a flow.
-
-\subsection{Adversarial Training}
-As alternative to omitting manipulable features, we attempted to make the classifier more robust against adversarial samples by augmenting the training set by adversarial flows generated using  \gls{cw}, labeled as additional attacks.
-For this purpose, we added one adversarial sample per attack sample to the training set.
-Since \gls{cw} is deterministic, this is the maximum number of adversarial samples possible.
-With this augmented training set we then alternated retraining of the neural network and optimization of the adversarial samples.
-% using gradient descent.
-
-A question which occurs in this process, is how many CPU cycles to spend on network training and adversarial sample optimization.
-As a natural approach, we decided to use as many backpropagation steps for neural network training as we use for adversarial sample optimization.
-For each adversarial sample optimization step, we performed 10 iterations of gradient descent.
-Hence, all adversarial samples were optimized each 10 epochs of neural network training.
-
-\autoref{fig:avdistance} shows the \gls{ars} throughout  adversarial training  for several attack categories and clearly indicates that adversarial training is effective, as the distance gradually increases.
-Hence, an attacker would have to modify attack samples more and more, eventually rendering the attack unpractical.
-
-%However, the figure also shows that an increase of distance can only be achieved after a considerable amount of training.
-It is, hence, questionable if the gained increase in robustness is enough for reliable attack detection.
-
-For both datasets, after a small number of epochs, it was no longer possible to create a significant number of adversarial samples for most attack categories.
-%For the datasets we considered, adversarial training left classification performance unaffected.
-%\todo{Max: Insert results that show that accuracy on normal data is still ok after adversarial training.}
-%\autoref{tab:adv_training_performance} depicts classification on non-adversarial data after 50 epochs of adversarial training.
-After 50 epochs of training, accuracy is essentially identical to the results presented in \autoref{tab:performance_results}.
-Recall and Youden increased and precision slightly decreased.
-However, this is due to the higher proportion of attack samples in training data and, hence, expected.
-We conclude that adversarial training is effective for reducing the attack surface for adversarial attacks in our scenario.
-
-% \begin{table}
-% \caption{Detection performance after adversarial training.} \label{tab:adv_training_performance}
-% \todo{Probably no space for table}
-% \centering
-% \begin{tabular}{l r r r r} \toprule
-% & \multicolumn{2}{c}{CIC-IDS-17} & \multicolumn{2}{c}{UNSW-NB-15} \\
-	% &	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
-% Accuracy	&	0.991	&	0.997	&	0.994	&	0.981	\\
-% Precision	&	0.971	&	0.996	&	0.761	&	0.704	\\
-% Recall	&	0.979	&	0.992	&	0.970	&	0.824	\\
-% F1	&	0.975	&	0.994	&	0.853	&	0.759	\\
-% Youden	&	0.973	&	0.991	&	0.965	&	0.811	\\
-
-% \bottomrule
-% \end{tabular}
-% \end{table}
-
-% AH. guess this is no longer true
-%However, against our expectations, this approach did not lead to satisfactory results in the sense that the network would learn to classify adversarial samples as attacks.
-In fact, classification accuracy stayed about constant during the training of the network.
-
-%A possible explanation for this observation might be that the adversarial samples assumed too much the shape of normal flows, so that the network could not reliable tell them apart.
-
-%\begin{table*}
-%\caption{Comparing performance with all features, only safe features (\gls{iat} and packet size removed) and safe features in one direction (\gls{iat} and packet size removed for packets coming from source).
-\textit{packets} means that we consider the neural network output of all packets of the flow while \textit{flows} means that we only consider the last packet.
-At the last packet the classifier is usually already a lot more confident than at the first packets of a flow and thus metrics are always better in this case.
-\todo{Max: Is this that interesting?
-Remove tables and inline the most interesting results?}}  \label{tab:performance_results_no_manipulable}
-%\newcommand{\cmidrulespace}{6pt}
-%\centering
-%\begin{tabular}{l l l l l l l} \toprule
-%& \multicolumn{2}{l}{all} & \multicolumn{2}{l}{safe features} & \multicolumn{2}{l}{safe features one direction} \\
-%\cmidrule(r){2-3} \cmidrule(lr){4-5} \cmidrule(l){6-7}
-%& packets & flows & packets & flows & packets & flows \\
-%\midrule
-%Accuracy & 0.9857054 & 0.99472587 & 0.96678023 & 0.99337596 & 0.98056073 & 0.99353386 \\
-%Precision & 0.94963342 & 0.9894422 & 0.87175275 & 0.98798596 & 0.91359787 & 0.98994595 \\
-%Recall & 0.96735895 & 0.98972051 & 0.94374958 & 0.98581328 & 0.97836672 & 0.9844478 \\
-%F1 & 0.95841423 & 0.98958134 & 0.90632359 & 0.98689842 & 0.94487366 & 0.98718922 \\
-%Youden & 0.95682951 & 0.98614231 & 0.91525628 & 0.98175163 & 0.95937772 & 0.9810602 \\
-%\bottomrule
-%\end{tabular}
-%\end{table*}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=0.98\columnwidth]{./plots/ars/ars_15.pdf}
-\caption{\gls{ars} during adversarial training for UNSW-NB15.}
-\label{fig:avdistance}
-\end{figure}
-
 \renewcommand*{\bibfont}{\small}

### 2020-01-21 learn.py
Add option to have variable tradeoff

@@ -571,6 +571,13 @@ def train_rl():
 			outputs_rl_critic = []
 
 			padded_input_data = torch.nn.utils.rnn.pad_sequence(input_data).to(device)
+
+			if not opt.variableTradeoff:
+				tradeoff = opt.accuracySparsityTradeoff
+			else:
+				tradeoff = torch.FloatTensor(batch_size).uniform_(0.0, 1.0).unsqueeze(0).to(device)
+				tradeoff_for_input = tradeoff.repeat(padded_input_data.shape[0],1).unsqueeze(2)
+				padded_input_data = torch.cat((padded_input_data, tradeoff_for_input), dim=-1)
 			batch_indices = torch.arange(batch_size)
 
 			while (remaining_seq_lens > torch.zeros_like(remaining_seq_lens)).any():
@@ -722,6 +729,9 @@ def train_rl():
 			mask_rl_two = mask_rl.repeat(1,1,2)
 			effective_output_rl_critic = outputs_catted_rl_critic[mask_rl_two].view(-1)
 
+			if opt.variableTradeoff:
+				tradeoff = tradeoff.repeat(outputs_catted_rl_critic.shape[0],1)
+
 			rewards_sparsity_catted = torch.stack(rewards_sparsity).to(device)
 			rewards_classification_catted = torch.stack(rewards_classification).to(device)
 
@@ -750,16 +760,22 @@ def train_rl():
 				effective_rl_actor_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
 
 			effective_rewards_sparsity = rewards_sparsity_catted.detach()[mask_rl_first].view(-1)
+			effective_rewards_sparsity_with_tradeoff = (tradeoff*rewards_sparsity_catted.detach())[mask_rl_first].view(-1)
 			effective_rewards_classification = rewards_classification_catted.detach()[mask_rl_first].view(-1)
 			# assert (effective_rewards_sparsity.shape == effective_rewards_classification.shape)
 
-			output_rl_critic_added = opt.accuracySparsityTradeoff*outputs_catted_rl_critic.detach()[:,:,0] + outputs_catted_rl_critic.detach()[:,:,1]
+			outputs_catted_rl_critic_sparsity = outputs_catted_rl_critic.detach()[:,:,0]
+			outputs_catted_rl_critic_classification = outputs_catted_rl_critic.detach()[:,:,1]
+
+			assert type(tradeoff) is float or outputs_catted_rl_critic_sparsity.shape == tradeoff.shape
+
[…]

### 2020-01-22 paper_rl/paper.tex
Add related work

@@
 \section{Introduction}
+\label{sec:introduction}
 
@@
 However, for practical use, not only classification accuracy and security are important but also computational efficiency.
-This follows from the fact that a rationally-acting company will only deploy an \gls{ids} if the cost that is saved by avoiding additional security breaches is higher than the cost of the \gls{ids} itself.
+This follows from the fact that a rationally-acting company will only deploy an \gls{ids} if the cost that is saved by avoiding security breaches is higher than the cost of the \gls{ids} itself.
 The cost of the \gls{ids} can be split in the cost of purchase/installation, the cost of operating it and the cost of maintaining it.
 The goal we pursue here is to develop sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data.
-For this purpose, we develop the \gls{rl}-based \gls{ids} \textit{\ours{}} that has a tunable parameter that allows trading off accuracy for sparsity of the data.
+Specifically, the sampling technique should
+\begin{enumerate}
+\item \textbf{choose optimally}, taking only the data that contain the most information and skip the less relevant ones. 
+\item have a parameter that allows to \textbf{trade off} classification performance for sparsity.
+\item be \textbf{independent of the classifier} so that any classifier can be used and that the classifier doesn't have to be aware of the sampling strategy. 
+\item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape. 
+\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets can be skipped altogether.
+\end{enumerate}
+
+For this purpose, we develop the \gls{rl}-based \gls{ids} \textit{\ours{}} fulfills the above properties.
 We show that a significant number of packets can be skipped while accuracy doesn't drop significantly.
@@
 However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.
+
+To encourage reproducibility and facilitate experimentation, we publicly release the source code, the data and the figures of this work at \url{https://github.com/CN-TU/adversarial-recurrent-ids/tree/rl}
+
+\section{Related Work}
+
+\cite{hartl_explainability_2019} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
+
+\cite{yu_learning_2017} propose using \gls{rl} to let an LSTM learn to skim text.
+Their technique does not aim to maximize sparsity but only lets the \gls{rl} optimize classification performance for the sequence as a whole.
+Thus, if it learns to skip elements it is only because it is better for achieving good classification performance, not because it wants to increase sparsity.
+
+\cite{campos_skip_2018} develop a technique that includes a \textit{skip gate} into LSTM or GRU \todo{citation needed} cells.
+It works without \gls{rl} but on the downside, their method cannot only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier.
+
+\cite{seo_neural_2018} use an LSTM with one full state vector and one reduced one.
+At each step they decide whether to use the reduced or the full state vector and can reduce computational cost by using the reduced one.
+For network traffic we think that it is often not even necessary to have a reduced state vector since from some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all.
+
+\cite{gui_long_2018} use \gls{rl} but their intention is not to skip samples but instead to make the \gls{rl} learn to choose a previous state that can aid at the current step.
+Thus it tries to help the \gls{rnn} with memorizing information from the past but does not actually take less data, which is the goal we pursue in this work.
+
+These proposed techniques focus their evaluation on text and hence do not fulfill certain criteria we set out in \autoref{sec:introduction}.
+Furthermore, all techniques that use \gls{rl} use discrete actions, meaning that there is a hyperparameter $k$, which influences the maximum jump that is possible.
+However, for a network flow, already after the second packet it might be obvious that a flow is an attack and thus the remaining packets can be completely ignored.
+It is thus beneficial to not have a fixed maximum step size but a continuous one.
 

### 2020-01-22 paper_rl/paper.tex
Small fix

@@
 \cite{campos_skip_2018} develop a technique that includes a \textit{skip gate} into LSTM or GRU \todo{citation needed} cells.
-It works without \gls{rl} but on the downside, their method cannot only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier.
+It works without \gls{rl} but on the downside, their method can only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier, while we want to be independent of it.
 

### 2020-01-22 paper_rl/paper.tex
add sampling techniques, framework structure and some description

@@
 \newacronym{rl}{RL}{Reinforcement Learning}
+\newacronym{dpi}{DPI}{Deep Packet Inspection}
 
@@
 
+\subsection{Classifier}
+\label{classifier}
+We implemented a three-layer LSTM-based classifier with 128 neurons at each layer.
+As the input features we use source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e.
+forward or reverse path) and all TCP flags (0 if the flow is not TCP).
+We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.
+Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
+During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
+We use Z-score normalization and a train/test split of 2:1.
+
+\subsection{RL for Sampling}
+Reinforcement Learning algorithms have shown that, given enough training instances, they perform optimally in decision-making situations.
+To this end, we introduce a \gls{rl}-based architecture that enables packet sampling decisions based on the current state/confidence of the LSTM classifier.
+We opt for an \emph{Actor-Critic} structure (as shown in \autoref{rnn:sampling}) which is a mixture of \emph{value-based} and \emph{policy-based} \gls{rl} strategies in which the decision of which packet to choose next (exactly how many packets to skip from the current packet) is made by the Actor deep network which, at the same time, constructs a probabilistic distribution of how many packets to skip and, tries to learn the best policy, i.e.
+action to be taken on the basis of the current state.
+The Critic network subsequently evaluates the behavior of the Actor network using the computed value function and gives it feedback based on the performance obtained by the LSTM classifier.
+The jump is taken from the Actor output distribution and fed into the packet parser, which is responsible for counting the packets and selecting the appropriate one.
+
+In what follows, we explore two technical details on how decisions are performed, either continuously or gradually.
+
+\begin{figure}
+\center
+  \includegraphics[width=0.3\textwidth]{img/rnn-sampling}
+  \caption{SparseIDS framework.
+Three deep networks are shown: an LSTM classifier, An Actor network and a Critic network.
+The Actor takes actions based on the feedback from the the Critic network and the classifier.
+The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.}
+  \label{rnn:sampling}
+\end{figure}
+
+\subsubsection{Continuous Actions}
+% here goes the math
+
+\subsubsection{Discrete Actions}
+
+\subsection{Rule-based Sampling Strategies}
+Traditionally, network packet sampling is performed on backbone infrastructures where, due to large bandwidths, not every packet can be analyzed.
+Sampling consequently helps experts understand the behavior of their network without the need to capture each and every packet.
+Assuming that the behavior of the network follows a predefined underlying process, sampling is therefore not detrimental on average to statistical analysis or measurement.
+Although the dataset in this work does not come from a backbone network, sampling reduces the high computational effort due to additional processing  required by the complex \gls{ml} based \gls{ids}.
+As a consequence, several intuitive but efficient sampling techniques can be applied in practice.
+In this study, we compare a few techniques to our sparseIDS approach.
+We opt for three rule-based sampling families that are widely used in practice, such rule-based approaches do not require feedback from higher layers (i.e.
+sampling rates are constant and not correlated with subsequent analysis).
+The strategy of each family is summarized as follows:
+
+\subsubsection{Random Sampling}
+Given the fraction $\alpha$ of packets to be selected in a set of $N$ packets, such techniques take exactly $n = \alpha \times N$ samples that are randomly sampled from a given probability mass function, such as a uniform distribution that keeps the pick-up probability similar for all packets.
+Practically, these techniques work properly if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is low.
+
+\subsubsection{First 'n' Sampling}
+As its name suggests, this technique only samples the first $n$ packets and ignores the rest of the flow.
+In such a case, the length of the entire flow is not needed and the capture can be stopped once enough packets are captured.
+Again, similar techniques work properly if the important information is present in the first few packets of a flow, i.e.
+a large data transfer in which the initial packets represent a hand-shake after which the majority of packets contain the raw data and are blind to non-\gls{dpi} systems.
+
+\subsubsection{Every 'i'th Sampling}
+Last but not least, a periodic sampling technique that takes a sample each $i= 1/ \alpha$ packet.
+Similar to the last technique, the length of the flow is not needed in this case.
+In practice, this technique is best suited to keep the behavioral distribution of the sampled set similar to the original distribution, thus suited for scenarios where information is heterogeneously distributed between packets.
+
+To begin with, the first packet of each flow is always taken for which a record is created in memory and after which the sampling routine is used.
+
+In each of the above techniques, the sequence of selected packets is then fed to the classifier discussed in \ref{classifier}, where features are extracted and a decision is made after each packet.
+Finally, the overall accuracy is calculated after the last packet in the sequence is received.
+
+There exists many more rule-based sampling techniques i.e., temporal, adaptive etc. However, studying their behaviors is out of the scope of this paper.
+
+\subsection{Dataset}
 % \begin{table}[b]
@@
 
-We implemented a three-layer LSTM-based classifier with 128 neurons at each layer. As the input features we use
-source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e.
-forward or reverse path) and all TCP flags (0 if the flow is not TCP).
-We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behaviour \cite{bachl_walling_2019}.
-Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
-During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
-We use Z-score normalization and a train/test split of 2:1.
+%For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin2018toward} dataset which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
 
-%For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
-For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
-
-Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
+The distribution of different attacks in the datasets are shown in \autoref{tab:occurrence}.
 
@@
 %\end{table}
+
+\section{Results and Discussion}
 

### 2020-01-23 paper_rl/paper.tex
edit sampling strategies and framework figure

@@
 \center
-  \includegraphics[width=0.3\textwidth]{img/rnn-sampling}
+  \includegraphics[width=0.3\textwidth]{img/rnn-sampling.pdf}
   \caption{SparseIDS framework.
@@
 \subsubsection{Random Sampling}
-Given the fraction $\alpha$ of packets to be selected in a set of $N$ packets, such techniques take exactly $n = \alpha \times N$ samples that are randomly sampled from a given probability mass function, such as a uniform distribution that keeps the pick-up probability similar for all packets.
-Practically, these techniques work properly if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is low.
+%Given the fraction $\alpha$ of packets to be selected in a set of $N$ packets, such techniques take exactly $n = \alpha \times N$ samples that are randomly sampled from a given probability mass function, such as a uniform distribution that keeps the pick-up probability similar for all packets.
+%Practically, these techniques work properly if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is low.
 
-\subsubsection{First 'n' Sampling}
+Given the fraction $\alpha$ of packets to be selected in a sequence of $N$ packets, these techniques can sequentially sample or skip a packet with a probability of $p_{sample}=\alpha$ and $p_{skip}=1-\alpha$ respectively.
+If the sequence is large enough, $n$ packets will be randomly sampled from a uniform distribution whereby $n= \alpha \times N$.
+Practically, these techniques work properly if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
+
+\subsubsection{(Equal) First 'n' Sampling}
 As its name suggests, this technique only samples the first $n$ packets and ignores the rest of the flow.
-In such a case, the length of the entire flow is not needed and the capture can be stopped once enough packets are captured.
+Nevertheless, there are two variants of this technique: (a) taking the first $n$ packets, where $n$ is the same for all flows, and (b) taking the first ``Equal'' proportion of packets from each flow (e.g.
+half of the flow, regardless of its length).
+In the first example, the size of the entire flow is not necessary and the capture can be stopped once enough packets are captured.
+In the second scenario, however, the length of each flow needs to be known.
 Again, similar techniques work properly if the important information is present in the first few packets of a flow, i.e.

### 2020-01-23 paper_rl/neural_network.tex
Add neural_network overview

\documentclass[crop]{standalone}
\usepackage{libertine}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,backgrounds}
\usepackage{amsmath}
\begin{document}
\newcommand{\nodedistance}{1.2}%
\newcommand{\numUnits}{128}%
%\pgfmathsetmacro{\shortside}{sqrt((\nodedistance * \nodedistance ) /2)}%
\pgfmathsetmacro{\shortside}{sqrt(2)/2 * \nodedistance}%
%\pgfmathsetmacro{\solutionx}{ \shortside }%
%\pgfmathsetmacro{\solutiony}{ - (3 * \shortside + 3 * \nodedistance ) }%
%\pgfmathsetmacro{\solutiondurationx}{ \shortside }%
%\pgfmathsetmacro{\solutiondurationy}{ - (3 * \shortside + 1 * \nodedistance ) }%
\tikzstyle{texto} = [above, text width=6em, text centered]%
\newcommand{\backgroundna}[3]{%%
  \begin{pgfonlayer}{background}%
    \path (#1.south -| #1.west)+(-0.5,-0.5) node (a1) {};%
    \path (#2.north -| #2.east)+(+0.5,+0.5) node (a2) {};%
    \path[rounded corners, draw=black!50]%
      (a1) rectangle (a2);%
    \path (a1.west |- a2.north)+(0.25,-0.25) node (u1)[anchor=north west]%
      {\textit{#3}};%
  \end{pgfonlayer}}%
\newcommand{\backgroundnc}[4]{%%
  \begin{pgfonlayer}{background}%
    \path (#1.south -| #4.west)+(-0.5,-0.5) node (a1) {};%
    \path (#2.north -| #2.east)+(+0.5,+0.5) node (a2) {};%
    \path[rounded corners, draw=black!50]%
      (a1) rectangle (a2);%
    \path (a1.east |- a1.south)+(2.5,0.25) node (u1)[anchor=south west]%
      {\textit{#3}};%
  \end{pgfonlayer}}%
\newcommand{\backgroundnclass}[3]{%%
  \begin{pgfonlayer}{background}%
    \path (#1.south -| #1.west)+(-0.5,-0.8) node (a1) {};%
    \path (#2.north -| #2.east)+(+0.5,+0.5) node (a2) {};%
    \path[rounded corners, draw=black!50]%
      (a1) rectangle (a2);%
    \path (a2.west |- a1.north)+(0.,-0.) node (u1)[anchor=south east]%
      {\textit{#3}};%
  \end{pgfonlayer}}%
  \tikzstyle{place}=[circle,thick,draw=blue!75,fill=blue!20,minimum size=6mm]%
  \tikzstyle{input place}=[circle,thick,draw=olive!75,fill=olive!20,minimum size=6mm]%
  \tikzstyle{output place}=[circle,thick,draw=magenta!75,fill=magenta!20,minimum size=6mm]%
   \tikzstyle{lstm place}=[circle,thick,draw=red!75,fill=red!20,minimum size=6mm]%
   \tikzstyle{softplus place}=[circle,thick,draw=green!75,fill=green!20,minimum size=6mm]%
  \tikzstyle{transition}=[rectangle,thick,draw=black!75,%
  			  fill=black!20,minimum size=4mm]%
\begin{tikzpicture}[every label/.append style={font=\tiny}, node distance=\nodedistance cm, >=stealth',bend angle=45]
\begin{scope}
\node [input place] (s1) [label={[align=center]left: State (8)}] {};
\node [place] (s2) [above right of=s1, label={[align=center]above: Linear}] {};
\path[->] (s1) edge node {} (s2);
\node [lstm place] (s3) [right of=s2, label={[align=center]below: LSTM (\numUnits{})}] {};
\path[->] (s2) edge node {} (s3); 
\path[->] (s3) edge [loop above] node {} (s3);   
\node [lstm place] (s4) [right of=s3, label={[align=center]below: LSTM (\numUnits{})}] {};    
\path[->] (s3) edge node {} (s4);
\path[->] (s4) edge [loop above] node {} (s4);
\node [lstm place] (s9) [right of=s4, label={[align=center]below: LSTM (\numUnits{})}] {};    
\path[->] (s4) edge node {} (s9);
\path[->] (s9) edge [loop above] node {} (s9);   
\node [place] (s7) [right of=s9, label={[align=center]below:{ Linear}}] {};    
\path[->] (s9) edge node {} (s7);    
\node [output place] (s8) [right of=s7, label={[align=center]below: $\sigma$~(1)}] {};    
\path[->] (s7) edge node {} (s8);
\node [place] (s5) [above of=s7, label={[align=center]above: Linear}] {};
\path[->] (s9) edge node {} (s5);
\node [output place] (s6) [right of=s5, label={[align=center]above: $\mu$ (1)}] {};        
\backgroundna{s2}{s6}{Actor network}
\path[->] (s5) edge node {} (s6);        
\node [place] (s10) [below right of=s1, label={[align=center]below: Linear}] {};
\path[->] (s1) edge node {} (s10);    
\node [lstm place] (s11) [right of=s10, label={[align=center]above: LSTM (\numUnits{})}] {};
\path[->] (s10) edge node {} (s11);
\path[->] (s11) edge [loop below] node {} (s11);
\node [lstm place] (s12) [right of=s11, label={[align=center]above: LSTM (\numUnits{})}] {};
\path[->] (s11) edge node {} (s12);
\path[->] (s12) edge [loop below] node {} (s12);    
\node [lstm place] (s21) [right of=s12, label={[align=center]above: LSTM (\numUnits{})}] {};
\path[->] (s12) edge node {} (s21);
\path[->] (s21) edge [loop below] node {} (s21);    
\node [place] (s13) [right of=s21, label={[align=center]above: Linear}] {};
\path[->] (s21) edge node {} (s13);
\node [output place] (s17) [right of=s13, label={[align=center]above: $v_\text{classification}$ (1)}] {};
\path[->] (s13) edge node {} (s17);
\node [place] (s16) [below of=s13, label={[align=center]below: Linear}] {};
\path[->] (s21) edge node {} (s16);    
\node [output place] (s20) [right of=s16, label={[align=center]below: $v_\text{sparsity}$ (1)}] {};    
\path[->] (s16) edge node {} (s20);
\backgroundnc{s20}{s17}{Critic network}{s10}
\node [place] (s30) at (\shortside,-3*\shortside-\nodedistance) [label={[align=center]below: Linear}] {};
\path[->] (s1) edge node {} (s30);
\node [lstm place] (s31) [right of=s30, label={[align=center]above: LSTM (\numUnits{})}] {};
\path[->] (s30) edge node {} (s31);
\path[->] (s31) edge [loop below] node {} (s31);
\node [lstm place] (s32) [right of=s31, label={[align=center]above: LSTM (\numUnits{})}] {};
\path[->] (s31) edge node {} (s32);
\path[->] (s32) edge [loop below] node {} (s32);    
\node [lstm place] (s33) [right of=s32, label={[align=center]above: LSTM (\numUnits{})}] {};
\path[->] (s32) edge node {} (s33);
\path[->] (s33) edge [loop below] node {} (s33);  
\node [place] (s34) [right of=s33, label={[align=center]above: Linear}] {};
\path[->] (s33) edge node {} (s34);    
\node [output place] (s35) [right of=s34, label={[align=center]above: confidence (1)}] {};    
\path[->] (s34) edge node {} (s35);
\backgroundnclass{s30}{s35}{Classifier};
\end{scope}
\end{tikzpicture}
\end{document}

### 2020-01-23 paper_rl/paper.tex
Changed detail in draw.io illustration

@@
 Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
+Additionally, we add the number of skipped packets since the last packet as a feature, which can help the classifier because otherwise it might be confused because of missing packets. 
 During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
@@
 
-In what follows, we explore two technical details on how decisions are performed, either continuously or gradually.
+In what follows, we explore two technical details on how decisions are performed, either continuously or descretely.
 
 \begin{figure}
-\center
-  \includegraphics[width=0.3\textwidth]{img/rnn-sampling.pdf}
+\centering
+  \includegraphics[width=\columnwidth]{img/rnn-sampling.pdf}
   \caption{SparseIDS framework.
@@
   \label{rnn:sampling}
+\end{figure}
+
+\begin{figure}
+\centering
+  \includegraphics[width=\columnwidth]{neural_network.pdf}
+  \caption{An overview of the complete neural network being used in our experiments.
+Ocher stands for the input, blue for linear layers, red for LSTM layers and purple for the outputs of the neural network.
+The numbers in parentheses next to each label stand for the width of that layer (i.e.~the number of neurons (except for the inputs and outputs)).
+Parts of the neural network are shared between $v_\text{classification}$ and $v_\text{sparsity}$ as well as $\mu$ and $\sigma$ as they output quantities of similar magnitude and thus the gradients that are computed when updating the neural network are of the same order of magnitude as well.
+Our implementation also supports sharing the LSTM between all three neural network.
+However, as the five outputs can produce losses that have different order of magnitude, larger gradients or larger losses would ``overwrite'' the gradients of the smaller ones.}
+  \label{rnn:neuralNetworkArchitecture}
 \end{figure}

### 2020-01-23 learn.py

@@ -2197,7 +2197,9 @@ if __name__=="__main__":
 	dataset = OurDataset(x, y, categories)
 
 	batchSize = 1 if opt.function == 'pred_plots' else opt.batchSize # FIXME Max: Why? What's wrong?
-	lstm_module = OurLSTMModule(x[0].shape[-1] + int(opt.variableTradeoff), y[0].shape[-1], opt.hidden_size, opt.n_layers, batchSize, device, num_outputs_actor=opt.lookaheadSteps if not opt.continuous else 2, num_outputs_critic=2).to(device)
+	input_dim = x[0].shape[-1] + int(opt.variableTradeoff)
+	print("input_dim", input_dim)
+	lstm_module = OurLSTMModule(input_dim, y[0].shape[-1], opt.hidden_size, opt.n_layers, batchSize, device, num_outputs_actor=opt.lookaheadSteps if not opt.continuous else 2, num_outputs_critic=2).to(device)
 
 	if opt.net != '':
 		print("Loading", opt.net)
@@ -2210,8 +2212,8 @@ if __name__=="__main__":
 			lstm_module.load_state_dict(model_dict)
 
 	if ("rl" in opt.function or opt.sampling=="rl") and not opt.shareNet:
-		lstm_module_rl_actor = OurLSTMModule(x[0].shape[-1] + int(opt.variableTradeoff), opt.lookaheadSteps if not opt.continuous else 2, opt.hidden_size, opt.n_layers, batchSize, device).to(device)
-		lstm_module_rl_critic = OurLSTMModule(x[0].shape[-1] + int(opt.variableTradeoff), 2, opt.hidden_size, opt.n_layers, batchSize, device).to(device)
+		lstm_module_rl_actor = OurLSTMModule(input_dim, opt.lookaheadSteps if not opt.continuous else 2, opt.hidden_size, opt.n_layers, batchSize, device).to(device)
+		lstm_module_rl_critic = OurLSTMModule(input_dim, 2, opt.hidden_size, opt.n_layers, batchSize, device).to(device)
 
 		if opt.net_actor != '':
 			print("Loading actor", opt.net_actor)

### 2020-01-23 paper_rl/paper.tex
Modifications to draw.io fig

@@
 \section{Setup}
-
+%ci
 \subsection{Classifier}
@@
 The Actor takes actions based on the feedback from the the Critic network and the classifier.
-The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.}
+The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.
+\todo{Max: I don't like the icon that is used for the distribution.
+It's a bit pixelated.}}
   \label{rnn:sampling}

### 2020-01-24 paper_rl/paper.tex
rephrased sentences in Intro and left them commented

@@
 
+% Fares:
+%As the need for security mechanisms in today's infrastructure communication networks grows, a substantial body of scientific work focuses on developing reliable, explainable frameworks for the identification of unwanted behaviors.
+In the past, a viable way to detect intrusion was to analyze the contents of the packets themselves and determine, for example, whether a packet contains potentially harmful content by matching patterns.
+More recently, with the increasing deployment of encryption, the focus is now on features that are always available for network monitoring devices such as port numbers, protocol flags, packet sizes when encrypting above the transport layer.
+
 Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties.
@@
 Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
-While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows.
+%While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows.
 Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
-At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
+%At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
+%\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
+
+% Fares:
+%A network communication is typically aggregated into \textit{flows}, which are commonly defined as a sequence of packets that share certain properties.
+When analyzing flows, not only the aforementioned features are available but also features representing the temporal behavior of the individual packets.
+Various approaches have been proposed to extract flows features after which anomaly detection can be performed, in \cite{meghdouri_analysis_2018} Meghdouri et al.
+compare such approaches.
+%While these approaches often work well, it is troublesome because, in the worst case, the whole flow must be received first and only then anomaly detection can be applied; revealing malicious flows.
+Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
+%At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
+%In \cite{hartl_explainability_2019}, Hartl et al.
+show that such an architecture has similar performance to other flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends and analyzed its robustness with respect to adversaries extensively.
 
@@
 Specifically, the sampling technique should
+
+% Fares:
+%Nevertheless, not only the accuracy and reliability of classification, but also the computational efficiency, are critical for practical use.
+This follows from the fact that a rationally-acting organization will only implement an \gls{ids} if the cost saved by preventing security breaches is higher than the cost of the \gls{ids} itself.
+Noteworthy that the cost of an \gls{ids} can be divided into (a) the cost of purchase/installation, (b) the cost of operation and (c) the cost of maintenance.
+The goal we pursue here is to develop autonomous sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data.
+Specifically, the sampling technique should:
+
 \begin{enumerate}
 \item \textbf{choose optimally}, taking only the data that contain the most information and skip the less relevant ones. 
+% Fares:
+%\item \textbf{choose optimally}, taking only samples that contain the most information and skip the less relevant ones. 
 \item have a parameter that allows to \textbf{trade off} classification performance for sparsity.
@@
 
-As \ours{} uses a parameter to trade off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
+% Fares:
+%For this purpose we develop a \gls{rl}-based \gls{ids}: \textit{\ours{}} that fulfills the above properties.
+We show that a significant number of packets can be skipped while the accuracy does not drop
+considerably. Compared to other common sampling techniques, \ours{} performs better when being trained on the same number of packets.
+
+As \ours{} uses a parameter to trade-off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
 However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.
+
+% Fares:
+%The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both later quantities.
+However, in certain situations, a certain sparsity is to be achieved, hence we further develop and propose a steering system that fluctuates such a trade-off parameter in order to obtain a coveted sparsity in a closed control loop.
 

### 2020-01-24 paper_rl/paper.tex
uncomment some sections

@@
 Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
-%While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows.
+While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows.
 Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
-%At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-%\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
+At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
+\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
 

### 2020-01-24 paper_rl/paper.tex
comment a section
Changes for RL section

--- 2020-01-23 paper_rl/paper.tex
@@
 
+% Adds a "given" symbol (vertical bar) that rescales in height (like \left and \right)
+\usepackage{mathtools}
+\newcommand\givenbase[1][]{\:#1\lvert\:}
+\let\given\givenbase
+\newcommand\sgiven{\givenbase[\delimsize]}
+\DeclarePairedDelimiterX\Basics[1](){\let\given\sgiven #1}
+\newcommand\Average{E\Basics}
+
 %%
@@
 \newacronym{dpi}{DPI}{Deep Packet Inspection}
+\newacronym{ac}{AC}{Actor-Critic}
 
@@
 
-\section{Setup}
-%ci
-\subsection{Classifier}
-\label{classifier}
+%\section{Setup}
+%%ci
+\section{Classifier}
+\label{sec:classifier}
 We implemented a three-layer LSTM-based classifier with 128 neurons at each layer.
@@
 
-\subsection{RL for Sampling}
-Reinforcement Learning algorithms have shown that, given enough training instances, they perform optimally in decision-making situations.
-To this end, we introduce a \gls{rl}-based architecture that enables packet sampling decisions based on the current state/confidence of the LSTM classifier.
-We opt for an \emph{Actor-Critic} structure (as shown in \autoref{rnn:sampling}) which is a mixture of \emph{value-based} and \emph{policy-based} \gls{rl} strategies in which the decision of which packet to choose next (exactly how many packets to skip from the current packet) is made by the Actor deep network which, at the same time, constructs a probabilistic distribution of how many packets to skip and, tries to learn the best policy, i.e.
-action to be taken on the basis of the current state.
-The Critic network subsequently evaluates the behavior of the Actor network using the computed value function and gives it feedback based on the performance obtained by the LSTM classifier.
-The jump is taken from the Actor output distribution and fed into the packet parser, which is responsible for counting the packets and selecting the appropriate one.
+\section{RL for Sampling}
+\label{sec:rlForSampling}
+Given enough training instances, \gls{rl} algorithms have been shown to perform very well in complex decision-making situations such as playing video games \cite{mnih_playing_2013} as well as networking problems such as congestion control \cite{bachl_rax_2019}.
+We thus consider an \gls{rl}-based architecture to be suitable for taking packet sampling decisions.
+The system should look at the current packet as well as the history of the conversation flow and output how many packets should be skipped until the next packet is considered in the current flow.
+The reward that the \gls{rl} tries to maximize should be a combination of classification accuracy and the number of packets that could be skipped (sparsity).
+An operator should be able to specify how much accuracy is allowed to be traded off for achieving higher sparsity.% and the objective of the \gls{rl} at a timestep t of a flow is defined as:
 
-In what follows, we explore two technical details on how decisions are performed, either continuously or descretely.
+%\begin{align}
+%\begin{split}
+%U_t = & \text{future average classification performance} \\
+%+ & \alpha\times\text{future average sparsity}
+%\end{split}
+%\end{align}
+
+For a flow of length $T$ at each timestep $t$ we have two reward metrics which we compute in a fashion that is known from \textit{R-learning} \cite{schwartz_reinforcement_1993}:
+
+\todo{Max: Add figure}
+The classification reward metric captures how correct the classifier is on average for all future packets of a flow.
+The label of each packet in a flow has a value of 0 for non-attack traffic and 1 for attack traffic with one flow being only attack or non-attack but not mixed.
+The confidence is the sigmoided ($\frac{e^x}{1+e^x}$) output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack.
+\begin{align}
+\begin{split}
+R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |\text{label}_t - \text{confidence}_t|
+\end{split}
+\end{align}
+
+The sparsity reward metric captures how much sparsity is achieved on average for all future packets of a flow.
+E.g.
+for a flow of length 10 if after timestep 5, 3 packets are chosen and 2 are skipped, the sparsity reward metric at timestep 5 is $\frac{2}{2+3}=\frac{2}{5}$.
+\begin{align}
+\begin{split}
+R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},t}}{\sum_{i=t+1}^{T} 1_{\text{chosen},t}+\sum_{i=t+1}^{T} 1_{\text{skipped},t}}
+\end{split}
+\end{align}
+
+We opt for an \emph{\gls{ac}} approach (as shown in \autoref{fig:sampling}) modeled after \cite{mnih_asynchronous_2016} because we consider it being more interpretable than a Deep \gls{rl} approach based on a \textit{Deep Q Network} like \cite{mnih_playing_2013} for reasons we explain later \todo{Max: Add ref to the section where we explain this}.
+
+Our proposed framework consists of three independent neural networks: the classifier, the critic and the actor.
+
+The classifier's goal is to guess correctly if a flow is an attack or not.
+For this, it outputs its confidence for non-attack/attack after reading a packet.
+As stated in \autoref{sec:classifier}, we choose an LSTM-based supervised classifier for this work, but our framework can work together with any classifier as long as it outputs a confidence for non-attack/attack at each timestep and can handle skipped packets.
+For instance it would also be possible to replace the LSTM with an unsupervised autoencoder like in \cite{mirsky_kitsune_2018}.
+
+The Critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a timestep $t$.
+For this it outputs $v_{\text{classification},t}$ and $v_{\text{sparsity},t}$ at each timestep $t$.
+It aims to predict the classification performance and sparsity correctly for each timestep by minimizing the following loss function:
+\begin{align}
+\begin{split}
+l_{\text{v},t} =& \left(R_{\text{classification},t} - v_{\text{sparsity},t}\right)^2 + \left(R_{\text{sparsity},t} - v_{\text{sparsity},t}\right)^2
+\end{split}
+\end{align}
+
+Now that the reward metrics and the Critic's outputs are defined, we can define the overall utility that we want to maximize at each timestep $t$ as follows:
+\begin{align}
+\begin{split}
+U_t =& \left(R_{\text{classification},t} + \alpha \cdot R_{\text{sparsity},t}\right) \\
+- & \left(v_{\text{classification},t} + \alpha \cdot v_{\text{sparsity},t}\right)
+\end{split}
+\end{align}
+
+This means that we want that we want the reward to be higher than the Critic's expectation.
+
+The Actor outputs a probability distribution at timestep $t$, which is sampled to determine the number of packets that should be skipped.
+It aims to change the distribution so that actions which result in a higher reward are chosen more frequently, maximizing the utility function.
+However, this strategy could lead to the Actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward.
+Thus, the Actor not only tries to change the distribution so that the actions become more optimal but at the same time also aims to maximize the entropy of the distribution so that alternative choices are still explored reasonably often.
+
+Specifically, at timestep $t$, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t$, the number of packets to be skipped, is sampled.
+Besides wanting to make actions which result in a high reward more likely, the actor also seeks to optimize the entropy $H(\cdot)$ of the probability distribution so that it keeps exploring and doesn't get stuck with a suboptimal policy.
+
+Combining these objectives, the actor network aims to minimize the loss function
+\begin{align}
+\begin{split}
+l_{\text{a},t} =& -\log \left( \pi \left( a_t \given s_t ; \theta_\text{a} \right) \right) U_t \\ 
+&- \beta H\left( \pi\left( s_t; \theta_a \right)\right)
+\end{split}
+\end{align}
+
+%The Critic network evaluates the behavior of the Actor network and computes the classification performance and sparsity that it thinks is going to be achieved.% using the computed value function and gives it feedback based on the performance obtained by the LSTM classifier.
+%The number of packets to skip is taken from the Actor's output distribution and fed into the packet parser, which is responsible for counting the packets and selecting the appropriate one.
+
+In related work regarding \gls{rl}-based sampling for sequences, only discrete actions were evaluated so far: The actor would output a categorical probability distribution with a fixed maximum number of bins $k$.
+A drawback of this approach is that only a predefined maximum number of packets can be skipped and that that for a very large number of $k$, it can take a long time to converge to an optimum policy as there are $k$ different options to explore at each timestep $t$.
+Besides discrete actions, we thus also experiment with continuous actions, which allow to skip arbitrarily many packets.
 
@@
 The Actor takes actions based on the feedback from the the Critic network and the classifier.
-The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.
-\todo{Max: I don't like the icon that is used for the distribution.
-It's a bit pixelated.}}
-  \label{rnn:sampling}
+The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.}
+  \label{fig:sampling}
 \end{figure}
@@
 However, as the five outputs can produce losses that have different order of magnitude, larger gradients or larger losses would ``overwrite'' the gradients of the smaller ones.}
-  \label{rnn:neuralNetworkArchitecture}
+  \label{fig:neuralNetworkArchitecture}
 \end{figure}
 
-\subsubsection{Continuous Actions}
-% here goes the math
-
-\subsubsection{Discrete Actions}
+\subsection{Continuous Actions}
+\todo{Max: Add formulas for the Log-Normal distribution}
 
@@
 
-In each of the above techniques, the sequence of selected packets is then fed to the classifier discussed in \ref{classifier}, where features are extracted and a decision is made after each packet.
+In each of the above techniques, the sequence of selected packets is then fed to the classifier discussed in \autoref{sec:classifier}, where features are extracted and a decision is made after each packet.
 Finally, the overall accuracy is calculated after the last packet in the sequence is received.
 
-There exists many more rule-based sampling techniques i.e., temporal, adaptive etc. However, studying their behaviors is out of the scope of this paper.
+There exist many more rule-based sampling techniques e.g., temporal, adaptive etc. However, studying their behaviors is out of the scope of this paper.
 
@@
 %For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
-For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin2018toward} dataset which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
 

### 2020-01-26 paper_rl/paper.tex
Added some text to the continuous section

@@
 
+% Fares:
+%A network communication is typically aggregated into \textit{flows}, which are commonly defined as a sequence of packets that share certain properties.
+When analyzing flows, not only the aforementioned features are available but also features representing the temporal behavior of the individual packets.
+Various approaches have been proposed to extract flows features after which anomaly detection can be performed, in \cite{meghdouri_analysis_2018} Meghdouri et al.
+compare such approaches.
+%While these approaches often work well, it is troublesome because, in the worst case, the whole flow must be received first and only then anomaly detection can be applied; revealing malicious flows.
+Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
+%At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
+%In \cite{hartl_explainability_2019}, Hartl et al.
+show that such an architecture has similar performance to other flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends and analyzed its robustness with respect to adversaries extensively.
+
 However, for practical use, not only classification accuracy and security are important but also computational efficiency.
@@
 
-As \ours{} uses a parameter to trade off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
+% Fares:
+%For this purpose we develop a \gls{rl}-based \gls{ids}: \textit{\ours{}} that fulfills the above properties.
+We show that a significant number of packets can be skipped while the accuracy does not drop
+%considerably. Compared to other common sampling techniques, \ours{} performs better when being trained on the same number of packets.
+
+As \ours{} uses a parameter to trade-off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
 However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.
+
+% Fares:
+%The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both later quantities.
+However, in certain situations, a certain sparsity is to be achieved, hence we further develop and propose a steering system that fluctuates such a trade-off parameter in order to obtain a coveted sparsity in a closed control loop.
 
@@
 
-\todo{Max: Add figure}
+\begin{figure}
+\centering
+  \includegraphics[width=\columnwidth]{img/rewards.pdf}
+  \caption{The computation of the rewards for a sample flow: Gray packets are skipped while white ones are chosen.
+At the top row the resulting rewards for classification for each chosen packet are depicted and in the bottom row the ones for sparsity.
+Numbers are rounded to two digits after the decimal point.}
+  \label{fig:rewards}
+\end{figure}
+
 The classification reward metric captures how correct the classifier is on average for all future packets of a flow.
@@
 &- \beta H\left( \pi\left( s_t; \theta_a \right)\right)
-\end{split}
+\end{split}\label{eq:actor}
 \end{align}
@@
 Besides discrete actions, we thus also experiment with continuous actions, which allow to skip arbitrarily many packets.
-
-\begin{figure}
-\centering
-  \includegraphics[width=\columnwidth]{img/rnn-sampling.pdf}
-  \caption{SparseIDS framework.
-Three deep networks are shown: an LSTM classifier, An Actor network and a Critic network.
-The Actor takes actions based on the feedback from the the Critic network and the classifier.
-The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.}
-  \label{fig:sampling}
-\end{figure}
 
@@
 
+An overview of the neural networks of the classifier, actor and critic that we use in this work is given in \autoref{fig:neuralNetworkArchitecture}.
+
+\begin{figure}
+\centering
+  \includegraphics[width=\columnwidth]{img/rnn-sampling.pdf}
+  \caption{SparseIDS framework.
+Three deep networks are shown: an LSTM classifier, An Actor network and a Critic network.
+The Actor takes actions based on the feedback from the the Critic network and the classifier.
+The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.}
+  \label{fig:sampling}
+\end{figure}
+
+A schematic overview over the \gls{rl} for sampling is given in \autoref{fig:sampling}.
+
 \subsection{Continuous Actions}
-\todo{Max: Add formulas for the Log-Normal distribution}
+
+For performing continious actions we need a probability distribution which can be parametrized to have the mean between 0 and positive infinity.
+Furthermore it must have at least two parameters so that the standard deviation, which is required for the entropy (\autoref{eq:actor}), can be changed independent of the mean.
+For example, the Exponential distribution can be parametrized to have its mean anywhere between 0 and positive infinity, however, it has only one parameter and thus the mean and the standard deviation cannot be independently changed.
+Conversely, a simple distribution which fits our requirements is the Log-normal distribution, which is defined as a normal distribution whose output is exponentiated:
+
+Let $Z$ be standard normally distributed and $\mu_{\text{normal}}$ and $\sigma_{\text{normal}} > 0$ be two real numbers. Then 
+\begin{align}
+X = e^{\mu_{\text{normal}}+\sigma_{\text{normal}} Z}
+\end{align}
+is log-normally distributed.
+
+The actor network (\autoref{fig:neuralNetworkArchitecture}) shows that the actor network has two outputs: $\mu$ and $\sigma$.
+These encode the mean and standard deviation of a log-normal distribution.
+Since the mean and the standard deviation of a log-normal distribution can never be smaller or equal to zero, we always apply the softplus function to them, so that the actor can only output $\mu$ and $\sigma$ that are larger than 0:
+\begin{align}
+\text{softplus}(x) = \log\left(1 + e^x \right)
+\end{align}
+To parametrize the normal distribution that underlies the log-normal distribution, we first have to compute the parameters $\mu_{\text{normal}}$ and $\sigma_{\text{normal}}$ from $\mu$ and $\sigma$.
+We can compute them as follows:
+
+\begin{align}
+\mu_{\text{normal}} = \frac{1}{2}\left( 2\log(\mu) - \log\left(\frac{\mu^2 + \sigma^2}{\mu^2}\right) \right)
+\end{align}
+
+\begin{align}
+\sigma_{\text{normal}} = \sqrt{\log\left(\frac{\mu^2 + \sigma^2}{\mu^2}\right)}
+\end{align}
 

### 2020-01-27 paper_rl/paper.tex
Fix typo
minor reorganisation and corrections

@@
 %%ci
-\section{Classifier}
-\label{sec:classifier}
+
+\section{\ours{}}
+
+Traditionally, network packet sampling is performed on backbone infrastructures where, due to large bandwidths, not every packet can be analyzed.
+Sampling consequently helps experts understand the behavior of their network without the need to capture each and every packet.
+Assuming that the behavior of the network follows a predefined underlying process, sampling is therefore not detrimental on average to statistical analysis or measurement.
+Although the dataset in this work does not come from a backbone network, sampling reduces the high computational effort due to additional processing required by the complex \gls{ml}-based \glspl{ids}.
+As a consequence, several intuitive but efficient sampling techniques can be applied in practice.
+In this Section, we discuss the architecture of our proposed sampling framework \ours.
+
+\subsection{Classifier}
+\label{subsec:classifier}
 We implemented a three-layer LSTM-based classifier with 128 neurons at each layer.
@@
 
-\section{RL for Sampling}
-\label{sec:rlForSampling}
+\subsection{Reinforcement Learning-based Sampling}
+\label{subsec:rlForSampling}
+
 Given enough training instances, \gls{rl} algorithms have been shown to perform very well in complex decision-making situations such as playing video games \cite{mnih_playing_2013} as well as networking problems such as congestion control \cite{bachl_rax_2019}.
@@
 
-The classification reward metric captures how correct the classifier is on average for all future packets of a flow.
+The classification reward metric (\autoref{eq:R_classification}) captures how correct the classifier is on average for all future packets of a flow.
 The label of each packet in a flow has a value of 0 for non-attack traffic and 1 for attack traffic with one flow being only attack or non-attack but not mixed.
@@
 \begin{split}
-R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |\text{label}_t - \text{confidence}_t|
+R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |\text{label}_i - \text{confidence}_i|
 \end{split}
+\label{eq:R_classification}
 \end{align}
 
-The sparsity reward metric captures how much sparsity is achieved on average for all future packets of a flow.
+The sparsity reward metric (\autoref{eq:R_sparsity}) captures how much sparsity is achieved on average for all future packets of a flow.
 E.g.
-for a flow of length 10 if after timestep 5, 3 packets are chosen and 2 are skipped, the sparsity reward metric at timestep 5 is $\frac{2}{2+3}=\frac{2}{5}$.
+for a flow of length 10 if after timestep 5, 3 packets are chosen and 2 are skipped from what is left, the sparsity reward metric at timestep 5 is $\frac{2}{2+3}=\frac{2}{5}$.
+\note{blue}{Fares: This equation can be shortened by diving by the total count of packets left.\\
+Fares: If the length of the flow is not known, how do we estimate T}.
 \begin{align}
 \begin{split}
-R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},t}}{\sum_{i=t+1}^{T} 1_{\text{chosen},t}+\sum_{i=t+1}^{T} 1_{\text{skipped},t}}
+R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},i}}{\sum_{i=t+1}^{T} 1_{\text{chosen},i}+\sum_{i=t+1}^{T} 1_{\text{skipped},i}}
 \end{split}
+\label{eq:R_sparsity}
 \end{align}
+
+\autoref{fig:rewards} shows how classification and sparsity rewards are computed for a toy example.
 
@@
 For this, it outputs its confidence for non-attack/attack after reading a packet.
-As stated in \autoref{sec:classifier}, we choose an LSTM-based supervised classifier for this work, but our framework can work together with any classifier as long as it outputs a confidence for non-attack/attack at each timestep and can handle skipped packets.
+As stated in \autoref{subsec:classifier}, we choose an LSTM-based supervised classifier for this work, but our framework can work together with any classifier as long as it outputs a confidence for non-attack/attack at each timestep and can handle skipped packets.
 For instance it would also be possible to replace the LSTM with an unsupervised autoencoder like in \cite{mirsky_kitsune_2018}.
@@
 For this it outputs $v_{\text{classification},t}$ and $v_{\text{sparsity},t}$ at each timestep $t$.
-It aims to predict the classification performance and sparsity correctly for each timestep by minimizing the following loss function:
+It aims to predict the classification performance and sparsity correctly for each timestep by minimizing the following value loss function:
 \begin{align}
 \begin{split}
-l_{\text{v},t} =& \left(R_{\text{classification},t} - v_{\text{sparsity},t}\right)^2 + \left(R_{\text{sparsity},t} - v_{\text{sparsity},t}\right)^2
+l_{\text{v},t} =& \left(R_{\text{classification},t} - v_{\text{classification},t}\right)^2 \\
++ & \left(R_{\text{sparsity},t} - v_{\text{sparsity},t}\right)^2
 \end{split}
@@
 
-Now that the reward metrics and the Critic's outputs are defined, we can define the overall utility that we want to maximize at each timestep $t$ as follows:
+Now that the reward metrics and the Critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each timestep $t$ as follows:
 \begin{align}
@@
 
-This means that we want that we want the reward to be higher than the Critic's expectation.
+This means that we want the reward to be higher than the Critic's expectation.
 
@@
 It aims to change the distribution so that actions which result in a higher reward are chosen more frequently, maximizing the utility function.
-However, this strategy could lead to the Actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward.
-Thus, the Actor not only tries to change the distribution so that the actions become more optimal but at the same time also aims to maximize the entropy of the distribution so that alternative choices are still explored reasonably often.
+However, this strategy could lead to the Actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward (local maxima).
+Thus, the Actor not only tries to change the distribution so that the actions become more optimal (higher reward) but at the same time also aims to maximize the entropy of the distribution so that alternative choices are still explored reasonably often.
 
-Specifically, at timestep $t$, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t$, the number of packets to be skipped, is sampled.
+Specifically, at timestep $t$, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t$ (the number of packets to be skipped) is sampled.
 Besides wanting to make actions which result in a high reward more likely, the actor also seeks to optimize the entropy $H(\cdot)$ of the probability distribution so that it keeps exploring and doesn't get stuck with a suboptimal policy.
@@
 \begin{split}
-l_{\text{a},t} =& -\log \left( \pi \left( a_t \given s_t ; \theta_\text{a} \right) \right) U_t \\ 
-&- \beta H\left( \pi\left( s_t; \theta_a \right)\right)
+l_{\text{a},t} =& \underbrace{ -\log \left( \pi \left( a_t \given s_t ; \theta_\text{a} \right) \right) U_t}_{\text{Policy Loss}} \\ 
+&- \beta \underbrace{H\left( \pi\left( s_t; \theta_a \right)\right) }_{\text{Enthropy}}
 \end{split}\label{eq:actor}
@@
 In related work regarding \gls{rl}-based sampling for sequences, only discrete actions were evaluated so far: The actor would output a categorical probability distribution with a fixed maximum number of bins $k$.
-A drawback of this approach is that only a predefined maximum number of packets can be skipped and that that for a very large number of $k$, it can take a long time to converge to an optimum policy as there are $k$ different options to explore at each timestep $t$.
+A drawback of this approach is that only a predefined maximum number of packets can be skipped and that for a very large number of $k$, it can take a long time to converge to an optimum policy as there are $k$ different options to explore at each timestep $t$.
 Besides discrete actions, we thus also experiment with continuous actions, which allow to skip arbitrarily many packets.
@@
 
+\note{blue}{Fares: can be moved one section above\\}
 An overview of the neural networks of the classifier, actor and critic that we use in this work is given in \autoref{fig:neuralNetworkArchitecture}.
@@
 \centering
-  \includegraphics[width=\columnwidth]{img/rnn-sampling.pdf}
+  \includegraphics[width=0.8\columnwidth]{img/rnn-sampling.pdf}
   \caption{SparseIDS framework.
@@
 
+\note{blue}{Fares: can be removed, it was already mentioned.\\}
 A schematic overview over the \gls{rl} for sampling is given in \autoref{fig:sampling}.
@@
 
-Let $Z$ be standard normally distributed and $\mu_{\text{normal}}$ and $\sigma_{\text{normal}} > 0$ be two real numbers. Then 
+Let $Z$ be a random variable from a standard normal distribution $\mathcal{N}(0, 1)$ and let $\mu_{\text{normal}}$ and $\sigma_{\text{normal}} > 0$ be two real numbers. Then 
 \begin{align}
@@
 \end{align}
-is log-normally distributed.
+is log-normally distributed variable from $\mathcal{LN}(\mu, \sigma^2)$.
 
-The actor network (\autoref{fig:neuralNetworkArchitecture}) shows that the actor network has two outputs: $\mu$ and $\sigma$.
-These encode the mean and standard deviation of a log-normal distribution.
-Since the mean and the standard deviation of a log-normal distribution can never be smaller or equal to zero, we always apply the softplus function to them, so that the actor can only output $\mu$ and $\sigma$ that are larger than 0:
+The architecture in \autoref{fig:neuralNetworkArchitecture} shows that the actor network has two outputs: $\mu$ and $\sigma$.
+These encode the mean and standard deviation of a log-normal distribution and can never be smaller or equal to zero hence, we always apply a rectifier i.e.
+a softplus function at the output:
 \begin{align}
@@
 
+\section{Experimentation}
+\todo{Fares: add intro}
+
 \subsection{Rule-based Sampling Strategies}
-Traditionally, network packet sampling is performed on backbone infrastructures where, due to large bandwidths, not every packet can be analyzed.
-Sampling consequently helps experts understand the behavior of their network without the need to capture each and every packet.
-Assuming that the behavior of the network follows a predefined underlying process, sampling is therefore not detrimental on average to statistical analysis or measurement.
-Although the dataset in this work does not come from a backbone network, sampling reduces the high computational effort due to additional processing  required by the complex \gls{ml} based \gls{ids}.
-As a consequence, several intuitive but efficient sampling techniques can be applied in practice.
-In this study, we compare a few techniques to our sparseIDS approach.
+In order to evaluate the performance of \ours, we compare it with few techniques used in practice. 
 We opt for three rule-based sampling families that are widely used in practice, such rule-based approaches do not require feedback from higher layers (i.e.
@@
 \subsubsection{Random Sampling}
-%Given the fraction $\alpha$ of packets to be selected in a set of $N$ packets, such techniques take exactly $n = \alpha \times N$ samples that are randomly sampled from a given probability mass function, such as a uniform distribution that keeps the pick-up probability similar for all packets.
-%Practically, these techniques work properly if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is low.
 
-Given the fraction $\alpha$ of packets to be selected in a sequence of $N$ packets, these techniques can sequentially sample or skip a packet with a probability of $p_{sample}=\alpha$ and $p_{skip}=1-\alpha$ respectively.
-If the sequence is large enough, $n$ packets will be randomly sampled from a uniform distribution whereby $n= \alpha \times N$.
-Practically, these techniques work properly if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
+Given the fraction $\alpha$ of packets to be selected in a sequence of $T$ packets, these techniques can sequentially sample or skip a packet with a probability of $p_{sample}=\alpha$ and $p_{skip}=1-\alpha$ respectively.
+If the sequence is large enough, $n$ packets will be randomly sampled from a uniform distribution whereby $n= \alpha \times T$.
+Practically, these techniques work properly (the effect on subsequent analysis is minimal) if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
 
@@
 
-In each of the above techniques, the sequence of selected packets is then fed to the classifier discussed in \autoref{sec:classifier}, where features are extracted and a decision is made after each packet.
+In each of the above techniques, the sequence of selected packets is then fed to the classifier discussed in \autoref{subsec:classifier}, where features are extracted and a decision is made after each packet.
 Finally, the overall accuracy is calculated after the last packet in the sequence is received.

### 2020-01-27 paper_rl/paper.tex
Fix --variableTradeoff option (hopefully)

--- 2020-01-26 paper_rl/paper.tex
@@
 The label of each packet in a flow has a value of 0 for non-attack traffic and 1 for attack traffic with one flow being only attack or non-attack but not mixed.
-The confidence is the sigmoided ($\frac{e^x}{1+e^x}$) output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack.
+The confidence is the sigmoided $\left(\frac{e^x}{1+e^x}\right)$ output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack.
 \begin{align}
 \begin{split}
-R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |\text{label}_t - \text{confidence}_t|
+R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |\text{label}_i - \text{confidence}_i
 \end{split}
@@
 \begin{split}
-R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},t}}{\sum_{i=t+1}^{T} 1_{\text{chosen},t}+\sum_{i=t+1}^{T} 1_{\text{skipped},t}}
+R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},i}}{\sum_{i=t+1}^{T} 1_{\text{chosen},i}+\sum_{i=t+1}^{T} 1_{\text{skipped},i}}
 \end{split}

### 2020-01-27 learn.py

@@ -313,12 +313,17 @@ def custom_collate(seqs, things=(True, True, True)):
 			lstm_module_rl_actor.init_hidden(batch_size)
 
 		padded_input_data = torch.nn.utils.rnn.pad_sequence(input_data).to(device)
+		if opt.variableTradeoff:
+			tradeoff = torch.tensor([opt.global_tradeoff]).repeat((batch_size)).unsqueeze(0).to(device)
+			tradeoff_for_input = tradeoff.repeat(padded_input_data.shape[0],1).unsqueeze(2)
+			padded_input_data = torch.cat((padded_input_data, tradeoff_for_input), dim=-1)
 		batch_indices = torch.arange(batch_size)
 
 		while (remaining_seq_lens > torch.zeros_like(remaining_seq_lens)).any():
 			current_collated_slice = padded_input_data[torch.min(seq_index, orig_seq_lens_minus_one), batch_indices, :].unsqueeze(0)
 			currently_skipped = chosen_indices[-1]-chosen_indices[-2]-1 if len(chosen_indices) > 1 else chosen_indices[0]
-			current_collated_slice[0,:,-1] = currently_skipped
+			current_collated_slice[0,:,-1-int(opt.variableTradeoff)] = currently_skipped
+			# print("current_collated_slice", current_collated_slice)
 
 			if not opt.shareNet:
 				current_slice_action_probs, _ = lstm_module_rl_actor(current_collated_slice)
@@ -352,8 +357,16 @@ def custom_collate(seqs, things=(True, True, True)):
 					break
 				masks[-1].append(True)
 			masks[-1] = torch.tensor(masks[-1])
-			assert len(masks[-1]) == current_seq_len
-		# TODO: Make sure that all_seqs becomes the new seqs variable and that everything goes smoothly :)
+
+		assert len(masks[-1]) == current_seq_len
+		if opt.variableTradeoff:
+			masks_catted = torch.cat(masks)
+			empirical_sparsity = float(torch.sum(masks_catted==0).item())/len(masks_catted)
+			print("opt.global_tradeoff", opt.global_tradeoff, "empirical_sparsity", empirical_sparsity)
+			if empirical_sparsity > opt.steering_target_sparsity:
+				opt.global_tradeoff -= opt.steering_step_size
+			elif empirical_sparsity < opt.steering_target_sparsity:
+				opt.global_tradeoff += opt.steering_step_size
 
 	assert len(seqs) == len(labels) == len(categories)
[…]

### 2020-01-27 paper_rl/paper.tex

--- 2020-01-27 paper_rl/paper.tex
@@
 The label of each packet in a flow has a value of 0 for non-attack traffic and 1 for attack traffic with one flow being only attack or non-attack but not mixed.
-The confidence is the sigmoided ($\frac{e^x}{1+e^x}$) output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack.
+The confidence is the sigmoided $\left(\frac{e^x}{1+e^x}\right)$ output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack.
 \begin{align}

### 2020-01-28 paper_rl/paper.tex
proposed new rewards figure. minore changes in the sampling section

@@
 
-\begin{figure}
-\centering
-  \includegraphics[width=\columnwidth]{img/rewards.pdf}
-  \caption{The computation of the rewards for a sample flow: Gray packets are skipped while white ones are chosen.
+%\begin{figure}
+%\centering
+%  \includegraphics[width=\columnwidth]{img/rewards.pdf}
+%  \caption{The computation of the rewards for a sample flow: Gray packets are skipped while white ones are chosen.
 At the top row the resulting rewards for classification for each chosen packet are depicted and in the bottom row the ones for sparsity.
 Numbers are rounded to two digits after the decimal point.}
+%  \label{fig:rewards}
+%\end{figure}
+
+\begin{figure*}
+\centering
+  \includegraphics[width=0.8\textwidth]{img/rewards_calc_example.pdf}
+  \caption{The computation of the rewards for a sample malicious flow: Gray packets (0) are skipped while black ones (1) are chosen.
+``\#Left\//\#Total'' and ``Confidence'' are the underlying parameters for rewards calculations (see \autoref{eq:R_classification} and \autoref{eq:R_sparsity}).
+Numbers are rounded to two digits after the decimal point.}
   \label{fig:rewards}
-\end{figure}
+\end{figure*}
 
@@
 \begin{split}
-R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |\text{label}_i - \text{confidence}_i|
+R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |1 - \text{label}_i - \text{confidence}_i|
 \end{split}
@@
 
-\section{Experimentation}
-\todo{Fares: add intro}
+\section{Comparative Study}
+In order to evaluate the performance of \ours, we compare it with few rule-based sampling techniques widely used in practice.
+We opt for three sampling families that do not require feedback from higher layers (i.e.
+sampling rates are constant and not correlated with subsequent analysis).
+The underlying principal of each family is summarized as follows:
 
-\subsection{Rule-based Sampling Strategies}
-In order to evaluate the performance of \ours, we compare it with few techniques used in practice.
-We opt for three rule-based sampling families that are widely used in practice, such rule-based approaches do not require feedback from higher layers (i.e.
-sampling rates are constant and not correlated with subsequent analysis).
-The strategy of each family is summarized as follows:
-
-\subsubsection{Random Sampling}
+\subsection{Random Sampling}
 
@@
 
-\subsubsection{(Equal) First 'n' Sampling}
+\subsection{(Equal) First 'n' Sampling}
 As its name suggests, this technique only samples the first $n$ packets and ignores the rest of the flow.
@@
 
-\subsubsection{Every 'i'th Sampling}
+\subsection{Every 'i'th Sampling}
 Last but not least, a periodic sampling technique that takes a sample each $i= 1/ \alpha$ packet.
@@
 There exist many more rule-based sampling techniques e.g., temporal, adaptive etc. However, studying their behaviors is out of the scope of this paper.
+
+\section{Results and Discussion}
 
@@
 
-\section{Results and Discussion}
-
 \section{Conclusion}

### 2020-01-28 paper_rl/paper.tex
Changes to tex and to py

@@
 \begin{split}
-R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |1 - \text{label}_i - \text{confidence}_i|
+R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |1 - \left(\text{label}_i - \text{confidence}_i\right)|
 \end{split}
@@
 Fares: If the length of the flow is not known, how do we estimate T}.
+\note{red}{Max: It's only relevant during training when you know the lengths of the flows. In deployment, you don't need the rewards anymore...}
 \begin{align}
 \begin{split}
-R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},i}}{\sum_{i=t+1}^{T} 1_{\text{chosen},i}+\sum_{i=t+1}^{T} 1_{\text{skipped},i}}
+R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},i}}{T-t-1}
 \end{split}
@@
 
-Specifically, at timestep $t$, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t$ (the number of packets to be skipped) is sampled.
+Specifically, at timestep $t$, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled.
+$a_t=1$ means that the next packet is inspected next, while, for example, $a_t=2$ means that one packet is skipped until the next that is chosen and so forth.
 Besides wanting to make actions which result in a high reward more likely, the actor also seeks to optimize the entropy $H(\cdot)$ of the probability distribution so that it keeps exploring and doesn't get stuck with a suboptimal policy.
@@
 The architecture in \autoref{fig:neuralNetworkArchitecture} shows that the actor network has two outputs: $\mu$ and $\sigma$.
-These encode the mean and standard deviation of a log-normal distribution and can never be smaller or equal to zero hence, we always apply a rectifier i.e.
-a softplus function at the output:
+These encode the mean and standard deviation of a log-normal distribution and can never be smaller or equal to zero hence, we always apply a rectifier function at the output.
+We choose the softplus, which is a smooth rectifier function:
 \begin{align}
@@
 
+When experimenting with continuous actions, we encountered one practical problem: At the beginning of training, the classifier would still perform badly and not give a useful reward to the actor.
+Thus the actor would only focus on the sparsity and maximize sparsity, as the classification reward would always be equally bad no matter how many packets are chosen.
+Thus the actor would learn to let $\mu$ go to infinity as this would increase the reward at the beginning when the interaction between the three components or our \gls{rl} system is still fragile.
+As a solution, we add a penalty when the actor jumps far behind the last packet of a flow.
+It only applies to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet.
+For this, we modify the formula of the sparsity reward (\autoref{eq:R_sparsity}) for the last packet as follows, where \textit{last} refers to the index of the last chosen packet in the flow:
+
+\begin{align}
+\begin{split}
+R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{T} 1_{\text{skipped},i}\right) - \left(\textit{last}+a_{\textit{last}} - T - 1\right)}{T-\textit{last}-1}
+\end{split}
+\label{eq:R_sparsity_last}
+\end{align}
+
+Specifically, it adds the term $\left(\textit{last}+a_{\textit{last}} - T - 1\right)$ to the numerator which penalizes the actor jumping more than one packet behind the flow.
+Jumping just right after the flow does not get penalized because if the actor considers the rest of the flow as irrelevant, ignoring the rest of the flow and jumping to the first packet behind the flow is a useful action.
+
 \section{Comparative Study}
@@
 
-\subsection{(Equal) First 'n' Sampling}
+\subsection{(Equal) First \emph{n} Sampling}
 As its name suggests, this technique only samples the first $n$ packets and ignores the rest of the flow.
-Nevertheless, there are two variants of this technique: (a) taking the first $n$ packets, where $n$ is the same for all flows, and (b) taking the first ``Equal'' proportion of packets from each flow (e.g.
+Nevertheless, there are two variants of this technique: (a) taking the first $n$ packets, where $n$ is the same for all flows, and (b) taking the first ``equal'' proportion of packets from each flow (e.g.
 half of the flow, regardless of its length).
-In the first example, the size of the entire flow is not necessary and the capture can be stopped once enough packets are captured.
+In the first technique, the size of the entire flow is not necessary and the capture can be stopped once enough packets are captured.
 In the second scenario, however, the length of each flow needs to be known.
@@
 
-\subsection{Every 'i'th Sampling}
+\subsection{Every \emph{i}th Sampling}
 Last but not least, a periodic sampling technique that takes a sample each $i= 1/ \alpha$ packet.
@@
 
-We have implemented a recurrent classifier based on LSTMs to detect network attacks, %While attack detection performance is not necessarily better than that of non-recurrent approaches,
-which can already detect attacks before they are over.
-Furthermore, the recurrent approach allows us to inspect the influence of single packets on the detection performance and shows which packets are \textit{characteristic} for attacks.
-Even though the interpretation of \glspl{rnn} poses several difficulties, we have demonstrated methods for gaining insights into the model's functioning.
-
-We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified.
-We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.
-Deploying an adversarial training procedure, we could significantly reduce the adversarial threat.
-
 \section*{Acknowledgements}

### 2020-01-28 learn.py

@@ -362,11 +362,13 @@ def custom_collate(seqs, things=(True, True, True)):
 		if opt.variableTradeoff:
 			masks_catted = torch.cat(masks)
 			empirical_sparsity = float(torch.sum(masks_catted==0).item())/len(masks_catted)
+			sparsities.append(empirical_sparsity)
 			print("opt.global_tradeoff", opt.global_tradeoff, "empirical_sparsity", empirical_sparsity)
-			if empirical_sparsity > opt.steering_target_sparsity:
-				opt.global_tradeoff -= opt.steering_step_size
-			elif empirical_sparsity < opt.steering_target_sparsity:
+			if empirical_sparsity < opt.steering_target_sparsity:
 				opt.global_tradeoff += opt.steering_step_size
+			elif empirical_sparsity > opt.steering_target_sparsity:
+				opt.global_tradeoff -= opt.steering_step_size
+			opt.global_tradeoff = min(max(opt.global_tradeoff, 0.0), 1.0)
 
 	assert len(seqs) == len(labels) == len(categories)
 	return [collate_things(item, index==0, sampling_masks=masks if opt.sampling!="" else None) for index, (item, thing) in enumerate(zip((seqs, labels, categories), things)) if thing]
@@ -863,6 +865,8 @@ def train_rl():
 					torch.save(lstm_module_rl_actor.state_dict(), '%s/%s' % (writer.log_dir, filename_rl_actor))
 					torch.save(lstm_module_rl_critic.state_dict(), '%s/%s' % (writer.log_dir, filename_rl_critic))
 
+
+sparsities = []
 
 @torch.no_grad()
 def test():
@@ -2161,9 +2165,9 @@ if __name__=="__main__":
 	parser.add_argument('--sampling', type=str, default="", help='technique employed to sample packets of flows; blank if no sampling is applied')
 	parser.add_argument('--samplingProbability', type=float, default=0.5, help='desired probability of choosing a packet when sampling; the first packet of a flow is always chosen')
 	parser.add_argument('--variableTradeoff', action='store_true', help='whether the tradeoff between accuracy and sparsity for RL should be randomly chosen for each flow')
-	parser.add_argument('--global_tradeoff', type=float, default=0.0, help='the initial tradeoff to use when using the control loop')
+	parser.add_argument('--global_tradeoff', type=float, default=1.0, help='the initial tradeoff to use when using the control loop')
 	parser.add_argument('--steering_step_size', type=float, default=0.01, help='step size for steering')
[…]

### 2020-01-29 learn.py
Implement steering and plotting

@@ -162,6 +162,12 @@ def get_one_hot_vector(class_indices, num_classes, batch_size):
 	y_onehot.zero_()
 	return y_onehot.scatter_(1, class_indices.unsqueeze(1), 1)
 
+chosen_packets_per_epoch_sampling_rl = [[]]
+all_packets_per_epoch_sampling_rl = [[]]
+batch_counter_sampling_rl = 0
+tradeoffs_sampling_rl = []
+initial_global_tradeoff = None
+
 def custom_collate(seqs, things=(True, True, True)):
 	seqs, labels, categories = zip(*seqs)
 	assert not opt.sampling!="" or opt.rl
@@ -360,15 +366,37 @@ def custom_collate(seqs, things=(True, True, True)):
 
 		assert len(masks[-1]) == current_seq_len
 		if opt.variableTradeoff:
+			global initial_global_tradeoff, batch_counter_sampling_rl
+			if batch_counter_sampling_rl == 0:
+				initial_global_tradeoff = opt.global_tradeoff
 			masks_catted = torch.cat(masks)
-			empirical_sparsity = float(torch.sum(masks_catted==0).item())/len(masks_catted)
-			sparsities.append(empirical_sparsity)
-			print("opt.global_tradeoff", opt.global_tradeoff, "empirical_sparsity", empirical_sparsity)
-			if empirical_sparsity < opt.steering_target_sparsity:
-				opt.global_tradeoff += opt.steering_step_size
-			elif empirical_sparsity > opt.steering_target_sparsity:
-				opt.global_tradeoff -= opt.steering_step_size
-			opt.global_tradeoff = min(max(opt.global_tradeoff, 0.0), 1.0)
+			all_chosen_packets = int(torch.sum(masks_catted==0).item())
+			all_current_packtes = len(masks_catted)
+
+			chosen_packets_per_epoch_sampling_rl[-1].append(all_chosen_packets)
+			all_packets_per_epoch_sampling_rl[-1].append(all_current_packtes)
+			batch_counter_sampling_rl += 1
+
+			if batch_counter_sampling_rl > 0 and batch_counter_sampling_rl % opt.batches_to_consider_for_steering == 0:
+				chosen_packets_per_epoch_sampling_rl[-1] = sum(chosen_packets_per_epoch_sampling_rl[-1])
+				all_packets_per_epoch_sampling_rl[-1] = sum(all_packets_per_epoch_sampling_rl[-1])
[…]

### 2020-01-29 plot_steering.py

@@ -1,0 +1,44 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import json
+import sys
+import matplotlib.pyplot as plt
+
+plt.rcParams["font.family"] = "serif"
+plt.rcParams['pdf.fonttype'] = 42
+
+colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
+
+path = sys.argv[1]
+f = open(path, "r")
+chosen_packets_per_epoch_sampling_rl, all_packets_per_epoch_sampling_rl, tradeoffs_sampling_rl = json.loads(f.read())
+f.close()
+
+chosen_packets_per_epoch_sampling_rl = np.array(chosen_packets_per_epoch_sampling_rl[:-1])
+all_packets_per_epoch_sampling_rl = np.array(all_packets_per_epoch_sampling_rl[:-1])
+tradeoffs_sampling_rl = np.array(tradeoffs_sampling_rl)
+
+assert len(chosen_packets_per_epoch_sampling_rl) == len(all_packets_per_epoch_sampling_rl) == len(tradeoffs_sampling_rl)
+
+print("Epochs", len(chosen_packets_per_epoch_sampling_rl))
+
+x = list(range(1,len(chosen_packets_per_epoch_sampling_rl)+1))
+
+# print("x", x, "chosen_packets_per_epoch_sampling_rl", chosen_packets_per_epoch_sampling_rl, "all_packets_per_epoch_sampling_rl", all_packets_per_epoch_sampling_rl, "tradeoffs_sampling_rl", tradeoffs_sampling_rl)
+
+plt.figure(figsize=(5,2))
+lines = []
+lines += plt.plot(x, 100*chosen_packets_per_epoch_sampling_rl/all_packets_per_epoch_sampling_rl, color=colors[0])
+plt.xlabel("Epochs")
+plt.ylabel('Sparsity (%)')
+
+plt.twinx()
+lines += plt.plot(x, tradeoffs_sampling_rl, color=colors[1])
+plt.legend(lines, ["Sparsity", "Tradeoff"], loc='upper right')
+plt.ylabel('Tradeoff')
+plt.tight_layout()
+plt.savefig("paper_rl/img/" + path[:-5] + '.pdf', bbox_inches = 'tight', pad_inches = 0)
+# plt.show()
+
+plt.close()

### 2020-01-29 paper_rl/paper.tex
Revise dataset section

@@
 %%ci
+
+\section{Dataset}
+
+\begin{table}[h]
+\caption{Flow occurrence frequency of attack types.}
+\label{tab:occurrence}
+\centering
+\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
+]{
+\begin{tabular}{l r}
+\toprule
+Attack type & \hspace*{-4mm}Proportion \\ \midrule
+DoS Hulk	&	10.10\%	\\
+PortScan, Firewall	&	6.90\%	\\
+DDoS LOIT	&	4.08\%	\\
+Infiltration	&	3.30\%	\\
+DoS GoldenEye	&	0.32\%	\\
+DoS SlowHTTPTest	&	0.18\%	\\
+DoS Slowloris	&	0.17\%	\\
+Brute-force SSH	&	0.11\%	\\
+Botnet ARES	&	0.03\%	\\
+XSS attack	&	0.03\%	\\
+PortScan, no Fw.	&	0.02\%	\\
+Brute-force FTP	&	0.01\%	\\
+SQL injection	&	$<$0.01\%	\\
+Heartbleed	&	$<$0.01\%	\\
+\bottomrule
+\end{tabular}
+}{}
+\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
+]{
+\begin{tabular}{l r}
+\toprule
+Attack type & \hspace*{-4mm}Proportion \\ \midrule
+Exploits	&	1.42\%	\\
+Fuzzers	&	1.01\%	\\
+Reconnaissance	&	0.58\%	\\
+Generic	&	0.21\%	\\
+DoS	&	0.19\%	\\
+Shellcode	&	0.08\%	\\
+Analysis	&	0.03\%	\\
+Backdoors	&	0.02\%	\\
+Worms	&	0.01\%	\\
+\bottomrule
+\end{tabular}
+}{}
+\end{table}
+
+For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+The distribution of different attacks in the datasets are shown in \autoref{tab:occurrence}.
 
@@
 
-\subsection{Dataset}
-% \begin{table}[b]
-% \caption{occurrence frequency of attack types.} \label{tab:occurrence}
-% \centering
-% \begin{tabular}{l r} \toprule
-% Attack type & Proportion \\
-% \midrule
-% Normal                                                         & $0.747468$ \\
-% DoS / DDoS:DoS Hulk                                            & $0.101014$ \\
-% PortScan:PortScan - Firewall off                               & $0.069020$ \\
-% DDoS:LOIT                                                      & $0.040784$ \\
-% Infiltration:Dropbox download                                  & $0.032982$ \\
-% DoS / DDoS:DoS GoldenEye                                       & $0.003224$ \\
-% DoS / DDoS:DoS Slowhttptest                                    & $0.001819$ \\
-% DoS / DDoS:DoS slowloris                                       & $0.001680$ \\
-% Brute Force:SSH-Patator                                        & $0.001107$ \\
-% Botnet:ARES                                                    & $0.000327$ \\
-% Web Attack:XSS                                                 & $0.000293$ \\
-% PortScan:PortScan - Firewall on                                & $0.000165$ \\
-% Brute Force:FTP-Patator                                        & $0.000110$ \\
-% Web Attack:Sql Injection                                       & $0.000006$ \\
-% DoS / DDoS:Heartbleed                                          & $0.000001$ \\
-% \bottomrule
-% \end{tabular}
-% \label{tab:occurrence}
-% \end{table}
-
-\begin{table}[b]
-\caption{Flow occurrence frequency of attack types.}
-\label{tab:occurrence}
-\centering
-%\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
-%]{
-\begin{tabular}{l r}
-\toprule
-Attack type & \hspace*{-4mm}Proportion \\ \midrule
-DoS Hulk	&	10.10\%	\\
-PortScan, Firewall	&	6.90\%	\\
-DDoS LOIT	&	4.08\%	\\
-Infiltration	&	3.30\%	\\
-DoS GoldenEye	&	0.32\%	\\
-DoS SlowHTTPTest	&	0.18\%	\\
-DoS Slowloris	&	0.17\%	\\
-Brute-force SSH	&	0.11\%	\\
-Botnet ARES	&	0.03\%	\\
-XSS attack	&	0.03\%	\\
-PortScan, no Fw.	&	0.02\%	\\
-Brute-force FTP	&	0.01\%	\\
-SQL injection	&	$<$0.01\%	\\
-Heartbleed	&	$<$0.01\%	\\
-\bottomrule
-\end{tabular}
-%}{}
-%\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
-%]{
-%\begin{tabular}{l r}
-%\toprule
-%Attack type & \hspace*{-4mm}Proportion \\ \midrule
-%Exploits	&	1.42\%	\\
-%Fuzzers	&	1.01\%	\\
-%Reconnaissance	&	0.58\%	\\
-%Generic	&	0.21\%	\\
-%DoS	&	0.19\%	\\
-%Shellcode	&	0.08\%	\\
-%Analysis	&	0.03\%	\\
-%Backdoors	&	0.02\%	\\
-%Worms	&	0.01\%	\\
-%\bottomrule
-%\end{tabular}
-%}{}
-\end{table}
-
-%For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
-For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
-
-The distribution of different attacks in the datasets are shown in \autoref{tab:occurrence}.
-
 %\begin{table}

### 2020-01-29 plot_steering.py
Minor

@@ -21,7 +21,7 @@ tradeoffs_sampling_rl = np.array(tradeoffs_sampling_rl)
 
 assert len(chosen_packets_per_epoch_sampling_rl) == len(all_packets_per_epoch_sampling_rl) == len(tradeoffs_sampling_rl)
 
-print("Epochs", len(chosen_packets_per_epoch_sampling_rl))
+# print("Epochs", len(chosen_packets_per_epoch_sampling_rl))
 
 x = list(range(1,len(chosen_packets_per_epoch_sampling_rl)+1))
 
@@ -30,7 +30,7 @@ x = list(range(1,len(chosen_packets_per_epoch_sampling_rl)+1))
 plt.figure(figsize=(5,2))
 lines = []
 lines += plt.plot(x, 100*chosen_packets_per_epoch_sampling_rl/all_packets_per_epoch_sampling_rl, color=colors[0])
-plt.xlabel("Epochs")
+plt.xlabel("Time steps")
 plt.ylabel('Sparsity (%)')
 
 plt.twinx()

### 2020-01-30 paper_rl/paper.tex
Add results draft

@@
 \newcommand\Average{E\Basics}
+
+\usepackage{tabularx}
+
+% Macro to make entire row bold from https://tex.stackexchange.com/questions/309833/format-whole-row-of-table-as-bold
+\newcommand\setrow[1]{\gdef\rowmac{#1}#1\ignorespaces}
+\newcommand\clearrow{\global\let\rowmac\relax}
+\clearrow
 
@@
 
-\section{Comparative Study}
-In order to evaluate the performance of \ours, we compare it with few rule-based sampling techniques widely used in practice.
+\section{Other sampling techniques}
+\label{sec:other_sampling_techniques}
+In order to evaluate the performance of \ours, we compare it with a few rule-based sampling techniques widely used in practice.
 We opt for three sampling families that do not require feedback from higher layers (i.e.
-sampling rates are constant and not correlated with subsequent analysis).
-The underlying principal of each family is summarized as follows:
+sampling rates are constant and not correlated with subsequent analysis): %The underlying principle of each family is summarized as follows:
 
@@
 
-Given the fraction $\alpha$ of packets to be selected in a sequence of $T$ packets, these techniques can sequentially sample or skip a packet with a probability of $p_{sample}=\alpha$ and $p_{skip}=1-\alpha$ respectively.
-If the sequence is large enough, $n$ packets will be randomly sampled from a uniform distribution whereby $n= \alpha \times T$.
+Given the fraction $p$ of packets to be selected in a sequence of $T$ packets, this technique will sample each individual packet with probability of $p$.
+If the sequence is large enough, $n$ packets will be randomly sampled from a uniform distribution whereby $n= p \times T$.
 Practically, these techniques work properly (the effect on subsequent analysis is minimal) if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
 
-\subsection{(Equal) First \emph{n} Sampling}
-As its name suggests, this technique only samples the first $n$ packets and ignores the rest of the flow.
-Nevertheless, there are two variants of this technique: (a) taking the first $n$ packets, where $n$ is the same for all flows, and (b) taking the first ``equal'' proportion of packets from each flow (e.g.
-half of the flow, regardless of its length).
-In the first technique, the size of the entire flow is not necessary and the capture can be stopped once enough packets are captured.
-In the second scenario, however, the length of each flow needs to be known.
-Again, similar techniques work properly if the important information is present in the first few packets of a flow, i.e.
-a large data transfer in which the initial packets represent a hand-shake after which the majority of packets contain the raw data and are blind to non-\gls{dpi} systems.
+\subsection{First \emph{n} Sampling}
+As its name suggests, this technique only samples the first packets from each flow and ignores the rest of the flow.
+For a sampling probability $p$ it will take the first $n$ packets from each flow so that $n \approx T * p$.
+For example, for a flow with length 10 and $p=0.2$, it will take the first 2 packets.
+For this procedure the length of a flow has to be known beforehand and thus it is not applicable in a streaming scenario.
+This technique is successful under the assumption that most information for determining maliciousness is present in the first few packets of a flow, i.e.
+a large data transfer in which the initial packets represent a hand-shake after which the majority of packets contain the raw data and do not reveal significant information to non-\gls{dpi} systems.
+
+\subsection{First \emph{n} Equal Sampling}
+This technique is similar to the one above, with the difference that the length of a flow doesn't have to be known beforehand.
+If flows have an average length of 10 packets, and the sampling probability $p=0.2$ it would take up to two packets from each flow.
 
 \subsection{Every \emph{i}th Sampling}
-Last but not least, a periodic sampling technique that takes a sample each $i= 1/ \alpha$ packet.
+Finally we use a periodic sampling technique that takes a sample each $i= 1/ \alpha$ packet.
 Similar to the last technique, the length of the flow is not needed in this case.
-In practice, this technique is best suited to keep the behavioral distribution of the sampled set similar to the original distribution, thus suited for scenarios where information is heterogeneously distributed between packets.
+In practice, this technique is best suited to keep the behavioral distribution of the sampled set similar to the original distribution, thus suited for scenarios where information is heterogeneously distributed between packets over an entire network flow.
 
-To begin with, the first packet of each flow is always taken for which a record is created in memory and after which the sampling routine is used.
+In each of the above techniques, the first packet of each flow is always taken because otherwise one could get undefined behavior: If a flow only consists of 1 packet and this packet is not chosen, it is not even defined if the prediction of the classifier was correct for this flow or not.
+Thus, in our implementations of the above sampling techniques, we make sure that even when the first packet is always chosen, the overall fraction of chosen packets corresponds to the sampling probability $p$.
 
-In each of the above techniques, the sequence of selected packets is then fed to the classifier discussed in \autoref{subsec:classifier}, where features are extracted and a decision is made after each packet.
-Finally, the overall accuracy is calculated after the last packet in the sequence is received.
+After applying the sampling techniques, the sequence of selected packets is then fed to the classifier discussed in \autoref{subsec:classifier}, where features are extracted and a decision is made after each packet.
+Also, the classifier gets information on how many packets were skipped prior to each packet.
 
-There exist many more rule-based sampling techniques e.g., temporal, adaptive etc. However, studying their behaviors is out of the scope of this paper.
+% Max: If we write things like this they might think we were too lazy to implement them :D
+%There exist many more rule-based sampling techniques e.g., temporal, adaptive etc. However, studying their behaviors is out of the scope of this paper.
 
-\section{Results and Discussion}
+\section{Results}
 
-%\begin{table}
-%% Evaluated using
-%% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
-%% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
-%% on Dec. 4th
-%\caption{Performance metrics per packet and per flow.} \label{tab:performance_results}
-%\centering
-%\begin{tabular}{l r r r r} \toprule
-%& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
-%	&	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
-%Accuracy	&	99.1\%	&	99.7\%	&	99.5\%	&	98.3\%	\\
-%Precision	&	97.0\%	&	99.7\%	&	83.4\%	&	78.6\%	\\
-%Recall	&	97.8\%	&	99.1\%	&	87.6\%	&	72.6\%	\\
-%F1	&	97.4\%	&	99.4\%	&	85.5\%	&	75.5\%	\\
-%Youden	&	97.2\%	&	99.0\%	&	87.3\%	&	71.9\%	\\
-%\bottomrule
-%\end{tabular}
-%\end{table}
+\begin{table}
+% Evaluated using
+% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
+% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
+% on Dec. 4th
+\caption{Comparing classification metrics per flow with different tradeoffs for both CIC-IDS-2017 and UNSW-NB15.} \label{tab:results_tradeoff}
+\centering
+\begin{tabular}{>{\rowmac}l>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
+Tradeoff & 0.0 & 0.1 & 0.5 & 0.0 & 0.1 & 0.5 \\	\midrule
+\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% & 98.9\% & 98.4\% & 98.5\% \\
+Precision & 99.4\% & 98.5\% & 98.7\% & 81.1\% & 69.5\% & 70.8\% \\
+Recall & 98.5\% & 99.0\% & 98.6\% & 84.0\% & 89.9\% & 89.6\% \\
+F1 & 99.0\% & 98.7\% & 98.6\% & 82.5\% & 78.4\% & 79.1\% \\
+Youden & 98.4\% & 98.5\% & 98.2\% & 83.3\% & 88.6\% & 88.4\% \\ 
+\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 0\% & 83.5\% & 84.5\% \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\subsection{Influence of the tradeoff}
+
+\autoref{tab:results_tradeoff} shows the influence of the tradeoff parameter $\alpha$ for both datasets.
+For the columns with a tradeoff of 0.0, we train the classifier regularly with all packets without using \gls{rl}.
+For tradeoffs 0.1 and 0.5 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff fixed for the whole training.
+We train for 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.
+Results show that the higher the tradeoff, the fewer packets are sampled and also the lower accuracy.
+The only exception for this is that for tradeoffs $0.1$ and $0.5$ for UNSW-NB15, the accuracy is almost the same or even a little bit higher for the larger tradeoff, which is counterintuitive.
+We consider this to be coincidence that can happen due to the stochasticity of the training process.
+
+\todo{Write that the sparsity is higher for UNSW-NB15 because it contains longer flows}
+
+\subsection{Discrete vs.~Continuous actions}
+
+\begin{table}
+\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions for both CIC-IDS-2017 and UNSW-NB15.
+The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
+\centering
+\begin{tabular}{>{\rowmac}l>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
+Actions & cont. & disc. & cont. & disc. \\	\midrule
+\setrow{\bfseries}Accuracy & 99.4\% & 99.2\% & 98.4\% & 98.4\% \\
+Precision & 98.5\% & 98.7\% & 69.5\% & 68.0\% \\
+Recall & 99.0\% & 98.3\% & 89.9\% & 94.6\% \\
+F1 & 99.7\% & 98.5\% & 78.4\% & 79.2\% \\
+Youden & 98.5\% & 97.8\% & 88.6\% & 93.2\% \\ 
+\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% & 83.5\% & 86.6\% \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions shows that for CIC-IDS-2017 continuous actions result in a higher accuracy with fewer samples being chosen.
+Conversely, for UNSW-NB-15, accuracy is the same for both continuous and discrete actions.
+However, with discrete actions, fewer packets are chosen, which leads to the conclusion that in this case, discrete actions perform better.
+This leads to the conclusion, that at least for our use case and the two datasets that we consider, there is no consistent difference in performance for continuous and discrete actions.
+Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while for continuous actions the only possibility is to increase or decrease the mean of the log-normal distribution, which can results in worse performance.
+On the other side, with discrete actions, the maximum number of samples to skip is fixed, which we consider a major drawback.
+Furthermore, training time is potentially higher since, for example, for a discrete action space of 20, all 20 different actions are treated as being completely different while actually there is not a large difference between skipping 18 or 19 packets while there is a big difference between skipping 0 packets compared to skipping 19 packets.
+This means that for discrete actions, all actions are viewed as being completely independent, which is not true as the reward function considers sparsity, which increases gradually the more packets are skipped and for which similar actions result in similar rewards.
+
+\subsection{Comparing with other sampling methods}
+
+\begin{table}
+\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies for CIC-IDS-2017.} \label{tab:results_others}
+\centering
+\begin{tabular}{>{\rowmac}l>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+& \gls{rl} & random & every $i$th & first $n$ & first $n$ equal \\	\midrule
+\setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
+Precision & 98.5\% & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
+Recall & 99.0\% & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
+F1 & 99.7\% & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
+Youden & 98.5\% & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
+\setrow{\bfseries}Sampled & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Next we compared the \gls{rl} technique usedy by \ours{} with several intuitive sampling strategies we proposed in \autoref{sec:other_sampling_techniques}. 
+\autoref{tab:results_others} shows that with the same sparsity, \ours{} achieves the highest accuracy.
+Not surprisingly, random sampling performs the worst, as it is possible, that a flow is ``unlucky'' and doesn't get enough packets sampled while another one might get more packets than necessary.
+The best strategy besides the \gls{rl} approach seems to be ``first $n$ equal'' sampling, which matches the conclusion from \cite{hartl_explainability_2019} that the first couple of packets are the most important ones for determining maliciousness of a flow.
+``first $n$ equal'' performs better than ``first $n$'' because for ``first $n$'', longer flows will have more packets sampled from them.
+Conversely, very short flows will get few packets.
+But since the first packets of a flow determine its maliciousness, the short flows get misclassified because they don't get enough packets sampled.
+
+\section{Steering}
+
+So far we have always trained \ours{} with a fixed tradeoff parameter $\alpha$.
+However, there is also the important use case in which an operator wants a certain minimum sparsity, for example 50\%, to limit the required computation power and thus limit cost.
+We thus propose a method that enables steering the tradeoff to achieve a certain wanted sparsity.
+\begin{enumerate}
+\item A model is trained where the tradeoff is randomly chosen for each flow during training.
+We set the probability distribution to be a uniform distribution between 0 and 1.
+We add a new feature to each packet which indicates the tradeoff chosen for this flow so that it can adapt its sampling strategy in anticipation of the tradeoff that will determine the reward.
+\item During deployment, the tradeoff is set to the maximum that was used during training (1 in our case), which results in maximum sparsity.
+Then, the tradeoff is continuously incremented if the current sparsity is still under the specified minimum.
+It is important that the tradeoff can never fall below 0 since this is a case that was not encountered during training and which can lead to anomalous behavior.
+\end{enumerate}
+
+%{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_5_DoS - DDoS-DoS Hulk"}
+\begin{figure}
+\centering
+  \includegraphics[width=\columnwidth]{{"img/lstm_module_8.pth_global_tradeoff_1.0_steering_step_size_0.05_steering_target_sparsity_0.5_batches_to_consider_for_steering_500_sampling_rl"}.pdf}
+  \caption{Steering the tradeoff with an allowed minimum sparsity of 50\%.
+The tradeoff is initialized as 1 in the beginning and continuously lowered as long as the sparsity is higher than 50\% for CIC-IDS-2017.
+The procedure stops if the minimum sparsity is reached or the tradeoff reached its minimum of 0, which is the case here.
+Sparsity falls from around 74.5\% to 73.5\%.}
+  \label{fig:steering}
+\end{figure}
+
+\autoref{fig:steering} shows the steering in action.
+It stops when the tradeoff reaches 0 and the specified minimum sparsity of 50\% has not been reached yet.
+It might seem surprising that even with a parameter of 0, which means that only classification matters and sparsity not at all, still a sparsity of 73.5\% is achieved.
+We attribute this to the fact that \ours{} learns that some flows contain may contain packets that do not contain any useful information about determining maliciousness.
+Thus they are even skipped when there is no reward for skipping them because they do not increase the classification score at all.
 
 \section{Conclusion}
+
+We saw that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity works as expected and that higher tradeoffs resulted in higher sparsity and potentially lower classification performance.
+Surprisingly, we saw that our solution \ours{} chooses very few packets: Typically $\approx \frac{1}{4}$ for CIC-IDS-2017 and $\approx \frac{1}{5}$ for UNSW-NB-15.
+
+Furthermore, our results showed that \gls{rl}-based sampling can intelligently choose relevant packets for intrusion detection and outperform non-smart sampling techniques. 
+Besides discrete actions for our \gls{rl} system we also implemented continuous ones.
+A naive implementation did not work as the training was too fragile in the beginning and the \gls{rl} system would diverge and the number of skipped packets would go to infinity.
+Adding a penalty for jumping behind the end of the sequence proved effective.
+Comparing with discrete actions, continuous actions seemed to perform on par.
+This is still an advantage, as for continuous actions the neural network only needs to have two outputs (mean and standard deviation) while for discrete ones it needs many outputs as there are actions.
+The continuous action framework is thus more lightweight.
 

### 2020-01-30 plot_steering.py

@@ -27,7 +27,7 @@ x = list(range(1,len(chosen_packets_per_epoch_sampling_rl)+1))
 
 # print("x", x, "chosen_packets_per_epoch_sampling_rl", chosen_packets_per_epoch_sampling_rl, "all_packets_per_epoch_sampling_rl", all_packets_per_epoch_sampling_rl, "tradeoffs_sampling_rl", tradeoffs_sampling_rl)
 
-plt.figure(figsize=(5,2))
+plt.figure(figsize=(5,3))
 lines = []
 lines += plt.plot(x, 100*chosen_packets_per_epoch_sampling_rl/all_packets_per_epoch_sampling_rl, color=colors[0])
 plt.xlabel("Time steps")

### 2020-01-31 learn.py
Add script that plots which packets are chosen

@@ -341,6 +341,7 @@ def custom_collate(seqs, things=(True, True, True)):
 			else:
 				current_slice_action_probs = torch.nn.functional.softplus(current_slice_action_probs)
 				decisions = current_slice_action_probs[:,:,0]
+				# print("decisions", decisions, "stds", current_slice_action_probs[:,:,1])
 
 			if not opt.continuous:
 				increment = decisions.view(-1)+1
@@ -352,6 +353,8 @@ def custom_collate(seqs, things=(True, True, True)):
 			chosen_indices.append(seq_index)
 			remaining_seq_lens = torch.max(orig_seq_lens - seq_index, torch.zeros_like(orig_seq_lens))
 
+		# chosen_indices_stacked = torch.stack(chosen_indices)
+		# print("chosen_indices_stacked", chosen_indices_stacked.transpose(1,0))
 		masks = []
 		for seq_index in range(batch_size):
 			current_seq_len = len(seqs[seq_index])
@@ -439,6 +442,8 @@ def collate_things(seqs, is_seqs=False, sampling_masks=None):
 		skipped = []
 		if is_seqs:
 			for mask in sampling_masks:
+				# print("mask", mask)
+				# assert not len(mask) > 1 or not mask[1]
 				seq_range = torch.arange(len(mask))[mask]
 				skipped.append(torch.cat((torch.tensor([0]), seq_range[1:]-seq_range[:-1]-1)))
 				assert (skipped[-1] >= 0).all()
@@ -915,6 +920,7 @@ def test():
 	results_by_attack_number = [list() for _ in range(min(attack_numbers), max(attack_numbers)+1)]
 	label_by_attack_number = [list() for _ in range(min(attack_numbers), max(attack_numbers)+1)]
 	sample_indices_by_attack_number = [list() for _ in range(min(attack_numbers), max(attack_numbers)+1)]
+	orig_lens = [list() for _ in range(min(attack_numbers), max(attack_numbers)+1)]
 
 	for input_data, labels, categories in tqdm(test_loader):
 
@@ -958,10 +964,11 @@ def test():
 			results_by_attack_number[flow_category].append(np.concatenate((flow_input, flow_output), axis=-1))
 			label_by_attack_number[flow_category] = flow_label
 			sample_indices_by_attack_number[flow_category].append(test_indices[samples])
+			orig_lens[flow_category].append(len(test_data[samples][0]))
 
 			samples += 1
 
[…]

### 2020-01-31 plot.py

@@ -48,12 +48,12 @@ for attack_type, seqs in enumerate(results_by_attack_number):
 	else:
 		all_seqs.extend(seqs)
 		all_attacks.extend(seqs)
-	
+
 reverse_mapping[len(results_by_attack_number)] = 'All samples'
 results_by_attack_number.append(all_seqs)
 reverse_mapping[len(results_by_attack_number)] = 'All attacks'
 results_by_attack_number.append(all_attacks)
-	
+
 for attack_type, seqs in enumerate(results_by_attack_number):
 	print (attack_type, reverse_mapping[attack_type], len(seqs))
 	if len(seqs) <= 0:
@@ -72,7 +72,7 @@ for attack_type, seqs in enumerate(results_by_attack_number):
 		for seq in seqs:
 			if len(seq) < i+1:
 				break
-				
+
 			if len(seq.shape) == 1:
 				seq = seq[:,None]
 
@@ -137,13 +137,13 @@ for attack_type, seqs in enumerate(results_by_attack_number):
 	all_labels = [item.get_label() for item in all_legends]
 	ax1.legend(all_legends, all_labels, loc='upper right', bbox_to_anchor=(1,0.95))
 	#  plt.ylim((min(first_quartiles), max(third_quartiles)))
-	
+
 	ax2.set_ylabel_legend(all_legends[0])
 	ax1.set_ylabel_legend(Rectangle((0,0), 1, 1, fc='gray', alpha=0.2), handlelength=0.7, handletextpad=0.4)
 
 	ticks = plt.xticks()
 	plt.xticks([ tick for tick in ticks[0][1:-1] if tick.is_integer() ])
-	
+
 	if SHOW_TITLE:
 		plt.title(reverse_mapping[attack_type])
 	# plt.legend(all_legends)

### 2020-01-31 plot_rl.py

@@ -1,0 +1,194 @@
+#!/usr/bin/env python3
+
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+import numpy as np
+import sys
+import os
+import json
+import pickle
+
+def numpy_sigmoid(x):
+	return 1/(1+np.exp(-x))
+
+DIR_NAME = "plots/plot_rl"
+ORDERING = ['', 'Length min', 'IAT min', 'Length max', 'IAT max']
+
+MAX_X = 20
+SHOW_TITLE = False
+
+plt.rcParams["font.family"] = "serif"
+
+
+dataroot_basename = sys.argv[1].split('_')[0]
+
+with open(dataroot_basename + "_categories_mapping.json", "r") as f:
+	categories_mapping_content = json.load(f)
+categories_mapping, mapping = categories_mapping_content["categories_mapping"], categories_mapping_content["mapping"]
+reverse_mapping = {v: k for k, v in mapping.items()}
+print("reverse_mapping", reverse_mapping)
+
+file_name = sys.argv[1]
+with open(file_name, "rb") as f:
+	loaded = pickle.load(f)
+results_by_attack_number, sample_indices_by_attack_number, orig_seq_lens = loaded["results_by_attack_number"], loaded["sample_indices_by_attack_number"], loaded["orig_lens"]
+
+# print("results", results_by_attack_number)
+# print("sample_indices", sample_indices_by_attack_number)
+lens_results = [len(item) for item in results_by_attack_number]
+lens_indices = [len(item) for item in sample_indices_by_attack_number]
+lens_lens = [len(item) for item in orig_seq_lens]
+
+print("lens_indices", "\n".join(["{}: {}".format(reverse_mapping[attack], length) for attack, length in zip(range(len(lens_indices)), lens_indices)]))
+assert lens_results == lens_indices == lens_lens
+
+colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
+
+all_seqs = []
+all_attacks = []
+all_seqs_lens = []
+all_attacks_lens = []
+for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_seq_lens)):
+	for seq in seqs:
+		seq[:,-1] = numpy_sigmoid(seq[:,-1])
+
+	if reverse_mapping[attack_type] == 'Normal':
[…]

### 2020-01-31 paper_rl/neural_network.tex
Changed detail in draw.io illustration
Changes for RL section
Change plots

@@
       (a1) rectangle (a2);%
-    \path (a1.east |- a1.south)+(2.5,0.25) node (u1)[anchor=south west]%
+    \path (a1.east |- a2.north)+(2.5,-0.75) node (u1)[anchor=south west]%
       {\textit{#3}};%
@@
 \begin{scope}
-\node [input place] (s1) [label={[align=center]left: State (8)}] {};
+\node [input place] (s1) [label={[align=center]below right: Sample (16)}] {};
 \node [place] (s2) [above right of=s1, label={[align=center]above: Linear}] {};
@@
 \path[->] (s5) edge node {} (s6);        
-\node [place] (s10) [below right of=s1, label={[align=center]below: Linear}] {};
+\node [place] (s10) at (\shortside,-\shortside-\nodedistance) [label={[align=center]below: Linear}] {};
 \path[->] (s1) edge node {} (s10);    
@@
 \path[->] (s21) edge [loop below] node {} (s21);
-\node [place] (s13) [right of=s21, label={[align=center]above: Linear}] {};
+\node [place] (s16) [right of=s21, label={[align=center]below: Linear}] {};
+\path[->] (s21) edge node {} (s16);    
+\node [output place] (s20) [right of=s16, label={[align=center]below: $v_\text{sparsity}$ (1)}] {};    
+\path[->] (s16) edge node {} (s20);
+\node [place] (s13) [above of=s16, label={[align=center]above: Linear}] {};
 \path[->] (s21) edge node {} (s13);
@@
 \path[->] (s13) edge node {} (s17);
-\node [place] (s16) [below of=s13, label={[align=center]below: Linear}] {};
-\path[->] (s21) edge node {} (s16);    
-\node [output place] (s20) [right of=s16, label={[align=center]below: $v_\text{sparsity}$ (1)}] {};    
-\path[->] (s16) edge node {} (s20);
 \backgroundnc{s20}{s17}{Critic network}{s10}

### 2020-01-31 plot_rl.py

@@ -88,7 +88,7 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 	filled_in_confidences = []
 	for seq, l in zip(seqs, lens):
 		skip_pattern = seq[:,-2]
-		print("skip_pattern", skip_pattern)
+		# print("skip_pattern", skip_pattern)
 		# print("skip_pattern", skip_pattern)
 		assert (np.round(skip_pattern) == skip_pattern).all(), skip_pattern
 		confidence = seq[:,-1]

### 2020-02-03 plot_rl.py
Fix plot_rl.py

@@ -88,8 +88,6 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 	filled_in_confidences = []
 	for seq, l in zip(seqs, lens):
 		skip_pattern = seq[:,-2]
-		# print("skip_pattern", skip_pattern)
-		# print("skip_pattern", skip_pattern)
 		assert (np.round(skip_pattern) == skip_pattern).all(), skip_pattern
 		confidence = seq[:,-1]
 		# print("confidence", confidence)
@@ -118,9 +116,9 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 	assert len(lens) == len(filled_ins) == len(filled_in_confidences)
 
 	lens = lens_to_plot
-	lens /= len(seqs)
+	# lens /= len(seqs)
 	print("lens", lens)
-	assert lens[0] == 1, lens[0]
+	# assert lens[0] == 1, lens[0]
 
 	filled_ins_equal_lengths = [np.concatenate((item, np.zeros(max_length-len(item)))) for item in filled_ins]
 	filled_ins_equal_lengths_stacked = np.stack(filled_ins_equal_lengths)
@@ -128,7 +126,7 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 	summed_chosen_packets = np.sum(filled_ins_equal_lengths_stacked, axis=0)
 	fraction_of_chosen_packets_by_flows = summed_chosen_packets/len(seqs)
 	print("fraction_of_chosen_packets_by_flows", fraction_of_chosen_packets_by_flows)
-	assert fraction_of_chosen_packets_by_flows[0] == 1, fraction_of_chosen_packets_by_flows[0]
+	# assert fraction_of_chosen_packets_by_flows[0] == 1, fraction_of_chosen_packets_by_flows[0]
 	assert len(fraction_of_chosen_packets_by_flows) == max_length
 
 	values_by_length = [list() for _ in range(max_length)]
@@ -150,12 +148,14 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 	fig, ax1 = plt.subplots(figsize=(5,3))
 	x_values = list(range(min(len(lens), MAX_X)))
[…]

### 2020-02-03 paper_rl/paper.tex
Intro rewording + minor changes

@@
 
-There is a significant body of scientific work focusing on the detection of unwanted behavior in networks.
+%There is a significant body of scientific work focusing on the detection of unwanted behavior in networks.
 In the past, a viable way of performing intrusion detection was to inspect the content of packets themselves and detect if a packet delivers potentially harmful content.
@@
 % Fares:
-%As the need for security mechanisms in today's infrastructure communication networks grows, a substantial body of scientific work focuses on developing reliable, explainable frameworks for the identification of unwanted behaviors.
+As the need for security mechanisms in today's infrastructure communication networks grows, a substantial body of scientific work focuses on developing reliable, explainable frameworks for the identification of unwanted behaviors.
 In the past, a viable way to detect intrusion was to analyze the contents of the packets themselves and determine, for example, whether a packet contains potentially harmful content by matching patterns.
-More recently, with the increasing deployment of encryption, the focus is now on features that are always available for network monitoring devices such as port numbers, protocol flags, packet sizes when encrypting above the transport layer.
+More recently, with the increasing deployment of encryption, the focus is now on features that are always available for network monitoring devices such as port numbers, protocol flags and packet sizes when encrypting above the transport layer.
 
-Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties.
+%Network communication is usually aggregated into \textit{flows}, which are commonly defined as a sequence of packets sharing certain properties.
 When analyzing flows, not only the aforementioned features are available but also features related to the timing of the individual packets.
 Various approaches have been proposed to extract features from flows and then perform anomaly detection with the extracted flows \cite{meghdouri_analysis_2018}.
-While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows.
+%While these approaches frequently work well, it is problematic that the whole flow has to be received first and only afterwards anomaly detection is applied, revealing attack flows.
 Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
-At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
+%At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
+%\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
 
 % Fares:
-%A network communication is typically aggregated into \textit{flows}, which are commonly defined as a sequence of packets that share certain properties.
+A network communication is typically aggregated into \textit{flows}, which are commonly defined as a sequence of packets that share certain properties.
 When analyzing flows, not only the aforementioned features are available but also features representing the temporal behavior of the individual packets.
-Various approaches have been proposed to extract flows features after which anomaly detection can be performed, in \cite{meghdouri_analysis_2018} Meghdouri et al.
-compare such approaches.
-%While these approaches often work well, it is troublesome because, in the worst case, the whole flow must be received first and only then anomaly detection can be applied; revealing malicious flows.
+Various approaches have been proposed to extract flows features after which anomaly detection can be performed, in \cite{meghdouri_analysis_2018} authors compare such approaches.
+While these approaches often work well, it is troublesome because, in the worst case, the whole flow must be received first and only then anomaly detection can be applied; revealing malicious flows.
 Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
-%At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-%In \cite{hartl_explainability_2019}, Hartl et al.
-show that such an architecture has similar performance to other flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends and analyzed its robustness with respect to adversaries extensively.
+At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
+In \cite{hartl_explainability_2019}, authors show that such an architecture has similar performance to other flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends and analyzed its robustness with respect to adversaries extensively.
 
-However, for practical use, not only classification accuracy and security are important but also computational efficiency.
+%However, for practical use, not only classification accuracy and security are important but also computational efficiency.
 This follows from the fact that a rationally-acting company will only deploy an \gls{ids} if the cost that is saved by avoiding security breaches is higher than the cost of the \gls{ids} itself.
@@
 % Fares:
-%Nevertheless, not only the accuracy and reliability of classification, but also the computational efficiency, are critical for practical use.
+Nevertheless, not only the accuracy and reliability of classification, but also the computational efficiency, are critical for practical use.
 This follows from the fact that a rationally-acting organization will only implement an \gls{ids} if the cost saved by preventing security breaches is higher than the cost of the \gls{ids} itself.
@@
 \begin{enumerate}
-\item \textbf{choose optimally}, taking only the data that contain the most information and skip the less relevant ones.
+%\item \textbf{choose optimally}, taking only the data that contain the most information and skip the less relevant ones.
 % Fares:
-%\item \textbf{choose optimally}, taking only samples that contain the most information and skip the less relevant ones.
+\item \textbf{choose optimally}, taking only samples that contain the most information and skip the less relevant ones.
 \item have a parameter that allows to \textbf{trade off} classification performance for sparsity.
@@
 
-For this purpose, we develop the \gls{rl}-based \gls{ids} \textit{\ours{}} fulfills the above properties.
+%For this purpose, we develop the \gls{rl}-based \gls{ids} \textit{\ours{}} fulfills the above properties.
 We show that a significant number of packets can be skipped while accuracy doesn't drop significantly.
@@
 % Fares:
-%For this purpose we develop a \gls{rl}-based \gls{ids}: \textit{\ours{}} that fulfills the above properties.
+For this purpose we develop a \gls{rl}-based \gls{ids}: \textit{\ours{}} that fulfills the above properties.
 We show that a significant number of packets can be skipped while the accuracy does not drop
-%considerably. Compared to other common sampling techniques, \ours{} performs better when being trained on the same number of packets.
+considerably. Compared to other common sampling techniques, \ours{} performs better when being trained on the same number of packets.
 
-As \ours{} uses a parameter to trade-off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
+%As \ours{} uses a parameter to trade-off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
 However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.
@@
 % Fares:
-%The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both later quantities.
+The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both later quantities.
 However, in certain situations, a certain sparsity is to be achieved, hence we further develop and propose a steering system that fluctuates such a trade-off parameter in order to obtain a coveted sparsity in a closed control loop.
 
-To encourage reproducibility and facilitate experimentation, we publicly release the source code, the data and the figures of this work at \url{https://github.com/CN-TU/adversarial-recurrent-ids/tree/rl}
+To encourage reproducibility and facilitate experimentation, we publicly release the source code, the data and the figures of this work\footnote{\url{https://github.com/CN-TU/adversarial-recurrent-ids/tree/rl}}.
 
@@
 
+\note{blue}{Fares: if we submit to CNS (page limit is 9), we can remove this sections as it is not relevant and statistics can be found online}
+
 \begin{table}[h]
-\caption{Flow occurrence frequency of attack types.}
+\caption{Occurrence frequency of the attack flows.}
 \label{tab:occurrence}
@@
 
-For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15:_2015} datasets which each include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15_2015} datasets, each includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
 The distribution of different attacks in the datasets are shown in \autoref{tab:occurrence}.
@@
 E.g.
-for a flow of length 10 if after timestep 5, 3 packets are chosen and 2 are skipped from what is left, the sparsity reward metric at timestep 5 is $\frac{2}{2+3}=\frac{2}{5}$.
-\note{blue}{Fares: This equation can be shortened by diving by the total count of packets left.\\
-Fares: If the length of the flow is not known, how do we estimate T}.
-\note{red}{Max: It's only relevant during training when you know the lengths of the flows. In deployment, you don't need the rewards anymore...}
+for a flow of length 10 if after timestep 5, 3 packets are chosen and 2 are skipped from what is left, the sparsity reward metric at timestep 5 is $\frac{2}{10-(4+1)}=\frac{2}{5}$.
+%\note{blue}{Fares: This equation can be shortened by diving by the total count of packets left.\\
+%Fares: If the length of the flow is not known, how do we estimate T}.
+%\note{red}{Max: It's only relevant during training when you know the lengths of the flows. In deployment, you don't need the rewards anymore...}
 \begin{align}
 \begin{split}
-R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},i}}{T-t-1}
+R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},i}}{T-(t+1)}
 \end{split}
@@
 
-Our proposed framework consists of three independent neural networks: the classifier, the critic and the actor.
+Our proposed framework (see \autoref{fig:neuralNetworkArchitecture}) consists of three independent neural networks: the classifier, the critic and the actor.
 
@@
 
-\note{blue}{Fares: can be moved one section above\\}
-An overview of the neural networks of the classifier, actor and critic that we use in this work is given in \autoref{fig:neuralNetworkArchitecture}.
+%\note{blue}{Fares: can be moved one section above\\}
+%An overview of the neural networks of the classifier, actor and critic that we use in this work is given in \autoref{fig:neuralNetworkArchitecture}.
 
@@
 
-\note{blue}{Fares: can be removed, it was already mentioned.\\}
-A schematic overview over the \gls{rl} for sampling is given in \autoref{fig:sampling}.
+%\note{blue}{Fares: can be removed, it was already mentioned.\\}
+%A schematic overview over the \gls{rl} for sampling is given in \autoref{fig:sampling}.
 
@@
 
-When experimenting with continuous actions, we encountered one practical problem: At the beginning of training, the classifier would still perform badly and not give a useful reward to the actor.
-Thus the actor would only focus on the sparsity and maximize sparsity, as the classification reward would always be equally bad no matter how many packets are chosen.
-Thus the actor would learn to let $\mu$ go to infinity as this would increase the reward at the beginning when the interaction between the three components or our \gls{rl} system is still fragile.
-As a solution, we add a penalty when the actor jumps far behind the last packet of a flow.
-It only applies to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet.
-For this, we modify the formula of the sparsity reward (\autoref{eq:R_sparsity}) for the last packet as follows, where \textit{last} refers to the index of the last chosen packet in the flow:
+When experimenting with continuous actions, we encountered one practical problem: At the beginning of training, the classifier would still perform badly and would not give the actor a useful reward.
+As a result, the actor would only focus on and maximize sparsity, as the classification reward would always be equally bad no matter how many packets are chosen.
+Thus, the actor would learn to let $\mu$ go to infinity as this would increase the reward at the beginning when the interaction between the three components of our \gls{rl} system is still fragile.
+As a solution, we add a penalty to the reward calculation when the actor jumps far behind the last flow packet.The modified reward applies only to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet.
+The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow:
 
@@
 \begin{split}
-R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{T} 1_{\text{skipped},i}\right) - \left(\textit{last}+a_{\textit{last}} - T - 1\right)}{T-\textit{last}-1}
+R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{T} 1_{\text{skipped},i}\right) - \left(\textit{last}+a_{\textit{last}} - T - 1\right)}{T- (\textit{last}+1)}
 \end{split}
@@
 
+The results of the sampling part are discussed in this section.
+We study the sparsity-accuracy tradeoff first, then we evaluate both discrete and continuous actions and, finally, we compare our results with other sampling techniques.
+Overall, sampling based on \gls{rl} reveals that a high number of packets can be skipped while maintaining high performance.
+
 \begin{table}
@@
 \centering
-\begin{tabular}{>{\rowmac}l>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
 & \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
 Tradeoff & 0.0 & 0.1 & 0.5 & 0.0 & 0.1 & 0.5 \\	\midrule
+\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 0\% & 83.5\% & 84.5\% \\ \midrule
 \setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% & 98.9\% & 98.4\% & 98.5\% \\
@@
 Youden & 98.4\% & 98.5\% & 98.2\% & 83.3\% & 88.6\% & 88.4\% \\ 
-\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 0\% & 83.5\% & 84.5\% \\
 \bottomrule
-\end{tabular}
+\end{tabular*}
 \end{table}
@@
 \autoref{tab:results_tradeoff} shows the influence of the tradeoff parameter $\alpha$ for both datasets.
-For the columns with a tradeoff of 0.0, we train the classifier regularly with all packets without using \gls{rl}.
-For tradeoffs 0.1 and 0.5 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff fixed for the whole training.
-We train for 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.
-Results show that the higher the tradeoff, the fewer packets are sampled and also the lower accuracy.
-The only exception for this is that for tradeoffs $0.1$ and $0.5$ for UNSW-NB15, the accuracy is almost the same or even a little bit higher for the larger tradeoff, which is counterintuitive.
-We consider this to be coincidence that can happen due to the stochasticity of the training process.
+For columns with a tradeoff of 0.0, we train the classifier regularly with all packets without using \gls{rl}.
+For tradeoffs 0.1 and 0.5 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff constant for the whole training.
+We train for 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.\par
+Results show that the higher the tradeoff, the less packets are sampled and the lower the accuracy.
+The only exception to this is that for tradeoffs of $0.1$ and $0.5$ for UNSW-NB15, the accuracy is almost the same or even slightly higher for the larger tradeoff, which is counter-intuitive.
+We consider this to be a coincidence that can happen because of the stochasticity of the training process.
 
-\todo{Write that the sparsity is higher for UNSW-NB15 because it contains longer flows}
+\note{blue}{Fares: my suggestion: 
+The only exception to this is that for tradeoffs of $0.1$ and $0.5$ for UNSW-NB15, the accuracy is almost the same or even slightly higher for the larger tradeoff, which is counter-intuitive.
+We assume that due to the nature of the UNSW-NB15 data set (which is highly heterogeneous), fewer packets yield better information (and therefore the memory of the LSTM classifier is cleaner) and hence better classification decision and higher accuracy.
+}
+
+Furthermore, due to the fact that UNSW-NB15 contains more packets per flow, its sparsity is higher.
+This, on the other hand, holds since almost the same number of packets is sampled from each flow for both datasets (see \autoref{sec:conclusion}).
+
+%\todo{Write that the sparsity is higher for UNSW-NB15 because it contains longer flows}
 
@@
 \centering
-\begin{tabular}{>{\rowmac}l>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
 & \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
 Actions & cont. & disc. & cont. & disc. \\	\midrule
+\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% & 83.5\% & 86.6\% \\ \midrule
 \setrow{\bfseries}Accuracy & 99.4\% & 99.2\% & 98.4\% & 98.4\% \\
@@
 Youden & 98.5\% & 97.8\% & 88.6\% & 93.2\% \\ 
-\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% & 83.5\% & 86.6\% \\
 \bottomrule
-\end{tabular}
+\end{tabular*}
 \end{table}
 
-\autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions shows that for CIC-IDS-2017 continuous actions result in a higher accuracy with fewer samples being chosen.
+\autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions and shows that for CIC-IDS-2017 continuous actions result in a higher accuracy with fewer samples being selected.
 Conversely, for UNSW-NB-15, accuracy is the same for both continuous and discrete actions.
@@
 This leads to the conclusion, that at least for our use case and the two datasets that we consider, there is no consistent difference in performance for continuous and discrete actions.
-Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while for continuous actions the only possibility is to increase or decrease the mean of the log-normal distribution, which can results in worse performance.
-On the other side, with discrete actions, the maximum number of samples to skip is fixed, which we consider a major drawback.
-Furthermore, training time is potentially higher since, for example, for a discrete action space of 20, all 20 different actions are treated as being completely different while actually there is not a large difference between skipping 18 or 19 packets while there is a big difference between skipping 0 packets compared to skipping 19 packets.
-This means that for discrete actions, all actions are viewed as being completely independent, which is not true as the reward function considers sparsity, which increases gradually the more packets are skipped and for which similar actions result in similar rewards.
+Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean log-normal distribution, which can lead to worse performance.
+On the other hand, with discrete actions, the maximum number of samples to skip is fixed, which we consider to be a major drawback.
+Furthermore, training time is potentially higher since, for example, for a discrete space of 20 actions, each option is treated as being completely different while actually there is not a large difference between skipping 18 or 19 packets, there is a big difference between skipping 0 packets compared to skipping 19 packets.
+This means that for discrete actions, all actions are perceived as being completely independent, which is not true as the reward function considers sparsity, which increases gradually as more packets are skipped and for which similar actions result in similar rewards.
 
@@
 \centering
-\begin{tabular}{>{\rowmac}l>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
 & \gls{rl} & random & every $i$th & first $n$ & first $n$ equal \\	\midrule
+\setrow{\bfseries}Sampled & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
 \setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
@@
 Youden & 98.5\% & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
-\setrow{\bfseries}Sampled & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\
 \bottomrule
-\end{tabular}
+\end{tabular*}
 \end{table}
 
-Next we compared the \gls{rl} technique usedy by \ours{} with several intuitive sampling strategies we proposed in \autoref{sec:other_sampling_techniques}. 
-\autoref{tab:results_others} shows that with the same sparsity, \ours{} achieves the highest accuracy.
-Not surprisingly, random sampling performs the worst, as it is possible, that a flow is ``unlucky'' and doesn't get enough packets sampled while another one might get more packets than necessary.
+Next we compared the \gls{rl} technique usedy by \ours{} with several intuitive sampling strategies suggested in \autoref{sec:other_sampling_techniques}. 
+\autoref{tab:results_others} reveals that \ours{} achieves the highest accuracy with the same sparsity.
+Not surprisingly, random sampling performs the worst, as it is likely that a flow is either ``misfortune'' and doesn't get enough packets sampled while another  one gets more or, the random packets chosen carry less useful information.
 The best strategy besides the \gls{rl} approach seems to be ``first $n$ equal'' sampling, which matches the conclusion from \cite{hartl_explainability_2019} that the first couple of packets are the most important ones for determining maliciousness of a flow.
+\note{blue}{Fares: The next sentence is a bit confusing.}
 ``first $n$ equal'' performs better than ``first $n$'' because for ``first $n$'', longer flows will have more packets sampled from them.
@@
 
-So far we have always trained \ours{} with a fixed tradeoff parameter $\alpha$.
-However, there is also the important use case in which an operator wants a certain minimum sparsity, for example 50\%, to limit the required computation power and thus limit cost.
-We thus propose a method that enables steering the tradeoff to achieve a certain wanted sparsity.
+So far, we have always trained \ours{} with a fixed tradeoff parameter $\alpha$.
+Nevertheless, there is also an important use case in which an operator wants a certain minimum sparsity, for example 50\%, to limit the necessary computation power and thus reduce costs.
+We thus propose a method that enables steering the tradeoff to achieve a certain desired sparsity.
 \begin{enumerate}
@@
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2020-02-03 paper_rl/paper.tex
changes to steering section and conclusion

@@
 As a solution, we add a penalty to the reward calculation when the actor jumps far behind the last flow packet.The modified reward applies only to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet.
-The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow:
+The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow ($t=last$):
 
@@
 Nevertheless, there is also an important use case in which an operator wants a certain minimum sparsity, for example 50\%, to limit the necessary computation power and thus reduce costs.
-We thus propose a method that enables steering the tradeoff to achieve a certain desired sparsity.
+We thus propose a method that enables steering the tradeoff to achieve a certain desired sparsity. The steering algorithm works as follows:
 \begin{enumerate}
-\item A model is trained where the tradeoff is randomly chosen for each flow during training.
-We set the probability distribution to be a uniform distribution between 0 and 1.
-We add a new feature to each packet which indicates the tradeoff chosen for this flow so that it can adapt its sampling strategy in anticipation of the tradeoff that will determine the reward.
+\item A model is trained where the trade-off is chosen randomly for each flow during training from a uniform distribution yielding values between 0 and 1.
+We add a new feature to each packet that indicates the tradeoff chosen for the respective flow so that it can adjust its sampling strategy in anticipation of the tradeoff that will determine the reward.
 \item During deployment, the tradeoff is set to the maximum that was used during training (1 in our case), which results in maximum sparsity.
-Then, the tradeoff is continuously incremented if the current sparsity is still under the specified minimum.
-It is important that the tradeoff can never fall below 0 since this is a case that was not encountered during training and which can lead to anomalous behavior.
+The tradeoff is then continuously decremented if the current sparsity is still above the specified minimum.
+It is important that the tradeoff can never fall below 0 because this is a case that has not been encountered during training and can lead to abnormal behavior.
 \end{enumerate}
@@
 
-\autoref{fig:steering} shows the steering in action.
+%\autoref{fig:steering} shows the steering in action.
 It stops when the tradeoff reaches 0 and the specified minimum sparsity of 50\% has not been reached yet.
@@
 
+\autoref{fig:steering} shows the steering in action.
+We start with a maximum sparsity of ~74.4\% and try to reach 50\%.
+In order to do so, the steering system start decreasing the tradoff parameter until it reaches 0.
+It might seem surprising that even with such value, which means that only classification matters and not sparsity at all, a sparsity of 73.5\% is still achieved.
+We attribute this (1) to the fact that \ours{} learns that some flows may contain packets that do not contribute in determining maliciousness and are skipped since they do not affect the accuracy and (2) the actor still in any case attempts to maximize the entropy of the output which induces skipping packets for fat-tailed distributions.
+%Nothworthy is the fact that the goal sparsity (50\%) is not reached due to the previous assumptions, however, a decreasing slope is clearly shown in figure 4
+
 \section{Conclusion}
 \label{sec:conclusion}
+ In this paper we try to solve the practical problem of accurate network traffic classification with less effort.
+To this end, we propose \ours{} which is a framework for packet sampling based on \gls{rl}.
 
-We saw that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity works as expected and that higher tradeoffs resulted in higher sparsity and potentially lower classification performance.
-Surprisingly, we saw that our solution \ours{} chooses very few packets: Typically $\approx \frac{1}{4}$ for CIC-IDS-2017 and $\approx \frac{1}{5}$ for UNSW-NB-15.
+We prove that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity works as expected and that higher tradeoffs resulted in higher sparsity and potentially lower classification performance.
+Surprisingly, we saw that our solution \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}th$ for CIC-IDS-2017 and $\approx \frac{1}{5}th$ for UNSW-NB-15.
+Furthermore, our results showed that \gls{rl}-based sampling can intelligently choose relevant packets for intrusion detection and outperform non-smart sampling techniques.
 
-Furthermore, our results showed that \gls{rl}-based sampling can intelligently choose relevant packets for intrusion detection and outperform non-smart sampling techniques. 
+A naive implementation would not work as the training is too fragile in the beginning and the \gls{rl} system suprisingly learned to maximize only one of the rewards (sparsity), this was solved by adding a penalty for jumping behind the end of the sequence which was proved effective.
+
 Besides discrete actions for our \gls{rl} system we also implemented continuous ones.
-A naive implementation did not work as the training was too fragile in the beginning and the \gls{rl} system would diverge and the number of skipped packets would go to infinity.
-Adding a penalty for jumping behind the end of the sequence proved effective.
 Comparing with discrete actions, continuous actions seemed to perform on par.
@@
 The continuous action framework is thus more lightweight.
+
+\note{blue}{Fares: add steering...}
 

### 2020-02-03 paper_rl/paper.tex
Changes

--- 2020-02-03 paper_rl/paper.tex
@@
 
+\begin{figure}
+\centering
+  \includegraphics[width=\columnwidth]{{"img/events.out.tfevents.1579352256.gpu"}.pdf}
+  \caption{Training with a tradeoff of 0.1 with continuous actions for the CIC-IDS-2017 dataset.
+Training accuracy increases sharply in the beginning while becoming slower the longer training lasts.
+Sparsity increases sharply in the beginning and then slower.
+As the classifier gets better, the \gls{rl} starts skipping more packets.}
+  \label{fig:training}
+\end{figure}
+
 The results of the sampling part are discussed in this section.
@@
 \item During deployment, the tradeoff is set to the maximum that was used during training (1 in our case), which results in maximum sparsity.
-Then, the tradeoff is continuously incremented if the current sparsity is still under the specified minimum.
+Then, the tradeoff is continuously decremented if the current sparsity is still under the specified minimum.
 It is important that the tradeoff can never fall below 0 since this is a case that was not encountered during training and which can lead to anomalous behavior.
@@
 
-%{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_5_DoS - DDoS-DoS Hulk"}
 \begin{figure}

### 2020-02-03 paper_rl/paper.tex
Add results

@@
 Overall, sampling based on \gls{rl} reveals that a high number of packets can be skipped while maintaining high performance.
+Then, we inspect how the \gls{rl} chooses packets and if some patterns can be spotted.
+
+For all our experiments we train for at least 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.
+\autoref{fig:training} shows how during training, accuracy increases quickly in the beginning and then gradually afterwards.
+At the same time, sparsity gets higher as the \gls{rl} learns which packets can be skipped since they do not influence classification performance.
 
@@
 For tradeoffs 0.1 and 0.5 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff constant for the whole training.
-We train for 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.\par
+
 Results show that the higher the tradeoff, the less packets are sampled and the lower the accuracy.
@@
 
+\subsection{Inspecting the \gls{rl}'s behavior}
+\label{subsec:inspecting}
+
+\begin{figure*}[h]
+\centering
+\subfloat[Benign Traffic\label{fig:inspectingNormal}
+]{
+\includegraphics[width=0.98\columnwidth]{{"../plots/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_10_Normal"}.pdf}
+}{}
+\subfloat[SSH Patator\label{fig:inspectingPatator}
+]{
+\includegraphics[width=0.98\columnwidth]{{"../plots/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_2_Brute Force-SSH-Patator"}.pdf}
+}{}
+\subfloat[DDoS Slowloris\label{fig:inspectingSlowloris}
+]{
+\includegraphics[width=0.98\columnwidth]{{"../plots/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_7_DoS - DDoS-DoS slowloris"}.pdf}
+}{}
+\subfloat[All samples\label{fig:inspectingAll}
+]{
+\includegraphics[width=0.98\columnwidth]{{"../plots/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_15_All samples"}.pdf}
+}{}
+\caption{Inspecting the sampling strategy of \ours{} for various attack types.
+The blue line shows the confidence.
+The gray bars show how many samples had this number of packets.
+For example, the gray bar in Figure \ref{fig:inspectingNormal} at position 19 with height 10000 signifies that 10000 flows of benign traffic had a length of at least 20 packets (since we say that the first packet is at position 0).
+The first packet is always chosen for each flow by design.}
+\label{fig:inspecting}
+\end{figure*}
+
+\autoref{fig:inspecting} shows the sampling strategy for different attack types.
+Figure \ref{fig:inspectingNormal} shows that for benign traffic, the sampling strategy is quite heterogeneous: Not specific packets get always chosen, but different packets get chosen, depending on the flow.
+This contrasts with SSH Patator (Figure \ref{fig:inspectingPatator}), where the 0th, the 3rd, the 13th and the 15th packet are always chosen, since they appear to be especially indicative of an attack.
+Also, the flows seems to contain no useful information between the 3rd and the 13th packet, so that this part is always skipped altogether.
+Slowloris (Figure \ref{fig:inspectingSlowloris}) is more similar to what we saw for benign traffic: The sampling strategy varies a lot for different flows and not always the same pattern is used.
+When looking at all flows (Figure \ref{fig:inspectingAll}) it becomes apparent that the second packet (index 1) is almost never chosen, and the third is also chosen quite rarely.
+We attribute this to the fact that most communication flows are TCP and use handshakes, meaning that the second packet is simply a reply a part of the handshake, revealing no important information.
+In general, we see that the sampling strategy varies significantly between different attack types and for some attack types even for different flows within the same attack type, showing that \ours{} learns to choose the packets effectively, which reveal the malicious nature of a flow.
+
 \section{Steering}
@@
 \label{sec:conclusion}
- In this paper we try to solve the practical problem of accurate network traffic classification with less effort.
+In this paper we try to solve the practical problem of accurate network traffic classification with less computational power.
 To this end, we propose \ours{} which is a framework for packet sampling based on \gls{rl}.
 
+% Max: 1/4 is already pronounced one fourth
 We prove that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity works as expected and that higher tradeoffs resulted in higher sparsity and potentially lower classification performance.
-Surprisingly, we saw that our solution \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}th$ for CIC-IDS-2017 and $\approx \frac{1}{5}th$ for UNSW-NB-15.
+Surprisingly, we saw that our solution \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}$ for CIC-IDS-2017 and $\approx \frac{1}{5}$ for UNSW-NB-15.
 Furthermore, our results showed that \gls{rl}-based sampling can intelligently choose relevant packets for intrusion detection and outperform non-smart sampling techniques.
 
-A naive implementation would not work as the training is too fragile in the beginning and the \gls{rl} system suprisingly learned to maximize only one of the rewards (sparsity), this was solved by adding a penalty for jumping behind the end of the sequence which was proved effective.
-
 Besides discrete actions for our \gls{rl} system we also implemented continuous ones.
-Comparing with discrete actions, continuous actions seemed to perform on par.
+We showed that a naive implementation would not work as the training is too fragile in the beginning and the \gls{rl} system learned to maximize only one of the rewards (sparsity) in the beginning.
+This was solved by adding a penalty for jumping behind the end of the sequence which was proved effective.
+Comparing with discrete actions, continuous actions seem to perform on par.
 This is still an advantage, as for continuous actions the neural network only needs to have two outputs (mean and standard deviation) while for discrete ones it needs many outputs as there are actions.
@@
 
-\note{blue}{Fares: add steering...}
+Inspecting the decisions of our \gls{rl} framework showed that the sampling strategy is clearly different for different attack strategies and also within some attack families, indicating that \ours{} indeed adopts different sampling strategies for different conversation flows.
+
+Finally, we implement a steering method with which a certain sparsity threshold can be specified which limits the computational expenses of an operator.
+
+In general, we think that \gls{rl}-based packet sampling is not only applicable to \gls{ids} but we are optimistic that other domains of networking can benefit from similar solutions to save computational power.
 

### 2020-02-03 plot_rl.py

@@ -173,7 +173,7 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 	all_legends.append(ret3)
 
 	all_labels = [item.get_label() for item in all_legends]
-	ax1.legend(all_legends, all_labels, loc='upper right', bbox_to_anchor=(1,0.95))
+	plt.legend(all_legends, all_labels, loc='upper right', bbox_to_anchor=(1,0.95))
 
 	ax2.set_ylabel_legend(all_legends[0])
 	# ax1.set_ylabel_legend(Rectangle((0,0), 1, 1, fc='gray', alpha=0.2), handlelength=0.7, handletextpad=0.4)

### 2020-02-04 paper_rl/paper.tex
minor text changes. Minor improvements to the framework figure

@@
 Overall, sampling based on \gls{rl} reveals that a high number of packets can be skipped while maintaining high performance.
-Then, we inspect how the \gls{rl} chooses packets and if some patterns can be spotted.
+To this end, we inspect how the \gls{rl} chooses packets and if some patterns can be noticed
 
-For all our experiments we train for at least 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.
-\autoref{fig:training} shows how during training, accuracy increases quickly in the beginning and then gradually afterwards.
+For all our experiments we train our networks for at least 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.
+\autoref{fig:training} shows how, during training, accuracy increases quickly at the beginning and then gradually afterwards.
 At the same time, sparsity gets higher as the \gls{rl} learns which packets can be skipped since they do not influence classification performance.
@@
 For example, the gray bar in Figure \ref{fig:inspectingNormal} at position 19 with height 10000 signifies that 10000 flows of benign traffic had a length of at least 20 packets (since we say that the first packet is at position 0).
+The red bars represent the count of packets sampled at this position for all flows.
 The first packet is always chosen for each flow by design.}
@@
 
+Eventually, we implement a steering method to help achieving a certain sparsity threshold. The steering mechanism can be used to limit the computational expenses of an operator.
+
 In general, we think that \gls{rl}-based packet sampling is not only applicable to \gls{ids} but we are optimistic that other domains of networking can benefit from similar solutions to save computational power.

### 2020-02-04 paper/paper.tex
Added MLP values to performance table

@@
 Since payload is no longer available,
-the focus now lies on features that are always available to network monitoring equipment like port numbers, protocol flags or packet sizes when encrypting above the transport layer.
+the focus now lies on features that are always available to network monitoring equipment like packet sizes, protocol flags or port numbers when encrypting above the transport layer.
 
@@
 Thus, we design a network \gls{ids} that operates on a per-packet basis and decides if a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
-At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
+At the same time, an \gls{rnn}-based \gls{ids} has the benefit of providing any available information to the classifier while avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features.
 We show that our system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates.
@@
 % on Dec. 4th
-\caption{Performance metrics per packet and per flow.} \label{tab:performance_results}
+\caption{Performance metrics per packet and per flow. MLP values from \cite{bachl_walling_2019} are presented for comparison.} \label{tab:performance_results}
 \centering
-\begin{tabular}{l r r r r} \toprule
-& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
-	&	Packet	&	Flow	&	Packet	&	Flow	\\	\midrule
-Accuracy	&	99.1\%	&	99.7\%	&	99.5\%	&	98.3\%	\\
-Precision	&	97.0\%	&	99.7\%	&	83.4\%	&	78.6\%	\\
-Recall	&	97.8\%	&	99.1\%	&	87.6\%	&	72.6\%	\\
-F1	&	97.4\%	&	99.4\%	&	85.5\%	&	75.5\%	\\
-Youden	&	97.2\%	&	99.0\%	&	87.3\%	&	71.9\%	\\
+\begin{tabular}{l r r r r r r} \toprule
+& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
+	&	Packet	&	Flow	& MLP &	Packet	&	Flow & MLP	\\	\midrule
+Accuracy	&	99.1\%	&	99.7\%	&	99.8\%	&	99.5\%	&	98.3\%	&	98.9\%	\\
+Precision	&	97.0\%	&	99.7\%	&	99.9\%	&	83.4\%	&	78.6\%	&	84.5\%	\\
+Recall	&	97.8\%	&	99.1\%	&	99.2\%	&	87.6\%	&	72.6\%	&	82.9\%	\\
+F1	&	97.4\%	&	99.4\%	&	99.5\%	&	85.5\%	&	75.5\%	&	83.7\%	\\
+Youden	&	97.2\%	&	99.0\%	&	99.1\%	&	87.3\%	&	71.9\%	&	82.3\%	\\
 \bottomrule
 \end{tabular}
+\vspace*{-.3mm}
 \end{table}
@@
 
-\section{Adversarial Robustness Metric}
+\section{Evaluating Adversarial Robustness}
 
@@
 \subsection{Reducing Features}
-The most obvious defense, which is only possible because of the nature of the dataset, is to simply omitting features an attacker can manipulate. We try two different approaches:
+The most obvious defense, which is only possible because of the nature of the dataset, is to simply omit features an attacker can manipulate. We try two different approaches:
 \begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]

### 2020-02-04 paper_rl/paper.tex
remove duplicate paragraph
Max's changes to the paper

--- 2020-02-03 paper_rl/paper.tex
@@
 A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of packets like current approaches but can possibly skip arbitrarily many packets, saving even more computational resources for long sequences.
-Finally we build an automatic steering mechanism that can guide \ours{} to achieve a desired level of sparsity.
+Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows.
+Finally we build an automatic steering mechanism that can guide \ours{} in deployment to achieve a desired level of sparsity.
 
@@
 % Fares:
-As the need for security mechanisms in today's infrastructure communication networks grows, a substantial body of scientific work focuses on developing reliable, explainable frameworks for the identification of unwanted behaviors.
+As the need for security mechanisms in today's infrastructure's communication networks grows, a substantial body of scientific work focuses on developing reliable and explainable frameworks for the identification of unwanted behavior.
 In the past, a viable way to detect intrusion was to analyze the contents of the packets themselves and determine, for example, whether a packet contains potentially harmful content by matching patterns.
@@
 % Fares:
-A network communication is typically aggregated into \textit{flows}, which are commonly defined as a sequence of packets that share certain properties.
+Network communication is typically aggregated into \textit{flows}, which are commonly defined as a sequence of packets that share certain properties.
 When analyzing flows, not only the aforementioned features are available but also features representing the temporal behavior of the individual packets.
-Various approaches have been proposed to extract flows features after which anomaly detection can be performed, in \cite{meghdouri_analysis_2018} authors compare such approaches.
-While these approaches often work well, it is troublesome because, in the worst case, the whole flow must be received first and only then anomaly detection can be applied; revealing malicious flows.
-Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
-At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
+Various approaches have been proposed to extract flow features after which anomaly detection can be performe: In \cite{meghdouri_analysis_2018} authors compare such approaches and shows their efficacy.
+While these approaches often work well, it is troublesome because the whole flow must be received first and only then anomaly detection can be applied, revealing malicious flows.
+Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
+At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
 In \cite{hartl_explainability_2019}, authors show that such an architecture has similar performance to other flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends and analyzed its robustness with respect to adversaries extensively.
@@
 This follows from the fact that a rationally-acting organization will only implement an \gls{ids} if the cost saved by preventing security breaches is higher than the cost of the \gls{ids} itself.
-Noteworthy that the cost of an \gls{ids} can be divided into (a) the cost of purchase/installation, (b) the cost of operation and (c) the cost of maintenance.
-The goal we pursue here is to develop autonomous sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data.
-Specifically, the sampling technique should:
+The cost of an \gls{ids} can be divided into (a) the cost of purchase/installation, (b) the cost of operation and (c) the cost of maintenance.
+The goal we pursue here is to develop optimal sampling strategies that do not significantly degrade classification performance while being able to only process a small fraction of the original data.
+Specifically, a good sampling technique should:
 
@@
 % Fares:
-The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both later quantities.
-However, in certain situations, a certain sparsity is to be achieved, hence we further develop and propose a steering system that fluctuates such a trade-off parameter in order to obtain a coveted sparsity in a closed control loop.
+The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both quantities.
+However, as in some deployments, a certain sparsity has to be minimally achieved to limit the computational expense, we also develop a steering system that adapts the trade-off parameter in a closed control loop to not exceed the given budget of computation power.
 
-To encourage reproducibility and facilitate experimentation, we publicly release the source code, the data and the figures of this work\footnote{\url{https://github.com/CN-TU/adversarial-recurrent-ids/tree/rl}}.
+To encourage reproducibility and facilitate experimentation, we publicly release the source code, the trained ML models, the data and the figures of this work\footnote{\url{https://github.com/CN-TU/adversarial-recurrent-ids/tree/rl}}.
 
@@
 
-\cite{hartl_explainability_2019} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
+\cite{hartl_explainability_2019} build an LSTM-based \cite{hochreiter_long_1997} recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
 
@@
 
-\cite{campos_skip_2018} develop a technique that includes a \textit{skip gate} into LSTM or GRU \todo{citation needed} cells.
-It works without \gls{rl} but on the downside, their method can only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier, while we want to be independent of it.
+\cite{campos_skip_2018} develop a technique that includes a \textit{skip gate} into LSTM or GRU \cite{cho_learning_2014} cells.
+It works without \gls{rl} but on the downside, their method can only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier, while we want a solution in which the sampling procedure and the classifier are independent.
 
@@
 At each step they decide whether to use the reduced or the full state vector and can reduce computational cost by using the reduced one.
-For network traffic we think that it is often not even necessary to have a reduced state vector since from some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all.
+For network traffic we think that it is often not even necessary to have a reduced state vector since for some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all.
 
@@
 
-\section{Dataset}
-
-\note{blue}{Fares: if we submit to CNS (page limit is 9), we can remove this sections as it is not relevant and statistics can be found online}
-
-\begin{table}[h]
-\caption{Occurrence frequency of the attack flows.}
-\label{tab:occurrence}
-\centering
-\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
-]{
-\begin{tabular}{l r}
-\toprule
-Attack type & \hspace*{-4mm}Proportion \\ \midrule
-DoS Hulk	&	10.10\%	\\
-PortScan, Firewall	&	6.90\%	\\
-DDoS LOIT	&	4.08\%	\\
-Infiltration	&	3.30\%	\\
-DoS GoldenEye	&	0.32\%	\\
-DoS SlowHTTPTest	&	0.18\%	\\
-DoS Slowloris	&	0.17\%	\\
-Brute-force SSH	&	0.11\%	\\
-Botnet ARES	&	0.03\%	\\
-XSS attack	&	0.03\%	\\
-PortScan, no Fw.	&	0.02\%	\\
-Brute-force FTP	&	0.01\%	\\
-SQL injection	&	$<$0.01\%	\\
-Heartbleed	&	$<$0.01\%	\\
-\bottomrule
-\end{tabular}
-}{}
-\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
-]{
-\begin{tabular}{l r}
-\toprule
-Attack type & \hspace*{-4mm}Proportion \\ \midrule
-Exploits	&	1.42\%	\\
-Fuzzers	&	1.01\%	\\
-Reconnaissance	&	0.58\%	\\
-Generic	&	0.21\%	\\
-DoS	&	0.19\%	\\
-Shellcode	&	0.08\%	\\
-Analysis	&	0.03\%	\\
-Backdoors	&	0.02\%	\\
-Worms	&	0.01\%	\\
-\bottomrule
-\end{tabular}
-}{}
-\end{table}
-
-For evaluation, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15_2015} datasets, each includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
-The distribution of different attacks in the datasets are shown in \autoref{tab:occurrence}.
-
 \section{\ours{}}
 
-Traditionally, network packet sampling is performed on backbone infrastructures where, due to large bandwidths, not every packet can be analyzed.
-Sampling consequently helps experts understand the behavior of their network without the need to capture each and every packet.
+%Commonly, network packet sampling is performed on backbone infrastructures where, due to large bandwidths, inspecting every packet in detail is expensive.
+Sampling is thus required to save computational power and expenses or even because current \gls{ml} and specifically \gls{dl} methods might even require more computational power than can be delivered by present-day hardware.
 Assuming that the behavior of the network follows a predefined underlying process, sampling is therefore not detrimental on average to statistical analysis or measurement.
 Although the dataset in this work does not come from a backbone network, sampling reduces the high computational effort due to additional processing required by the complex \gls{ml}-based \glspl{ids}.
-As a consequence, several intuitive but efficient sampling techniques can be applied in practice.
-In this Section, we discuss the architecture of our proposed sampling framework \ours.
+To fulfill the goals outlined in \autoref{sec:introduction}, in this section we propose the architecture of our proposed sampling framework \ours.
 
@@
 Additionally, we add the number of skipped packets since the last packet as a feature, which can help the classifier because otherwise it might be confused because of missing packets.
-During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
-We use Z-score normalization and a train/test split of 2:1.
+We used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
 
@@
 
-We opt for an \emph{\gls{ac}} approach (as shown in \autoref{fig:sampling}) modeled after \cite{mnih_asynchronous_2016} because we consider it being more interpretable than a Deep \gls{rl} approach based on a \textit{Deep Q Network} like \cite{mnih_playing_2013} for reasons we explain later \todo{Max: Add ref to the section where we explain this}.
+We opt for an \emph{\gls{ac}} approach (as shown in \autoref{fig:sampling}) modeled after \cite{mnih_asynchronous_2016} because we consider it being more interpretable than a Deep \gls{rl} approach based on a \textit{Deep Q Network} like \cite{mnih_playing_2013}, as outputs the reward that was expected and the reward that was actually achieved, which we explain in more detail in the next paragraphs.
 
@@
 This means that we want the reward to be higher than the Critic's expectation.
+Note that it would be also possible to unify both reward metrics in one and have the Critic have only one output, but we consider it to be more interpretable to have two separate outputs, which can be inspected by an operator.
 
@@
 It aims to change the distribution so that actions which result in a higher reward are chosen more frequently, maximizing the utility function.
-However, this strategy could lead to the Actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward (local maxima).
+However, this strategy could lead to the Actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward.
+This would be equivalent to the Actor being stuck in a local maximum.
 Thus, the Actor not only tries to change the distribution so that the actions become more optimal (higher reward) but at the same time also aims to maximize the entropy of the distribution so that alternative choices are still explored reasonably often.
 
-Specifically, at timestep $t$, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled.
-$a_t=1$ means that the next packet is inspected next, while, for example, $a_t=2$ means that one packet is skipped until the next that is chosen and so forth.
+Specifically, at packet $t$ of a network flow, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled.
+$a_t=1$ means that the next packet is inspected next, while, for example, $a_t=2$ means that one packet is skipped until the next that is chosen, $a_t = 3$ means that two are skipped and so forth.
 Besides wanting to make actions which result in a high reward more likely, the actor also seeks to optimize the entropy $H(\cdot)$ of the probability distribution so that it keeps exploring and doesn't get stuck with a suboptimal policy.
@@
 For performing continious actions we need a probability distribution which can be parametrized to have the mean between 0 and positive infinity.
-Furthermore it must have at least two parameters so that the standard deviation, which is required for the entropy (\autoref{eq:actor}), can be changed independent of the mean.
+Furthermore it must have at least two parameters so that the standard deviation, which is required for the entropy (\autoref{eq:actor}), can be changed independently of the mean.
 For example, the Exponential distribution can be parametrized to have its mean anywhere between 0 and positive infinity, however, it has only one parameter and thus the mean and the standard deviation cannot be independently changed.
@@
 \end{align}
-is log-normally distributed variable from $\mathcal{LN}(\mu, \sigma^2)$.
+is a log-normally distributed variable.
+Note that the Log-normal distribution's mean $\mu$ and its standard deviation $\sigma$ are different from $\mu_\text{normal}$ and $\sigma_\text{normal}$, which are used to parametrize the distribution.
 
 The architecture in \autoref{fig:neuralNetworkArchitecture} shows that the actor network has two outputs: $\mu$ and $\sigma$.
-These encode the mean and standard deviation of a log-normal distribution and can never be smaller or equal to zero hence, we always apply a rectifier function at the output.
+These encode the mean and standard deviation of a log-normal distribution and can never be smaller or equal to zero and hence we always apply a rectifier function at the output.
 We choose the softplus, which is a smooth rectifier function:
@@
 When experimenting with continuous actions, we encountered one practical problem: At the beginning of training, the classifier would still perform badly and would not give the actor a useful reward.
-As a result, the actor would only focus on and maximize sparsity, as the classification reward would always be equally bad no matter how many packets are chosen.
+As a result, the actor would only focus on sparsity and aim to maximize it, as the classification reward would always be equally bad no matter how many packets are chosen.
 Thus, the actor would learn to let $\mu$ go to infinity as this would increase the reward at the beginning when the interaction between the three components of our \gls{rl} system is still fragile.
@@
 \label{sec:other_sampling_techniques}
-In order to evaluate the performance of \ours, we compare it with a few rule-based sampling techniques widely used in practice.
-We opt for three sampling families that do not require feedback from higher layers (i.e.
+In order to evaluate the performance of \ours, we compare it with a few straightforward sampling techniques.
+We opt for three sampling families that do not require feedback from higher layers.% (i.e.
 sampling rates are constant and not correlated with subsequent analysis): %The underlying principle of each family is summarized as follows:
+% Max: Don't understand
 
@@
 If the sequence is large enough, $n$ packets will be randomly sampled from a uniform distribution whereby $n= p \times T$.
-Practically, these techniques work properly (the effect on subsequent analysis is minimal) if the candidate packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
+In practice, these technique works well (the negative effect on subsequent analysis is minimal) if several packets share have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
 
@@
 \subsection{Every \emph{i}th Sampling}
-Finally we use a periodic sampling technique that takes a sample each $i= 1/ \alpha$ packet.
+Finally we use a periodic sampling technique that takes a sample each $i= \frac{1}{p}$ packet.
 Similar to the last technique, the length of the flow is not needed in this case.
-In practice, this technique is best suited to keep the behavioral distribution of the sampled set similar to the original distribution, thus suited for scenarios where information is heterogeneously distributed between packets over an entire network flow.
+In practice, this technique is best suited to keep the distribution of the sampled set similar to the original distribution, and it is thus suited for scenarios where information is heterogeneously distributed between packets over an entire network flow.
 
@@
 %There exist many more rule-based sampling techniques e.g., temporal, adaptive etc. However, studying their behaviors is out of the scope of this paper.
+
+\section{Dataset}
+
+%\note{blue}{Fares: if we submit to CNS (page limit is 9), we can remove this sections as it is not relevant and statistics can be found online}
+% Max: Should be fine :)
+
+\begin{table}[h]
+\caption{Occurrence frequency of the attack flows.}
+\label{tab:occurrence}
+\centering
+\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
+]{
+\begin{tabular}{l r}
+\toprule
+Attack type & \hspace*{-4mm}Proportion \\ \midrule
+DoS Hulk	&	10.10\%	\\
+PortScan, Firewall	&	6.90\%	\\
+DDoS LOIT	&	4.08\%	\\
+Infiltration	&	3.30\%	\\
+DoS GoldenEye	&	0.32\%	\\
+DoS SlowHTTPTest	&	0.18\%	\\
+DoS Slowloris	&	0.17\%	\\
+Brute-force SSH	&	0.11\%	\\
+Botnet ARES	&	0.03\%	\\
+XSS attack	&	0.03\%	\\
+PortScan, no Fw.	&	0.02\%	\\
+Brute-force FTP	&	0.01\%	\\
+SQL injection	&	$<$0.01\%	\\
+Heartbleed	&	$<$0.01\%	\\
+\bottomrule
+\end{tabular}
+}{}
+\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
+]{
+\begin{tabular}{l r}
+\toprule
+Attack type & \hspace*{-4mm}Proportion \\ \midrule
+Exploits	&	1.42\%	\\
+Fuzzers	&	1.01\%	\\
+Reconnaissance	&	0.58\%	\\
+Generic	&	0.21\%	\\
+DoS	&	0.19\%	\\
+Shellcode	&	0.08\%	\\
+Analysis	&	0.03\%	\\
+Backdoors	&	0.02\%	\\
+Worms	&	0.01\%	\\
+\bottomrule
+\end{tabular}
+}{}
+\end{table}
+
+For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15_2015} datasets, which both include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+The distribution of different attacks in the datasets are shown in \autoref{tab:occurrence}.
+
+We use Z-score normalization and a train/test split of 2:1 for both datasets.
+To speed up experimentation, we cut flows at a maximum length of 20.
+As Figure \ref{fig:inspectingAll} shows, there are only few flows that exceed that length.
 
@@
 
-The results of the sampling part are discussed in this section.
-We study the sparsity-accuracy tradeoff first, then we evaluate both discrete and continuous actions and, finally, we compare our results with other sampling techniques.
-Overall, sampling based on \gls{rl} reveals that a high number of packets can be skipped while maintaining high performance.
+%The results of the sampling part are discussed in this section. 
+We study the effect of the sparsity-accuracy tradeoff first, then we evaluate both discrete and continuous actions and, finally, we compare our results with other sampling techniques.
+Overall, sampling based on \gls{rl} reveals that a high number of packets can be skipped while maintaining good accuracy.
 Then, we inspect how the \gls{rl} chooses packets and if some patterns can be spotted.
@@
 At the same time, sparsity gets higher as the \gls{rl} learns which packets can be skipped since they do not influence classification performance.
+An important implementation detail is that we always start with the first packet of a flow, meaning that it cannot be skipped.
+Also, we set the entropy parameter to $0.01$.
+
+Opposed to the behavior during training, we do not randomly sample from the probability distribution of the actor during testing but instead take the mean of the distribution for the continuous case and the mode for the discrete one.
+This is reasonable and best practice \cite{mnih_asynchronous_2016} since during testing we do not aim to explore the possible choices anymore since they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore.
 
@@
 
-\note{blue}{Fares: my suggestion:
-The only exception to this is that for tradeoffs of $0.1$ and $0.5$ for UNSW-NB15, the accuracy is almost the same or even slightly higher for the larger tradeoff, which is counter-intuitive.
+%\note{blue}{Fares: my suggestion:
+%The only exception to this is that for tradeoffs of $0.1$ and $0.5$ for UNSW-NB15, the accuracy is almost the same or even slightly higher for the larger tradeoff, which is counter-intuitive.
 We assume that due to the nature of the UNSW-NB15 data set (which is highly heterogeneous), fewer packets yield better information (and therefore the memory of the LSTM classifier is cleaner) and hence better classification decision and higher accuracy.
-}
+%}
+% Max: Could be but it is too speculative in my opinion. From what I saw, neural networks depend quite a lot on the initialization and on luck, so it might just be that it was lucky...
 
 Furthermore, due to the fact that UNSW-NB15 contains more packets per flow, its sparsity is higher.
-This, on the other hand, holds since almost the same number of packets is sampled from each flow for both datasets (see \autoref{sec:conclusion}).
+%This, on the other hand, holds since almost the same number of packets is sampled from each flow for both datasets (see \autoref{sec:conclusion}).
+% Max: I don't understand the sentence commented out above...
 
@@
 This leads to the conclusion, that at least for our use case and the two datasets that we consider, there is no consistent difference in performance for continuous and discrete actions.
-Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean log-normal distribution, which can lead to worse performance.
+Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean of the log-normal distribution, which can lead to worse performance.
 On the other hand, with discrete actions, the maximum number of samples to skip is fixed, which we consider to be a major drawback.
-Furthermore, training time is potentially higher since, for example, for a discrete space of 20 actions, each option is treated as being completely different while actually there is not a large difference between skipping 18 or 19 packets, there is a big difference between skipping 0 packets compared to skipping 19 packets.
-This means that for discrete actions, all actions are perceived as being completely independent, which is not true as the reward function considers sparsity, which increases gradually as more packets are skipped and for which similar actions result in similar rewards.
+Furthermore, training time is potentially higher since, for example, for a discrete action space of 20 actions, each option is treated as being completely different and, while actually there is not a large difference between skipping 18 or 19 packets, there is a big difference between skipping 0 packets compared to skipping 19 packets.
+This means that for discrete actions, all actions are perceived as being completely independent, which is not true as the reward function considers sparsity, which increases gradually as more packets are skipped and thus similar actions result in similar rewards.
 
@@
 
-Next we compared the \gls{rl} technique usedy by \ours{} with several intuitive sampling strategies suggested in \autoref{sec:other_sampling_techniques}.
+Next we compared the \gls{rl} technique used by \ours{} with several intuitive sampling strategies suggested in \autoref{sec:other_sampling_techniques}.
 \autoref{tab:results_others} reveals that \ours{} achieves the highest accuracy with the same sparsity.
-Not surprisingly, random sampling performs the worst, as it is likely that a flow is either ``misfortune'' and doesn't get enough packets sampled while another  one gets more or, the random packets chosen carry less useful information.
+Not surprisingly, random sampling performs the worst, as it is likely that a flow is either ``misfortunate'' and doesn't get enough packets sampled while another  one gets more packets or the packets that carry more useful information.
 The best strategy besides the \gls{rl} approach seems to be ``first $n$ equal'' sampling, which matches the conclusion from \cite{hartl_explainability_2019} that the first couple of packets are the most important ones for determining maliciousness of a flow.
-\note{blue}{Fares: The next sentence is a bit confusing.}
 ``first $n$ equal'' performs better than ``first $n$'' because for ``first $n$'', longer flows will have more packets sampled from them.
 Conversely, very short flows will get few packets.
-But since the first packets of a flow determine its maliciousness, the short flows get misclassified because they don't get enough packets sampled.
+For example, with $p=0.2$, a flow of length 20 will get 4 packets while one with 4 packets will get only 1.
+However, with `first $n$ equal'', the short and the long flow will get the same number of packets.
+Apparently it is better if all flows get the same number of packets than if long flows get more packets.
+`
 
@@
 
+One concern could be the overly deterministic nature of the decisions of \ours{} for some attack types: An adversary could see that only certain packets of a flow are looked at by the \gls{rl} system and thus only manipulate the few packets that the classifier looks at to generate adversarial samples with methods as studied by \cite{hartl_explainability_2019}.
+One potential remedy for this is to also randomly sample from the distribution of the actor during testing like it is done during training.
+This could result in slightly worse accuracy but lead to behavior that is harder to predict for an adversary.
+Currently we take the mean of the probability distribution for continuous actions and the mode for the discrete one, resulting in deterministic behavior.
+
 \section{Steering}
@@
 We start with a maximum sparsity of ~74.4\% and try to reach 50\%.
-In order to do so, the steering system start decreasing the tradoff parameter until it reaches 0.
-It might seem surprising that even with such value, which means that only classification matters and not sparsity at all, a sparsity of 73.5\% is still achieved.
-We attribute this (1) to the fact that \ours{} learns that some flows may contain packets that do not contribute in determining maliciousness and are skipped since they do not affect the accuracy and (2) the actor still in any case attempts to maximize the entropy of the output which induces skipping packets for fat-tailed distributions.
-%Nothworthy is the fact that the goal sparsity (50\%) is not reached due to the previous assumptions, however, a decreasing slope is clearly shown in figure 4
+To do so, the steering system starts decreasing the tradeoff parameter until it reaches 0.
+It might seem surprising that even with a tradeoff of 0, which means that only classification matters and not sparsity not at all, a sparsity of 73.5\% is still achieved.
+We attribute this to the fact that \ours{} learns that some flows may contain packets that do not contribute in determining maliciousness and are skipped since they do not affect the accuracy.
+%\item the actor still in any case attempts to maximize the entropy of its distribution which means that when no packets are skipped, entropy is very low.
+\end{enumerate*}%Nothworthy is the fact that the goal sparsity (50\%) is not reached due to the previous assumptions, however, a decreasing slope is clearly shown in figure 4
 
@@
 \label{sec:conclusion}
-In this paper we try to solve the practical problem of accurate network traffic classification with less computational power.
-To this end, we propose \ours{} which is a framework for packet sampling based on \gls{rl}.
+In this paper we aimed to develop smart packet sampling methods for the use case of \glspl{ids} to save computational power.
+To this end, we proposed \ours{} which is a framework for packet sampling based on \gls{rl}.
 
 % Max: 1/4 is already pronounced one fourth
-We prove that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity works as expected and that higher tradeoffs resulted in higher sparsity and potentially lower classification performance.
-Surprisingly, we saw that our solution \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}$ for CIC-IDS-2017 and $\approx \frac{1}{5}$ for UNSW-NB-15.
+We show that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity indeed results in higher sparsity the larger the tradeoff parameter while potentially sacrificing classification performance.
+Surprisingly, we saw that \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}$ for CIC-IDS-2017 and $\approx \frac{1}{5}$ for UNSW-NB-15.
 Furthermore, our results showed that \gls{rl}-based sampling can intelligently choose relevant packets for intrusion detection and outperform non-smart sampling techniques.
@@
 We showed that a naive implementation would not work as the training is too fragile in the beginning and the \gls{rl} system learned to maximize only one of the rewards (sparsity) in the beginning.
-This was solved by adding a penalty for jumping behind the end of the sequence which was proved effective.
+This was solved by adding a penalty for jumping behind the end of the sequence which proved effective.
 Comparing with discrete actions, continuous actions seem to perform on par.
 This is still an advantage, as for continuous actions the neural network only needs to have two outputs (mean and standard deviation) while for discrete ones it needs many outputs as there are actions.
-The continuous action framework is thus more lightweight.
+The continuous action framework is thus more lightweight and also can learn to skip arbitrarily many packets.
 

### 2020-02-04 learn.py

@@ -23,6 +23,7 @@ import matplotlib.pyplot as plt
 from sklearn.metrics import accuracy_score,recall_score, precision_score, f1_score, balanced_accuracy_score
 
 ADVERSARIAL_THRESH = 50
+CLAMPING_CONSTANT = math.log(1e-20)
 
 def output_scores(y_true, y_pred):
 	accuracy = accuracy_score(y_true, y_pred)
@@ -828,7 +829,7 @@ def train_rl():
 			effective_output_rl_critic_added = output_rl_critic_added[mask_rl_first].view(-1)
 			# assert (effective_output_rl_critic_added.shape == effective_rewards_sparsity.shape)
 			current_entropy = effective_rl_actor_dists.entropy()
-			log_prob = effective_rl_actor_dists.log_prob(effective_output_rl_actor_decisions)
+			log_prob = torch.clamp(effective_rl_actor_dists.log_prob(effective_output_rl_actor_decisions), CLAMPING_CONSTANT, float("inf"))
 			assert effective_output_rl_critic_added.shape == log_prob.shape == current_entropy.shape == effective_rewards_sparsity_with_tradeoff.shape == effective_rewards_classification.shape
 			loss_rl_actor = torch.mean(- log_prob*((effective_rewards_sparsity_with_tradeoff + effective_rewards_classification) - effective_output_rl_critic_added) - opt.entropyRegularizationMultiplier*current_entropy)
 

### 2020-02-04 paper_rl/paper.tex

@@
 
-For all our experiments we train for at least 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.
-\autoref{fig:training} shows how during training, accuracy increases quickly in the beginning and then gradually afterwards.
+For all our experiments we train our networks for at least 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.
+\autoref{fig:training} shows how, during training, accuracy increases quickly at the beginning and then gradually afterwards.
 At the same time, sparsity gets higher as the \gls{rl} learns which packets can be skipped since they do not influence classification performance.
@@
 
-Finally, we implement a steering method with which a certain sparsity threshold can be specified which limits the computational expenses of an operator.
+Eventually, we implement a steering method to help achieving a certain sparsity threshold. The steering mechanism can be used to limit the computational expenses of an operator.
 

### 2020-02-04 learn.py
Fix packet sampling figures' legends

@@ -23,7 +23,7 @@ import matplotlib.pyplot as plt
 from sklearn.metrics import accuracy_score,recall_score, precision_score, f1_score, balanced_accuracy_score
 
 ADVERSARIAL_THRESH = 50
-CLAMPING_CONSTANT = math.log(1e-20)
+CLAMPING_CONSTANT = math.log(1e-10)
 
 def output_scores(y_true, y_pred):
 	accuracy = accuracy_score(y_true, y_pred)

### 2020-02-04 paper/paper.tex
Introduce feature sensitivity to abstract, introduction and conclusion

@@
 Feature importance methods show that even relatively unimportant features can be effectively abused to generate adversarial samples.
-Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids}, capturing a common notion of adversarial robustness, and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}.
+We thus introduce the concept of \textit{feature sensitivity} which quantifies how much potential a feature has to cause misclassification.
+
+Since traditional evaluation metrics such as accuracy are not sufficient for quantifying the adversarial threat, we propose the \gls{ars} for comparing \glspl{ids} %, capturing a common notion of adversarial robustness,
+and show that an adversarial training procedure can significantly and successfully reduce the attack surface.% and improve \gls{ars}.
 %as a countermeasure.
@@
 are not even particularly important for the \gls{rnn}'s classification outcome. %Especially the first packets of a flow are critical for determining the decision of a recurrent \gls{ids}.
+Thus, we propose \textit{feature sensitivity} methods, which show how prone a feature is to cause misclassification.
 
@@
 \end{tabular}
-\vspace*{-.3mm}
+%\vspace*{-0.3cm}
 \end{table}
@@
 \subsection{Reducing Features}
-The most obvious defense, which is only possible because of the nature of the dataset, is to simply omit features an attacker can manipulate. We try two different approaches:
+The most obvious defense is to simply omit features an attacker can manipulate. We try two different approaches:
 \begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
 \item Leaving out all manipulable features: packet size and \gls{iat}
-\item Leaving out manipulable features only when the attacker can manipulate them, i.e.
-only in the direction from the attacker to the victim.
-This, however, does not prevent adversarial samples for botnets, for which both sides are malicious.
+\item Leaving them out only in the direction from the attacker to the victim. This, however, does not prevent adversarial samples for botnets, for which the attacker has control over both sides.
 \end{itemize}
@@
 A question which occurs in this process, is how many CPU cycles to spend on network training and adversarial sample optimization.
-As a natural approach, we decided to use as many backpropagation steps for neural network training as we use for adversarial sample optimization.
+We decided to use as many backpropagation steps for training as we use for adversarial sample optimization.
 For each adversarial sample optimization step, we performed 10 iterations of gradient descent.
@@
 We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified.
+Newly developed \textit{Feature sensitivity} methods can show which features can easily be manipulated by an adversary.
 We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.

### 2020-02-04 paper_rl/paper.tex
Cleanup

@@
 
-The Critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a timestep $t$.
+The critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a timestep $t$.
 For this it outputs $v_{\text{classification},t}$ and $v_{\text{sparsity},t}$ at each timestep $t$.
@@
 
-Now that the reward metrics and the Critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each timestep $t$ as follows:
+Now that the reward metrics and the critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each timestep $t$ as follows:
 \begin{align}
@@
 
-This means that we want the reward to be higher than the Critic's expectation.
-Note that it would be also possible to unify both reward metrics in one and have the Critic have only one output, but we consider it to be more interpretable to have two separate outputs, which can be inspected by an operator.
+This means that we want the reward to be higher than the critic's expectation.
+Note that it would be also possible to unify both reward metrics in one and have the critic have only one output, but we consider it to be more interpretable to have two separate outputs, which can be inspected by an operator.
 
-The Actor outputs a probability distribution at timestep $t$, which is sampled to determine the number of packets that should be skipped.
+The actor outputs a probability distribution at timestep $t$, which is sampled to determine the number of packets that should be skipped.
 It aims to change the distribution so that actions which result in a higher reward are chosen more frequently, maximizing the utility function.
-However, this strategy could lead to the Actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward.
-This would be equivalent to the Actor being stuck in a local maximum.
-Thus, the Actor not only tries to change the distribution so that the actions become more optimal (higher reward) but at the same time also aims to maximize the entropy of the distribution so that alternative choices are still explored reasonably often.
+However, this strategy could lead to the actor getting stuck with what it considers the best decision and never try alternatives even though they might result in a higher reward.
+This would be equivalent to the actor being stuck in a local maximum.
+Thus, the actor not only tries to change the distribution so that the actions become more optimal (higher reward) but at the same time also aims to maximize the entropy of the distribution so that alternative choices are still explored reasonably often.
 
-Specifically, at packet $t$ of a network flow, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the Actor $\theta_a$ the Actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled.
+Specifically, at packet $t$ of a network flow, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the actor $\theta_a$ the actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled.
 $a_t=1$ means that the next packet is inspected next, while, for example, $a_t=2$ means that one packet is skipped until the next that is chosen, $a_t = 3$ means that two are skipped and so forth.
@@
   \caption{SparseIDS framework.
-Three deep networks are shown: an LSTM classifier, An Actor network and a Critic network.
-The Actor takes actions based on the feedback from the the Critic network and the classifier.
-The Critic network tracks the Actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the Actor network.}
+Three deep networks are shown: an LSTM classifier, An actor network and a critic network.
+The actor takes actions based on the feedback from the the critic network and the classifier.
+The critic network tracks the actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the actor network.}
   \label{fig:sampling}
@@
 This is reasonable and best practice \cite{mnih_asynchronous_2016} since during testing we do not aim to explore the possible choices anymore since they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore.
+Furthermore, the critic is only necessary during training for the reward of the actor.
+During testing, it can be disabled, saving computational power.
 

### 2020-02-05 paper_rl/paper.tex
Paper draft
Former-commit-id: d123f9f0682384e9159dde997d326157673054ce [formerly 7e507fbecfb0474707bdf96ca9995181ca955ddf]
Former-commit-id: 461dc62c394e691999898683d0300471af15b7af

@@
 \glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data.
-They allow analyzing network flows and already determining if a flow is malicious or not before it is over, making it possible to take action immediately.
+They allow already determining if a flow is malicious or not before it is over, making it possible to take action immediately.
 However, considering the large number of packets that have to be inspected, the question of computational as well as energy efficiency arises.
@@
 Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy with the same number of sampled packets.
-A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of packets like current approaches but can possibly skip arbitrarily many packets, saving even more computational resources for long sequences.
+A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can possibly skip arbitrarily many packets, saving even more computational resources for long sequences.
 Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows.
@@
 As the need for security mechanisms in today's infrastructure's communication networks grows, a substantial body of scientific work focuses on developing reliable and explainable frameworks for the identification of unwanted behavior.
-In the past, a viable way to detect intrusion was to analyze the contents of the packets themselves and determine, for example, whether a packet contains potentially harmful content by matching patterns.
+In the past, a viable way to detect intrusions was to analyze the contents of the packets themselves and determine, for example, whether a packet contains potentially harmful content by matching patterns.
 More recently, with the increasing deployment of encryption, the focus is now on features that are always available for network monitoring devices such as port numbers, protocol flags and packet sizes when encrypting above the transport layer.
@@
 When analyzing flows, not only the aforementioned features are available but also features representing the temporal behavior of the individual packets.
-Various approaches have been proposed to extract flow features after which anomaly detection can be performe: In \cite{meghdouri_analysis_2018} authors compare such approaches and shows their efficacy.
-While these approaches often work well, it is troublesome because the whole flow must be received first and only then anomaly detection can be applied, revealing malicious flows.
+Various approaches have been proposed to extract flow features with which anomaly detection can be performed: In \cite{meghdouri_analysis_2018} compare such approaches and show their efficacy.
+While these approaches often work well, a major drawback is that the whole flow must be received first and only then anomaly detection can be applied, revealing malicious flows.
 Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
-At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-In \cite{hartl_explainability_2019}, authors show that such an architecture has similar performance to other flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends and analyzed its robustness with respect to adversaries extensively.
+In this way, an \gls{rnn}-based \gls{ids} has the benefit of avoiding feature engineering procedures and let the classifier build high level features by itself.
+\cite{hartl_explainability_2019} show that such an architecture has similar performance to traditional flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends.
+Furthermore they analyzed its robustness with respect to adversaries extensively.
 
@@
 \item \textbf{choose optimally}, taking only samples that contain the most information and skip the less relevant ones.
-\item have a parameter that allows to \textbf{trade off} classification performance for sparsity.
+\item have a parameter that allows to \textbf{trade off} classification performance for sparsity (choosing fewer packets).
 \item be \textbf{independent of the classifier} so that any classifier can be used and that the classifier doesn't have to be aware of the sampling strategy.
@@
 % Fares:
-For this purpose we develop a \gls{rl}-based \gls{ids}: \textit{\ours{}} that fulfills the above properties.
+For this purpose we develop an \gls{rl}-based \gls{ids}: \textit{\ours{}} that fulfills the above properties.
 We show that a significant number of packets can be skipped while the accuracy does not drop
@@
 The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both quantities.
-However, as in some deployments, a certain sparsity has to be minimally achieved to limit the computational expense, we also develop a steering system that adapts the trade-off parameter in a closed control loop to not exceed the given budget of computation power.
+However, as in some deployments, a certain sparsity has to be minimally achieved to limit the computational expense, we also develop a steering system that adapts the trade-off parameter in a closed control loop.
+In this way the given budget of computation power is not exceeded.
 
@@
 
-These proposed techniques focus their evaluation on text and hence do not fulfill certain criteria we set out in \autoref{sec:introduction}.
+These proposed techniques focus their evaluation on text and hence do not fulfill all criteria we set out in \autoref{sec:introduction}.
 Furthermore, all techniques that use \gls{rl} use discrete actions, meaning that there is a hyperparameter $k$, which influences the maximum jump that is possible.
@@
 It is thus beneficial to not have a fixed maximum step size but a continuous one.
+This allows skipping arbitrarily many packets.
 
@@
 We implemented a three-layer LSTM-based classifier with 128 neurons at each layer.
+We choose 128 neurons as we don't notice a significant degradation of performance compared to the results of \cite{hartl_explainability_2019} (check Table II, flow accuracy) and as fewer neurons allowed for more agility during the training phase and allowed us to run more experiments.
 As the input features we use source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e.
@@
 
+Intuitively, the classification reward makes sure, that packets are chosen that contain the maximum accuracy and thus result in the highest classification performance.
+
 The sparsity reward metric (\autoref{eq:R_sparsity}) captures how much sparsity is achieved on average for all future packets of a flow.
@@
 
-We opt for an \emph{\gls{ac}} approach (as shown in \autoref{fig:sampling}) modeled after \cite{mnih_asynchronous_2016} because we consider it being more interpretable than a Deep \gls{rl} approach based on a \textit{Deep Q Network} like \cite{mnih_playing_2013}, as outputs the reward that was expected and the reward that was actually achieved, which we explain in more detail in the next paragraphs.
+We opt for an \emph{\gls{ac}} approach (as shown in \autoref{fig:sampling}) modeled after \cite{mnih_asynchronous_2016} because we consider it being more interpretable than a Deep \gls{rl} approach based on a \textit{Deep Q Network} like \cite{mnih_playing_2013}, as it outputs the reward that was expected and the reward that was actually achieved, which we explain in more detail in the next paragraphs.
 
@@
 For this, it outputs its confidence for non-attack/attack after reading a packet.
-As stated in \autoref{subsec:classifier}, we choose an LSTM-based supervised classifier for this work, but our framework can work together with any classifier as long as it outputs a confidence for non-attack/attack at each timestep and can handle skipped packets.
-For instance it would also be possible to replace the LSTM with an unsupervised autoencoder like in \cite{mirsky_kitsune_2018}.
+As stated in \autoref{subsec:classifier}, we choose an LSTM-based supervised classifier for this work, but our framework can work together with any classifier as long as it outputs a confidence for non-attack/attack at each packet and can handle skipped packets.
+For instance it would also be possible to replace the LSTM with an unsupervised autoencoder akin to \cite{mirsky_kitsune_2018}.
 
-The critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a timestep $t$.
-For this it outputs $v_{\text{classification},t}$ and $v_{\text{sparsity},t}$ at each timestep $t$.
-It aims to predict the classification performance and sparsity correctly for each timestep by minimizing the following value loss function:
+The critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a timestep $t$ (a timestep is a packet).
+For this, given the current input vector and the state of the LSTM cells $s_t$ and the neural network weights of the critic $\theta_v$, it outputs $v_{\text{classification},t}$ and $v_{\text{sparsity},t}$ at each timestep $t$.
+It aims to predict the classification performance and sparsity correctly for each timestep by minimizing the following loss function:
 \begin{align}
 \begin{split}
-l_{\text{v},t} =& \left(R_{\text{classification},t} - v_{\text{classification},t}\right)^2 \\
-+ & \left(R_{\text{sparsity},t} - v_{\text{sparsity},t}\right)^2
+l_{\text{v},t} =& \left(R_{\text{classification},t} - v_{\text{classification},t}\left(s_t ; \theta_\text{c} \right)\right)^2 \\
++ & \left(R_{\text{sparsity},t} - v_{\text{sparsity},t}\left(s_t ; \theta_\text{c} \right)\right)^2
 \end{split}
@@
 
-Specifically, at packet $t$ of a network flow, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the actor $\theta_a$ the actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled.
+Specifically, at packet $t$ of a network flow, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the actor $\theta_a$, the actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled.
 $a_t=1$ means that the next packet is inspected next, while, for example, $a_t=2$ means that one packet is skipped until the next that is chosen, $a_t = 3$ means that two are skipped and so forth.
@@
   \caption{SparseIDS framework.
-Three deep networks are shown: an LSTM classifier, An actor network and a critic network.
+Three deep networks are shown: the LSTM classifier, the actor network and the critic network.
 The actor takes actions based on the feedback from the the critic network and the classifier.
-The critic network tracks the actor actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the actor network.}
+The critic network tracks the actor's actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the actor network.}
   \label{fig:sampling}
@@
 For example, the Exponential distribution can be parametrized to have its mean anywhere between 0 and positive infinity, however, it has only one parameter and thus the mean and the standard deviation cannot be independently changed.
-Conversely, a simple distribution which fits our requirements is the Log-normal distribution, which is defined as a normal distribution whose output is exponentiated:
+Conversely, a simple distribution which fits our requirements is the log-normal distribution, which is defined as a normal distribution whose output is exponentiated:
 
@@
 is a log-normally distributed variable.
-Note that the Log-normal distribution's mean $\mu$ and its standard deviation $\sigma$ are different from $\mu_\text{normal}$ and $\sigma_\text{normal}$, which are used to parametrize the distribution.
+Note that the log-normal distribution's mean $\mu$ and its standard deviation $\sigma$ are different from $\mu_\text{normal}$ and $\sigma_\text{normal}$, which are used to parametrize the distribution.
 
@@
 As a result, the actor would only focus on sparsity and aim to maximize it, as the classification reward would always be equally bad no matter how many packets are chosen.
-Thus, the actor would learn to let $\mu$ go to infinity as this would increase the reward at the beginning when the interaction between the three components of our \gls{rl} system is still fragile.
-As a solution, we add a penalty to the reward calculation when the actor jumps far behind the last flow packet.The modified reward applies only to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet.
-The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow ($t=last$):
+Thus, the actor would learn to let $\mu$ go to infinity.
+This would increase the reward at the beginning when the interaction between the three components of our \gls{rl} system is still fragile.
+As a solution, we add a penalty to the reward calculation when the actor jumps far behind the last flow packet.
+The modified reward applies only to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet.
+The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow ($t=\textit{last}$):
 
@@
 In order to evaluate the performance of \ours, we compare it with a few straightforward sampling techniques.
-We opt for three sampling families that do not require feedback from higher layers.% (i.e.
+We opt for three sampling families that, unlike \ours{} cannot adapt their strategy based on packet contents.
+% (i.e.
 sampling rates are constant and not correlated with subsequent analysis): %The underlying principle of each family is summarized as follows:
@@
 If the sequence is large enough, $n$ packets will be randomly sampled from a uniform distribution whereby $n= p \times T$.
-In practice, these technique works well (the negative effect on subsequent analysis is minimal) if several packets share have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
+In practice, these technique works well if several packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
 
@@
 
-For all our experiments we train our networks for at least 8 epochs ($>10$ million flows for both datasets) and then evaluate the resulting neural networks on the test dataset.
+For all our experiments we train our networks for at least 8 epochs ($>10$ million flows for both datasets) with the Adam optimizer with a learning rate of $0.001$.
+An important implementation detail is that we always start with the first packet of a flow, meaning that it cannot be skipped.
+Also, we set the entropy parameter to $0.01$.
 \autoref{fig:training} shows how, during training, accuracy increases quickly at the beginning and then gradually afterwards.
 At the same time, sparsity gets higher as the \gls{rl} learns which packets can be skipped since they do not influence classification performance.
-An important implementation detail is that we always start with the first packet of a flow, meaning that it cannot be skipped.
-Also, we set the entropy parameter to $0.01$.
+%We then evaluate the resulting neural networks on the test dataset.
 
-Opposed to the behavior during training, we do not randomly sample from the probability distribution of the actor during testing but instead take the mean of the distribution for the continuous case and the mode for the discrete one.
-This is reasonable and best practice \cite{mnih_asynchronous_2016} since during testing we do not aim to explore the possible choices anymore since they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore.
+Opposed to the behavior during training, we do not randomly sample from the probability distribution of the actor during testing.
+Instead take the mean of the distribution for the continuous case and the mode for the discrete one.
+This is reasonable and best practice \cite{mnih_asynchronous_2016} since during testing we do not aim to explore the possible choices anymore as they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore.
 Furthermore, the critic is only necessary during training for the reward of the actor.
-During testing, it can be disabled, saving computational power.
+During deployment, it can be disabled, saving computational power.
 
@@
 
-Furthermore, due to the fact that UNSW-NB15 contains more packets per flow, its sparsity is higher.
+The fact that for UNSW-NB15 sparsity is higher is caused by the fact that it has more packets per flow on average and thus more redundant information that the \gls{rl} learns to skip.
 %This, on the other hand, holds since almost the same number of packets is sampled from each flow for both datasets (see \autoref{sec:conclusion}).
@@
 \autoref{tab:results_others} reveals that \ours{} achieves the highest accuracy with the same sparsity.
-Not surprisingly, random sampling performs the worst, as it is likely that a flow is either ``misfortunate'' and doesn't get enough packets sampled while another  one gets more packets or the packets that carry more useful information.
+Not surprisingly, random sampling performs the worst, as it is likely that a flow is ``unfortunate'' and doesn't get enough packets sampled while another one gets more packets or the packets that carry more useful information.
 The best strategy besides the \gls{rl} approach seems to be ``first $n$ equal'' sampling, which matches the conclusion from \cite{hartl_explainability_2019} that the first couple of packets are the most important ones for determining maliciousness of a flow.
@@
 Apparently it is better if all flows get the same number of packets than if long flows get more packets.
-`
 
@@
 This contrasts with SSH Patator (Figure \ref{fig:inspectingPatator}), where the 0th, the 3rd, the 13th and the 15th packet are always chosen, since they appear to be especially indicative of an attack.
-Also, the flows seems to contain no useful information between the 3rd and the 13th packet, so that this part is always skipped altogether.
+Also, the flows seem to contain no useful information between the 3rd and the 13th packet, so that this part is always skipped altogether.
 Slowloris (Figure \ref{fig:inspectingSlowloris}) is more similar to what we saw for benign traffic: The sampling strategy varies a lot for different flows and not always the same pattern is used.
@@
 \begin{enumerate}
-\item A model is trained where the trade-off is chosen randomly for each flow during training from a uniform distribution yielding values between 0 and 1.
+\item A model is trained for which the trade-off is chosen randomly for each flow during training from a uniform distribution yielding values between 0 and 1.
 We add a new feature to each packet that indicates the tradeoff chosen for the respective flow so that it can adjust its sampling strategy in anticipation of the tradeoff that will determine the reward.
@@
 % Max: 1/4 is already pronounced one fourth
-We show that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity indeed results in higher sparsity the larger the tradeoff parameter while potentially sacrificing classification performance.
+We show that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity indeed results in higher sparsity the larger the tradeoff parameter, while potentially classification performance is sacrificed.
 Surprisingly, we saw that \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}$ for CIC-IDS-2017 and $\approx \frac{1}{5}$ for UNSW-NB-15.
@@
 Comparing with discrete actions, continuous actions seem to perform on par.
-This is still an advantage, as for continuous actions the neural network only needs to have two outputs (mean and standard deviation) while for discrete ones it needs many outputs as there are actions.
+This is still an advantage, as for continuous actions the neural network only needs to have two outputs (mean and standard deviation) while for discrete ones it needs as many outputs as there are actions.
 The continuous action framework is thus more lightweight and also can learn to skip arbitrarily many packets.
@@
 
-In general, we think that \gls{rl}-based packet sampling is not only applicable to \gls{ids} but we are optimistic that other domains of networking can benefit from similar solutions to save computational power.
+In general, we think that \gls{rl}-based packet sampling is not only applicable to \gls{ids} but we are confident that other domains of networking can benefit from similar solutions to save computational expenses.
 

### 2020-02-05 learn.py
"Fix" underflow of standard deviation by setting a minimum standard deviation
Former-commit-id: 586ffb6ff18186c5a351a952ab0caabc8cbb094a [formerly 3424326dd65e46d6d120a5eac7a1b2c86e5b3daf]
Former-commit-id: 63e76a56d035ca66a654da3f5536cf6a6c88791a

@@ -23,7 +23,6 @@ import matplotlib.pyplot as plt
 from sklearn.metrics import accuracy_score,recall_score, precision_score, f1_score, balanced_accuracy_score
 
 ADVERSARIAL_THRESH = 50
-CLAMPING_CONSTANT = math.log(1e-10)
 
 def output_scores(y_true, y_pred):
 	accuracy = accuracy_score(y_true, y_pred)
@@ -666,7 +665,24 @@ def train_rl():
 				else:
 					current_slice_action_probs = torch.nn.functional.softplus(current_slice_action_probs)
 					locs, scales = compute_normal_distribution_from_log_normal(current_slice_action_probs[:,:,0], current_slice_action_probs[:,:,1])
+
+					if (scales < CLAMPING_CONSTANT).any():
+						print("scales underflowed!")
+						print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
+						print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
+						print("locs", locs)
+						print("scales", scales)
+						scales = torch.max(scales, CLAMPING_CONSTANT)
+
 					current_slice_action_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
+					# try:
+					# 	current_slice_action_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
+					# except ValueError as e:
+					# 	print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
+					# 	print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
+					# 	print("locs", locs)
+					# 	print("scales", scales)
+					# 	raise e
 				outputs_rl_actor.append(current_slice_action_probs)
 				decisions = current_slice_action_dists.sample()
 				outputs_rl_actor_decisions.append(decisions)
@@ -829,10 +845,9 @@ def train_rl():
 			effective_output_rl_critic_added = output_rl_critic_added[mask_rl_first].view(-1)
 			# assert (effective_output_rl_critic_added.shape == effective_rewards_sparsity.shape)
 			current_entropy = effective_rl_actor_dists.entropy()
-			log_prob = torch.clamp(effective_rl_actor_dists.log_prob(effective_output_rl_actor_decisions), CLAMPING_CONSTANT, float("inf"))
[…]

### 2020-02-05 learn.py
"Fix" underflow of standard deviation by setting a minimum standard deviation for good
Former-commit-id: 8aa8f3b411df7caa377f83c5e7c0d69680539774 [formerly 005071ef92f7a1889d9a61cc4aaf5ab2dad2f78e]
Former-commit-id: 5cf32a089f73c5ce228f96cfd9d429a65e5834e7

@@ -675,14 +675,7 @@ def train_rl():
 						scales = torch.max(scales, CLAMPING_CONSTANT)
 
 					current_slice_action_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
-					# try:
-					# 	current_slice_action_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
-					# except ValueError as e:
-					# 	print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
-					# 	print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
-					# 	print("locs", locs)
-					# 	print("scales", scales)
-					# 	raise e
+
 				outputs_rl_actor.append(current_slice_action_probs)
 				decisions = current_slice_action_dists.sample()
 				outputs_rl_actor_decisions.append(decisions)
@@ -829,6 +822,15 @@ def train_rl():
 				effective_rl_actor_dists = torch.distributions.categorical.Categorical(logits=effective_output_rl_actor, validate_args=True)
 			else:
 				locs, scales = compute_normal_distribution_from_log_normal(effective_output_rl_actor[:,0], effective_output_rl_actor[:,1])
+
+				if (scales < CLAMPING_CONSTANT).any():
+					print("scales underflowed!")
+					print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
+					print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
+					print("locs", locs)
+					print("scales", scales)
+					scales = torch.max(scales, CLAMPING_CONSTANT)
+
 				effective_rl_actor_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
 
 			effective_rewards_sparsity = rewards_sparsity_catted.detach()[mask_rl_first].view(-1)

### 2020-02-06 learn.py
Fixes

@@ -667,14 +667,22 @@ def train_rl():
 					locs, scales = compute_normal_distribution_from_log_normal(current_slice_action_probs[:,:,0], current_slice_action_probs[:,:,1])
 
 					if (scales < CLAMPING_CONSTANT).any():
-						print("scales underflowed!")
+						print("problem at top!")
 						print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
 						print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
 						print("locs", locs)
 						print("scales", scales)
 						scales = torch.max(scales, CLAMPING_CONSTANT)
 
-					current_slice_action_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
+					try:
+						current_slice_action_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
+					except ValueError as e:
+						print("crash at top!")
+						print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
+						print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
+						print("locs", locs)
+						print("scales", scales)
+						raise e
 
 				outputs_rl_actor.append(current_slice_action_probs)
 				decisions = current_slice_action_dists.sample()
@@ -708,8 +716,8 @@ def train_rl():
 							overshoot[seq_index] = chosen_indices[step_index+1][seq_index] - orig_seq_lens[seq_index]
 						new_already_found_packets_per_sample[seq_index] += 1
 						already_skipped_packets_per_sample[seq_index] += torch.min(chosen_indices[step_index+1][seq_index] if step_index+1 < len(chosen_indices) else torch.tensor(float("inf")), orig_seq_lens[seq_index]) - chosen_indices[step_index][seq_index] - 1
-				# rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
-				rewards_sparsity.append((already_skipped_packets_per_sample-overshoot)/(already_skipped_packets_per_sample+already_found_packets_per_sample))
+				rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
[…]

### 2020-02-06 paper_rl/paper.tex
revise paper based on Tanja's comments: text modifications, figures improvements.

@@
 \glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data.
-They allow already determining if a flow is malicious or not before it is over, making it possible to take action immediately.
+They already allow determining if a flow is malicious or not before it is over, making it possible to take action immediately.
 However, considering the large number of packets that have to be inspected, the question of computational as well as energy efficiency arises.
@@
 \label{sec:introduction}
+\note{blue}{Fares: I fixed the paper based on Tanja's comments.
+still  (1) the abstract is not touched yet (2) figures with axes (time step) need to be changed...
+We need to be carefull with the 9 pages limit}
 
@@
 In the past, a viable way to detect intrusions was to analyze the contents of the packets themselves and determine, for example, whether a packet contains potentially harmful content by matching patterns.
-More recently, with the increasing deployment of encryption, the focus is now on features that are always available for network monitoring devices such as port numbers, protocol flags and packet sizes when encrypting above the transport layer.
+More recently, with the increasing deployment of encryption, the focus is now on features available for network monitoring devices even if packets are encrypted, such as packet sizes and temporal behavior.
+In our work we concentrate on features available when encrypting above transport layer (like for example TLS or QUIC), i.e.
+we also include port numbers and protocol flags.
 
@@
 When analyzing flows, not only the aforementioned features are available but also features representing the temporal behavior of the individual packets.
-Various approaches have been proposed to extract flow features with which anomaly detection can be performed: In \cite{meghdouri_analysis_2018} compare such approaches and show their efficacy.
-While these approaches often work well, a major drawback is that the whole flow must be received first and only then anomaly detection can be applied, revealing malicious flows.
-Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether a packet is anomalous based on features that are available even for traffic that is encrypted above the transport layer, like for example TLS or QUIC.
+Various approaches have been proposed to extract flow features with which anomaly detection can be performed: Authors in \cite{meghdouri_analysis_2018} compare such approaches and show their efficacy.
+While these approaches often work well, a major drawback is that the whole flow must be received first and only then anomaly detection can be applied, to reveal malicious flows.
+Thus, we design a network \gls{ids} that operates on a per-packet basis and determines whether a packet is anomalous based on the aforementioned features.
 In this way, an \gls{rnn}-based \gls{ids} has the benefit of avoiding feature engineering procedures and let the classifier build high level features by itself.
@@
 
-For a flow of length $T$ at each timestep $t$ we have two reward metrics which we compute in a fashion that is known from \textit{R-learning} \cite{schwartz_reinforcement_1993}:
+For a flow of length $N$ at each packet $n$ we have two reward metrics which we compute in a fashion that is known from \textit{R-learning} \cite{schwartz_reinforcement_1993}:
 
@@
   \caption{The computation of the rewards for a sample malicious flow: Gray packets (0) are skipped while black ones (1) are chosen.
-``\#Left\//\#Total'' and ``Confidence'' are the underlying parameters for rewards calculations (see \autoref{eq:R_classification} and \autoref{eq:R_sparsity}).
+``\#Skipped\//\#Total'' (number of packets skipped with respect to the total number of packets both starting from the current position) and ``Confidence'' are the underlying parameters for rewards calculations (see \autoref{eq:R_classification} and \autoref{eq:R_sparsity}).
 Numbers are rounded to two digits after the decimal point.}
@@
 \begin{split}
-R_{\text{classification},t} = \frac{1}{T-t} \sum_{i=t+1}^{T} |1 - \left(\text{label}_i - \text{confidence}_i\right)|
+R_{\text{classification},n} = \frac{1}{N-n} \sum_{i=n+1}^{N} |1 - \left(\text{label}_i - \text{confidence}_i\right)|
 \end{split}
@@
 E.g.
-for a flow of length 10 if after timestep 5, 3 packets are chosen and 2 are skipped from what is left, the sparsity reward metric at timestep 5 is $\frac{2}{10-(4+1)}=\frac{2}{5}$.
+for a flow of length 10 if after the 5th packet, 3 packets are chosen and 2 are skipped from what is left, the sparsity reward metric at packet 5 is $\frac{2}{10-(4+1)}=\frac{2}{5}$.
 %\note{blue}{Fares: This equation can be shortened by diving by the total count of packets left.\\
@@
 \begin{split}
-R_{\text{sparsity},t} = \frac{\sum_{i=t+1}^{T} 1_{\text{skipped},i}}{T-(t+1)}
+R_{\text{sparsity},n} = \frac{\sum_{i=n+1}^{N} 1_{\text{skipped},i}}{N-(n+1)}
 \end{split}
@@
 
-The critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a timestep $t$ (a timestep is a packet).
-For this, given the current input vector and the state of the LSTM cells $s_t$ and the neural network weights of the critic $\theta_v$, it outputs $v_{\text{classification},t}$ and $v_{\text{sparsity},t}$ at each timestep $t$.
-It aims to predict the classification performance and sparsity correctly for each timestep by minimizing the following loss function:
+The critic aims to estimate the future expected average classification performance and the future expected average sparsity that can be achieved at a packet $n$.
+For this, given the current input vector and the state of the LSTM cells $s_n$ and the neural network weights of the critic $\theta_v$, it outputs the value functions $v_{\text{classification},n}$ and $v_{\text{sparsity},n}$ at each packet $n$.
+It aims to predict the classification performance and sparsity correctly for each packet by minimizing the following loss function:
 \begin{align}
 \begin{split}
-l_{\text{v},t} =& \left(R_{\text{classification},t} - v_{\text{classification},t}\left(s_t ; \theta_\text{c} \right)\right)^2 \\
-+ & \left(R_{\text{sparsity},t} - v_{\text{sparsity},t}\left(s_t ; \theta_\text{c} \right)\right)^2
+l_{\text{v},n} =& \left(R_{\text{classification},n} - v_{\text{classification},n}\left(s_n ; \theta_\text{c} \right)\right)^2 \\
++ & \left(R_{\text{sparsity},n} - v_{\text{sparsity},n}\left(s_n ; \theta_\text{c} \right)\right)^2
 \end{split}
@@
 
-Now that the reward metrics and the critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each timestep $t$ as follows:
+Now that the reward metrics and the critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each packet $n$ as follows:
 \begin{align}
 \begin{split}
-U_t =& \left(R_{\text{classification},t} + \alpha \cdot R_{\text{sparsity},t}\right) \\
-- & \left(v_{\text{classification},t} + \alpha \cdot v_{\text{sparsity},t}\right)
+U_n =& \left(R_{\text{classification},n} + \alpha \cdot R_{\text{sparsity},n}\right) \\
+- & \left(v_{\text{classification},n} + \alpha \cdot v_{\text{sparsity},n}\right)
 \end{split}
@@
 
+With $\alpha$ being the sparsity-accuracy tradeoff parameter.
 This means that we want the reward to be higher than the critic's expectation.
@@
 
-The actor outputs a probability distribution at timestep $t$, which is sampled to determine the number of packets that should be skipped.
+The actor outputs a probability distribution at packet $n$, which is sampled to determine the number of packets that should be skipped next.
 It aims to change the distribution so that actions which result in a higher reward are chosen more frequently, maximizing the utility function.
@@
 
-Specifically, at packet $t$ of a network flow, given the current state $s_t$ (which consists of the current input as well as the LSTM state) and the neural network weights of the actor $\theta_a$, the actor outputs a probability distribution $\pi \left( a_t \given s_t ; \theta_\text{a} \right)$ from which the action $a_t \ge 1$ is sampled.
-$a_t=1$ means that the next packet is inspected next, while, for example, $a_t=2$ means that one packet is skipped until the next that is chosen, $a_t = 3$ means that two are skipped and so forth.
+Specifically, at packet $n$ of a network flow, given the current state $s_n$ (which consists of the current input as well as the LSTM state) and the neural network weights of the actor $\theta_a$, the actor outputs a probability distribution $\pi \left( a_n \given s_n ; \theta_\text{a} \right)$ from which the action $a_n \ge 1$ is sampled.
+$a_n=1$ means that the next packet is inspected next, while, for example, $a_n=2$ means that one packet is skipped until the next that is chosen, $a_n = 3$ means that two are skipped and so forth.
 Besides wanting to make actions which result in a high reward more likely, the actor also seeks to optimize the entropy $H(\cdot)$ of the probability distribution so that it keeps exploring and doesn't get stuck with a suboptimal policy.
@@
 \begin{split}
-l_{\text{a},t} =& \underbrace{ -\log \left( \pi \left( a_t \given s_t ; \theta_\text{a} \right) \right) U_t}_{\text{Policy Loss}} \\
-&- \beta \underbrace{H\left( \pi\left( s_t; \theta_a \right)\right) }_{\text{Enthropy}}
+l_{\text{a},n} =& \underbrace{ -\log \left( \pi \left( a_n \given s_n ; \theta_\text{a} \right) \right) U_n}_{\text{Policy Loss}} \\
+&- \beta \underbrace{H\left( \pi\left( s_n; \theta_a \right)\right) }_{\text{Entropy}}
 \end{split}\label{eq:actor}
@@
 In related work regarding \gls{rl}-based sampling for sequences, only discrete actions were evaluated so far: The actor would output a categorical probability distribution with a fixed maximum number of bins $k$.
-A drawback of this approach is that only a predefined maximum number of packets can be skipped and that for a very large number of $k$, it can take a long time to converge to an optimum policy as there are $k$ different options to explore at each timestep $t$.
-Besides discrete actions, we thus also experiment with continuous actions, which allow to skip arbitrarily many packets.
+A drawback of this approach is that only a predefined maximum number of packets can be skipped at once (the maximum of the actions space) and that for a very large number of $k$, it can take a long time to converge to an optimum policy as there are $k$ different options to explore at each packet $n$.
+Besides discrete actions, we thus also experiment with continuous actions, which allow to skip arbitrarily many packets with long-tailed distributions.
 
@@
 \end{align}
-To parametrize the normal distribution that underlies the log-normal distribution, we first have to compute the parameters $\mu_{\text{normal}}$ and $\sigma_{\text{normal}}$ from $\mu$ and $\sigma$.
-We can compute them as follows:
+Since the log-normal distribution can not be explicitly parametrized, we parametrize first the normal distribution that underlies the log-normal distribution by deriving the parameters $\mu_{\text{normal}}$ and $\sigma_{\text{normal}}$ from $\mu$ and $\sigma$ as follows:
 
@@
 The modified reward applies only to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet.
-The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow ($t=\textit{last}$):
+The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow ($n=\textit{last}$):
 
@@
 \begin{split}
-R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{T} 1_{\text{skipped},i}\right) - \left(\textit{last}+a_{\textit{last}} - T - 1\right)}{T- (\textit{last}+1)}
+R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{N} 1_{\text{skipped},i}\right) - \left(\textit{last}+a_{\textit{last}} - N - 1\right)}{N- (\textit{last}+1)}
 \end{split}
@@
 
-Specifically, it adds the term $\left(\textit{last}+a_{\textit{last}} - T - 1\right)$ to the numerator which penalizes the actor jumping more than one packet behind the flow.
+Specifically, it adds the term $\left(\textit{last}+a_{\textit{last}} - N - 1\right)$ to the numerator which penalizes the actor jumping more than one packet behind the flow.
 Jumping just right after the flow does not get penalized because if the actor considers the rest of the flow as irrelevant, ignoring the rest of the flow and jumping to the first packet behind the flow is a useful action.
@@
 
-Given the fraction $p$ of packets to be selected in a sequence of $T$ packets, this technique will sample each individual packet with probability of $p$.
-If the sequence is large enough, $n$ packets will be randomly sampled from a uniform distribution whereby $n= p \times T$.
+Given the fraction $p$ of packets to be selected in a sequence of $N$ packets, this technique will sample each individual packet with probability of $p$.
+Theoretically, if the sequence is large enough (infinitely large), $m$ packets will be randomly sampled from a uniform distribution whereby $m \approx \lfloor N \times p \rfloor$.
+Nevertheless, in the case of short flows, $m$ is not constant and varies to some degree.
 In practice, these technique works well if several packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
 
-\subsection{First \emph{n} Sampling}
+\subsection{Relative First \emph{m} Packets}
 As its name suggests, this technique only samples the first packets from each flow and ignores the rest of the flow.
-For a sampling probability $p$ it will take the first $n$ packets from each flow so that $n \approx T * p$.
-For example, for a flow with length 10 and $p=0.2$, it will take the first 2 packets.
+For a sampling rate $p$ it will take the first $m$ packets from each flow so that $m = \lfloor N \times p \rfloor$.
+For example, for a flow with length 10 and $p=0.2$, it will take the first 2 packets, for another flow of length 40 and for the same sampling rate ($p=0.2$), it will sample the first 8 packets.
 For this procedure the length of a flow has to be known beforehand and thus it is not applicable in a streaming scenario.
@@
 
-\subsection{First \emph{n} Equal Sampling}
+\subsection{First \emph{m} Packets}
 This technique is similar to the one above, with the difference that the length of a flow doesn't have to be known beforehand.
-If flows have an average length of 10 packets, and the sampling probability $p=0.2$ it would take up to two packets from each flow.
+If flows have an average length of 10 packets, and the sampling rate $p=0.2$ it would take up to two packets from each flow regardless of its length.
 
 \subsection{Every \emph{i}th Sampling}
-Finally we use a periodic sampling technique that takes a sample each $i= \frac{1}{p}$ packet.
+Finally we use a periodic sampling technique that for a sampling rate $p$, it takes a sample each $i= \lfloor \frac{1}{p} \rfloor$ packet(s).
 Similar to the last technique, the length of the flow is not needed in this case.
@@
 Opposed to the behavior during training, we do not randomly sample from the probability distribution of the actor during testing.
-Instead take the mean of the distribution for the continuous case and the mode for the discrete one.
+We instead take the mean of the distribution for the continuous case and the mode for the discrete one.
 This is reasonable and best practice \cite{mnih_asynchronous_2016} since during testing we do not aim to explore the possible choices anymore as they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore.
@@
 
-The fact that for UNSW-NB15 sparsity is higher is caused by the fact that it has more packets per flow on average and thus more redundant information that the \gls{rl} learns to skip.
+%The fact that for UNSW-NB15 sparsity is higher is caused by the fact that it has more packets per flow on average and thus more redundant information that the \gls{rl} learns to skip.
 %This, on the other hand, holds since almost the same number of packets is sampled from each flow for both datasets (see \autoref{sec:conclusion}).
 % Max: I don't understand the sentence commented out above...
+
+The higher sparsity in UNSW-NB15 is caused by the fact that it has more packets per flow on average and thus \ours{} achieves higher sparsity even by sampling the same number of packets as in CIC-IDS-2017.
 
@@
 \begin{enumerate}
-\item A model is trained for which the trade-off is chosen randomly for each flow during training from a uniform distribution yielding values between 0 and 1.
+\item A model is trained for which the tradeoff is chosen randomly for each flow during training from a uniform distribution yielding values between 0 and 1.
 We add a new feature to each packet that indicates the tradeoff chosen for the respective flow so that it can adjust its sampling strategy in anticipation of the tradeoff that will determine the reward.

### 2020-02-06 paper_rl/paper.tex
change 'testing' to 'deployement' because it confuses the reader.

@@
 
-Opposed to the behavior during training, we do not randomly sample from the probability distribution of the actor during testing.
+Opposed to the behavior during training, we do not randomly sample from the probability distribution of the actor during deployment.
 We instead take the mean of the distribution for the continuous case and the mode for the discrete one.
-This is reasonable and best practice \cite{mnih_asynchronous_2016} since during testing we do not aim to explore the possible choices anymore as they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore.
+This is reasonable and best practice \cite{mnih_asynchronous_2016} since during deployment we do not aim to explore the possible choices anymore as they were already learned during training: Only the optimal choice, which was learned, should be applied and no exploration is necessary since the actor does not train anymore.
 Furthermore, the critic is only necessary during training for the reward of the actor.
@@
 One concern could be the overly deterministic nature of the decisions of \ours{} for some attack types: An adversary could see that only certain packets of a flow are looked at by the \gls{rl} system and thus only manipulate the few packets that the classifier looks at to generate adversarial samples with methods as studied by \cite{hartl_explainability_2019}.
-One potential remedy for this is to also randomly sample from the distribution of the actor during testing like it is done during training.
+One potential remedy for this is to also randomly sample from the distribution of the actor during deployment like it is done during training.
 This could result in slightly worse accuracy but lead to behavior that is harder to predict for an adversary.

### 2020-02-06 paper_rl/paper.tex
Update related work

@@
 
-\cite{hartl_explainability_2019} build an LSTM-based \cite{hochreiter_long_1997} recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
+There is a significant body of research on sampling for networks: \cite{zseby_sampling_2005} give an overview of common methods for packet sampling in IP networks.
+\cite{estan_building_2004} develop a scheme that adapts the sampling rate depending on available resources.
+
+Several approaches have been proposed to implement sampling for \gls{ids}:  \cite{estan_new_2003} propose to single out large flows for \gls{ids} and sample fewer packets for increased computational efficiency.
+\cite{bakhoum_intrusion_2011} use an approach based on Markov chains to sample packets for \gls{ids}.
+\cite{murali_kodialam_detecting_2003} lay out a game-theoretic model for determining which network paths are more vulnerable and require more packet sampling. 
+\cite{ha_suspicious_2016} develop a system for Software Defined Networks, which aims to sample more packets from more vulnerable parts of the network.
+
+Regarding \gls{rnn} for \gls{ids}, \cite{hartl_explainability_2019} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
+
+Some works investigated ways to let \glspl{rnn} skip parts of sequences.
+The focus of these works was mostly Natural Language Processing and we know of no work that focuses on network traffic.
+The following papers provide an overview of this research domain:
 
@@
 
-\cite{campos_skip_2018} develop a technique that includes a \textit{skip gate} into LSTM or GRU \cite{cho_learning_2014} cells.
+\cite{campos_skip_2018} develop a technique that includes a \textit{skip gate} into LSTM and other recurrent cells.
 It works without \gls{rl} but on the downside, their method can only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier, while we want a solution in which the sampling procedure and the classifier are independent.

### 2020-02-06 learn.py

@@ -871,7 +871,7 @@ def train_rl():
 			total_loss.backward()
 			optimizer.step()
 
-			if batches_processed % 100 == 0:
+			if batches_processed % opt.loggingInterval == 0:
 				writer.add_scalar("loss", loss.item(), samples)
 				writer.add_scalar("loss_critic", loss_rl_critic.item(), samples)
 				writer.add_scalar("loss_actor", loss_rl_actor.item(), samples)
@@ -2235,6 +2235,7 @@ if __name__=="__main__":
 	parser.add_argument('--steering_step_size', type=float, default=0.01, help='step size for steering')
 	parser.add_argument('--steering_target_sparsity', type=float, default=0.5, help='target sparsity for steering')
 	parser.add_argument('--batches_to_consider_for_steering', type=int, default=100)
+	parser.add_argument('--loggingInterval', type=int, default=100)
 
 	# parser.add_argument('--nSamples', type=int, default=1, help='number of items to sample for the feature importance metric')
 

### 2020-02-06 paper_rl/paper.tex
Remove UNSW-NB15 for now and bring it under 9 pages

@@
 They already allow determining if a flow is malicious or not before it is over, making it possible to take action immediately.
-However, considering the large number of packets that have to be inspected, the question of computational as well as energy efficiency arises.
+However, considering the large number of packets that have to be inspected, the question of computational efficiency arises.
 We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high.
 Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy with the same number of sampled packets.
-A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can possibly skip arbitrarily many packets, saving even more computational resources for long sequences.
+A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can possibly skip arbitrarily many packets in one step, saving even more computational resources for long sequences.
 Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows.
@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 Our implementation also supports sharing the LSTM between all three neural network.
-However, as the five outputs can produce losses that have different order of magnitude, larger gradients or larger losses would ``overwrite'' the gradients of the smaller ones.}
+However, as the five outputs can have different orders of magnitude, larger gradients would ``overwrite'' the gradients of the smaller ones.}
   \label{fig:neuralNetworkArchitecture}
@@
 
-Let $Z$ be a random variable from a standard normal distribution $\mathcal{N}(0, 1)$ and let $\mu_{\text{normal}}$ and $\sigma_{\text{normal}} > 0$ be two real numbers. Then
-\begin{align}
-X = e^{\mu_{\text{normal}}+\sigma_{\text{normal}} Z}
-\end{align}
-is a log-normally distributed variable.
+%Let $Z$ be a random variable from a standard normal distribution $\mathcal{N}(0, 1)$ and let $\mu_{\text{normal}}$ and $\sigma_{\text{normal}} > 0$ be two real numbers. Then
+%\begin{align}
+%X = e^{\mu_{\text{normal}}+\sigma_{\text{normal}} Z}
+%\end{align}
+%is a log-normally distributed variable.
 Note that the log-normal distribution's mean $\mu$ and its standard deviation $\sigma$ are different from $\mu_\text{normal}$ and $\sigma_\text{normal}$, which are used to parametrize the distribution.
@@
 \text{softplus}(x) = \log\left(1 + e^x \right)
-\end{align}
-Since the log-normal distribution can not be explicitly parametrized, we parametrize first the normal distribution that underlies the log-normal distribution by deriving the parameters $\mu_{\text{normal}}$ and $\sigma_{\text{normal}}$ from $\mu$ and $\sigma$ as follows:
-
-\begin{align}
-\mu_{\text{normal}} = \frac{1}{2}\left( 2\log(\mu) - \log\left(\frac{\mu^2 + \sigma^2}{\mu^2}\right) \right)
-\end{align}
-
-\begin{align}
-\sigma_{\text{normal}} = \sqrt{\log\left(\frac{\mu^2 + \sigma^2}{\mu^2}\right)}
 \end{align}
@@
 \begin{split}
-R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{N} 1_{\text{skipped},i}\right) - \left(\textit{last}+a_{\textit{last}} - N - 1\right)}{N- (\textit{last}+1)}
+R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{N} 1_{\text{skipped},i}\right)}{N- (\textit{last}+1+\left(\textit{last}+a_{\textit{last}} - N - 1\right))}
 \end{split}
@@
 
-Specifically, it adds the term $\left(\textit{last}+a_{\textit{last}} - N - 1\right)$ to the numerator which penalizes the actor jumping more than one packet behind the flow.
+Specifically, it adds the term $\left(\textit{last}+a_{\textit{last}} - N - 1\right)$ to the denominator which penalizes the actor jumping more than one packet behind the flow.
 Jumping just right after the flow does not get penalized because if the actor considers the rest of the flow as irrelevant, ignoring the rest of the flow and jumping to the first packet behind the flow is a useful action.
@@
 \begin{table}[h]
-\caption{Occurrence frequency of the attack flows.}
+\caption{Occurrence frequency of the attack flows of CIC-IDS-2017.}
 \label{tab:occurrence}
 \centering
-\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
-]{
 \begin{tabular}{l r}
@@
 \end{tabular}
-}{}
-\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
-]{
-\begin{tabular}{l r}
-\toprule
-Attack type & \hspace*{-4mm}Proportion \\ \midrule
-Exploits	&	1.42\%	\\
-Fuzzers	&	1.01\%	\\
-Reconnaissance	&	0.58\%	\\
-Generic	&	0.21\%	\\
-DoS	&	0.19\%	\\
-Shellcode	&	0.08\%	\\
-Analysis	&	0.03\%	\\
-Backdoors	&	0.02\%	\\
-Worms	&	0.01\%	\\
-\bottomrule
-\end{tabular}
-}{}
 \end{table}
 
-For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} and \textit{UNSW-NB15} \cite{moustafa_unsw-nb15_2015} datasets, which both include more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
+%\begin{table}[h]
+%\caption{Occurrence frequency of the attack flows.}
+%\label{tab:occurrence}
+%\centering
+%\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
+%]{
+%\begin{tabular}{l r}
+%\toprule
+%Attack type & \hspace*{-4mm}Proportion \\ \midrule
+%DoS Hulk	&	10.10\%	\\
+%PortScan, Firewall	&	6.90\%	\\
+%DDoS LOIT	&	4.08\%	\\
+%Infiltration	&	3.30\%	\\
+%DoS GoldenEye	&	0.32\%	\\
+%DoS SlowHTTPTest	&	0.18\%	\\
+%DoS Slowloris	&	0.17\%	\\
+%Brute-force SSH	&	0.11\%	\\
+%Botnet ARES	&	0.03\%	\\
+%XSS attack	&	0.03\%	\\
+%PortScan, no Fw.	&	0.02\%	\\
+%Brute-force FTP	&	0.01\%	\\
+%SQL injection	&	$<$0.01\%	\\
+%Heartbleed	&	$<$0.01\%	\\
+%\bottomrule
+%\end{tabular}
+%}{}
+%\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
+%]{
+%\begin{tabular}{l r}
+%\toprule
+%Attack type & \hspace*{-4mm}Proportion \\ \midrule
+%Exploits	&	1.42\%	\\
+%Fuzzers	&	1.01\%	\\
+%Reconnaissance	&	0.58\%	\\
+%Generic	&	0.21\%	\\
+%DoS	&	0.19\%	\\
+%Shellcode	&	0.08\%	\\
+%Analysis	&	0.03\%	\\
+%Backdoors	&	0.02\%	\\
+%Worms	&	0.01\%	\\
+%\bottomrule
+%\end{tabular}
+%}{}
+%\end{table}
+
+For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset, which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
 The distribution of different attacks in the datasets are shown in \autoref{tab:occurrence}.
 
-We use Z-score normalization and a train/test split of 2:1 for both datasets.
+We use Z-score normalization and a train/test split of 2:1.
 To speed up experimentation, we cut flows at a maximum length of 20.
@@
 
-\begin{figure}
+\begin{figure}[h]
 \centering
   \includegraphics[width=\columnwidth]{{"img/events.out.tfevents.1579352256.gpu"}.pdf}
-  \caption{Training with a tradeoff of 0.1 with continuous actions for the CIC-IDS-2017 dataset.
-Training accuracy increases sharply in the beginning while becoming slower the longer training lasts.
-Sparsity increases sharply in the beginning and then slower.
+  \caption{Training with a tradeoff of 0.1 with continuous actions.
+Training accuracy increases sharply in the beginning while becoming slower later on.
+Sparsity increases sharply in the beginning and then more slowly.
 As the classifier gets better, the \gls{rl} starts skipping more packets.}
@@
 
+%\begin{table}
+%\caption{Comparing classification metrics per flow with different tradeoffs for both CIC-IDS-2017 and UNSW-NB15.} \label{tab:results_tradeoff}
+%\centering
+%\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+%& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
+%Tradeoff & 0.0 & 0.1 & 0.5 & 0.0 & 0.1 & 0.5 \\	\midrule
+%\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 0\% & 83.5\% & 84.5\% \\ \midrule
+%\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% & 98.9\% & 98.4\% & 98.5\% \\
+%Precision & 99.4\% & 98.5\% & 98.7\% & 81.1\% & 69.5\% & 70.8\% \\
+%Recall & 98.5\% & 99.0\% & 98.6\% & 84.0\% & 89.9\% & 89.6\% \\
+%F1 & 99.0\% & 98.7\% & 98.6\% & 82.5\% & 78.4\% & 79.1\% \\
+%Youden & 98.4\% & 98.5\% & 98.2\% & 83.3\% & 88.6\% & 88.4\% \\
+%\bottomrule
+%\end{tabular*}
+%\end{table}
+
 \begin{table}
-% Evaluated using
-% ./python learn.py --dataroot flows.pickle --net runs/Oct26_00-03-50_gpu/lstm_module_1284.pth --function test --batchSize 512
-% ./python learn.py --dataroot flows15.pickle --net runs/Oct28_15-41-46_gpu/lstm_module_997.pth --function test --batchSize 512
-% on Dec. 4th
-\caption{Comparing classification metrics per flow with different tradeoffs for both CIC-IDS-2017 and UNSW-NB15.} \label{tab:results_tradeoff}
+\caption{Comparing classification metrics per flow with different tradeoffs.} \label{tab:results_tradeoff}
 \centering
-\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
-& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
-Tradeoff & 0.0 & 0.1 & 0.5 & 0.0 & 0.1 & 0.5 \\	\midrule
-\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 0\% & 83.5\% & 84.5\% \\ \midrule
-\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% & 98.9\% & 98.4\% & 98.5\% \\
-Precision & 99.4\% & 98.5\% & 98.7\% & 81.1\% & 69.5\% & 70.8\% \\
-Recall & 98.5\% & 99.0\% & 98.6\% & 84.0\% & 89.9\% & 89.6\% \\
-F1 & 99.0\% & 98.7\% & 98.6\% & 82.5\% & 78.4\% & 79.1\% \\
-Youden & 98.4\% & 98.5\% & 98.2\% & 83.3\% & 88.6\% & 88.4\% \\
+\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+Tradeoff & 0.0 & 0.1 & 0.5 \\	\midrule
+\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% \\ \midrule
+\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% \\
+Precision & 99.4\% & 98.5\% & 98.7\% \\
+Recall & 98.5\% & 99.0\% & 98.6\% \\
+F1 & 99.0\% & 98.7\% & 98.6\% \\
+Youden & 98.4\% & 98.5\% & 98.2\% \\
 \bottomrule
-\end{tabular*}
+\end{tabular}
 \end{table}
@@
 For tradeoffs 0.1 and 0.5 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff constant for the whole training.
-
 Results show that the higher the tradeoff, the less packets are sampled and the lower the accuracy.
-The only exception to this is that for tradeoffs of $0.1$ and $0.5$ for UNSW-NB15, the accuracy is almost the same or even slightly higher for the larger tradeoff, which is counter-intuitive.
-We consider this to be a coincidence that can happen because of the stochasticity of the training process.
-
-%\note{blue}{Fares: my suggestion:
-%The only exception to this is that for tradeoffs of $0.1$ and $0.5$ for UNSW-NB15, the accuracy is almost the same or even slightly higher for the larger tradeoff, which is counter-intuitive.
-We assume that due to the nature of the UNSW-NB15 data set (which is highly heterogeneous), fewer packets yield better information (and therefore the memory of the LSTM classifier is cleaner) and hence better classification decision and higher accuracy.
-%}
-% Max: Could be but it is too speculative in my opinion. From what I saw, neural networks depend quite a lot on the initialization and on luck, so it might just be that it was lucky...
-
-%The fact that for UNSW-NB15 sparsity is higher is caused by the fact that it has more packets per flow on average and thus more redundant information that the \gls{rl} learns to skip.
-%This, on the other hand, holds since almost the same number of packets is sampled from each flow for both datasets (see \autoref{sec:conclusion}).
-% Max: I don't understand the sentence commented out above...
-
-The higher sparsity in UNSW-NB15 is caused by the fact that it has more packets per flow on average and thus \ours{} achieves higher sparsity even by sampling the same number of packets as in CIC-IDS-2017.
-
-%\todo{Write that the sparsity is higher for UNSW-NB15 because it contains longer flows}
 
@@
 
+%\begin{table}
+%\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions for CIC-IDS-2017 and UNSW-NB15.
+The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
+%\centering
+%\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+%& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
+%Actions & cont. & disc. & cont. & disc. \\	\midrule
+%\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% & 83.5\% & 86.6\% \\ \midrule
+%\setrow{\bfseries}Accuracy & 99.4\% & 99.2\% & 98.4\% & 98.4\% \\
+%Precision & 98.5\% & 98.7\% & 69.5\% & 68.0\% \\
+%Recall & 99.0\% & 98.3\% & 89.9\% & 94.6\% \\
+%F1 & 99.7\% & 98.5\% & 78.4\% & 79.2\% \\
+%Youden & 98.5\% & 97.8\% & 88.6\% & 93.2\% \\
+%\bottomrule
+%\end{tabular*}
+%\end{table}
+
 \begin{table}
-\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions for both CIC-IDS-2017 and UNSW-NB15.
-The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
+\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
 \centering
-\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
-& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
-Actions & cont. & disc. & cont. & disc. \\	\midrule
-\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% & 83.5\% & 86.6\% \\ \midrule
-\setrow{\bfseries}Accuracy & 99.4\% & 99.2\% & 98.4\% & 98.4\% \\
-Precision & 98.5\% & 98.7\% & 69.5\% & 68.0\% \\
-Recall & 99.0\% & 98.3\% & 89.9\% & 94.6\% \\
-F1 & 99.7\% & 98.5\% & 78.4\% & 79.2\% \\
-Youden & 98.5\% & 97.8\% & 88.6\% & 93.2\% \\
+\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+Actions & continuous & discrete \\	\midrule
+\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% \\ \midrule
+\setrow{\bfseries}Accuracy & 99.4\% & 99.2\% \\
+Precision & 98.5\% & 98.7\% \\
+Recall & 99.0\% & 98.3\% \\
+F1 & 99.7\% & 98.5\% \\
+Youden & 98.5\% & 97.8\% \\
 \bottomrule
-\end{tabular*}
+\end{tabular}
 \end{table}
 
-\autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions and shows that for CIC-IDS-2017 continuous actions result in a higher accuracy with fewer samples being selected.
-Conversely, for UNSW-NB-15, accuracy is the same for both continuous and discrete actions.
-However, with discrete actions, fewer packets are chosen, which leads to the conclusion that in this case, discrete actions perform better.
-This leads to the conclusion, that at least for our use case and the two datasets that we consider, there is no consistent difference in performance for continuous and discrete actions.
+\autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions and shows that continuous actions result in a higher accuracy with fewer samples being selected.
+This leads to the conclusion, that at least for our use case and for the dataset we consider, there is no consistent difference in performance for continuous and discrete actions.
 Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean of the log-normal distribution, which can lead to worse performance.
@@
 \begin{table}
-\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies for CIC-IDS-2017.} \label{tab:results_others}
+\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies.} \label{tab:results_others}
 \centering
-\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
 & \gls{rl} & random & every $i$th & first $n$ & first $n$ equal \\	\midrule
@@
 \bottomrule
-\end{tabular*}
+\end{tabular}
 \end{table}
@@
 }{}
-\caption{Inspecting the sampling strategy of \ours{} for various attack types.
+\caption{Inspecting the sampling strategy of \ours{} for various attack types with continuous actions and a tradeoff of $0.1$.
 The blue line shows the confidence.
@@
   \caption{Steering the tradeoff with an allowed minimum sparsity of 50\%.
-The tradeoff is initialized as 1 in the beginning and continuously lowered as long as the sparsity is higher than 50\% for CIC-IDS-2017.
+The tradeoff is initialized as 1 and continuously lowered as long as the sparsity is higher than 50\%.
 The procedure stops if the minimum sparsity is reached or the tradeoff reached its minimum of 0, which is the case here.
@@
 We show that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity indeed results in higher sparsity the larger the tradeoff parameter, while potentially classification performance is sacrificed.
-Surprisingly, we saw that \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}$ for CIC-IDS-2017 and $\approx \frac{1}{5}$ for UNSW-NB-15.
+Surprisingly, we saw that \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}$.
 Furthermore, our results showed that \gls{rl}-based sampling can intelligently choose relevant packets for intrusion detection and outperform non-smart sampling techniques.

### 2020-02-06 paper_rl/paper.tex
Changes to the text

@@
 \glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data.
-They already allow determining if a flow is malicious or not before it is over, making it possible to take action immediately.
+They allow determining if a flow is malicious or not before it is over in a streaming fashion, making it possible to take action immediately.
 However, considering the large number of packets that have to be inspected, the question of computational efficiency arises.
 We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high.
-Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy with the same number of sampled packets.
-A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can possibly skip arbitrarily many packets in one step, saving even more computational resources for long sequences.
+Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy by learning to sample only relevant packets.
+A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can even skip arbitrarily many packets in one step, saving even more computational resources for long sequences.
 Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows.
@@
 \label{sec:introduction}
-\note{blue}{Fares: I fixed the paper based on Tanja's comments.
+%\note{blue}{Fares: I fixed the paper based on Tanja's comments.
 still  (1) the abstract is not touched yet (2) figures with axes (time step) need to be changed...
@@
 As stated in \autoref{subsec:classifier}, we choose an LSTM-based supervised classifier for this work, but our framework can work together with any classifier as long as it outputs a confidence for non-attack/attack at each packet and can handle skipped packets.
-For instance it would also be possible to replace the LSTM with an unsupervised autoencoder akin to \cite{mirsky_kitsune_2018}.
+For instance, it would also be possible to replace the LSTM with an unsupervised autoencoder akin to \cite{mirsky_kitsune_2018}.
 
@@
 For tradeoffs 0.1 and 0.5 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff constant for the whole training.
-Results show that the higher the tradeoff, the less packets are sampled and the lower the accuracy.
+Results show that the higher the tradeoff, the fewer packets are sampled and the lower the accuracy.
 
@@
 \begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
-& \gls{rl} & random & every $i$th & first $n$ & first $n$ equal \\	\midrule
+& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
 \setrow{\bfseries}Sampled & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
@@
 Not surprisingly, random sampling performs the worst, as it is likely that a flow is ``unfortunate'' and doesn't get enough packets sampled while another one gets more packets or the packets that carry more useful information.
-The best strategy besides the \gls{rl} approach seems to be ``first $n$ equal'' sampling, which matches the conclusion from \cite{hartl_explainability_2019} that the first couple of packets are the most important ones for determining maliciousness of a flow.
-``first $n$ equal'' performs better than ``first $n$'' because for ``first $n$'', longer flows will have more packets sampled from them.
+The best strategy besides the \gls{rl} approach seems to be ``first $m$'' sampling, which matches the conclusion from \cite{hartl_explainability_2019} that the first couple of packets are the most important ones for determining maliciousness of a flow.
+``first $m$'' performs better than ``relative first $m$'' because for ``relative first $m$'', longer flows will have more packets sampled from them.
 Conversely, very short flows will get few packets.
 For example, with $p=0.2$, a flow of length 20 will get 4 packets while one with 4 packets will get only 1.
-However, with `first $n$ equal'', the short and the long flow will get the same number of packets.
+However, with `first $m$'', the short and the long flow will get the same number of packets.
 Apparently it is better if all flows get the same number of packets than if long flows get more packets.

### 2020-02-06 paper_rl/paper.tex
Add computational complexity

@@
 \item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape.
-\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets can be skipped altogether.
+\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets be lightweight, so that computation efficiency is not lower because of the sampling overhead.
 \end{enumerate}
@@
 We show that a significant number of packets can be skipped while the accuracy does not drop
-considerably. Compared to other common sampling techniques, \ours{} performs better when being trained on the same number of packets.
+considerably. Compared to other common sampling techniques, \ours{} performs better when being trained on the same number of flows.
 
@@
 In this way the given budget of computation power is not exceeded.
+
+To make \ours{} as lightweight as possible, our implementation allows to let the sampling and the classifier to be implemented in the same neural network, meaning that no overhead is introduced.
+Without this option enabled, the computational power required is $2\times$ what would be required without the sampling.
+Thus, for this option, less than $50\%$ of packets must be sampled to gain a computational efficiency advantage.
 
@@
 % Max: 1/4 is already pronounced one fourth
-We show that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity indeed results in higher sparsity the larger the tradeoff parameter, while potentially classification performance is sacrificed.
+We showed that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity indeed results in higher sparsity the larger the tradeoff parameter, while potentially classification performance is sacrificed.
 Surprisingly, we saw that \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}$.

### 2020-02-07 paper_rl/paper.tex
Update results

@@
 \centering
-  \includegraphics[width=\columnwidth]{{"img/events.out.tfevents.1579352256.gpu"}.pdf}
+%  \includegraphics[width=\columnwidth]{{"img/events.out.tfevents.1579352256.gpu"}.pdf}
+  \includegraphics[width=\columnwidth]{{"img/events.out.tfevents.1580988920.gpu"}.pdf}
   \caption{Training with a tradeoff of 0.1 with continuous actions.
@@
 \autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions and shows that continuous actions result in a higher accuracy with fewer samples being selected.
-This leads to the conclusion, that at least for our use case and for the dataset we consider, there is no consistent difference in performance for continuous and discrete actions.
+This leads to the conclusion, that at least for our use case and for the dataset we consider, continuous actions seem to perform better.
 Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean of the log-normal distribution, which can lead to worse performance.
@@
 \centering
-  \includegraphics[width=\columnwidth]{{"img/lstm_module_8.pth_global_tradeoff_1.0_steering_step_size_0.05_steering_target_sparsity_0.5_batches_to_consider_for_steering_500_sampling_rl"}.pdf}
+  \includegraphics[width=\columnwidth]{{"img/Jan27_18-32-15_gpu_lstm_module_8.pth_global_tradeoff_1.0_steering_step_size_0.05_steering_target_sparsity_0.5_batches_to_consider_for_steering_500_sampling_rl"}.pdf}
   \caption{Steering the tradeoff with an allowed minimum sparsity of 50\%.

### 2020-02-07 learn.py

@@ -391,7 +391,7 @@ def custom_collate(seqs, things=(True, True, True)):
 
 				print("opt.global_tradeoff", opt.global_tradeoff, "empirical_sparsity", empirical_sparsity)
 
-				new_file_name = f"{opt.net.split('/')[-1]}_global_tradeoff_{initial_global_tradeoff}_steering_step_size_{opt.steering_step_size}_steering_target_sparsity_{opt.steering_target_sparsity}_batches_to_consider_for_steering_{opt.batches_to_consider_for_steering}_sampling_rl.json"
+				new_file_name = f"{'_'.join(opt.net.split('/')[-2:])}_global_tradeoff_{initial_global_tradeoff}_steering_step_size_{opt.steering_step_size}_steering_target_sparsity_{opt.steering_target_sparsity}_batches_to_consider_for_steering_{opt.batches_to_consider_for_steering}_sampling_rl.json"
 				f = open(new_file_name, "w")
 				f.write(json.dumps([chosen_packets_per_epoch_sampling_rl, all_packets_per_epoch_sampling_rl, tradeoffs_sampling_rl]))
 				f.close()
@@ -850,6 +850,7 @@ def train_rl():
 					raise e
 
 			effective_rewards_sparsity = rewards_sparsity_catted.detach()[mask_rl_first].view(-1)
+			assert type(tradeoff) is float or tradeoff.shape == rewards_sparsity_catted.shape
 			effective_rewards_sparsity_with_tradeoff = (tradeoff*rewards_sparsity_catted.detach())[mask_rl_first].view(-1)
 			effective_rewards_classification = rewards_classification_catted.detach()[mask_rl_first].view(-1)
 			# assert (effective_rewards_sparsity.shape == effective_rewards_classification.shape)

### 2020-02-07 learn.py
Add results for tradeoff 1.0

@@ -667,7 +667,7 @@ def train_rl():
 					locs, scales = compute_normal_distribution_from_log_normal(current_slice_action_probs[:,:,0], current_slice_action_probs[:,:,1])
 
 					if (scales < CLAMPING_CONSTANT).any():
-						print("problem at top!")
+						print("Step", samples, "problem at top!")
 						print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
 						print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
 						print("locs", locs)
@@ -677,7 +677,7 @@ def train_rl():
 					try:
 						current_slice_action_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
 					except ValueError as e:
-						print("crash at top!")
+						print("Step", samples, "crash at top!")
 						print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
 						print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
 						print("locs", locs)
@@ -832,7 +832,7 @@ def train_rl():
 				locs, scales = compute_normal_distribution_from_log_normal(effective_output_rl_actor[:,0], effective_output_rl_actor[:,1])
 
 				if (scales < CLAMPING_CONSTANT).any():
-					print("problem at bottom!")
+					print("Step", samples, "problem at bottom!")
 					print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
 					print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
 					print("locs", locs)
@@ -842,7 +842,7 @@ def train_rl():
 				try:
 					effective_rl_actor_dists = torch.distributions.log_normal.LogNormal(locs, scales, validate_args=True)
 				except ValueError as e:
-					print("crash at bottom!")
+					print("Step", samples, "crash at bottom!")
 					print("current_slice_action_probs[:,:,0]", current_slice_action_probs[:,:,0])
 					print("current_slice_action_probs[:,:,1]", current_slice_action_probs[:,:,1])
 					print("locs", locs)

### 2020-02-07 plot_steering.py
Add some more data

@@ -38,7 +38,7 @@ lines += plt.plot(x, tradeoffs_sampling_rl, color=colors[1])
 plt.legend(lines, ["Sparsity", "Tradeoff"], loc='upper right')
 plt.ylabel('Tradeoff')
 plt.tight_layout()
-plt.savefig("paper_rl/img/" + path[:-5] + '.pdf', bbox_inches = 'tight', pad_inches = 0)
+plt.savefig("paper_rl/img/" + (path.split("/")[-1])[:-5] + '.pdf', bbox_inches = 'tight', pad_inches = 0)
 # plt.show()
 
 plt.close()

### 2020-02-07 learn.py
Remove softplus from critic

@@ -656,7 +656,7 @@ def train_rl():
 				else:
 					(current_slice_output, current_slice_action_probs, current_slice_values), _ = lstm_module(current_collated_slice)
 
-				current_slice_values = torch.nn.functional.softplus(current_slice_values)
+				# current_slice_values = torch.nn.functional.softplus(current_slice_values)
 
 				outputs.append(current_slice_output)
 
@@ -716,8 +716,8 @@ def train_rl():
 							overshoot[seq_index] = chosen_indices[step_index+1][seq_index] - orig_seq_lens[seq_index]
 						new_already_found_packets_per_sample[seq_index] += 1
 						already_skipped_packets_per_sample[seq_index] += torch.min(chosen_indices[step_index+1][seq_index] if step_index+1 < len(chosen_indices) else torch.tensor(float("inf")), orig_seq_lens[seq_index]) - chosen_indices[step_index][seq_index] - 1
-				rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
-				# rewards_sparsity.append((already_skipped_packets_per_sample-overshoot)/(already_skipped_packets_per_sample+already_found_packets_per_sample))
+				# rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
+				rewards_sparsity.append((already_skipped_packets_per_sample-overshoot)/(already_skipped_packets_per_sample+already_found_packets_per_sample))
 				overshoot = torch.zeros((batch_size), dtype=torch.float32).to(device)
 				already_found_packets_per_sample = new_already_found_packets_per_sample
 

### 2020-02-07 paper_rl/paper.tex
Add results for tradeoff 1.0
Tiny change in table
Small text change

@@
 \glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data.
-They allow determining if a flow is malicious or not before it is over in a streaming fashion, making it possible to take action immediately.
+They allow determining if a flow is malicious or already before it is over, making it possible to take action immediately.
 However, considering the large number of packets that have to be inspected, the question of computational efficiency arises.
@@
 \item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape.
-\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets be lightweight, so that computation efficiency is not lower because of the sampling overhead.
+\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets
+\item be \textbf{lightweight}, so that computation efficiency is not lower because of the sampling overhead. 
 \end{enumerate}
@@
 \begin{table}
-\caption{Comparing classification metrics per flow with different tradeoffs.} \label{tab:results_tradeoff}
+\caption{Comparing classification metrics per flow with different tradeoffs. The higher the tradeoff, the higher the sparsity but also the lower the accuracy.} \label{tab:results_tradeoff}
 \centering
-\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
-Tradeoff & 0.0 & 0.1 & 0.5 \\	\midrule
-\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% \\ \midrule
-\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% \\
-Precision & 99.4\% & 98.5\% & 98.7\% \\
-Recall & 98.5\% & 99.0\% & 98.6\% \\
-F1 & 99.0\% & 98.7\% & 98.6\% \\
-Youden & 98.4\% & 98.5\% & 98.2\% \\
+\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+Tradeoff & 0.0 & 0.1 & 0.5 & 1.0 \\	\midrule
+\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 76.8\% \\ \midrule
+\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% & 99.0\% \\
+Precision & 99.4\% & 98.5\% & 98.7\% & 97.9\% \\
+Recall & 98.5\% & 99.0\% & 98.6\% & 98.1\% \\
+F1 & 99.0\% & 98.7\% & 98.6\% & 98.0\% \\
+Youden & 98.4\% & 98.5\% & 98.2\% & 97.4\% \\
 \bottomrule

### 2020-02-07 paper_rl/paper.tex
Tiny fix
work on related work

@@
 \glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data.
-They allow determining if a flow is malicious or already before it is over, making it possible to take action immediately.
+They allow determining if a flow is malicious or not before it is over in a streaming fashion, making it possible to take action immediately.
 However, considering the large number of packets that have to be inspected, the question of computational efficiency arises.
@@
 \item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape.
-\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets
-\item be \textbf{lightweight}, so that computation efficiency is not lower because of the sampling overhead. 
+\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets be lightweight, so that computation efficiency is not lower because of the sampling overhead.
 \end{enumerate}
@@
 
-There is a significant body of research on sampling for networks: \cite{zseby_sampling_2005} give an overview of common methods for packet sampling in IP networks.
+%There is a significant body of research on sampling for communication networks: \cite{zseby_sampling_2005} give an overview of common methods for packet sampling in IP networks.
 \cite{estan_building_2004} develop a scheme that adapts the sampling rate depending on available resources.
 
-Several approaches have been proposed to implement sampling for \gls{ids}:  \cite{estan_new_2003} propose to single out large flows for \gls{ids} and sample fewer packets for increased computational efficiency.
+Sampling of network packets to minimize computational effort has been extensively studied and a significant body of reaserch has developed a variety of methods to cope with the complexity of traffic.
+In \cite{estan_building_2004}, authors have developed a scheme that varies the sampling rate depending on the available resources, making sampling framework more effective than conventional fixed-rate sampling.
+In view of the endless options of sampling techniques, the authors in \cite{zseby_sampling_2005} published an overview of common methods for packet sampling in IP networks.
+
+%Several approaches have been proposed to implement sampling for \gls{ids}:  \cite{estan_new_2003} propose to single out large flows for \gls{ids} and sample fewer packets for increased computational efficiency.
 \cite{bakhoum_intrusion_2011} use an approach based on Markov chains to sample packets for \gls{ids}.
-\cite{murali_kodialam_detecting_2003} lay out a game-theoretic model for determining which network paths are more vulnerable and require more packet sampling. 
-\cite{ha_suspicious_2016} develop a system for Software Defined Networks, which aims to sample more packets from more vulnerable parts of the network.
+%\cite{murali_kodialam_detecting_2003} lay out a game-theoretic model for determining which network paths are more vulnerable and require more packet sampling. 
+%\cite{ha_suspicious_2016} develop a system for Software Defined Networks, which aims to sample more packets from more vulnerable parts of the network.
 
-Regarding \gls{rnn} for \gls{ids}, \cite{hartl_explainability_2019} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
+As far as sampling for \glspl{ids} is concerned, several approaches have been proposed to sample packets retaining high detection performance.
+Authors in \cite{estan_new_2003} propose to single out large flows for \gls{ids} and sample fewer packets for increased computational efficiency.
+In \cite{bakhoum_intrusion_2011} an approach based on Markov chains to sample packets for \gls{ids} was used.
+\cite{murali_kodialam_detecting_2003} lay out a game-theoretic model for determining which network paths are more vulnerable and require more packet sampling and \cite{ha_suspicious_2016} develop a system for Software Defined Networks, which aims to sample more packets from more vulnerable parts of the network.
 
-Some works investigated ways to let \glspl{rnn} skip parts of sequences.
+% Regarding \gls{rnn} for \gls{ids}, \cite{hartl_explainability_2019} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
+
+Conversely, \gls{rnn}-based \glspl{ids} has been shown to perform very well.
+In \cite{hartl_explainability_2019}, authors build a LSTM-based recurrent classifier for Network Intrusion Detection and further develop explainability methods as well as recurrent-classifiers security assessment.
+
+%Some works investigated ways to let \glspl{rnn} skip parts of sequences.
 The focus of these works was mostly Natural Language Processing and we know of no work that focuses on network traffic.
@@
 
-\cite{yu_learning_2017} propose using \gls{rl} to let an LSTM learn to skim text.
+In this context, some works investigated ways to let \glspl{rnn} skip parts of sequences.
+The focus of these works was mostly Natural Language Processing and we know of no work that focuses on network traffic.
+The following papers provide an overview of this research domain:
+
+Authors in \cite{yu_learning_2017} propose the usage of \gls{rl} to make a LSTM network learn to skim text.
 Their technique does not aim to maximize sparsity but only lets the \gls{rl} optimize classification performance for the sequence as a whole.
@@
 
-\cite{campos_skip_2018} develop a technique that includes a \textit{skip gate} into LSTM and other recurrent cells.
+Another technique developed in \cite{campos_skip_2018} includes a \textit{skip gate} into LSTM and other recurrent cells.
 It works without \gls{rl} but on the downside, their method can only be trained once and not be adapted afterwards and also it explicitly depends on the the implementation of the underlying classifier, while we want a solution in which the sampling procedure and the classifier are independent.
 
-\cite{seo_neural_2018} use an LSTM with one full state vector and one reduced one.
-At each step they decide whether to use the reduced or the full state vector and can reduce computational cost by using the reduced one.
-For network traffic we think that it is often not even necessary to have a reduced state vector since for some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all.
+A LSTM network with two state vectors (one full state vector and one reduced one) was used in \cite{seo_neural_2018}.
+At each step the network decides whether to use the reduced or the full state vector and can reduce computational cost by using the reduced one.
+For network traffic it is often not even necessary to have a reduced state vector since for some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all \cite{hartl_explainability_2019}.
 
@@
 However, for a network flow, already after the second packet it might be obvious that a flow is an attack and thus the remaining packets can be completely ignored.
-It is thus beneficial to not have a fixed maximum step size but a continuous one.
-This allows skipping arbitrarily many packets.
+It is therefore beneficial not to have a fixed maximum step size but a continuous one.
+This allows skipping arbitrarily many packets and gives more flexibility  to the sampling framework.
 

### 2020-02-07 paper_rl/paper.tex

@@
 & \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
-\setrow{\bfseries}Sampled & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
+\setrow{\bfseries}Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
 \setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\

### 2020-02-07 parse.py
Change parse.py

@@ -6,8 +6,9 @@ import itertools
 
 import pickle
 import json
+import sys
 
-df = pd.read_csv('packet.csv')
+df = pd.read_csv(sys.argv[1])
 mapping = dict([(b, a) for a, b in enumerate(sorted(set(list(df["Attack"]))))])
 
 categories = ['Botnet', 'Brute Force', 'DoS', 'Infiltration', 'Normal', 'PortScan', 'Web Attack']

### 2020-02-07 paper_rl/paper.tex
Change caption of training figure
reposition tables + note about time steps

@@
   \caption{Training with a tradeoff of 0.1 with continuous actions.
-Training accuracy increases sharply in the beginning while becoming slower later on.
-Sparsity increases sharply in the beginning and then more slowly.
+Training accuracy increases sharply in the beginning while increasing more slowly later on.
+The behavior for sparsity is similar.
 As the classifier gets better, the \gls{rl} starts skipping more packets.}
@@
 
-\begin{table}
+\begin{table*}
+\parbox{.45\linewidth}{
 \caption{Comparing classification metrics per flow with different tradeoffs. The higher the tradeoff, the higher the sparsity but also the lower the accuracy.} \label{tab:results_tradeoff}
 \centering
-\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
 Tradeoff & 0.0 & 0.1 & 0.5 & 1.0 \\	\midrule
@@
 \bottomrule
-\end{tabular}
-\end{table}
+\end{tabular*}
+}
+\hspace{0.8cm}
+\parbox{.45\linewidth}{
+\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies.} \label{tab:results_others}
+\centering
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
+\setrow{\bfseries}Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
+\setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
+Precision & 98.5\% & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
+Recall & 99.0\% & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
+F1 & 99.7\% & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
+Youden & 98.5\% & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
+\bottomrule
+\end{tabular*}
+}
+\end{table*}
 
@@
 
-\begin{table}
-\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies.} \label{tab:results_others}
-\centering
-\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
-& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
-\setrow{\bfseries}Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
-\setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
-Precision & 98.5\% & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
-Recall & 99.0\% & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
-F1 & 99.7\% & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
-Youden & 98.5\% & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
-\bottomrule
-\end{tabular}
-\end{table}
+%\begin{table}
+%\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies.} \label{tab:results_others}
+%\centering
+%\begin{tabular}{>{\rowmac}l >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+%& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
+%\setrow{\bfseries}Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
+%\setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
+%Precision & 98.5\% & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
+%Recall & 99.0\% & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
+%F1 & 99.7\% & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
+%Youden & 98.5\% & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
+%\bottomrule
+%\end{tabular}
+%\end{table}
 
@@
 \end{figure*}
+
+% Fares: if figures are not changed add "The time steps represent the index of the sampled packets." to the caption
 

### 2020-02-08 plot_rl.py
Regenerate figures

@@ -183,7 +183,7 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 
 	if SHOW_TITLE:
 		plt.title(reverse_mapping[attack_type])
-	ax1.set_xlabel('Time step $t$')
+	ax1.set_xlabel('Packet $n$')
 	plt.xlim((-0.5,max(x_values)+0.5))
 	plt.tight_layout()
 

### 2020-02-08 plot_steering.py

@@ -30,7 +30,7 @@ x = list(range(1,len(chosen_packets_per_epoch_sampling_rl)+1))
 plt.figure(figsize=(5,3))
 lines = []
 lines += plt.plot(x, 100*chosen_packets_per_epoch_sampling_rl/all_packets_per_epoch_sampling_rl, color=colors[0])
-plt.xlabel("Time steps")
+plt.xlabel("Time")
 plt.ylabel('Sparsity (%)')
 
 plt.twinx()

### 2020-02-09 paper_rl/paper.tex
chages after comments

@@
 E.g.
-for a flow of length 10 if after the 5th packet, 3 packets are chosen and 2 are skipped from what is left, the sparsity reward metric at packet 5 is $\frac{2}{10-(4+1)}=\frac{2}{5}$.
+for a flow of length 10 if after 5 packets have been processed, 3 packets are chosen and 2 are skipped from the remaining 5 packets, the sparsity reward metric at packet 5 is $\frac{2}{10-(4+1)}=\frac{2}{5}$.
 %\note{blue}{Fares: This equation can be shortened by diving by the total count of packets left.\\
@@
 Theoretically, if the sequence is large enough (infinitely large), $m$ packets will be randomly sampled from a uniform distribution whereby $m \approx \lfloor N \times p \rfloor$.
-Nevertheless, in the case of short flows, $m$ is not constant and varies to some degree.
+Nevertheless, in the case of short flows, $m$ is not constant and varies to some degree (this effect is larger for short flows).
 In practice, these technique works well if several packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
@@
 For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset, which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
-The distribution of different attacks in the datasets are shown in \autoref{tab:occurrence}.
+The distribution of different attacks in the dataset are shown in \autoref{tab:occurrence}.
 
@@
 
-For all our experiments we train our networks for at least 8 epochs ($>10$ million flows for both datasets) with the Adam optimizer with a learning rate of $0.001$.
+For all our experiments we train our networks for at least 8 epochs ($>10$ million flows) with the Adam optimizer with a learning rate of $0.001$.
 An important implementation detail is that we always start with the first packet of a flow, meaning that it cannot be skipped.
@@
 \setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 76.8\% \\ \midrule
-\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% & 99.0\% \\
-Precision & 99.4\% & 98.5\% & 98.7\% & 97.9\% \\
-Recall & 98.5\% & 99.0\% & 98.6\% & 98.1\% \\
-F1 & 99.0\% & 98.7\% & 98.6\% & 98.0\% \\
-Youden & 98.4\% & 98.5\% & 98.2\% & 97.4\% \\
+Accuracy & \textbf{99.5\%} & 99.4\% & 99.3\% & 99.0\% \\
+Precision & \textbf{99.4\%} & 98.5\% & 98.7\% & 97.9\% \\
+Recall & 98.5\% & \textbf{99.0\%} & 98.6\% & 98.1\% \\
+F1 & \textbf{99.0\%} & 98.7\% & 98.6\% & 98.0\% \\
+Youden & 98.4\% & \textbf{98.5\%} & 98.2\% & 97.4\% \\
 \bottomrule
@@
 \setrow{\bfseries}Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
-\setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
-Precision & 98.5\% & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
-Recall & 99.0\% & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
-F1 & 99.7\% & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
-Youden & 98.5\% & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
+Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
+Precision & \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
+Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
+F1 & \textbf{99.7\%} & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
+Youden & \textbf{98.5\%} & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
 \bottomrule
@@
 
-\autoref{tab:results_tradeoff} shows the influence of the tradeoff parameter $\alpha$ for both datasets.
+\autoref{tab:results_tradeoff} shows the influence of the tradeoff parameter $\alpha$.
 For columns with a tradeoff of 0.0, we train the classifier regularly with all packets without using \gls{rl}.
@@
 \setrow{\bfseries}Sparsity & 76.3\% & 73.6\% \\ \midrule
-\setrow{\bfseries}Accuracy & 99.4\% & 99.2\% \\
-Precision & 98.5\% & 98.7\% \\
-Recall & 99.0\% & 98.3\% \\
-F1 & 99.7\% & 98.5\% \\
-Youden & 98.5\% & 97.8\% \\
+Accuracy & \textbf{99.4\%} & 99.2\% \\
+Precision & 98.5\% & \textbf{98.7\%} \\
+Recall & \textbf{99.0\%} & 98.3\% \\
+F1 & \textbf{99.7\%} & 98.5\% \\
+Youden & \textbf{98.5\%} & 97.8\% \\
 \bottomrule
@@
 To do so, the steering system starts decreasing the tradeoff parameter until it reaches 0.
-It might seem surprising that even with a tradeoff of 0, which means that only classification matters and not sparsity not at all, a sparsity of 73.5\% is still achieved.
+It might seem surprising that even with a tradeoff of 0, which means that only classification matters and sparsity has zero influence, a sparsity of 73.5\% is still achieved.
 We attribute this to the fact that \ours{} learns that some flows may contain packets that do not contribute in determining maliciousness and are skipped since they do not affect the accuracy.

### 2020-02-09 learn.py
Add some new results file

@@ -716,8 +716,11 @@ def train_rl():
 							overshoot[seq_index] = chosen_indices[step_index+1][seq_index] - orig_seq_lens[seq_index]
 						new_already_found_packets_per_sample[seq_index] += 1
 						already_skipped_packets_per_sample[seq_index] += torch.min(chosen_indices[step_index+1][seq_index] if step_index+1 < len(chosen_indices) else torch.tensor(float("inf")), orig_seq_lens[seq_index]) - chosen_indices[step_index][seq_index] - 1
-				# rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
-				rewards_sparsity.append((already_skipped_packets_per_sample-overshoot)/(already_skipped_packets_per_sample+already_found_packets_per_sample))
+				# XXX: Max: I think that the first clause makes more sense but it leads to super weird results in conjunction with opt.variableTradeoff
+				if not opt.variableTradeoff:
+					rewards_sparsity.append((already_skipped_packets_per_sample)/(already_skipped_packets_per_sample+already_found_packets_per_sample+overshoot))
+				else:
+					rewards_sparsity.append((already_skipped_packets_per_sample-overshoot)/(already_skipped_packets_per_sample+already_found_packets_per_sample))
 				overshoot = torch.zeros((batch_size), dtype=torch.float32).to(device)
 				already_found_packets_per_sample = new_already_found_packets_per_sample
 

### 2020-02-09 paper_rl/paper.tex
Fix refs and implement nasty hack for table top margin

@@
 %At the same time, an \gls{rnn}-based \gls{ids} has the benefit of avoiding tedious feature engineering procedures, which derive statistical measures from the sequence of packet features, but providing any available information to the classifier.
-%\cite{hartl_explainability_2019} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
+%\cite{hartl_explainability_2020} showed that such a system has similar performance to other flow-based anomaly detection systems but can detect anomalies \textit{before} the flow terminates and analyzed its security with respect to adversaries extensively.
 
@@
 In this way, an \gls{rnn}-based \gls{ids} has the benefit of avoiding feature engineering procedures and let the classifier build high level features by itself.
-\cite{hartl_explainability_2019} show that such an architecture has similar performance to traditional flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends.
+\cite{hartl_explainability_2020} show that such an architecture has similar performance to traditional flow-based anomaly detection systems, but can still detect anomalies \textit{before} the flow ends.
 Furthermore they analyzed its robustness with respect to adversaries extensively.
@@
 
-% Regarding \gls{rnn} for \gls{ids}, \cite{hartl_explainability_2019} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
+% Regarding \gls{rnn} for \gls{ids}, \cite{hartl_explainability_2020} build an LSTM-based recurrent classifier for Network Intrusion Detection and develop methods for explainability and to assess the security of recurrent classifiers.
 
 Conversely, \gls{rnn}-based \glspl{ids} has been shown to perform very well.
-In \cite{hartl_explainability_2019}, authors build a LSTM-based recurrent classifier for Network Intrusion Detection and further develop explainability methods as well as recurrent-classifiers security assessment.
+In \cite{hartl_explainability_2020}, authors build a LSTM-based recurrent classifier for Network Intrusion Detection and further develop explainability methods as well as recurrent-classifiers security assessment.
 
@@
 At each step the network decides whether to use the reduced or the full state vector and can reduce computational cost by using the reduced one.
-For network traffic it is often not even necessary to have a reduced state vector since for some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all \cite{hartl_explainability_2019}.
+For network traffic it is often not even necessary to have a reduced state vector since for some attack flows, already the first couple of packets might contain enough information and then the remaining packets of the flow do not have to be considered anymore at all \cite{hartl_explainability_2020}.
 
-\cite{gui_long_2018} use \gls{rl} but their intention is not to skip samples but instead to make the \gls{rl} learn to choose a previous state that can aid at the current step.
+\cite{gui_long_2019} use \gls{rl} but their intention is not to skip samples but instead to make the \gls{rl} learn to choose a previous state that can aid at the current step.
 Thus it tries to help the \gls{rnn} with memorizing information from the past but does not actually take less data, which is the goal we pursue in this work.
@@
 We implemented a three-layer LSTM-based classifier with 128 neurons at each layer.
-We choose 128 neurons as we don't notice a significant degradation of performance compared to the results of \cite{hartl_explainability_2019} (check Table II, flow accuracy) and as fewer neurons allowed for more agility during the training phase and allowed us to run more experiments.
+We choose 128 neurons as we don't notice a significant degradation of performance compared to the results of \cite{hartl_explainability_2020} (check Table II, flow accuracy) and as fewer neurons allowed for more agility during the training phase and allowed us to run more experiments.
 As the input features we use source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e.
@@
 
-\begin{table*}
-\parbox{.45\linewidth}{
+\begin{table}[h]
+%\parbox{.45\linewidth}{
 \caption{Comparing classification metrics per flow with different tradeoffs. The higher the tradeoff, the higher the sparsity but also the lower the accuracy.} \label{tab:results_tradeoff}
@@
 Tradeoff & 0.0 & 0.1 & 0.5 & 1.0 \\	\midrule
-\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 76.8\% \\ \midrule
+Sparsity & 0\% & 76.3\% & 76.5\% & \textbf{76.8\%} \\ \midrule
 Accuracy & \textbf{99.5\%} & 99.4\% & 99.3\% & 99.0\% \\
@@
 \end{tabular*}
-}
-\hspace{0.8cm}
-\parbox{.45\linewidth}{
-\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies.} \label{tab:results_others}
+%}
+\end{table}
+
+\begin{table}[h]
+% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
+\vspace{0.075in}
+\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff 0.1) and alternative strategies.}
+\label{tab:results_others}
 \centering
@@
 & \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
-\setrow{\bfseries}Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
+Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
 Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
@@
 \end{tabular*}
-}
-\end{table*}
+%}
+\end{table}
 
@@
 
-\begin{table}
+\begin{table}[h]
 \caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
@@
 Actions & continuous & discrete \\	\midrule
-\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% \\ \midrule
+Sparsity & \textbf{76.3\%} & 73.6\% \\ \midrule
 Accuracy & \textbf{99.4\%} & 99.2\% \\
@@
 Not surprisingly, random sampling performs the worst, as it is likely that a flow is ``unfortunate'' and doesn't get enough packets sampled while another one gets more packets or the packets that carry more useful information.
-The best strategy besides the \gls{rl} approach seems to be ``first $m$'' sampling, which matches the conclusion from \cite{hartl_explainability_2019} that the first couple of packets are the most important ones for determining maliciousness of a flow.
+The best strategy besides the \gls{rl} approach seems to be ``first $m$'' sampling, which matches the conclusion from \cite{hartl_explainability_2020} that the first couple of packets are the most important ones for determining maliciousness of a flow.
 ``first $m$'' performs better than ``relative first $m$'' because for ``relative first $m$'', longer flows will have more packets sampled from them.
@@
 
-One concern could be the overly deterministic nature of the decisions of \ours{} for some attack types: An adversary could see that only certain packets of a flow are looked at by the \gls{rl} system and thus only manipulate the few packets that the classifier looks at to generate adversarial samples with methods as studied by \cite{hartl_explainability_2019}.
+One concern could be the overly deterministic nature of the decisions of \ours{} for some attack types: An adversary could see that only certain packets of a flow are looked at by the \gls{rl} system and thus only manipulate the few packets that the classifier looks at to generate adversarial samples with methods as studied by \cite{hartl_explainability_2020}.
 One potential remedy for this is to also randomly sample from the distribution of the actor during deployment like it is done during training.

### 2020-02-09 paper_rl/paper.tex
Add missing figures

@@
 \centering
-  \includegraphics[width=\columnwidth]{{"img/Jan27_18-32-15_gpu_lstm_module_8.pth_global_tradeoff_1.0_steering_step_size_0.05_steering_target_sparsity_0.5_batches_to_consider_for_steering_500_sampling_rl"}.pdf}
-  \caption{Steering the tradeoff with an allowed minimum sparsity of 50\%.
-The tradeoff is initialized as 1 and continuously lowered as long as the sparsity is higher than 50\%.
+  \includegraphics[width=\columnwidth]{{"img/Feb07_17-08-56_gpu_lstm_module_8.pth_global_tradeoff_1.0_steering_step_size_0.1_steering_target_sparsity_0.5_batches_to_consider_for_steering_1000_sampling_rl"}.pdf}
+  \caption{Steering the tradeoff with an allowed minimum sparsity of 50\% over the test dataset.
+The tradeoff is initialized as 1 and continuously lowered in a streaming fashion as long as the sparsity is higher than 50\%.
 The procedure stops if the minimum sparsity is reached or the tradeoff reached its minimum of 0, which is the case here.
-Sparsity falls from around 74.5\% to 73.5\%.}
+Sparsity falls from a high of around 72.5\% to a low of 71\%.}
   \label{fig:steering}

### 2020-02-09 learn.py

@@ -10,6 +10,7 @@ import random
 import time
 from functools import reduce
 from tqdm import tqdm
+import statistics
 
 import collections
 import itertools
@@ -167,6 +168,7 @@ all_packets_per_epoch_sampling_rl = [[]]
 batch_counter_sampling_rl = 0
 tradeoffs_sampling_rl = []
 initial_global_tradeoff = None
+global_all_end_accuracies = [[]]
 
 def custom_collate(seqs, things=(True, True, True)):
 	seqs, labels, categories = zip(*seqs)
@@ -383,17 +385,20 @@ def custom_collate(seqs, things=(True, True, True)):
 			if batch_counter_sampling_rl > 0 and batch_counter_sampling_rl % opt.batches_to_consider_for_steering == 0:
 				chosen_packets_per_epoch_sampling_rl[-1] = sum(chosen_packets_per_epoch_sampling_rl[-1])
 				all_packets_per_epoch_sampling_rl[-1] = sum(all_packets_per_epoch_sampling_rl[-1])
+				# print("global_all_end_accuracies", global_all_end_accuracies)
+				global_all_end_accuracies[-1] = statistics.mean(global_all_end_accuracies[-1])
 				empirical_sparsity = float(chosen_packets_per_epoch_sampling_rl[-1]/all_packets_per_epoch_sampling_rl[-1])
 				chosen_packets_per_epoch_sampling_rl.append([])
 				all_packets_per_epoch_sampling_rl.append([])
+				global_all_end_accuracies.append([])
 
 				tradeoffs_sampling_rl.append(opt.global_tradeoff)
 
-				print("opt.global_tradeoff", opt.global_tradeoff, "empirical_sparsity", empirical_sparsity)
+				print("opt.global_tradeoff", opt.global_tradeoff, "empirical_sparsity", empirical_sparsity, "global_all_end_accuracies", global_all_end_accuracies[-2])
 
 				new_file_name = f"{'_'.join(opt.net.split('/')[-2:])}_global_tradeoff_{initial_global_tradeoff}_steering_step_size_{opt.steering_step_size}_steering_target_sparsity_{opt.steering_target_sparsity}_batches_to_consider_for_steering_{opt.batches_to_consider_for_steering}_sampling_rl.json"
 				f = open(new_file_name, "w")
[…]

### 2020-02-09 plot_rl.py

@@ -146,6 +146,7 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 	# lens = [item.shape[0] for item in values_by_length]
 
 	fig, ax1 = plt.subplots(figsize=(5,3))
+
 	x_values = list(range(min(len(lens), MAX_X)))
 	ret = ax1.bar(x_values, lens[:MAX_X], width=1, color="gray", alpha=0.2, label="fraction of samples")
 	# ret4 = ax1.bar(x_values, fraction_of_chosen_packets_by_flows[:MAX_X], width=1, color="red", alpha=0.2, label="fraction of chosen samples")

### 2020-02-09 plot_steering.py

@@ -12,7 +12,8 @@ colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
 
 path = sys.argv[1]
 f = open(path, "r")
-chosen_packets_per_epoch_sampling_rl, all_packets_per_epoch_sampling_rl, tradeoffs_sampling_rl = json.loads(f.read())
+chosen_packets_per_epoch_sampling_rl, all_packets_per_epoch_sampling_rl, tradeoffs_sampling_rl, end_accuracies = json.loads(f.read())
+# chosen_packets_per_epoch_sampling_rl, all_packets_per_epoch_sampling_rl, tradeoffs_sampling_rl = json.loads(f.read())
 f.close()
 
 chosen_packets_per_epoch_sampling_rl = np.array(chosen_packets_per_epoch_sampling_rl[:-1])
@@ -23,20 +24,27 @@ assert len(chosen_packets_per_epoch_sampling_rl) == len(all_packets_per_epoch_sa
 
 # print("Epochs", len(chosen_packets_per_epoch_sampling_rl))
 
-x = list(range(1,len(chosen_packets_per_epoch_sampling_rl)+1))
+# x = list(range(1,len(chosen_packets_per_epoch_sampling_rl)+1))
+maximum = 100
+x = np.linspace(maximum/len(chosen_packets_per_epoch_sampling_rl), maximum, num=len(chosen_packets_per_epoch_sampling_rl))
 
 # print("x", x, "chosen_packets_per_epoch_sampling_rl", chosen_packets_per_epoch_sampling_rl, "all_packets_per_epoch_sampling_rl", all_packets_per_epoch_sampling_rl, "tradeoffs_sampling_rl", tradeoffs_sampling_rl)
 
-plt.figure(figsize=(5,3))
+fig, ax1 = plt.subplots(figsize=(5,3))
+# plt.figure(figsize=(5,3))
 lines = []
-lines += plt.plot(x, 100*chosen_packets_per_epoch_sampling_rl/all_packets_per_epoch_sampling_rl, color=colors[0])
-plt.xlabel("Time")
-plt.ylabel('Sparsity (%)')
+lines += ax1.plot(x, 100*chosen_packets_per_epoch_sampling_rl/all_packets_per_epoch_sampling_rl, color=colors[0])
+ax1.set_xlabel("Fraction of the test dataset (%)")
[…]

### 2020-02-10 paper_rl/paper.tex
Finalize

@@
 \glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data.
-They allow determining if a flow is malicious or already before it is over, making it possible to take action immediately.
+They allow determining if a flow is malicious or not already before it is over, making it possible to take action immediately.
 However, considering the large number of packets that have to be inspected, the question of computational efficiency arises.
@@
 Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy by learning to sample only relevant packets.
-A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can even skip arbitrarily many packets in one step, saving even more computational resources for long sequences.
+A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can even skip arbitrarily many packets in one step.
+This enables saving even more computational resources for long sequences.
 Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows.
@@
 \item be \textbf{retrainable} so that the sampling can be continuously adapted depending on the current threat landscape.
-\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not and the remaining packets
+\item be able to \textbf{skip arbitrarily} many packets since network flows can be very long but only the first couple of packets are needed to decide whether an attack occurred or not. 
 \item be \textbf{lightweight}, so that computation efficiency is not lower because of the sampling overhead. 
@@
 % Fares:
-For this purpose we develop an \gls{rl}-based \gls{ids}: \textit{\ours{}} that fulfills the above properties.
+For this purpose we develop the \gls{rl}-based \gls{ids} \textit{\ours{}} that fulfills the above properties.
 We show that a significant number of packets can be skipped while the accuracy does not drop
@@
 
-%As \ours{} uses a parameter to trade-off accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
+%As \ours{} uses a parameter to tradeoff accuracy for sparsity, the resulting accuracy and sparsity are determined by this parameter.
 However, since it also might be desired to specify the sparsity itself, we propose a steering system which takes a desired sparsity as its input and modifies the parameter of the reinforcement learning in a control loop until the specified sparsity is reached.
@@
 % Fares:
-The accuracy-sparsity trade-off of \ours{} can be regulated by a reinforcement learning parameter that directly affects both quantities.
-However, as in some deployments, a certain sparsity has to be minimally achieved to limit the computational expense, we also develop a steering system that adapts the trade-off parameter in a closed control loop.
+The accuracy-sparsity tradeoff of \ours{} can be regulated by a reinforcement learning parameter that directly affects both quantities.
+However, as in some deployments, a certain sparsity has to be minimally achieved to limit the computational expense, we also develop a steering system that adapts the tradeoff parameter in a closed control loop.
 In this way the given budget of computation power is not exceeded.
@@
 
-Sampling of network packets to minimize computational effort has been extensively studied and a significant body of reaserch has developed a variety of methods to cope with the complexity of traffic.
-In \cite{estan_building_2004}, authors have developed a scheme that varies the sampling rate depending on the available resources, making sampling framework more effective than conventional fixed-rate sampling.
-In view of the endless options of sampling techniques, the authors in \cite{zseby_sampling_2005} published an overview of common methods for packet sampling in IP networks.
+Sampling of network packets to minimize computational effort has been extensively studied.
+%and a significant body of reaserch has developed a variety of methods to cope with the complexity of traffic.
+In \cite{estan_building_2004}, authors have developed a scheme that varies the sampling rate depending on the available resources, making the sampling framework more effective than conventional fixed-rate sampling.
+In view of the numerous sampling techniques that exist, the authors in \cite{zseby_sampling_2005} published an overview of common methods for packet sampling in IP networks.
 
@@
 
-Conversely, \gls{rnn}-based \glspl{ids} has been shown to perform very well.
-In \cite{hartl_explainability_2020}, authors build a LSTM-based recurrent classifier for Network Intrusion Detection and further develop explainability methods as well as recurrent-classifiers security assessment.
+Considering recurrent approaches for network traffic, \gls{rnn}-based \glspl{ids} have been shown to perform very well.
+In \cite{hartl_explainability_2020}, authors build an LSTM-based recurrent classifier for Network Intrusion Detection and further develop explainability methods as well as methods for security assessment of recurrent classifiers.
 
@@
 
-In this context, some works investigated ways to let \glspl{rnn} skip parts of sequences.
+Regarding skipping parts of sequences for \glspl{rnn}, some works investigated ways to let \glspl{rnn} skip parts of sequences.
 The focus of these works was mostly Natural Language Processing and we know of no work that focuses on network traffic.
@@
 
-Authors in \cite{yu_learning_2017} propose the usage of \gls{rl} to make a LSTM network learn to skim text.
+Authors in \cite{yu_learning_2017} propose the usage of \gls{rl} to make an LSTM network learn to skim text.
 Their technique does not aim to maximize sparsity but only lets the \gls{rl} optimize classification performance for the sequence as a whole.
@@
 
-A LSTM network with two state vectors (one full state vector and one reduced one) was used in \cite{seo_neural_2018}.
+An LSTM network with two state vectors (one full state vector and one reduced one) was used in \cite{seo_neural_2018}.
 At each step the network decides whether to use the reduced or the full state vector and can reduce computational cost by using the reduced one.
@@
 Given enough training instances, \gls{rl} algorithms have been shown to perform very well in complex decision-making situations such as playing video games \cite{mnih_playing_2013} as well as networking problems such as congestion control \cite{bachl_rax_2019}.
+% and video streaming \cite{mao_neural_2017}.
 We thus consider an \gls{rl}-based architecture to be suitable for taking packet sampling decisions.
@@
 
-Intuitively, the classification reward makes sure, that packets are chosen that contain the maximum accuracy and thus result in the highest classification performance.
+Intuitively, the classification reward makes sure that packets are chosen that contain the most information and thus result in the highest classification performance.
 
@@
 A drawback of this approach is that only a predefined maximum number of packets can be skipped at once (the maximum of the actions space) and that for a very large number of $k$, it can take a long time to converge to an optimum policy as there are $k$ different options to explore at each packet $n$.
-Besides discrete actions, we thus also experiment with continuous actions, which allow to skip arbitrarily many packets with long-tailed distributions.
+Besides discrete actions, we thus also experiment with continuous actions, which allow to skip arbitrarily many packets.
+% with long-tailed distributions.
 
@@
 
-For performing continious actions we need a probability distribution which can be parametrized to have the mean between 0 and positive infinity.
+For performing continuous actions we need a probability distribution which can be parametrized to have the mean between 0 and positive infinity.
 Furthermore it must have at least two parameters so that the standard deviation, which is required for the entropy (\autoref{eq:actor}), can be changed independently of the mean.
 For example, the Exponential distribution can be parametrized to have its mean anywhere between 0 and positive infinity, however, it has only one parameter and thus the mean and the standard deviation cannot be independently changed.
-Conversely, a simple distribution which fits our requirements is the log-normal distribution, which is defined as a normal distribution whose output is exponentiated:
+Conversely, a simple distribution which fits our requirements is the log-normal distribution, which is defined as a normal distribution whose output is exponentiated.
 
@@
 \end{align}
+
+When sampling from the log-normal distribution, a real number is obtained.
+We round it down and then add 1.
+This gives us a natural number between 1 and $\infty$, which states how many packets should be skipped.
 
@@
 Given the fraction $p$ of packets to be selected in a sequence of $N$ packets, this technique will sample each individual packet with probability of $p$.
-Theoretically, if the sequence is large enough (infinitely large), $m$ packets will be randomly sampled from a uniform distribution whereby $m \approx \lfloor N \times p \rfloor$.
-Nevertheless, in the case of short flows, $m$ is not constant and varies to some degree (this effect is larger for short flows).
-In practice, these technique works well if several packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
+Theoretically, if the sequence is large enough (infinitely large), $m$ packets will be randomly sampled from a uniform distribution whereby $m \approx N \times p$.
+Naturally, the number of chosen packets can vary to some degree.
+This variation is larger for short flows.
+In practice, random sampling works well if several packets have similar properties (redundancy) and therefore the probability of skipping important information is minimal.
 
@@
 As its name suggests, this technique only samples the first packets from each flow and ignores the rest of the flow.
-For a sampling rate $p$ it will take the first $m$ packets from each flow so that $m = \lfloor N \times p \rfloor$.
+For a sampling rate $p$ it will take the first $m$ packets from each flow so that $m \approx N \times p$.
 For example, for a flow with length 10 and $p=0.2$, it will take the first 2 packets, for another flow of length 40 and for the same sampling rate ($p=0.2$), it will sample the first 8 packets.
@@
 If flows have an average length of 10 packets, and the sampling rate $p=0.2$ it would take up to two packets from each flow regardless of its length.
+To determine the average length of flows, it is necessary to use experience from past flows.
 
 \subsection{Every \emph{i}th Sampling}
-Finally we use a periodic sampling technique that for a sampling rate $p$, it takes a sample each $i= \lfloor \frac{1}{p} \rfloor$ packet(s).
+Finally, we use a periodic sampling technique that, for a sampling rate $p$, takes a sample each $i \approx \frac{1}{p} $ packets.
 Similar to the last technique, the length of the flow is not needed in this case.
@@
 Overall, sampling based on \gls{rl} reveals that a high number of packets can be skipped while maintaining good accuracy.
-Then, we inspect how the \gls{rl} chooses packets and if some patterns can be spotted.
+Then, we inspect how the \gls{rl} chooses packets and show that certain patterns can be spotted.
 
-For all our experiments we train our networks for at least 8 epochs ($>10$ million flows) with the Adam optimizer with a learning rate of $0.001$.
+For all our experiments we train our neural networks for 8 epochs ($>10$ million flows) with the Adam optimizer with a learning rate of $0.001$.
 An important implementation detail is that we always start with the first packet of a flow, meaning that it cannot be skipped.
@@
 \begin{table}[h]
+% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
+\vspace{0.075in}
 %\parbox{.45\linewidth}{
@@
 \begin{table}[h]
-% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
-\vspace{0.075in}
 \caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff 0.1) and alternative strategies.}
@@
 For columns with a tradeoff of 0.0, we train the classifier regularly with all packets without using \gls{rl}.
-For tradeoffs 0.1 and 0.5 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff constant for the whole training.
+For tradeoffs 0.1, 0.5 and 1.0 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff constant for the whole training.
 Results show that the higher the tradeoff, the fewer packets are sampled and the lower the accuracy.
@@
 \begin{table}[h]
+% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
+\vspace{0.075in}
 \caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
@@
 \autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions and shows that continuous actions result in a higher accuracy with fewer samples being selected.
-This leads to the conclusion, that at least for our use case and for the dataset we consider, continuous actions seem to perform better.
+This leads to the conclusion that at least for our use case and for the dataset we consider, continuous actions seem to perform better.
 Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean of the log-normal distribution, which can lead to worse performance.
@@
 
-Next we compared the \gls{rl} technique used by \ours{} with several intuitive sampling strategies suggested in \autoref{sec:other_sampling_techniques}.
+Next we compared the \gls{rl} technique used by \ours{} with several common sampling strategies suggested in \autoref{sec:other_sampling_techniques}.
 \autoref{tab:results_others} reveals that \ours{} achieves the highest accuracy with the same sparsity.
@@
 ]{
-\includegraphics[width=0.98\columnwidth]{{"../plots/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_10_Normal"}.pdf}
+\includegraphics[width=0.98\columnwidth]{{"img/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_10_Normal"}.pdf}
 }{}
@@
 ]{
-\includegraphics[width=0.98\columnwidth]{{"../plots/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_2_Brute Force-SSH-Patator"}.pdf}
+\includegraphics[width=0.98\columnwidth]{{"img/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_2_Brute_Force-SSH-Patator"}.pdf}
 }{}
@@
 ]{
-\includegraphics[width=0.98\columnwidth]{{"../plots/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_7_DoS - DDoS-DoS slowloris"}.pdf}
+\includegraphics[width=0.98\columnwidth]{{"img/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_7_DoS_-_DDoS-DoS_slowloris"}.pdf}
 }{}
@@
 ]{
-\includegraphics[width=0.98\columnwidth]{{"../plots/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_15_All samples"}.pdf}
+\includegraphics[width=0.98\columnwidth]{{"img/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_15_All_samples"}.pdf}
 }{}
@@
 For example, the gray bar in Figure \ref{fig:inspectingNormal} at position 19 with height 10000 signifies that 10000 flows of benign traffic had a length of at least 20 packets (since we say that the first packet is at position 0).
-The red bars represent the count of packets sampled at this position for all flows.
+The red bars represent the number of packets sampled at this position for all flows.
 The first packet is always chosen for each flow by design.}
@@
 
-One concern could be the overly deterministic nature of the decisions of \ours{} for some attack types: An adversary could see that only certain packets of a flow are looked at by the \gls{rl} system and thus only manipulate the few packets that the classifier looks at to generate adversarial samples with methods as studied by \cite{hartl_explainability_2020}.
+One concern could be the overly deterministic nature of the decisions of \ours{} for some attack types: An adversary could see that only certain packets of a flow are looked at by the \gls{rl} system and thus only manipulate the few packets that the classifier looks at to generate adversarial samples with methods like the ones studied by \cite{hartl_explainability_2020}.
 One potential remedy for this is to also randomly sample from the distribution of the actor during deployment like it is done during training.
@@
 \item A model is trained for which the tradeoff is chosen randomly for each flow during training from a uniform distribution yielding values between 0 and 1.
-We add a new feature to each packet that indicates the tradeoff chosen for the respective flow so that it can adjust its sampling strategy in anticipation of the tradeoff that will determine the reward.
+We add a new feature to each packet that indicates the tradeoff chosen for the respective flow so that the \gls{rl} can adjust its sampling strategy in anticipation of the tradeoff that will determine the reward.
 \item During deployment, the tradeoff is set to the maximum that was used during training (1 in our case), which results in maximum sparsity.
@@
 \autoref{fig:steering} shows the steering in action.
-We start with a maximum sparsity of ~74.4\% and try to reach 50\%.
+We start with a maximum sparsity of approx.~72.5\% and try to reach 50\%.
 To do so, the steering system starts decreasing the tradeoff parameter until it reaches 0.
 It might seem surprising that even with a tradeoff of 0, which means that only classification matters and sparsity has zero influence, a sparsity of 73.5\% is still achieved.
-We attribute this to the fact that \ours{} learns that some flows may contain packets that do not contribute in determining maliciousness and are skipped since they do not affect the accuracy.
+We attribute this to the fact that \ours{} learns that some flows may contain packets that do not contribute to determining maliciousness at all and are skipped since they do not affect the accuracy.
 %\item the actor still in any case attempts to maximize the entropy of its distribution which means that when no packets are skipped, entropy is very low.
@@
 Surprisingly, we saw that \ours{} on average chooses very few packets per flow: Typically $\approx \frac{1}{4}$.
-Furthermore, our results showed that \gls{rl}-based sampling can intelligently choose relevant packets for intrusion detection and outperform non-smart sampling techniques.
+Furthermore, it outperforms non-smart sampling techniques.
 
@@
 This was solved by adding a penalty for jumping behind the end of the sequence which proved effective.
-Comparing with discrete actions, continuous actions seem to perform on par.
-This is still an advantage, as for continuous actions the neural network only needs to have two outputs (mean and standard deviation) while for discrete ones it needs as many outputs as there are actions.
+Comparing with discrete actions, continuous actions seem to perform slightly better given the same training time.
+Additionally it is also an advantage that for continuous actions the neural network only needs to have two outputs (mean and standard deviation) while for discrete ones it needs as many outputs as there are actions.
 The continuous action framework is thus more lightweight and also can learn to skip arbitrarily many packets.

### 2020-02-10 plot_rl.py

@@ -11,7 +11,7 @@ import pickle
 def numpy_sigmoid(x):
 	return 1/(1+np.exp(-x))
 
-DIR_NAME = "plots/plot_rl"
+DIR_NAME = "paper_rl/img/plot_rl"
 ORDERING = ['', 'Length min', 'IAT min', 'Length max', 'IAT max']
 
 MAX_X = 20
@@ -189,7 +189,7 @@ for attack_type, (seqs, lens) in enumerate(zip(results_by_attack_number, orig_se
 	plt.tight_layout()
 
 	os.makedirs(DIR_NAME, exist_ok=True)
-	plt.savefig(DIR_NAME+'/{}_{}_{}.pdf'.format(file_name.split("/")[-1], attack_type, reverse_mapping[attack_type].replace("/", "-").replace(":", "-")), bbox_inches = 'tight', pad_inches = 0)
+	plt.savefig(DIR_NAME+'/{}_{}_{}.pdf'.format(file_name.split("/")[-1], attack_type, reverse_mapping[attack_type].replace("/", "-").replace(":", "-").replace(" ", "_").replace("(", "_").replace(")", "_")), bbox_inches = 'tight', pad_inches = 0)
 	plt.clf()
 
 

### 2020-02-11 paper_rl/paper.tex
Fix pluralization

@@
 
-In general, we think that \gls{rl}-based packet sampling is not only applicable to \gls{ids} but we are confident that other domains of networking can benefit from similar solutions to save computational expenses.
+In general, we think that \gls{rl}-based packet sampling is not only applicable to \glspl{ids} but we are confident that other domains of networking can benefit from similar solutions to save computational expenses.
 

### 2020-02-13 paper/paper.tex
Added contributions list

@@
 For example, \cite{bachl_walling_2019} investigates remedies for poisoning attacks on \glspl{ids} and \cite{hashemi_towards_2019} investigate adversarial robustness of common network \glspl{ids}.
-In this research, we investigate whether adversarial samples can be found for our \gls{rnn}-based model, i.e.
-minimally different attack flows which are classified as benign.
+In this research, we investigate whether adversarial samples, i.e.
+minimally different attack flows which are classified as benign, can be found for our \gls{rnn}-based model.
 This is not a trivially answerable question since adversarial samples have mostly been analyzed in the context of computer vision.
@@
 Thus, we argue that traditional performance metrics like, e.g., accuracy are not sufficient for evaluating security-sensitive \gls{ml} systems.
-We propose the \acrfull{ars} as a new performance score, which captures the notion of how easily an adversary can generate adversarial samples.
+Despite the well-known threat of adversarial samples, we find that related literature lacks metrics for adversarial robustness and, in particular, no measure has been proposed for quantifying the robustness of \gls{ml}-based \glspl{ids}.
 
-Due to the threat of adversarial attacks, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans.
+Due to the threat of adversarial attacks, but also as a basic requirement for social acceptance of an \gls{ml}-based system, a further crucial requirement for \gls{ml}-based \glspl{ids} is that the classifier's decisions are interpretable by humans.
 This recently stirred up increased interest in explainability methods.
 Also in this case, common methods are designed to work with images or tabular data, but not with sequences.
-Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data.
-%These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
+
+In this paper, we attempt to provide a comprehensive discussion of these security-related topics in the context of \glspl{rnn}. Our main contributions are:
+\begin{itemize}[leftmargin=0.5cm]
+\item We analyze several methods for generating adversarial samples and show that adversarial samples can be generated efficiently for an \gls{rnn}-based classifier.
+\item Based on common robustness notions of related works, we propose the \acrfull{ars} as a new performance score for \glspl{ids}, which captures the notion of how easily an adversary can generate adversarial samples.
+We demonstrate that the \gls{ars} can be computed efficiently.
+\item We review methods for evaluating which features have a significant impact on the classifier's prediction, both picking up methods that have been proposed in the literature, extending them for \glspl{rnn}, and devising new methods.
 Astonishingly, feature importance methods reveal that
@@
 
+\item Going further, we investigate which packets have a significant contribution to the classifier's decision and which values of these features lead to classification as attack.
+Hence, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data.
+%These show the varying characteristics of different attacks and they aid with visualizing potentially vulnerable features for adversaries.
 %Since common explainability methods are designed to work with images or tabular data but not sequences, we extend existing explainability methods such as \glspl{pdp} \cite{friedman_greedy_2001} for sequential data.
@@
 
-Based on the insights gained by the feature importance and explainability methods, we propose two defense methods.
+\item
+Based on the insights gained by the feature importance and explainability methods, we finally propose two defense methods.
 %develop a simple method for defending against adversaries who can manipulate arbitrary packets by simply leaving out potentially manipulable information from the classification process.
@@
 we can reduce the attack surface and harden the resulting \gls{ids} while retaining all features and similar classification performance. % like for the original \gls{ml} model.
+\end{itemize}
 
@@
 But even these features cannot be manipulated due to problem-specific constraints:
-\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
+\begin{itemize}[leftmargin=0.5cm] %s[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
 \item Only packets can be manipulated which are transmitted by the attacker, except for botnet and backdoor traffic, which is entirely controlled by an attacker and thus only packets travelling in one direction can be manipulated.
@@
 %AH. changed notation and wording
-for classification as attack and a feature value $x$. Accuracy then is $\int f(A,x) dx$ for attacks and $1-\int f(A,x) dx$ for benign traffic.
+for classification as attack and a feature value $x$. Accuracy then is $\int_{\mathbb R} f(A,x) dx$ for attacks and $1-\int_{\mathbb R} f(A,x) dx$ for benign traffic.
 \autoref{fig:mutinfo_example} shows two different distributions
@@
 While at the first couple of packets the confidence is not very high, towards the end of the flow it reaches values close to 1 and stays there.
-Hence, not only is the classifier able to make a reasonable classification after just a few packets, the figure also suggests that indeed later packets have a neglectable influence on the prediction.
+Hence, not only is the classifier able to make a reasonable classification after just a few packets, the figure also suggests that indeed later packets have a negligible influence on the prediction.
 
@@
 We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified.
-Newly developed \textit{Feature sensitivity} methods can show which features can easily be manipulated by an adversary.
+Newly developed \textit{feature sensitivity} methods can show which features can easily be manipulated by an adversary.
 We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.

### 2020-02-14 paper/paper.tex
Readded PDP and feature dropout table

@@
 
-We implemented a three-layer LSTM-based classifier with 512 neurons at each layer. As the input features we use
+We implemented a three-layer LSTM-based classifier with 512 neurons at each layer.
+For a large-enough network, we do not expect these architectural parameters to have a severe impact on classification accuracy, so we chose these parameters to obtain a good performance while keeping training duration at an acceptable level.
+
+As the input features we use
 source port, destination port, protocol identifier, packet length, \gls{iat} to the previous packet in the flow, packet direction (i.e.
@@
 We omitted \gls{ttl} values, as they are likely to lead to unwanted prediction behavior \cite{bachl_walling_2019}.
-Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the others vary.
+Among the used features, source port, destination port and protocol identifier are constant over the whole flow while the other features vary.
 During flow extraction we used the usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address.
-We use Z-score normalization and a train/test split of 2:1.
+We use Z-score normalization to transform feature values to the range appropriate for neural network training.
+We ensured that our classifiers do not suffer from overfitting using a train/test split of 2:1.
 
@@
 Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
-\autoref{tab:performance_results} shows that our \gls{rnn}-based classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}.
+\autoref{tab:performance_results} shows the achieved classification performance when evalutuating metrics per packet and per flow and includes performance results for an MLP classifier from~\cite{bachl_walling_2019} for comparison.
+% \autoref{tab:performance_results} shows that
+As depicted, our \gls{rnn}-based classifiers achieve an accuracy that is similar to previous work based on these datasets~\cite{meghdouri_analysis_2018,bachl_walling_2019}.
 However, unlike these classifiers, our recurrent classifier has the advantage of being able to detect attacks already before the attack flows terminate.
@@
 F1	&	97.4\%	&	99.4\%	&	99.5\%	&	85.5\%	&	75.5\%	&	83.7\%	\\
-Youden	&	97.2\%	&	99.0\%	&	99.1\%	&	87.3\%	&	71.9\%	&	82.3\%	\\
+Youden's J	&	97.2\%	&	99.0\%	&	99.1\%	&	87.3\%	&	71.9\%	&	82.3\%	\\
 \bottomrule
 \end{tabular}
-%\vspace*{-0.3cm}
 \end{table}
@@
 
+\begin{figure}[b]
+\includegraphics[width=\columnwidth]{./plots/adv_comparison/tradeoff.pdf}
+\caption{Performance of \gls{cw} with different $\kappa$ values.}
+\label{fig:adv_cw}
+\end{figure}
+
 %Note, however, that f
@@
 % Hence, the depicted distance increase presumably is due to samples for which no adversarial sample could be found for lower $\kappa$ values.
-
-\begin{figure}[t]
-\includegraphics[width=\columnwidth]{./plots/adv_comparison/tradeoff.pdf}
-\caption{Performance of \gls{cw} with different $\kappa$ values.}
-\label{fig:adv_cw}
-\end{figure}
 
@@
 \subsection{Feature Importance Metrics}
-%\begin{table}
-%\caption{Accuracy drop for:}
-%\label{tab:feat_importance}
-%\centering
-%\subfloat[Random replacement method\label{fig:featRandom}
-%]{
-%\begin{tabular}{l r}
-%\toprule
-%Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
-%Protocol & 0.207 \\
-%Packet Length & 0.165 \\
-%SYN Flag & 0.099 \\
-%ACK Flag & 0.084 \\
-%Direction & 0.073 \\
-%Destination port & 0.071 \\
-%Source port & 0.060 \\
-%RST Flag & 0.057 \\
-%PSH Flag & 0.056 \\
-%Interarrival time & 0.024 \\
-%FIN Flag & 0.012 \\
-%URG Flag & 0 \\
-%ECE Flag & 0 \\
-%CWR Flag & 0 \\
-%NS Flag & 0 \\
-%\bottomrule
-%\end{tabular}
-%}{}
-%\subfloat[Feature dropout\label{fig:featDropout}
-%]{
-%\begin{tabular}{l r}
-%\toprule
-%Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
-%Destination port & 0.025 \\
-%Source port & 0.003 \\
-%RST Flag & 0.001 \\
-%ACK Flag & 0.001 \\
-%Protocol & 0.001 \\
-%Packet Length & 0.001 \\
-%Direction & 0.001 \\
-%SYN Flag & 0 \\
-%Interarrival time & 0 \\
-%FIN Flag & 0 \\
-%ECE Flag & 0 \\
-%URG Flag & 0 \\
-%CWR Flag & 0 \\
-%NS Flag & 0 \\
-%PSH Flag & 0 \\
-%\bottomrule
-%\end{tabular}
-%}{}
-%\end{table}
+\begin{table}
+\caption{Accuracy drop for:}
+\label{tab:feat_importance}
+\centering
+\subfloat[Input perturbation\label{fig:featRandom}
+]{
+\begin{tabular}{l r}
+\toprule
+Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
+Protocol & 0.207 \\
+Packet Length & 0.165 \\
+SYN Flag & 0.099 \\
+ACK Flag & 0.084 \\
+Direction & 0.073 \\
+Destination port & 0.071 \\
+Source port & 0.060 \\
+RST Flag & 0.057 \\
+PSH Flag & 0.056 \\
+Interarrival time & 0.024 \\
+FIN Flag & 0.012 \\
+URG Flag & 0 \\
+ECE Flag & 0 \\
+CWR Flag & 0 \\
+NS Flag & 0 \\
+\bottomrule
+\end{tabular}
+}{}
+\subfloat[Feature dropout\label{fig:featDropout}
+]{
+\begin{tabular}{l r}
+\toprule
+Feature & \hspace*{-4mm}Accuracy~drop \\ \midrule
+Destination port & 0.025 \\
+Source port & 0.003 \\
+RST Flag & 0.001 \\
+ACK Flag & 0.001 \\
+Protocol & 0.001 \\
+Packet Length & 0.001 \\
+Direction & 0.001 \\
+SYN Flag & 0 \\
+Interarrival time & 0 \\
+FIN Flag & 0 \\
+ECE Flag & 0 \\
+URG Flag & 0 \\
+CWR Flag & 0 \\
+NS Flag & 0 \\
+PSH Flag & 0 \\
+\bottomrule
+\end{tabular}
+}{}
+\end{table}
 As a first step to understanding the neural network's decisions, we  estimate how important individual features are for the model's predictions.
@@
 \subsubsection{Neural Network Weights}
-In previous works~\cite{olden_accurate_2004}, a simple method for determining feature importance in neural networks has been summing up neural network weights leading from a certain input to the prediction outcome.
+In previous works~\cite{olden_accurate_2004,olden_illuminating_2002}, a simple method for determining feature importance in neural networks has been summing up neural network weights leading from a certain input to the prediction outcome.
 The weights method can be considered for both feature importance and feature sensitivity, however, clearly, especially in the case of complex network architectures, this method is likely to provide wrong results.
@@
 The overall outcome is a classifier being able to deal with missing features.
-The results in \autoref{fig:feat_imp} show that, unlike input perturbation, feature dropout  does not vastly overestimate features' importance.
+The results in \autoref{tab:feat_importance} show that, unlike input perturbation, feature dropout  does not vastly overestimate features' importance.
 With feature dropout, it becomes apparent that only very few features  actually contain unique information, affecting accuracy when left out: the destination port and the source port.
@@
 We thus recommend training two models: One regular one and one with feature dropout to use for the feature importance.
+
+\begin{figure*}
+\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
+\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2015.pdf}
+\caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
+\label{fig:feat_imp}
+\end{figure*}
+
+\begin{figure}[b]
+\includegraphics[width=\columnwidth]{./plots/mutinfo_example.pdf}
+\caption{Two distributions yielding an identical accuracy drop.}
+\label{fig:mutinfo_example}
+\end{figure}
 
@@
 \subsubsection{Mutual Information}
-\begin{figure}
-\includegraphics[width=\columnwidth]{./plots/mutinfo_example.pdf}
-\caption{Two distributions yielding an identical accuracy drop.}
-\label{fig:mutinfo_example}
-\end{figure}
 Input perturbation and feature dropout mainly address feature importance.
@@
 
-\begin{figure*}
-\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2017.pdf}\hspace{.02\textwidth}
-\includegraphics[width=.49\textwidth]{./plots/importance/feat_imp_flow_2015.pdf}
-\caption{Feature importance metrics for the flow prediction for CIC-IDS-2017 (left side) and UNSW-NB15 (right side).}
-\label{fig:feat_imp}
-\end{figure*}
-Furthermore, let $f(A,x)$  denote the joint probability
+Furthermore, we argue that accuracy drop depicts an importance measure which might be misleading for evaluating feature sensitivity.
+To see this, let $f(A,x)$  denote the joint probability
 %\todo{Max: Ist joint probability nicht immer wenn man mehrere Ereignisse hat? Ich hätte hier probability density function geschrieben...}
@@
 yielding the same accuracy, but clearly the right-side distribution has a larger influence on the prediction, as in the right-side case the prediction deterministically depends on the feature value.
+
 To capture such dependencies, we propose to use mutual information %instead of accuracy drop
@@
 Considering its functioning of completely removing individual features, feature dropout is the most reliable method for evaluating importance for removing features.
-It is remarkable that none of the other methods is able to depict the distinct peak of importance for the destination port.
+It is remarkable that none of the other methods is able to depict the distinct peak of importance for the destination port visible for feature dropout in \autoref{fig:feat_imp} and \autoref{tab:feat_importance}.
 
@@
 We consider the question of model transferability of substantial importance for \gls{ids} applications, but out of scope for the present research.
+
+\begin{figure}
+\includegraphics[width=\columnwidth]{{"../plots/plot_pdp/flows_pdp_outcomes_0_3.pickle_10_Normal"}.pdf}
+\caption{PD plot for the source and destination port features.}
+\label{fig:pdps}
+\end{figure}
 
@@
 \glspl{pdp}, which have been proposed for non-recurrent classification methods, by plotting the \gls{rnn}'s flow prediction outcome over one of these  features.
-The results (omitted for brevity) show that for some attack types the port numbers play an important role. When looking at the \gls{pdp} for benign traffic samples, %(non-attack traffic)
+The results show that for some attack types the port numbers play an important role. When looking at the \gls{pdp} for benign traffic samples in \autoref{fig:pdps}, %(non-attack traffic)
 it becomes apparent that traffic destined to a high destination port is generally indicative of an attack. We argue that this is because most services that regular users use have low port numbers.
@@
 \end{figure}
+
 To evaluate this hypothesis, \autoref{fig:confidence} depicts the classifier's prediction confidence for each time step, along with the number of samples having at least this length, which were used for evaluating the figure.
@@
 
-\autoref{fig:pred_plots2} shows an example together with the mean values for both unmodified samples and adversarial samples.
-Also in this figure we notice that mainly the first few packets are able to influence the prediction outcome and modifying features at a later time step does not change its confidence any longer.
-In many cases, the adversarial sample generation indeed moves packet features to areas where the network is less likely to be classified as attacks, confirming the effectiveness of \glspl{pdp}.
-In other cases, however, we did not observe an agreement between \gls{pdp} and adversarial modifications, hinting at dependencies which cannot be presented by \glspl{pdp}.
-Since the \gls{iat} is undefined for the first packet, we always set it to 0.
-Still, interestingly, the plot shows, that the classifier considers packets with a higher IAT to be more likely to be attacks than those with a smaller one.
-
 \begin{figure}[h]
@@
 \end{figure}
+
+\autoref{fig:pred_plots2} shows an example together with the mean values for both unmodified samples and adversarial samples.
+Also in this figure we notice that mainly the first few packets are able to influence the prediction outcome and modifying features at a later time step does not change its confidence any longer.
+In many cases, the adversarial sample generation indeed moves packet features to areas where the network is less likely to be classified as attacks, confirming the effectiveness of \glspl{pdp}.
+In other cases, however, we did not observe an agreement between \gls{pdp} and adversarial modifications, hinting at dependencies which cannot be presented by \glspl{pdp}.
+Since the \gls{iat} is undefined for the first packet, we always set it to 0.
+Still, interestingly, the plot shows, that the classifier considers packets with a higher IAT to be more likely to be attacks than those with a smaller one.
 
@@
 
-\begin{figure}
-%AH. I think the 'Firewall off' plot is not a good example as it shows only 4 packets
-\includegraphics[width=\columnwidth]{{"./plots/plot_features/flows_pred_plots2_outcomes_0_3.pickle_2_Brute_Force-SSH-Patator"}.pdf}
-\caption{\gls{iat} and packet length for SSH brute-force attacks in CIC-IDS-2017.}% The solid line shows the median value of the feature for all flows of a specific attack type.
-The shaded region are the first and third quartile.} % think legend is enough for unterstanding
-\label{fig:plot_features}
-\end{figure}
 Finally, we investigated whether our classification task involves recognizing complex patterns in the feature space.
@@
 We now investigate two approaches to increase robustness of the \glspl{rnn} against adversarial attacks.
+Several methods have been proposed to improve robustness in the context of computer vision~\cite{akhtar_threat_2018}.
+We chose the methods presented below, because they are simple and can be applied to our recurrent setting in a straight-forward fashion.
 \subsection{Reducing Features}
-The most obvious defense is to simply omit features an attacker can manipulate. We try two different approaches:
-\begin{itemize}[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
-\item Leaving out all manipulable features: packet size and \gls{iat}
+The most obvious defense is to simply omit features an attacker can manipulate. We try two different approaches of this defense strategy:
+\begin{itemize}[leftmargin=0.5cm] %[topsep=0pt,wide,labelwidth=!,labelindent=0pt]
+\item Leaving out all manipulable features, i.e. packet size and \gls{iat}, in both directions.
 \item Leaving them out only in the direction from the attacker to the victim. This, however, does not prevent adversarial samples for botnets, for which the attacker has control over both sides.
@@
 As alternative to omitting manipulable features, we attempted to make the classifier more robust against adversarial samples by augmenting the training set by adversarial flows generated using  \gls{cw}, labeled as additional attacks.
-For this purpose, we added one adversarial sample per attack sample to the training set.
+This approach can be considered an adversarial training procedure, which is a common defense method in the related literature~\cite{akhtar_threat_2018,madry_towards_2018}.
+We added one adversarial sample per attack sample to the training set.
 Since \gls{cw} is deterministic, this is the maximum number of adversarial samples possible.
@@
 
+\begin{figure}[t]
+%AH. I think the 'Firewall off' plot is not a good example as it shows only 4 packets
+\includegraphics[width=\columnwidth]{{"./plots/plot_features/flows_pred_plots2_outcomes_0_3.pickle_2_Brute_Force-SSH-Patator"}.pdf}
+\caption{\gls{iat} and packet length for SSH brute-force attacks in CIC-IDS-2017.}% The solid line shows the median value of the feature for all flows of a specific attack type.
+The shaded region are the first and third quartile.} % think legend is enough for unterstanding
+\label{fig:plot_features}
+\end{figure}
+
+\begin{figure}[b]
+\centering
+\includegraphics[width=0.98\columnwidth]{./plots/ars/ars_15.pdf}
+\caption{\gls{ars} during adversarial training for UNSW-NB15.}
+\label{fig:avdistance}
+\end{figure}
+
 For both datasets, after a small number of epochs, it was no longer possible to create a significant number of adversarial samples for most attack categories.
@@
 After 50 epochs of training, accuracy is essentially identical to the results presented in \autoref{tab:performance_results}.
-Recall and Youden increased and precision slightly decreased.
+Recall and informedness increased and precision slightly decreased.
 However, this is due to the higher proportion of attack samples in training data and, hence, expected.
@@
 We have implemented a recurrent classifier based on LSTMs to detect network attacks, %While attack detection performance is not necessarily better than that of non-recurrent approaches,
-which can already detect attacks before they are over.
-Furthermore, the recurrent approach allows us to inspect the influence of single packets on the detection performance and shows which packets are \textit{characteristic} for attacks.
+which is able to  detect attacks already before they are over.
+The recurrent approach allowed us to inspect the influence of single packets on the detection performance and shows which packets are \textit{characteristic} for attacks.
 Even though the interpretation of \glspl{rnn} poses several difficulties, we have demonstrated methods for gaining insights into the model's functioning.
@@
 We showed that even though our use case is very different from computer vision, adversarial samples can be found efficiently, even if only ostensibly unimportant features can be modified.
-Newly developed \textit{feature sensitivity} methods can show which features can easily be manipulated by an adversary.
+We introduced \textit{feature sensitivity} methods to show which features can easily be manipulated by an adversary to cause a wrong classification.
 We proposed the \gls{ars} for quantifying and comparing the adversarial threat for \glspl{ids}.
@@
 The Titan Xp used for this research was donated by the NVIDIA Corporation.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.98\columnwidth]{./plots/ars/ars_15.pdf}
-\caption{\gls{ars} during adversarial training for UNSW-NB15.}
-\label{fig:avdistance}
-\end{figure}
 

### 2020-02-14 plot_pdp.py

@@ -7,8 +7,12 @@ import os
 import json
 import pickle
 from collections import Counter
+from matplotlib.lines import Line2D
+from matplotlib.patches import Rectangle
 
 DIR_NAME = "plots/plot_pdp"
+
+plt.rcParams["font.family"] = "serif"
 
 dataroot_basename = sys.argv[1].split('_')[0]
 
@@ -27,15 +31,17 @@ results_by_attack_number, feature_names, feature_values_by_attack_number = loade
 
 colors = plt.rcParams['axes.prop_cycle'].by_key()['color']
 
+display_names = {'srcPort': 'Source port', 'dstPort': 'Destination port'}
+
 for attack_type, (all_features, all_features_values) in enumerate(zip(results_by_attack_number, feature_values_by_attack_number)):
 
 	print("attack_type", attack_type)
-	fig, ax1 = plt.subplots()
+	fig, ax1 = plt.subplots(figsize=(5,2.4))
 
 	ax2 = ax1.twinx()
 
 	ax2.set_ylabel('Prediction')
-	ax1.set_ylabel("Number of samples")
+	ax1.set_ylabel("Flow number")
 
 	ax1.yaxis.tick_right()
 	ax1.yaxis.set_label_position("right")
@@ -46,6 +52,7 @@ for attack_type, (all_features, all_features_values) in enumerate(zip(results_by
 		continue
 	# print("all_features.shape", all_features.shape)
 	all_legends = []
+	all_labels = []
 	for feature_name, feature_index in zip(feature_names, range(all_features.shape[0])):
 
 		as_ints = list(all_features_values[feature_index].astype(np.int32))
@@ -58,24 +65,30 @@ for attack_type, (all_features, all_features_values) in enumerate(zip(results_by
 		values = counted.values()
 
 		# print("keys", keys, "values", values)
-		ret1 = ax1.bar(keys, values, width=1000, color=colors[feature_index], alpha=0.5, label="{} occurrence".format(feature_name))
+		ret1 = ax1.bar(keys, values, width=1000, color=colors[feature_index], alpha=0.2, label="{} occurrence".format(feature_name))
 
 		ret2 = ax2.plot(all_features[feature_index,0,:], all_features[feature_index,1,:], color=colors[feature_index], label="{} confidence".format(feature_name))
[…]

### 2020-02-14 paper/paper.tex
Fix typo

@@
 \usepackage{booktabs}
-\usepackage{hyperref}
+\usepackage[pdfauthor={Alexander Hartl, Maximilian Bachl, Joachim Fabini, Tanja Zseby},
+pdftitle={Explainability and Adversarial Robustness for RNNs}]{hyperref}
 \usepackage{amsmath}
@@
 Attacks contained in the datasets are shown in \autoref{tab:occurrence}.
-\autoref{tab:performance_results} shows the achieved classification performance when evalutuating metrics per packet and per flow and includes performance results for an MLP classifier from~\cite{bachl_walling_2019} for comparison.
+\autoref{tab:performance_results} shows the achieved classification performance when evaluating metrics per packet and per flow and includes performance results for an MLP classifier from~\cite{bachl_walling_2019} for comparison.
 % \autoref{tab:performance_results} shows that

### 2020-02-14 paper_rl/paper.tex
Paper fixes

@@
 They allow determining if a flow is malicious or not already before it is over, making it possible to take action immediately.
-However, considering the large number of packets that have to be inspected, the question of computational efficiency arises.
+However, considering the large number of packets that has to be inspected, the question of computational efficiency arises.
 We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high.
@@
 \end{align}
-
 Intuitively, the classification reward makes sure that packets are chosen that contain the most information and thus result in the highest classification performance.
@@
 \end{align}
-
 \autoref{fig:rewards} shows how classification and sparsity rewards are computed for a toy example.
@@
 \begin{split}
-l_{\text{a},n} =& \underbrace{ -\log \left( \pi \left( a_n \given s_n ; \theta_\text{a} \right) \right) U_n}_{\text{Policy Loss}} \\
-&- \beta \underbrace{H\left( \pi\left( s_n; \theta_a \right)\right) }_{\text{Entropy}}
+l_{\text{a},n} =& \underbrace{ -\log \left( \pi \left( a_n \given s_n ; \theta_\text{a} \right) \right) U_n}_{\text{Policy Loss}} - \beta \underbrace{H\left( \pi\left( s_n; \theta_a \right)\right) }_{\text{Entropy}}
 \end{split}\label{eq:actor}
@@
   \caption{SparseIDS framework.
-Three deep networks are shown: the LSTM classifier, the actor network and the critic network.
+Three deep networks are shown: the classifier, the actor network and the critic network.
 The actor takes actions based on the feedback from the the critic network and the classifier.
-The critic network tracks the actor's actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the actor network.}
+The critic network tracks the actor's actions based on the classifier confidence.
+The packet selector chooses packets based on the decision of the actor network.}
   \label{fig:sampling}
@@
 \end{align}
-
 When sampling from the log-normal distribution, a real number is obtained.
@@
 The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow ($n=\textit{last}$):
-
 \begin{align}
-\begin{split}
 R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{N} 1_{\text{skipped},i}\right)}{N- (\textit{last}+1+\left(\textit{last}+a_{\textit{last}} - N - 1\right))}
-\end{split}
 \label{eq:R_sparsity_last}
 \end{align}
-
 Specifically, it adds the term $\left(\textit{last}+a_{\textit{last}} - N - 1\right)$ to the denominator which penalizes the actor jumping more than one packet behind the flow.
@@
 %\end{table}
-
 For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset, which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
@@
 %\end{table}
-
+\subsection{Influence of the tradeoff}
 \begin{table}[h]
-% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
-\vspace{0.075in}
+%% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
+%\vspace{0.075in}
 %\parbox{.45\linewidth}{
@@
 \end{table}
-
-\begin{table}[h]
-\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff 0.1) and alternative strategies.}
-\label{tab:results_others}
-\centering
-\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
-& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
-Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
-Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
-Precision & \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
-Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
-F1 & \textbf{99.7\%} & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
-Youden & \textbf{98.5\%} & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
-\bottomrule
-\end{tabular*}
-%}
-\end{table}
-
-\subsection{Influence of the tradeoff}
-
 \autoref{tab:results_tradeoff} shows the influence of the tradeoff parameter $\alpha$.
@@
 
+\begin{table}[h]
+%% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
+%\vspace{0.075in}
+\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
+\centering
+\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+Actions & continuous & discrete \\	\midrule
+Sparsity & \textbf{76.3\%} & 73.6\% \\ \midrule
+Accuracy & \textbf{99.4\%} & 99.2\% \\
+Precision & 98.5\% & \textbf{98.7\%} \\
+Recall & \textbf{99.0\%} & 98.3\% \\
+F1 & \textbf{99.7\%} & 98.5\% \\
+Youden & \textbf{98.5\%} & 97.8\% \\
+\bottomrule
+\end{tabular}
+\end{table}
 %\begin{table}
@@
 %\end{table}
-
-\begin{table}[h]
-% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
-\vspace{0.075in}
-\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
-\centering
-\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
-Actions & continuous & discrete \\	\midrule
-Sparsity & \textbf{76.3\%} & 73.6\% \\ \midrule
-Accuracy & \textbf{99.4\%} & 99.2\% \\
-Precision & 98.5\% & \textbf{98.7\%} \\
-Recall & \textbf{99.0\%} & 98.3\% \\
-F1 & \textbf{99.7\%} & 98.5\% \\
-Youden & \textbf{98.5\%} & 97.8\% \\
-\bottomrule
-\end{tabular}
-\end{table}
-
 \autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions and shows that continuous actions result in a higher accuracy with fewer samples being selected.
-This leads to the conclusion that at least for our use case and for the dataset we consider, continuous actions seem to perform better.
+We chose 20 discrete actions as flows have a maximum length of 20, so with 20 actions, the \gls{rl} can decide to skip the entire flow at the first packet.
+When we tried with fewer discrete actions (like 8), we saw that sparsity was limited as the \gls{rl} would have liked to skip more packets but couldn't.
+The results lead to the conclusion that at least for our use case and for the dataset we consider, continuous actions seem to perform better.
 Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean of the log-normal distribution, which can lead to worse performance.
@@
 
+\begin{table}[h]
+% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
+\vspace{0.075in}
+\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff 0.1) and alternative strategies.}
+\label{tab:results_others}
+\centering
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
+Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
+Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
+Precision & \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
+Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
+F1 & \textbf{99.7\%} & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
+Youden & \textbf{98.5\%} & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
+\bottomrule
+\end{tabular*}
+%}
+\end{table}
 %\begin{table}
@@
 For example, with $p=0.2$, a flow of length 20 will get 4 packets while one with 4 packets will get only 1.
-However, with `first $m$'', the short and the long flow will get the same number of packets.
+However, with ``first $m$'', the short and the long flow will get the same number of packets.
 Apparently it is better if all flows get the same number of packets than if long flows get more packets.

### 2020-02-14 paper/paper.tex
Removed page numbering
Small changes

@@
 
-\thispagestyle{plain}
-\pagestyle{plain}
+% \thispagestyle{plain}
+% \pagestyle{plain}
 
@@
 we can reduce the attack surface and harden the resulting \gls{ids} while retaining all features and similar classification performance. % like for the original \gls{ml} model.
+We show that the \gls{ars} is significantly higher for the hardened model. 
 \end{itemize}
 
-We make all the source code, data and figures freely available to enable reproducibility and encourage experimentation at \url{https://github.com/CN-TU/adversarial-recurrent-ids}.
+We make all the source code, data, trained machine learning models and figures freely available to enable reproducibility and encourage experimentation at \url{https://github.com/CN-TU/adversarial-recurrent-ids}.
 

### 2020-02-19 paper/paper.tex
Add IEEE copyright notice

@@
 \usepackage{subfig}
+\usepackage{tikz}
 
@@
 
+\newcommand\copyrighttext{%
+  \footnotesize \textcopyright 2020 IEEE. Personal use of this material is permitted.
+  Permission from IEEE must be obtained for all other uses, in any current or future
+  media, including reprinting/republishing this material for advertising or promotional
+  purposes, creating new collective works, for resale or redistribution to servers or
+  lists, or reuse of any copyrighted component of this work in other works.}
+\newcommand\copyrightnotice{%
+\begin{tikzpicture}[remember picture,overlay]
+\node[anchor=south,yshift=10pt] at (current page.south) {\fbox{\parbox{\dimexpr\textwidth-\fboxsep-\fboxrule\relax}{\copyrighttext}}};
+\end{tikzpicture}%
+}
+
 \maketitle%
+\copyrightnotice
 
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{{"../plots/plot_pdp/flows_pdp_outcomes_0_3.pickle_10_Normal"}.pdf}
+\includegraphics[width=\columnwidth]{{"plots/plot_pdp/flows_pdp_outcomes_0_3.pickle_10_Normal"}.pdf}
 \caption{PD plot for the source and destination port features.}

### 2020-03-26 learn.py
Truly random seed

@@ -657,7 +657,7 @@ def train_rl():
 				if not opt.shareNet:
 					current_slice_output, _ = lstm_module(current_collated_slice)
 					current_slice_action_probs, _ = lstm_module_rl_actor(current_collated_slice)
-					current_slice_values = lstm_module_rl_critic(current_collated_slice)[0]
+					current_slice_values, _ = lstm_module_rl_critic(current_collated_slice)
 				else:
 					(current_slice_output, current_slice_action_probs, current_slice_values), _ = lstm_module(current_collated_slice)
 
@@ -971,7 +971,10 @@ def test():
 		assert batch_size <= opt.batchSize, "batch_size: {}, opt.batchSize: {}".format(batch_size, opt.batchSize)
 		lstm_module.init_hidden(batch_size)
 
-		output, seq_lens = lstm_module(input_data)
+		if not opt.shareNet:
+			output, seq_lens = lstm_module(input_data)
+		else:
+			(output, _, _), seq_lens = lstm_module(input_data)
 
 		index_tensor = torch.arange(0, output.shape[0], dtype=torch.int64).unsqueeze(1).unsqueeze(2).repeat(1, output.shape[1], output.shape[2])
 
@@ -2211,7 +2214,7 @@ if __name__=="__main__":
 	parser.add_argument('--net_actor', default='', help="path to net (to continue training) for RL actor")
 	parser.add_argument('--net_critic', default='', help="path to net (to continue training) for RL critic")
 	parser.add_argument('--function', default='train', help='the function that is going to be called')
-	parser.add_argument('--manualSeed', default=0, type=int, help='manual seed')
+	parser.add_argument('--manualSeed', default=None, help='manual seed')
 	parser.add_argument('--maxLength', type=int, default=20, help='max length')
 	parser.add_argument('--maxSize', type=int, default=sys.maxsize, help='limit of samples to consider')
 	parser.add_argument('--removeChangeable', action='store_true', help='when training remove all features that an attacker could manipulate easily without changing the attack itself')
@@ -2252,10 +2255,11 @@ if __name__=="__main__":
 
 	opt = parser.parse_args()
 	print(opt)
-	SEED = opt.manualSeed
-	random.seed(SEED)
-	np.random.seed(SEED)
-	torch.manual_seed(SEED)
[…]

### 2020-03-26 paper_rl/paper.tex
Include results for shared weights

@@
 The numbers in parentheses next to each label stand for the width of that layer (i.e.~the number of neurons (except for the inputs and outputs)).
-Parts of the neural network are shared between $v_\text{classification}$ and $v_\text{sparsity}$ as well as $\mu$ and $\sigma$ as they output quantities of similar magnitude and thus the gradients that are computed when updating the neural network are of the same order of magnitude as well.
-Our implementation also supports sharing the LSTM between all three neural network.
-However, as the five outputs can have different orders of magnitude, larger gradients would ``overwrite'' the gradients of the smaller ones.}
+Our implementation also supports sharing the LSTM between all three neural networks, which greatly reduces computational complexity: In this case, the first linear layers as well as the three layers of LSTM cells are shared while the last linear layers are still separate.
+We later show that shared weights perform almost on par with separate ones.}
   \label{fig:neuralNetworkArchitecture}
@@
 
+\subsection{Shared weights}
+\begin{table}[h]
+%% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
+%\vspace{0.075in}
+%\parbox{.45\linewidth}{
+\caption{Using shared weights for all components of \ours{}.
+This means that computational complexity in deployment is exactly the same compared to a classifier without our \gls{rl} framework.} \label{tab:results_shared}
+\centering
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+Tradeoff & 0.0 & 0.1 & 0.5 & 1.0 \\	\midrule
+Sparsity & 0\% & 77.6\% & 77.6\% & \textbf{78.2\%} \\ \midrule
+Accuracy & \textbf{99.5\%} & 99.4\% & 99.3\% & 99.0\% \\
+Precision & \textbf{99.4\%} & 99.2\% & 98.6\% & 98.9\% \\
+Recall & \textbf{98.5\%} & 98.2\% & 98.7\% & 96.9\% \\
+F1 & \textbf{99.0\%} & 98.7\% & 98.6\% & 97.9\% \\
+Youden & \textbf{98.4\%} & 97.9\% & 98.2\% & 96.6\% \\
+\bottomrule
+\end{tabular*}
+%}
+\end{table}
+\autoref{tab:results_shared} shows that when sharing the neural network weights of the classifier, the actor and the critic, \ours{} achieves the same very high accuracy as in the case in which all every component has its own weights (\autoref{tab:results_tradeoff}).
+The achieved sparsity is even higher when using separate weights.
+One explanation for this might be that sharing weights has a regularizing effect.
+With shared weights, the \gls{rl} mechanism adds no computational complexity since all the operations in the neural network are the same for the actor and the classifier (the critic is not needed during deployment).
+The only part that is not shared is the last linear layer, which connects the last LSTM layer with the outputs.
+
 \subsection{Inspecting the \gls{rl}'s behavior}

### 2020-03-26 learn.py
Add new neural network weights

@@ -2255,7 +2255,7 @@ if __name__=="__main__":
 
 	opt = parser.parse_args()
 	print(opt)
-	seed = opt.manualSeed if opt.manualSeed is not None else int.from_bytes(os.urandom(2), "big")
+	seed = int(opt.manualSeed) if opt.manualSeed is not None else int.from_bytes(os.urandom(2), "big")
 	print("seed", seed)
 	random.seed(seed)
 	np.random.seed(seed)

### 2020-04-15 paper_rl/paper.tex
Incorporate reviewers' comments and minor text fixes

@@
 We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high.
+To minimize the computational expenses of the \gls{rl}-based sampling we show that a shared neural network can be used for both the classifier and the \gls{rl} logic.
+Thus, no additional resources are consumed by the sampling in deployment.
 Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy by learning to sample only relevant packets.
@@
 To make \ours{} as lightweight as possible, our implementation allows to let the sampling and the classifier to be implemented in the same neural network, meaning that no overhead is introduced.
-Without this option enabled, the computational power required is $2\times$ what would be required without the sampling.
+With this option disabled, the computational power required is $2\times$ what would be required without the sampling.
 Thus, for this option, less than $50\%$ of packets must be sampled to gain a computational efficiency advantage.
+Fortunately, we can show that with shared weights the same accuracy can be achieved while having no computational overhead.
 
@@
 \begin{split}
-R_{\text{classification},n} = \frac{1}{N-n} \sum_{i=n+1}^{N} |1 - \left(\text{label}_i - \text{confidence}_i\right)|
+R_{\text{classification},n} = \frac{1}{N-n} \sum_{i=n+1}^{N} 1 - |\text{label}_i - \text{confidence}_i|
 \end{split}
@@
 
-Now that the reward metrics and the critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each packet $n$ as follows:
+Now that the reward metrics and the critic's outputs are defined, we can define the overall utility $U$ that we want to maximize at each packet $n$ using the sparsity-accuracy tradeoff parameter $\alpha$ as follows:
 \begin{align}
@@
 
-With $\alpha$ being the sparsity-accuracy tradeoff parameter.
 This means that we want the reward to be higher than the critic's expectation.
@@
 Ocher stands for the input, blue for linear layers, red for LSTM layers and purple for the outputs of the neural network.
-The numbers in parentheses next to each label stand for the width of that layer (i.e.~the number of neurons (except for the inputs and outputs)).
+The numbers in parentheses stand for the width of that layer.
 Our implementation also supports sharing the LSTM between all three neural networks, which greatly reduces computational complexity: In this case, the first linear layers as well as the three layers of LSTM cells are shared while the last linear layers are still separate.
-We later show that shared weights perform almost on par with separate ones.}
+We later show that shared weights perform on par with separate ones.}
   \label{fig:neuralNetworkArchitecture}
@@
 These encode the mean and standard deviation of a log-normal distribution and can never be smaller or equal to zero and hence we always apply a rectifier function at the output.
-We choose the softplus, which is a smooth rectifier function:
-\begin{align}
-\text{softplus}(x) = \log\left(1 + e^x \right)
-\end{align}
+We choose the softplus ($\log\left(1 + e^x \right)$), which is a smooth rectifier function:
 When sampling from the log-normal distribution, a real number is obtained.
@@
 
-\begin{table}[h]
-\caption{Occurrence frequency of the attack flows of CIC-IDS-2017.}
-\label{tab:occurrence}
-\centering
-\begin{tabular}{l r}
-\toprule
-Attack type & \hspace*{-4mm}Proportion \\ \midrule
-DoS Hulk	&	10.10\%	\\
-PortScan, Firewall	&	6.90\%	\\
-DDoS LOIT	&	4.08\%	\\
-Infiltration	&	3.30\%	\\
-DoS GoldenEye	&	0.32\%	\\
-DoS SlowHTTPTest	&	0.18\%	\\
-DoS Slowloris	&	0.17\%	\\
-Brute-force SSH	&	0.11\%	\\
-Botnet ARES	&	0.03\%	\\
-XSS attack	&	0.03\%	\\
-PortScan, no Fw.	&	0.02\%	\\
-Brute-force FTP	&	0.01\%	\\
-SQL injection	&	$<$0.01\%	\\
-Heartbleed	&	$<$0.01\%	\\
-\bottomrule
-\end{tabular}
-\end{table}
+%\begin{table}[h]
+%\caption{Occurrence frequency of the attack flows of CIC-IDS-2017.}
+%\label{tab:occurrence}
+%\centering
+%\begin{tabular}{l r}
+%\toprule
+%Attack type & \hspace*{-4mm}Proportion \\ \midrule
+%DoS Hulk	&	10.10\%	\\
+%PortScan, Firewall	&	6.90\%	\\
+%DDoS LOIT	&	4.08\%	\\
+%Infiltration	&	3.30\%	\\
+%DoS GoldenEye	&	0.32\%	\\
+%DoS SlowHTTPTest	&	0.18\%	\\
+%DoS Slowloris	&	0.17\%	\\
+%Brute-force SSH	&	0.11\%	\\
+%Botnet ARES	&	0.03\%	\\
+%XSS attack	&	0.03\%	\\
+%PortScan, no Fw.	&	0.02\%	\\
+%Brute-force FTP	&	0.01\%	\\
+%SQL injection	&	$<$0.01\%	\\
+%Heartbleed	&	$<$0.01\%	\\
+%\bottomrule
+%\end{tabular}
+%\end{table}
 
@@
 %\end{table}
-For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset, which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks.
-The distribution of different attacks in the dataset are shown in \autoref{tab:occurrence}.
+For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset, which includes more than 2 million flows of network data, containing both benign (74.75\%) traffic and a large number of different attacks (25.25\%, 14 attack types).
+% The distribution of different attacks in the dataset are shown in \autoref{tab:occurrence}.
 
@@
   \includegraphics[width=\columnwidth]{{"img/events.out.tfevents.1580988920.gpu"}.pdf}
-  \caption{Training with a tradeoff of 0.1 with continuous actions.
+  \caption{Training with a tradeoff of 0.1 with continuous actions for several epochs.
 Training accuracy increases sharply in the beginning while increasing more slowly later on.
@@
 %\parbox{.45\linewidth}{
-\caption{Using shared weights for all components of \ours{}.
-This means that computational complexity in deployment is exactly the same compared to a classifier without our \gls{rl} framework.} \label{tab:results_shared}
+\caption{Using shared weights for all components of \ours{} resulting in no overhead in deployment compared to a classifier without our \gls{rl} framework.} \label{tab:results_shared}
 \centering
@@
 
-One concern could be the overly deterministic nature of the decisions of \ours{} for some attack types: An adversary could see that only certain packets of a flow are looked at by the \gls{rl} system and thus only manipulate the few packets that the classifier looks at to generate adversarial samples with methods like the ones studied by \cite{hartl_explainability_2020}.
+%One concern could be the overly deterministic nature of the decisions of \ours{} for some attack types: An adversary could see that only certain packets of a flow are looked at by the \gls{rl} system and thus only manipulate the few packets that the classifier looks at to generate adversarial samples with methods like the ones studied by \cite{hartl_explainability_2020}.
 One potential remedy for this is to also randomly sample from the distribution of the actor during deployment like it is done during training.
@@
 The tradeoff is initialized as 1 and continuously lowered in a streaming fashion as long as the sparsity is higher than 50\%.
-The procedure stops if the minimum sparsity is reached or the tradeoff reached its minimum of 0, which is the case here.
-Sparsity falls from a high of around 72.5\% to a low of 71\%.}
+The procedure stops if the minimum sparsity is reached or the tradeoff reached its minimum of 0, which is the case here.}
   \label{fig:steering}
@@
 To this end, we proposed \ours{} which is a framework for packet sampling based on \gls{rl}.
-
-% Max: 1/4 is already pronounced one fourth
 We showed that the \gls{rl}-based approach with a tradeoff between classification accuracy and sparsity indeed results in higher sparsity the larger the tradeoff parameter, while potentially classification performance is sacrificed.
@@
 The continuous action framework is thus more lightweight and also can learn to skip arbitrarily many packets.
+
+We show that using a shared neural network for both the \gls{ids} as well as the sampling can achieve the same classification accuracy and sparsity while not resulting in any computational overhead during deployment.
 

### 2020-04-17 paper_rl/paper.tex
Small text fixes

@@
 
-To make \ours{} as lightweight as possible, our implementation allows to let the sampling and the classifier to be implemented in the same neural network, meaning that no overhead is introduced.
-With this option disabled, the computational power required is $2\times$ what would be required without the sampling.
+To make \ours{} as lightweight as possible, our implementation allows to let the sampling and the classifier be implemented in the same neural network, meaning that no overhead is introduced.
+With this option disabled, the computational power required is $2\times$ of what would be required without the sampling.
 Thus, for this option, less than $50\%$ of packets must be sampled to gain a computational efficiency advantage.
@@
 %The results of the sampling part are discussed in this section.
-We study the effect of the sparsity-accuracy tradeoff first, then we evaluate both discrete and continuous actions and, finally, we compare our results with other sampling techniques.
+%We study the effect of the sparsity-accuracy tradeoff first, then we evaluate both discrete and continuous actions, finally, we compare our results with other sampling techniques.
 Overall, sampling based on \gls{rl} reveals that a high number of packets can be skipped while maintaining good accuracy.
@@
 % XXX: Max: Without the vspace, the table violates the margins. WTF!!!
-\vspace{0.075in}
+%\vspace{0.075in}
 \caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff 0.1) and alternative strategies.}
@@
 %% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
-%\vspace{0.075in}
+\vspace{0.075in}
 %\parbox{.45\linewidth}{
@@
 One explanation for this might be that sharing weights has a regularizing effect.
+
 With shared weights, the \gls{rl} mechanism adds no computational complexity since all the operations in the neural network are the same for the actor and the classifier (the critic is not needed during deployment).
@@
 
-We show that using a shared neural network for both the \gls{ids} as well as the sampling can achieve the same classification accuracy and sparsity while not resulting in any computational overhead during deployment.
+We showed that using a shared neural network for both the \gls{ids} as well as the sampling can achieve the same classification accuracy and sparsity while not resulting in any computational overhead during deployment.
 

### 2020-05-04 paper_rl/paper.tex
Add motivation for intrusion detection in the context of cloud/fog/edge computing

@@
 They allow determining if a flow is malicious or not already before it is over, making it possible to take action immediately.
-However, considering the large number of packets that has to be inspected, the question of computational efficiency arises.
+However, considering the large number of packets that has to be inspected, for example in cloud/fog and edge computing, the question of computational efficiency arises.
 We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high.
@@
 % Fares:
-As the need for security mechanisms in today's infrastructure's communication networks grows, a substantial body of scientific work focuses on developing reliable and explainable frameworks for the identification of unwanted behavior.
+With constant growth of today's cloud, fog and edge computing infrastructure the need for reliable and explainable anomaly and intrusion detection methods goes hand in hand.
+This is especially important for cloud, fog and edge computing since hosting companies commonly run third parties' code and hosting illegal activity can come with legal repercussions and is ethically discouraged.
+
 In the past, a viable way to detect intrusions was to analyze the contents of the packets themselves and determine, for example, whether a packet contains potentially harmful content by matching patterns.

### 2020-05-18 learn.py
rl argument is not required, fix ident

@@ -201,9 +201,8 @@ def train():
 	lstm_module.train()
 
 	if opt.rl:
-
-	train_indices, _ = get_nth_split(dataset, n_fold, fold)
-	train_data = torch.utils.data.Subset(dataset, train_indices)
+		train_indices, _ = get_nth_split(dataset, n_fold, fold)
+		train_data = torch.utils.data.Subset(dataset, train_indices)
 	if opt.advTraining:
 		train_data = AdvDataset(train_data)
 		adv_generator = adv_internal(in_training=True, iterations=10)
@@ -1563,7 +1562,7 @@ if __name__=="__main__":
 	parser.add_argument('--n_layers', type=int, default=3, help='number of LSTM layers')
 	parser.add_argument('--adjustFeatImpDistribution', action='store_true', help='adjust randomization feature importance distributions to a practically relevant shape')
 	parser.add_argument('--skipArsDistanceCheck', action='store_true', help='stop ARS computation as soon as 50% theshold is reached')
-	parser.add_argument('--rl', required=True, help='do RL')
+	parser.add_argument('--rl', required=False, help='do RL')
 	parser.add_argument('--lookaheadSteps', type=int, default=8, help='number of steps to look into the future for RL')
 
 	# parser.add_argument('--nSamples', type=int, default=1, help='number of items to sample for the feature importance metric')

### 2020-05-18 paper_rl/paper.tex
Make paper IEEE compliant
Paper changes

@@
 %% The first command in your LaTeX source must be the \documentclass command.
+\newcommand{\CLASSINPUTtoptextmargin}{0.8in}
 \documentclass[conference]{IEEEtran}
@@
 \usepackage{booktabs}
+% Use the draft option of hyperref for IEEE compliance
 \usepackage{hyperref}
@@
 \begin{figure*}
+
 \centering
@@
 %% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
-\vspace{0.075in}
+%\vspace{0.075in}
 %\parbox{.45\linewidth}{

### 2020-05-18 learn.py
Cleaned up code

@@ -200,9 +200,8 @@ def train():
 	fold = opt.fold
 	lstm_module.train()
 
-	if opt.rl:
-		train_indices, _ = get_nth_split(dataset, n_fold, fold)
-		train_data = torch.utils.data.Subset(dataset, train_indices)
+	train_indices, _ = get_nth_split(dataset, n_fold, fold)
+	train_data = torch.utils.data.Subset(dataset, train_indices)
 	if opt.advTraining:
 		train_data = AdvDataset(train_data)
 		adv_generator = adv_internal(in_training=True, iterations=10)
@@ -1539,7 +1538,6 @@ if __name__=="__main__":
 	parser.add_argument('--nFold', type=int, default=3, help='total number of folds')
 	parser.add_argument('--batchSize', type=int, default=128, help='input batch size')
 	parser.add_argument('--net', default='', help="path to net (to continue training)")
-	parser.add_argument('--net_rl', default='', help="path to net (to continue training) for RL")
 	parser.add_argument('--function', default='train', help='the function that is going to be called')
 	parser.add_argument('--manualSeed', default=0, type=int, help='manual seed')
 	parser.add_argument('--maxLength', type=int, default=100, help='max length')
@@ -1562,8 +1560,6 @@ if __name__=="__main__":
 	parser.add_argument('--n_layers', type=int, default=3, help='number of LSTM layers')
 	parser.add_argument('--adjustFeatImpDistribution', action='store_true', help='adjust randomization feature importance distributions to a practically relevant shape')
 	parser.add_argument('--skipArsDistanceCheck', action='store_true', help='stop ARS computation as soon as 50% theshold is reached')
-	parser.add_argument('--rl', required=False, help='do RL')
-	parser.add_argument('--lookaheadSteps', type=int, default=8, help='number of steps to look into the future for RL')
 
 	# parser.add_argument('--nSamples', type=int, default=1, help='number of items to sample for the feature importance metric')
 
@@ -1621,12 +1617,5 @@ if __name__=="__main__":
 	if opt.net != '':
 		print("Loading", opt.net)
 		lstm_module.load_state_dict(torch.load(opt.net, map_location=device))
-
-	if opt.rl:
-		lstm_module_rl = OurLSTMModule(x[0].shape[-1], opt.lookaheadSteps, opt.hidden_size, opt.n_layers, batchSize, device).to(device)
-
-		if opt.net_rl != '':
[…]

### 2020-05-27 paper_rl/presentation.tex
Add presentation

%
% Beispieldokument für TU beamer theme
%
% v1.0: 12.10.2014
% 
\documentclass[xcolor={dvipsnames}]{beamer}
\usetheme{TU}

\title{SparseIDS: Learning Packet Sampling with Reinforcement Learning}

\author[Maximilian Bachl]{%
	Maximilian Bachl\email{maximilian.bachl@tuwien.ac.at} \and Fares Meghdouri\email{fares.meghdouri@tuwien.ac.at} \and Tanja Zseby\email{tanja.zseby@tuwien.ac.at} \and Joachim Fabini\email{joachim.fabini@tuwien.ac.at}
}

\institute{%
	Technische Universität Wien, Vienna, Austria
}

%\session{XXXX \#}

%Kann angepasst werden, wie es beliebt. Entweder das Datum der Präsentation oder das Datum der aktuellen Präsentations-Version.
%\date[\the\day.\the\month.\the\year]{\today}
\date[July 1, 2020]{July 1, 2020}

\begin{document}

\maketitle

\section{Introduction}

\begin{frame}{Recurrent Neural Networks (RNNs)}
\begin{itemize}
\item RNNs are neural networks for \textbf{sequences}
\item Good performance for network traffic (\textit{Explainability and Adversarial Robustness for RNNs}, Hartl, Bachl, Fabini, Zseby, 2020) with \textbf{supervised} training
\item Can detect intrusion \textbf{before} attack is over
\end{itemize}
\end{frame}

\begin{frame}{Sampling for RNNs}
\begin{itemize}
\item Processing every packet by a neural network is \textbf{resource-intensive}
\item Often, attack is apparent after the first packets of a flow $\rightarrow$ not \textbf{necessary} to continue inspecting packets
\item Some packets might contain \textbf{no information} that is useful for classification
\end{itemize}

\pause
\begin{block}{Insight:}
Only process packets if the are useful for the classifier! 
\end{block}

\end{frame}

\section{Concept}

\begin{frame}{Defining the goal}
Classifier should
\begin{itemize}
\item be \textbf{reasonably accurate}
\item \textbf{skip} packets if they \textbf{don't provide substantial benefit}
\end{itemize}

\pause
\begin{block}{Insight:}
Specify tradeoff between accuracy/sparsity 
\end{block}

\end{frame}

\begin{frame}{Defining the goal}
Optimize
\begin{align*}
\text{Objective} = \text{Accuracy} + \alpha \times \text{Sparsity}
\end{align*}
\end{frame}

\begin{frame}{Neural Network architecture}

\centering
\includegraphics[width=0.6\columnwidth]{{neural_network}.pdf}

\end{frame}

% -------------
% Last page
% -------------
\makelastslide

\end{document}

### 2020-05-28 paper_rl/presentation.tex
Finish presentation draft

@@
 \usetheme{TU}
+
+\usepackage{booktabs}
+\usepackage{tabularx}
+% Macro to make entire row bold from https://tex.stackexchange.com/questions/309833/format-whole-row-of-table-as-bold
+\newcommand\setrow[1]{\gdef\rowmac{#1}#1\ignorespaces}
+\newcommand\clearrow{\global\let\rowmac\relax}
+\clearrow
 
@@
 \end{itemize}
-
 \pause
@@
 \end{block}
-
 \end{frame}
@@
 \end{block}
-
 \end{frame}
 
-\begin{frame}{Defining the goal}
+\begin{frame}{Objective function}
 Optimize
@@
 \begin{frame}{Neural Network architecture}
+Implement using \textbf{Reinforcement Learning} based on the \textbf{Actor Critic} approach:
+\begin{itemize}
+\item \textbf{Classifier:} Decides whether a flow is benign or an attack at each packet
+\item \textbf{Critic Network:} \textbf{Estimates} the \textbf{accuracy} and \textbf{sparsity} for the current flow
+\item \textbf{Actor network:} Outputs the number of \textbf{packets to skip} to maximize objective based on Critic's output
+\end{itemize}
+\end{frame}
 
+\begin{frame}{Neural Network architecture 2}
 \centering
 \includegraphics[width=0.6\columnwidth]{{neural_network}.pdf}
+\end{frame}
 
+\section{Results}
+
+\begin{frame}{Training}
+\centering
+\includegraphics[width=1\columnwidth]{{"img/events.out.tfevents.1580988920.gpu"}.pdf}
+\end{frame}
+
+\begin{frame}{Influence of tradeoff $\alpha$}
+\centering
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+Tradeoff & 0.0 & 0.1 & 0.5 & 1.0 \\	\midrule
+Sparsity & 0\% & 76.3\% & 76.5\% & \textbf{76.8\%} \\ \midrule
+Accuracy & \textbf{99.5\%} & 99.4\% & 99.3\% & 99.0\% \\
+Precision & \textbf{99.4\%} & 98.5\% & 98.7\% & 97.9\% \\
+Recall & 98.5\% & \textbf{99.0\%} & 98.6\% & 98.1\% \\
+F1 & \textbf{99.0\%} & 98.7\% & 98.6\% & 98.0\% \\
+Youden & 98.4\% & \textbf{98.5\%} & 98.2\% & 97.4\% \\
+\bottomrule
+\end{tabular*}
+\end{frame}
+
+\begin{frame}{Continuous vs discrete actions}
+\begin{itemize}
+\item \textbf{Current techniques} for Natural Language Processing allow skipping up to \textbf{fixed number} of steps (like 20).
+\begin{itemize}
+\item Using a categorical probability distribution.
+\end{itemize}
+\item Our goal: \textbf{skip arbitrarily many} packets. 
+\begin{itemize}
+\item Need a continuous distribution that can't output $<0$ because e.g.~skipping -1 packet is impossible.
+\end{itemize}
+\end{itemize}
+\pause
+\begin{block}{Insight:}
+Use log-normal probability distribution! 
+\end{block}
+\end{frame}
+
+\begin{frame}{Log-normal distribution}
+\centering
+\includegraphics[width=0.65\columnwidth]{{img/2000px-PDF-log_normal_distributions.svg}.png}
+\end{frame}
+
+\begin{frame}{Continuous vs discrete actions}
+\centering
+\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
+Actions & continuous & discrete \\	\midrule
+Sparsity & \textbf{76.3\%} & 73.6\% \\ \midrule
+Accuracy & \textbf{99.4\%} & 99.2\% \\
+Precision & 98.5\% & \textbf{98.7\%} \\
+Recall & \textbf{99.0\%} & 98.3\% \\
+F1 & \textbf{99.7\%} & 98.5\% \\
+Youden & \textbf{98.5\%} & 97.8\% \\
+\bottomrule
+\end{tabular}
+\end{frame}
+
+\begin{frame}{Comparing with other sampling methods}
+Is RL-based sampling better than other simpler techniques?
+
+We implement:
+\begin{itemize}
+\item \textbf{random}: take packets with equal probability
+\item \textbf{every \textit{i}th}: take every $i$th packet
+\item \textbf{relative first \textit{m}}: for example: take the first 20\% of packets of each flow
+\item \textbf{first \textit{m}}: for example: take the first 10 packets of each flow
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Comparing with other sampling methods}
+\centering
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+& RL & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
+Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
+Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
+Precision & \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
+Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
+F1 & \textbf{99.7\%} & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
+Youden & \textbf{98.5\%} & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
+\bottomrule
+\end{tabular*}
+\end{frame}
+
+\begin{frame}{Shared weights}
+\begin{itemize}
+\item Having \textbf{three neural networks} is a large \textbf{overhead}.
+\item During deployment the critic isn't needed so there are two left.
+\item What if the classifier, the critic and the actor \textbf{share all weights}?
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Shared weights}
+\centering
+\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+Tradeoff & 0.0 & 0.1 & 0.5 & 1.0 \\	\midrule
+Sparsity & 0\% & 77.6\% & 77.6\% & \textbf{78.2\%} \\ \midrule
+Accuracy & \textbf{99.5\%} & 99.4\% & 99.3\% & 99.0\% \\
+Precision & \textbf{99.4\%} & 99.2\% & 98.6\% & 98.9\% \\
+Recall & \textbf{98.5\%} & 98.2\% & 98.7\% & 96.9\% \\
+F1 & \textbf{99.0\%} & 98.7\% & 98.6\% & 97.9\% \\
+Youden & \textbf{98.4\%} & 97.9\% & 98.2\% & 96.6\% \\
+\bottomrule
+\end{tabular*}
+\end{frame}
+
+\begin{frame}{Visualizing RL's behavior}
+Questions:
+\begin{itemize}
+\item Does the RL pick the \textbf{same packets} for \textbf{each attack} type?
+\item Does it pick \textbf{fewer packets} towards the \textbf{end of a flow}?
+\item How does the \textbf{confidence increase} during a flow?
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Benign traffic}
+\centering
+\includegraphics[width=0.98\columnwidth]{{"img/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_10_Normal"}.pdf}
+\end{frame}
+
+\begin{frame}{SSH Patator}
+\centering
+\includegraphics[width=0.98\columnwidth]{{"img/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_2_Brute_Force-SSH-Patator"}.pdf}
+\end{frame}
+
+\begin{frame}{DDoS Slowloris}
+\centering
+\includegraphics[width=0.98\columnwidth]{{"img/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_7_DoS_-_DDoS-DoS_slowloris"}.pdf}
+\end{frame}
+
+\begin{frame}{All samples}
+\centering
+\includegraphics[width=0.98\columnwidth]{{"img/plot_rl/flows_Jan18_14-06-48_gpu_lstm_module_8.pth_prediction_outcomes_0_3.pickle_15_All_samples"}.pdf}
+\end{frame}
+
+\begin{frame}{Steering}
+Set a maximum percentage of packets to be sampled (like 50\%). Then
+\begin{itemize}
+\item \textbf{initialize} RL with \textbf{high tradeoff} $\alpha$
+\item \textbf{continuously decrease} $\alpha$ towards 0
+\item \textbf{stop} if maximum percentage reached
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Steering}
+\centering
+  \includegraphics[width=\columnwidth]{{"img/Feb07_17-08-56_gpu_lstm_module_8.pth_global_tradeoff_1.0_steering_step_size_0.1_steering_target_sparsity_0.5_batches_to_consider_for_steering_1000_sampling_rl"}.pdf}
+\end{frame}
+
+\section{Discussion}
+\begin{frame}{Conclusions}
+\begin{itemize}
+\item Reinforcement Learning \textbf{works} very well for sampling
+\item \textbf{Sharing} all neural network weights makes our approach \textbf{very lightweight}
+\item \textbf{Continuous actions} enable skipping \textbf{arbitrarily many} packets and work well
+\item \textbf{Steering} allows using a \textbf{fixed cost} budget and then guide the RL to achieve \textbf{maximum accuracy} for that budget
+\item \textbf{Other applications} for traffic classification and other networking tasks?
+\end{itemize}
 \end{frame}

### 2020-05-28 paper_rl/presentation.tex
Incorporate Joachim's remarks

@@
 
-\author[Maximilian Bachl]{%
-	Maximilian Bachl\email{maximilian.bachl@tuwien.ac.at} \and Fares Meghdouri\email{fares.meghdouri@tuwien.ac.at} \and Tanja Zseby\email{tanja.zseby@tuwien.ac.at} \and Joachim Fabini\email{joachim.fabini@tuwien.ac.at}
+\author[M. Bachl et al.]{%
+	\underline{Maximilian Bachl}\email{maximilian.bachl@tuwien.ac.at} \and Fares Meghdouri \and Tanja Zseby \and Joachim Fabini
 }
@@
 \item RNNs are neural networks for \textbf{sequences}
-\item Good performance for network traffic (\textit{Explainability and Adversarial Robustness for RNNs}, Hartl, Bachl, Fabini, Zseby, 2020) with \textbf{supervised} training
-\item Can detect intrusion \textbf{before} attack is over
+\item Good performance for network traffic \footnote{\textit{Explainability and Adversarial Robustness for RNNs}, Hartl, Bachl, Fabini, Zseby, 2020} with \textbf{supervised} training
+\item Can detect intrusion \textbf{before} an attack is over
 \end{itemize}
@@
 \pause
-\begin{block}{Insight:}
-Only process packets if the are useful for the classifier! 
+\begin{block}{Goal:}
+Only process packets if they are useful for the classifier! 
 \end{block}
@@
 \end{align*}
+where Sparsity is $\frac{\text{sampled packets}}{\text{all packets}}$ and $\alpha$ is a configurable tradeoff parameter. 
 \end{frame}
@@
 \item \textbf{Critic Network:} \textbf{Estimates} the \textbf{accuracy} and \textbf{sparsity} for the current flow
-\item \textbf{Actor network:} Outputs the number of \textbf{packets to skip} to maximize objective based on Critic's output
+\item \textbf{Actor network:} Outputs the number of subsequent \textbf{packets to skip} to maximize objective based on Critic's output
 \end{itemize}
@@
 \begin{itemize}
-\item Need a continuous distribution that can't output $<0$ because e.g.~skipping -1 packet is impossible.
+\item Need a continuous distribution that can't output $<0$ because e.g.~skipping -1 packet is not useful.
 \end{itemize}
@@
 \pause
-\begin{block}{Insight:}
+\begin{block}{Solution:}
 Use log-normal probability distribution! 
@@
 
-We implement:
+Comparison against:
 \begin{itemize}
@@
 \item \textbf{every \textit{i}th}: take every $i$th packet
-\item \textbf{relative first \textit{m}}: for example: take the first 20\% of packets of each flow
+\item \textbf{relative first \textit{m}}: for example: take the first 20\% of packets of each flow (not possible if flow length isn't known in advance)
 \item \textbf{first \textit{m}}: for example: take the first 10 packets of each flow
@@
 \centering
-\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
-& RL & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
+\begin{tabular*}{1.05\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
+& RL & random & every $i$th & rel.~first \emph{m} & first \emph{m} \\	\midrule
 Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
 Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
-Precision & \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
+Precision& \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
 Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
@@
 \begin{frame}{Steering}
-Set a maximum percentage of packets to be sampled (like 50\%). Then
+Set a maximum percentage of packets to be sampled (like 50\%) as a \textbf{budget}. Then
 \begin{itemize}
@@
 \item \textbf{Steering} allows using a \textbf{fixed cost} budget and then guide the RL to achieve \textbf{maximum accuracy} for that budget
-\item \textbf{Other applications} for traffic classification and other networking tasks?
+\item \textbf{Other applications} of RL-based packet sampling for traffic classification and other networking tasks?
 \end{itemize}

### 2020-05-28 plot_steering.py

@@ -41,7 +41,7 @@ ax2 = ax1.twinx()
 
 lines += ax2.plot(x, tradeoffs_sampling_rl, color=colors[1])
 # plt.legend(lines, ["Sparsity", "Tradeoff"], loc='upper right')
-ax2.set_ylabel('Tradeoff')
+ax2.set_ylabel(r'Tradeoff $\alpha$')
 ax1.set_ylabel_legend(lines[0])
 ax2.set_ylabel_legend(lines[1])
 

### 2020-07-07 paper/presentation/slides.tex
Added slides

%
% Beispieldokument für TU beamer theme
%
% v1.0: 12.10.2014
% 
\documentclass{beamer}
\usetheme{TU}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{svg}
\def\xcolorversion{2.00}
\def\xkeyvalversion{1.8}
\usepackage{pgf}
\usepackage{tikz}
\usetikzlibrary{arrows,shapes,snakes,automata,backgrounds,petri}

\usepackage{mathtools}
\newcommand\givenbase[1][]{\:#1\lvert\:}
\let\given\givenbase
\newcommand\sgiven{\givenbase[\delimsize]}
\DeclarePairedDelimiterX\Basics[1](){\let\given\sgiven #1}
\newcommand\Average{E\Basics}

\usepackage{subfig}

\title{Explainability and Adversarial Robustness for RNNs}

\author[Alexander Hartl]{
	Alexander Hartl\inst{a} \and Maximilian Bachl\inst{a} \and \\ Joachim Fabini\inst{a} \and Tanja Zseby\inst{a}
}

\institute{%
	\inst{a}Technische Universität Wien, Vienna, Austria
}

%\session{XXXX \#}

%Kann angepasst werden, wie es beliebt. Entweder das Datum der Präsentation oder das Datum der aktuellen Präsentations-Version.
\date[\the\day.\the\month.\the\year]{\today}

\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}

% -------------
% Titlepage
% -------------
\maketitle

\section{Introduction}

\begin{frame}{Intrusion Detection Systems}
\includegraphics[width=\columnwidth]{figures/scenario.pdf}
\end{frame}

\begin{frame}{Intrusion Detection Systems}
\includegraphics[width=\columnwidth]{figures/scenario2.pdf}
\end{frame}

\begin{frame}{Intrusion Detection Systems}
\includegraphics[width=\columnwidth]{figures/ids.pdf}
\end{frame}

\section{Intrusion Detection using RNNs}
\begin{frame}{Intrusion Detection using RNNs}
\begin{itemize}
\item Detect attacks before they are over
\item Classifier obtains all available information
\end{itemize}
\vspace{10pt}
\centering
\includegraphics[width=0.7\columnwidth]{figures/rnn.pdf}
\end{frame}

\begin{frame}{Results}
\includegraphics[width=\columnwidth]{figures/results.pdf}
\end{frame}

\begin{frame}{Results}
\vspace*{-49pt}
\includegraphics[width=\columnwidth]{figures/results.pdf}

\vspace*{-130pt}
\fcolorbox{black}{white}{
\begin{minipage}{0.9\textwidth}
\vspace{10pt}
\begin{itemize}
\item How does the classifier come to a decision?
\item Is it possible to trick the classifier?
\end{itemize}
\vspace{10pt}
\end{minipage}
}
\end{frame}

\section{Adversarial ML}
\begin{frame}{Adversarial Samples}
\begin{itemize}
\item Craft attack samples which are classified as benign \vspace{10pt}
\item Real-world constraints
\begin{itemize}
\item Only packets from adversary manipulable
\item Only few features can be manipulated
\item Packet length cannot decrease
\item IATs cannot decrease
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Finding Adversarial Samples}
\begin{itemize}
\item Carlini-Wagner {\footnotesize ($X$...orig. sample, $Z$...logit output, $\kappa,\delta$...parameters) \hspace*{-1cm} }
\begin{equation*} \label{eq:carliniWagner}
\argmin_{\tilde X}d(X,\tilde X) + \kappa  \max(Z(\tilde X), \delta)
\end{equation*}
\item FGSM {\footnotesize ($L$...neural network loss, $\epsilon$...parameter) }
\begin{equation*}
\tilde X = X + \epsilon \, \text{sgn}( \nabla_X L(X))
\end{equation*}
\item $L_\infty$-bounded Projected Gradient Descent

\end{itemize}
\end{frame}

\begin{frame}{Finding Adversarial Samples}
\begin{figure}[h]
\includegraphics[width=\columnwidth]{../plots/adv_comparison/adv_comparison_17.pdf}
\end{figure}
\end{frame}

\begin{frame}{The Adversarial Robustness Score}
\begin{itemize}
\item Quantify ease of generating adversarial samples
\item Carlini-Wagner {\footnotesize ($X$...orig. sample, $Z$...logit output, $\kappa,\delta$...parameters) \hspace*{-1cm} }
\begin{equation*} \label{eq:carliniWagner}
\argmin_{\tilde X} d(X,\tilde X) + \kappa  \max(Z(\tilde X), \delta)
\end{equation*}
\item ARS $\approx$ avg. distance of the 50\% successful adversarial samples with lowest distance
\end{itemize}
\end{frame}

\section{Explainability}
\begin{frame}{Explainability}
\begin{enumerate}
\item Feature Importance \vspace{10pt}
\item Explainability Plots \vspace{10pt}
\item Plots for Sequences
\end{enumerate}
\end{frame}

\begin{frame}{Feature Importance}
\begin{itemize}
\item Neural Network Weights
\begin{itemize}
\item simple, fast, but inaccurate
\end{itemize}
\item Input Perturbation
\begin{itemize}
\item how important is feature for accuracy
\end{itemize}
\item Feature Dropout {\footnotesize (introduced in this research) }
\begin{itemize}
\item solves problem that classifier hasn't been trained for perturbed variables
\end{itemize}
\item Mutual Information
\begin{itemize}
\item for feature \textit{sensitivity}
\end{itemize}
\end{itemize}
\vspace*{-1cm}\hspace{5cm}\includegraphics[width=0.6\columnwidth]{../plots/importance/feat_imp_flow_2017.pdf}
\end{frame}

\begin{frame}{Explainability Plots}

\begin{equation*}
\text{PDP}_{c,i}(w) = \mathbb E_{\boldsymbol X | C} \Big(f \left( X_1,\ldots,X_{i-1},w,X_{i+1},\ldots X_n \right) | c\Big)
\end{equation*}
\begin{figure}
\includegraphics[width=\columnwidth]{{"../../plots/plot_pdp/flows_pdp_outcomes_0_3.pickle_10_Normal"}.pdf}
\end{figure}

\end{frame}

\begin{frame}{Classifier Confidence}
How fast does RNN come to a decision?

\vspace{20pt}
\includegraphics[width=\columnwidth]{{"../plots/plot/flows_pred_plots2_outcomes_0_3.pickle_15_All_samples"}.pdf}
\end{frame}

\begin{frame}{Sequential Partial Dependence}
\begin{align*}
\text{se}&\text{qPDP}_{c,i}(t,w)= \\ &\mathbb E_{X | C} \Big(f \left(h_{t-1}( X), X_1^t,\ldots,X_{i-1}^t,w,X_{i+1}^t,\ldots X_n^t \right) | c\Big) \nonumber
\end{align*}
\centering
\includegraphics[width=0.75\columnwidth]{{"../plots/plot2_adv/flows_pred_plots2_outcomes_flows_adv_1.0_notBidirectional_outcomes_0_3.pickle_0_3.pickle_7_DoS_-_DDoS-DoS_slowloris"}.pdf}
\end{frame}

\section{Defenses}
\begin{frame}{Defenses}
\begin{itemize}
\item Omit manipulable features
\begin{itemize}
\item Either all packet lengths and IATs
\item or just in one direction
\end{itemize}
$\rightarrow$ Surprisingly good accuracy \vspace{10pt}
\item Adversarial Training

\vspace{5pt}
\includegraphics[width=0.7\columnwidth]{../plots/ars/ars_15.pdf}
\end{itemize}
\end{frame}

\section{Conclusions}
\begin{frame}{Conclusions}
\begin{itemize}
\item RNNs usable for IDSs
\item Adversarial ML algorithms work
\begin{itemize}
\item Defenses work, too
\item ARS for quantifying robustness
\end{itemize}
\item Explainability
\begin{itemize}
\item Feature importance and sensitivity, dependent on purpose 
\item Visualization, e.g. sequential PDPs
\item Find packets which are characteristic for attack
\end{itemize}
\end{itemize}
\end{frame}
% -------------
% Last page
% -------------
\makelastslide

\end{document}

### 2020-07-08 paper/presentation/slides.tex
Changed slides

@@
 
-\section{Intrusion Detection using RNNs}
-\begin{frame}{Intrusion Detection using RNNs}
+\section{An RNN for Intrusion Detection}
+\begin{frame}{Using a Recurrent Neural Network}
 \begin{itemize}
@@
 \centering
-\includegraphics[width=0.7\columnwidth]{figures/rnn.pdf}
+\includegraphics[width=0.8\columnwidth]{figures/rnn.pdf}
 \end{frame}
@@
 \begin{itemize}
-\item Carlini-Wagner {\footnotesize ($X$...orig. sample, $Z$...logit output, $\kappa,\delta$...parameters) \hspace*{-1cm} }
+\item Carlini-Wagner {\footnotesize ($X$...orig.~sample; $Z$...logit output; $\kappa,\delta$...parameters) \hspace*{-1cm} }
 \begin{equation*} \label{eq:carliniWagner}
@@
 \end{equation*}
-\item FGSM {\footnotesize ($L$...neural network loss, $\epsilon$...parameter) }
+\item FGSM {\footnotesize ($L$...neural network loss; $\epsilon$...parameter) }
 \begin{equation*}
@@
 \item Quantify ease of generating adversarial samples
-\item Carlini-Wagner {\footnotesize ($X$...orig. sample, $Z$...logit output, $\kappa,\delta$...parameters) \hspace*{-1cm} }
+\item Carlini-Wagner {\footnotesize ($X$...orig.~sample; $Z$...logit output; $\kappa,\delta$...parameters) \hspace*{-1cm} }
 \begin{equation*} \label{eq:carliniWagner}
@@
 \begin{itemize}
-\item Either all packet lengths and IATs
-\item or just in one direction
+\item Either omit all packet lengths and IATs
+\item or retain features in non-manipulable direction
 \end{itemize}
@@
 \item Defenses work, too
-\item ARS for quantifying robustness
+\item \textit{Adversarial Robustness Score} for quantifying robustness
 \end{itemize}
@@
 \item Feature importance and sensitivity, dependent on purpose 
-\item Visualization, e.g. sequential PDPs
+\item Visualization\\$\rightarrow$ e.g. \textit{sequential Partial Dependence Plots}
 \item Find packets which are characteristic for attack
